---
title: "Introduction to markeR"
author: "Rita M. Silva"
date: "`r Sys.Date()`"
output:
  BiocStyle::html_document:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Introduction to markeR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(
echo = TRUE,
message = FALSE,
warning = FALSE,
results = 'markup',
dev = 'png',     # raster images to reduce HTML size
dpi = 72,
fig.retina = 1
)
```
 
# Introduction

**markeR** is an R package that provides a modular and extensible framework for the systematic evaluation of gene sets as phenotypic markers using transcriptomic data. The package is designed to support both quantitative analyses and visual exploration of gene set behaviour across experimental and clinical phenotypes.

In this vignette, we demonstrate the core functionalities of markeR using a pre-processed gene expression dataset and associated metadata from Marthandan et al. (2016) (GSE63577). This dataset comprises human fibroblast samples cultured under two experimental conditions: replicative senescence and proliferative control.

## Primary Objectives

The primary objectives of **markeR** are to:

- Establish a reproducible and standardised framework for quantifying gene sets as phenotypic markers across transcriptomic datasets.
- Provide a unified interface for complementary analytical strategies, including score-based and enrichment-based methods.
- Enable systematic comparison of user-defined gene sets with curated reference collections (e.g., MSigDB) to enhance biological interpretability.
- Deliver modular tools for visualisation and evaluation, allowing exploration of both gene set behaviour and individual gene-level contributions. 

## Features and Capabilities

The package integrates multiple analytical strategies for quantifying phenotypes using gene sets, supporting both unidirectional and bidirectional gene set definitions:

- **Score-based methods**:  log2-median expression, ranking approaches, and single-sample gene set enrichment analysis (ssGSEA) to quantify coordinated expression within a gene set.
- **Enrichment-based methods**: GSEA using moderated t- or B-statistics.
- **Gene-level exploration**: expression heatmaps, violin plots, ROC curve analysis, area under the curve (AUC) calculations, effect size estimation, and principal component analysis (PCA) to characterise the contribution of individual genes.
- **Gene set similarity assessment**: Jaccard index and log odds ratio (logOR) calculations to compare user-defined gene sets with reference or user-defined collections.

The package is implemented with flexibility and extensibility as core design principles. Visualisation functionality leverages `ggplot2`, `ComplexHeatmap`, `ggpubr`, `cowplot`, and `grid`, enabling a wide range of customisation options. The modular structure ensures compatibility with additional methods in future versions of the package, providing a robust framework for reproducible and standardised phenotypic characterisation of gene sets.

# Installation

Install the latest release from Bioconductor:

```{r, eval=FALSE}
# Install from Bioconductor
if (!requireNamespace("BiocManager", quietly = TRUE))
  install.packages("BiocManager")
BiocManager::install("markeR")
```

Or install the latest development release of markeR from [GitHub](https://github.com/) with:

```r
#devtools::install_github("DiseaseTranscriptomicsLab/markeR@*release")
```

```{r, echo=FALSE} 
library(markeR)
```
# Common Workflow

## Input Requirements
 
### Gene Sets

A named list where each element is a gene set.

- Use a **character vector** when direction is unknown (unidirectional).
- Use a **data frame** with columns `gene` and `direction` (values `+1` for up, `-1` for down) when direction is known (bidirectional).

```{r example-gene-sets-vector, echo=FALSE}
# Gene set without direction
gene_set1 <- c("GeneA", "GeneB", "GeneC", "GeneD")

# Gene set with direction
gene_set2 <- data.frame(
  gene = c("GeneX", "GeneY", "GeneZ"),
  direction = c(1, -1, 1),
  stringsAsFactors = FALSE
)

# Combine both into a named list
gene_sets <- list(
  Set1 = gene_set1,
  Set2 = gene_set2
)
```

```{r show-gene-set, echo=TRUE}
# Example
gene_sets
```

For this vignette, we will use three senescence-related gene sets that ship with the package: **LiteratureMarkers** (bidirectional), **REACTOME_CELLULAR_SENESCENCE** (undirected), and **HernandezSegura** (bidirectional). See `?genesets_example`.

```{r loadsig}
# Load example gene sets
data(genesets_example)
```

### Expression Data Frame

A filtered and normalised gene expression matrix (genes × samples). Row names must be gene identifiers; column names must match sample IDs in the metadata.

In this vignette, we use a pre‑processed dataset from Marthandan et al. (2016, GSE63577) with human fibroblasts under **replicative senescence** and **proliferative control**. See `?counts_example` for structure.

```{r loaddata}
# Load example expression data
data(counts_example)
counts_example[1:5, 1:5]
```

### Sample Metadata

A data frame with annotations per sample. Row names must match the expression matrix column names; the first column contains sample IDs.

We use the accompanying metadata from the Marthandan et al. (2016) study. See `?metadata_example`.

```{r load_metadata}
# Load example metadata
data(metadata_example)
head(metadata_example)
```

## Select Mode of Analysis

markeR provides two modes of operation:
* **Benchmarking Mode**:
Evaluates one or more gene sets against a metadata variable, returning comparative visualisations across scoring and enrichment methods.

* **Discovery Mode**:
examines the relationship between a single, well-characterised gene set and one or more variables of interest, suitable for exploratory or hypothesis-generating analyses. 
 
## Choose a Quantification Approach

Two complementary strategies are implemented for quantifying associations between gene sets and phenotypes:

### Score-Based Approach

Score-based methods assign a **single numeric value to each sample**, summarising expression within the gene set. These scores can be visualised using built-in functions, or used directly in downstream analyses (*e.g.*, group comparisons, correlations with phenotypes). 

Available methods:

* **Log2-median**: median log2 expression of all genes in the set; sensitive to absolute expression shifts.

* **Ranking**: mean rank of gene set members within each sample; reflects relative ordering independent of expression magnitude.

* **ssGSEA**: single-sample gene set enrichment score using the ssGSEA algorithm; captures coordinated up- or down-regulation across the set.

Robust gene sets are expected to yield consistent patterns across scoring methods.

### Enrichment-Based Approach

Enrichment-based methods implement a **Gene Set Enrichment Analysis (GSEA)** framework. Genes are ranked according to differential statistics, and enrichment of gene set members at the extremes of the ranked list is assessed.

* **GSEA**: computes a Normalised Enrichment Score (NES) per variable of interest, with adjustment for gene set size and multiple hypothesis testing.

This strategy is most appropriate when the aim is to evaluate coordinated gene set behaviour relative to ranked differential signals.

## Visualisation and Evaluation

### Benchmarking Mode

In **Benchmarking Mode**, `markeR` offers a range of visual summaries:

* Violin or scatter plots showing score distributions by phenotype
* Volcano plots and heatmaps based on effect sizes (Cohen’s *d* or *f*)
* ROC curves and AUC values 
* Null distribution testing using random gene sets matched for size and directionality
* Lollipop plots summarising NES with adjusted p-values
* Enrichment plots showing running enrichment scores across ranked gene lists
* Volcano plots 

Example of common workflows in Benchmarking Mode (full tutorial [here][tutorial-benchmarking]):

#### Score-based approach

Score-based methods quantify gene set activity by assigning a numeric value to each sample, representing the coordinated expression of the genes in a set. These scores facilitate direct comparisons between phenotypic groups and can be used in downstream analyses, including for example correlation with other variables of interest.
We first compute scores using the log2-median method, which calculates median-centered expression for each gene and averages across the set. In this dataset, the HernandezSegura and LiteratureMarkers signatures yield large effect sizes (|Cohen’s d|), whereas REACTOME_CELLULAR_SENESCENCE shows more modest separation between conditions. 

```{r fig.width=8, fig.height=4, warning=FALSE, message=FALSE}
# Quantification approach: scores
# Method: log2-median 
PlotScores(data = counts_example, 
           metadata = metadata_example, 
           gene_sets = genesets_example, 
           Variable="Condition",  
           method ="logmedian",    
           nrow = 1,    
           pointSize=4,  
           title="Marthandan et al. 2016", 
           widthTitle = 24,
           labsize=12, 
           titlesize = 12)  

```


Next, scores are calculated using multiple methods (log2-median, ranking, ssGSEA) to assess consistency across strategies. Outputs include heatmaps and volcano plots of effect sizes (|Cohen’s d|). Across methods, HernandezSegura and LiteratureMarkers consistently discriminate Senescent from Proliferative samples, while REACTOME_CELLULAR_SENESCENCE shows weaker and less consistent separation.


```{r warning=FALSE, message=FALSE}
Overall_Scores <- PlotScores(data = counts_example, 
                             metadata = metadata_example,  
                             gene_sets=genesets_example, 
                             Variable="Condition",  
                             method ="all",   
                             ncol = NULL, 
                             nrow = 1, 
                             widthTitle=30, 
                             limits = c(0,3.5),   
                             title="Marthandan et al. 2016", 
                             titlesize = 10,
                             ColorValues = list(heatmap=c("#F9F4AE", "#B44141"),
                                                volcano=signature_colors <- c(
                                                  HernandezSegura = "#A07395",               
                                                  REACTOME_CELLULAR_SENESCENCE = "#6B8E9E",  
                                                  LiteratureMarkers = "#CA7E45"             
                                                )
                             ),
                             mode="simple",
                             widthlegend=30, 
                             sig_threshold=0.05, 
                             cohen_threshold=0.6,
                             pointSize=6,
                             colorPalette="Paired")  
```

```{r Overall_Scores_heatmap, fig.width=10, fig.height=2, warning=FALSE, message=FALSE}
Overall_Scores$heatmap
```

```{r Overall_Scores_volcano, fig.width=6, fig.height=3, warning=FALSE, message=FALSE}
Overall_Scores$volcano
```

The discriminatory capacity of gene set scores can also be assessed using ROC curves and corresponding AUC values. HernandezSegura and LiteratureMarkers exhibit consistently high AUCs across scoring methods, while REACTOME_CELLULAR_SENESCENCE shows more heterogeneous performance.

```{r roc_scores, fig.width=10, fig.height=3, warning=FALSE, message=FALSE}
ROC_Scores(data = counts_example, 
           metadata = metadata_example, 
           gene_sets=genesets_example, 
           method = "all", 
           variable ="Condition",
           colors = c(logmedian = "#3E5587", ssGSEA = "#B65285", ranking = "#B68C52"), 
           grid = TRUE, 
           spacing_annotation=0.3, 
           ncol=NULL, 
           nrow=1,
           mode = "simple",
           widthTitle = 28,
           titlesize = 10,  
           title="Marthandan et al. 2016") 

```

Finally, false positive rates are estimated by simulating random gene sets of equal size. This step provides a null distribution against which observed signals can be compared. In this example, the LiteratureMarkers signature demonstrates the strongest performance, with only two random sets exceeding its effect size under the ranking method. Increasing the number of simulations would yield finer resolution at the cost of additional computational time.

```{r FDRSim, fig.width=12, fig.height=3,warning=FALSE, message=FALSE}

FPR_Simulation(data = counts_example,
               metadata = metadata_example,
               original_signatures = genesets_example,
               gene_list = row.names(counts_example),
               number_of_sims = 10,
               title = "Marthandan et al. 2016",
               widthTitle = 30,
               Variable = "Condition",
               titlesize = 12,
               pointSize = 5,
               labsize = 10,
               mode = "simple",
               ColorValues=NULL,
               ncol=NULL, 
               nrow=1 ) 

```


#### Enrichment-based approach


Enrichment-based methods assess whether members of a gene set are consistently overrepresented among genes ranked by differential expression statistics. Unlike score-based approaches, which assign a per-sample score, enrichment-based analyses explicitly leverage contrasts between levels of a given variable and provide formal significance testing. This strategy is therefore complementary, offering population-level insight into coordinated gene set behaviour.


The first step is the estimation of differential expression. Here, a design matrix is automatically constructed from the `Condition` variable in the metadata, defining the contrast `Senescent - Proliferative`. Internally, this fits a linear model without an intercept, enabling quick comparison between levels of a categorical variable. 

```{r DEGs, fig.width=10, fig.height=3, warning=FALSE, message=FALSE}
# Build design matrix from variables (Condition) and apply a contrast.
# The design matrix is constructed automatically using the variable "Condition".
DEGs <- calculateDE(data = counts_example,
                    metadata = metadata_example,
                    variables = "Condition",
                    contrasts = c("Senescent - Proliferative"))
DEGs$`Senescent-Proliferative`[1:5,]
```

Differential expression results can be visualised with volcano plots. These highlight the magnitude of expression changes (log2 fold change) against statistical evidence (adjusted p-value), with genes from predefined sets emphasised (Up- in green and downregulated in red). In this example:

* Genes in the HernandezSegura set annotated as upregulated (green) display positive log2 fold changes, whereas those annotated as downregulated (red) show negative log2 fold changes. This pattern is consistent with the annotation of the set, although these genes are not necessarily those exhibiting the largest absolute fold changes.
* In the LiteratureMarkers set, **LMNB1** and **MKI67** exhibit strongly negative log2FC, consistent with their roles as proliferation markers absent in senescent cells.
* Genes from the REACTOME_CELLULAR_SENESCENCE set are undirected and appear across the full range of log2FC values, diluting discriminatory power.


```{r DEGsvolcano, fig.width=10, fig.height=3,warning=FALSE, message=FALSE} 
# Change order: gene sets in columns, contrast in rows
plotVolcano(DEGs, genes = genesets_example, 
            N = NULL,
            x = "logFC", y = "-log10(adj.P.Val)", pointSize = 2,
            color = "#6489B4", highlightcolor = "#05254A", highlightcolor_upreg = "#038C65", highlightcolor_downreg = "#8C0303",nointerestcolor = "#B7B7B7",
            threshold_y = NULL, threshold_x = NULL,
            xlab = NULL, ylab = NULL, ncol = NULL, nrow = NULL, title = "Marthandan et al. 2016",
            labsize = 10, widthlabs = 24, invert = TRUE)
```

We next apply GSEA to the differential expression results. This evaluates whether genes from each set are non-randomly concentrated at the top or bottom of the ranked list. Results are reported as NES with multiple-testing adjusted p-values. Gene ranking can be based on alternative statistics (e.g. moderated t or B-statistic), which influence whether results are interpreted as enrichment/depletion or as altered directionality. For a full example and explanation, see the online tutorial [here][tutorial-benchmarking].


```{r GSEA, warning=FALSE, message=FALSE}
GSEAresults <- runGSEA(DEGList = DEGs, 
                       gene_sets = genesets_example,
                       stat = NULL,
                       ContrastCorrection = FALSE)

GSEAresults
```
 
Enrichment curves from `fgsea` show the running enrichment score across the ranked list, indicating the distribution of set members.


```{r GSEA_plotenrichment, fig.width=10, fig.height=3, warning=FALSE, message=FALSE}
plotGSEAenrichment(GSEA_results=GSEAresults, 
                   DEGList=DEGs, 
                   gene_sets=genesets_example, 
                   widthTitle=40, grid = TRUE, titlesize = 10, nrow=1, ncol=3) 
```

Enrichment results can also be summarised in a lollipop plot, which compactly shows the NES and highlights statistically significant gene sets. This provides a quick overview of which pathways are most strongly altered. Here, we can see that the HernandezSegura gene set clearly exhibits the strongest enrichment signal.
 
```{r GSEA_lollypop, fig.width=5, fig.height=4, out.width="60%", warning=FALSE, message=FALSE}
plotNESlollipop(GSEA_results=GSEAresults, 
                saturation_value=NULL, 
                nonsignif_color = "#F4F4F4", 
                signif_color = "red",
                sig_threshold = 0.05, 
                grid = FALSE, 
                nrow = NULL, ncol = NULL, 
                widthlabels=13, 
                title=NULL, titlesize=12) 
```
 
A scatter (volcano-style) plots integrate NES and adjusted p-value, enabling rapid identification of the most robustly altered sets.

```{r GSEA_volcano, fig.width=8, fig.height=3, warning=FALSE, message=FALSE}
plotCombinedGSEA(GSEAresults, sig_threshold = 0.05, PointSize=6, widthlegend = 26 )
```

From this exercise in Benchmarking Mode, we can see that two gene sets clearly perform best at discriminating between Senescent and Proliferative conditions: HernandezSegura and LiteratureMarkers. The REACTOME_CELLULAR_SENESCENCE gene set does not show a strong signal; since it is undirected and lacks information on up- or downregulation, this also dilutes the signal, highlighting the importance of providing directionality when available.

Across methods, scoring and enrichment approaches provide complementary insights. Score-based methods offer sample-level resolution, capturing strong contributions from individual genes, while enrichment-based methods evaluate coordinated behaviour across the set, typically at the group level. Even among scoring methods, rank-based approaches (e.g., ssGSEA or rank scoring) are generally more robust to technical noise, whereas magnitude-based methods (e.g., log2-median) better detect shifts in well-controlled data. Performance also depends on sample size and gene set size, explaining why HernandezSegura may appear stronger in enrichment analyses and LiteratureMarkers in scoring analyses. Integrating both approaches provides the most comprehensive view of gene set behaviour.


### Discovery Mode


In **Discovery Mode** (see full tutorial [here][tutorial-discovery]), analyses focus on a single gene set of interest, providing a targeted view of its behaviour across experimental variables. The output includes:

* Score distributions stratified by variable
* Pairwise contrasts (Cohen’s *d*) and overall effect sizes (Cohen’s *f*)
* Enrichment score summaries (NES) with adjusted p-values (*e.g.*, lollipop plots)

We illustrate this using the HernandezSegura gene set from our example collection:

```{r}
HernandezSegura_GeneSet <- list(HernandezSegura=genesets_example$HernandezSegura) 
```

For illustration purposes, two synthetic variables were added to the data:

* `DaysToSequencing`, the number of days between sample preparation and sequencing
* `Researcher`, identifying the person who processed each sample

This enables exploration of associations between gene set activity and both categorical and continuous variables.

```{r load_metadata_discovery}
data(metadata_example)
set.seed("123456")
metadata_example$Researcher <- sample(c("John","Ana","Francisca"),39, replace = TRUE)
metadata_example$DaysToSequencing <- sample(c(1:20),39, replace = TRUE)
head(metadata_example)
```

We can then examine how the selected gene set associates with these variables using enrichment-based approaches in Discovery Mode. The resulting plot highlights significant associations across variables and visually summarises the direction and strength of the effect. The “simple” mode provides comparison of effect sizes for pairwise contrasts between only two levels of the variable, but can be changed to more levels of comparison (see `?VariableAssociation`). 

In this example, the HernandezSegura gene set shows significant enrichment in samples sequenced by Francisca relative to those processed by Ana or John, suggesting that this gene set may be particularly relevant to her experimental conditions or sample processing. This gene set has also a strong enrichment for the comparison between proliferative and senescent, which is expected given the nature of the gene set and the results from the Benchmarking Mode.  

```{r GSEA_varassoc, fig.width=6, fig.height=6,warning=FALSE, message=FALSE , out.width="70%"}
VariableAssociation(
  data = counts_example,
  metadata = metadata_example,
  method = "GSEA",
  cols = c("Condition","Researcher","DaysToSequencing"),
  mode = "simple",
  gene_set = HernandezSegura_GeneSet,
  saturation_value = NULL,
  nonsignif_color = "white",
  signif_color = "red",
  sig_threshold = 0.05,
  widthlabels = 30,
  labsize = 10,
  titlesize = 14,
  pointSize = 5
) $plot

```
 
Alternatively, **score-based methods** provide a per-sample metric of gene set activity, which can then be summarised across variables. Here, scores are computed using the log2-median method. The `Condition` variable shows a large effect size (Cohen’s f), confirming strong discrimination between Senescent and Proliferative samples. In contrast, `Researcher` does not show a detectable association, in contrast to enrichment-based results. This divergence illustrates the value of applying both enrichment- and score-based approaches in a complementary manner.


```{r variableassoc_score_sen, fig.width=7, fig.height=7,warning=FALSE, message=FALSE , out.width="70%" } 
VariableAssociation(data = counts_example, 
                    metadata = metadata_example, 
                    method = "logmedian",
                    cols = c("Condition","Researcher","DaysToSequencing"),  
                    gene_set = HernandezSegura_GeneSet,
                    mode="simple",
                    nonsignif_color = "white", signif_color = "red", saturation_value=NULL,sig_threshold = 0.05,
                    widthlabels=30, labsize=10, titlesize=14, pointSize=5, discrete_colors=NULL,
                    continuous_color = "#8C6D03", color_palette = "Set2")$Overall 

```


Benchmarking Mode offers the most comprehensive set of features and allows users to seamlessly move from Discovery to Benchmarking Mode once a variable of interest has been identified and further testing is required. The main difference from Discovery Mode is that Benchmarking is designed to evaluate multiple gene sets simultaneously, whereas Discovery Mode focuses on quantifying a single, robust gene set.
 

## Individual Gene Exploration  

To better understand the contribution of individual genes within a gene set, and identify whether specific genes drive the overall signal,  `markeR` provides dedicated functionality via `VisualiseIndividualGenes.` This enables the identification of genes that may disproportionately contribute to the observed signal. Available options include:

* Expression heatmaps of genes across samples and groups 
* Violin plots showing expression distributions of individual genes 
* Correlation heatmaps to reveal co-expression patterns among genes in the set
* ROC curves and AUC values to evaluate discriminatory ability of single genes.
* Effect size estimation (Cohen’s *d*) to quantify group-level differences.
* Principal Component Analysis (PCA) restricted to the gene set, to evaluate variance explained and sample clustering.

For a complete overview, see `?VisualiseIndividualGenes` and the extended online tutorial [here](https://diseasetranscriptomicslab.github.io/markeR/articles/Tutorial_BenchmarkingMode.html#visualise-individual-gene-behaviour).
 
## Compare with Reference Gene Sets  

markeR also supports comparison of user-defined gene sets against reference collections (e.g., MSigDB), enabling systematic contextualisation. Two complementary similarity metrics are implemented:

* **Jaccard Index**:
Measures gene overlap relative to union size.

* **Log Odds Ratio (logOR)**:
Enrichment assessment using Fisher’s exact test against a specified gene universe.

Filters can be applied based on similarity thresholds (e.g., minimum Jaccard, OR, or p-value).

Example of Gene Set Similarity (full tutorial [here][tutorial-signaturesimilarity]): Comparing two user-defined signatures against other user signatures and the MSigDB C2:CP:KEGG_LEGACY collection. Similarity is summarised in a heatmap based on the log odds ratio, highlighting associations above a defined threshold (e.g., log10OR > 2). 


```{r, fig.width=6, fig.height=8}

# Example data
signature1 <- c("TP53", "BRCA1", "MYC", "EGFR", "CDK2")
signature2 <- c("ATXN2", "FUS", "MTOR", "CASP3")

signature_list <- list(
  "User_Apoptosis" = c("TP53", "CASP3", "BAX"),
  "User_CellCycle" = c("CDK2", "CDK4", "CCNB1", "MYC"),
  "User_DNARepair" = c("BRCA1", "RAD51", "ATM"),
  "User_MTOR" = c("MTOR", "AKT1", "RPS6KB1")
)

geneset_similarity(
  signatures = list(Sig1 = signature1, Sig2 = signature2),
  other_user_signatures = signature_list,
  collection = "C2",
  subcollection = "CP:KEGG_LEGACY",
  metric = "odds_ratio",
  # Define gene universe (e.g., genes from HPA or your dataset)
  universe = unique(c(
    signature1, signature2,
    unlist(signature_list),
    msigdbr::msigdbr(species = "Homo sapiens", category = "C2")$gene_symbol
  )),
  or_threshold = 100, #log10OR = 2
  pval_threshold = 0.05, 
  width_text=50
)$plot

```


# Further Reading

Full tutorials with extended examples:

- [Benchmarking Mode](https://diseasetranscriptomicslab.github.io/markeR/articles/Article_BenchmarkingMode.html)  
- [Discovery Mode](https://diseasetranscriptomicslab.github.io/markeR/articles/Article_DiscoveryMode.html)  
- [Signature Similarity](https://diseasetranscriptomicslab.github.io/markeR/articles/Article_GeneSetSimilarity.html)

# Contact

📩 For questions, contact:

**Rita Martins-Silva**  
Email: [rita.silva@medicina.ulisboa.pt](mailto:rita.silva@medicina.ulisboa.pt)



[tutorial-benchmarking]: https://diseasetranscriptomicslab.github.io/markeR/articles/Article_BenchmarkingMode.html
[tutorial-discovery]: https://diseasetranscriptomicslab.github.io/markeR/articles/Article_DiscoveryMode.html
[tutorial-signaturesimilarity]: https://diseasetranscriptomicslab.github.io/markeR/articles/Article_GeneSetSimilarity.html


# Session Information

```{r}
sessionInfo()
```