---
title: "Gene Set Similarity Tutorial"
output: rmarkdown::html_document
vignette: >
  %\VignetteIndexEntry{Introduction to markeR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Installation 

The user can install the development version of markeR from [GitHub](https://github.com/) with:
 
```{r, include = FALSE }
library(markeR)
```

``` r
# install.packages("devtools")
devtools::install_github("DiseaseTranscriptomicsLab/markeR")
```


# Signature Similarity 

Even if a user-defined gene signature demonstrates strong discriminatory power between conditions, it may reflect known biological pathways rather than novel mechanisms. To address this, the `geneset_similarity()` function computes pairwise **Jaccard indices** or **log odds ratios (logOR)** between user-provided gene signatures and a reference set, quantifying their overlap as a percentage or a statistical enrichment.

Users can compare their signatures to:

* **Custom gene sets**, defined manually, or  
* **MSigDB collections**, via the [`msigdbr`](https://cran.r-project.org/package=msigdbr) package.

The function provides options to:

* Filter by **Jaccard index threshold**, using `jaccard_threshold`  
* Filter by **odds ratio and p-value**, using `or_threshold` and `pval_threshold`  
* Limit the number of top-matching reference signatures shown, using `num_sigs_toplot`  

---

# Similarity via Jaccard Index

The **Jaccard index** measures raw set overlap:

\[
\text{Jaccard}(A, B) = \frac{|A \cap B|}{|A \cup B|}
\]

## Example 1: Compare against user-defined and MSigDB gene sets
  

```{r, fig.width=6, fig.height=6, out.width="60%", warning=FALSE, message=FALSE}
# Example data
signature1 <- c("TP53", "BRCA1", "MYC", "EGFR", "CDK2")
signature2 <- c("ATXN2", "FUS", "MTOR", "CASP3")

signature_list <- list(
  "User_Apoptosis" = c("TP53", "CASP3", "BAX"),
  "User_CellCycle" = c("CDK2", "CDK4", "CCNB1", "MYC"),
  "User_DNARepair" = c("BRCA1", "RAD51", "ATM"),
  "User_MTOR" = c("MTOR", "AKT1", "RPS6KB1")
)

geneset_similarity(
  signatures = list(Sig1 = signature1, Sig2 = signature2),
  other_user_signatures = signature_list,
  collection = "C2",
  subcollection = "CP:KEGG_LEGACY",
  num_sigs_toplot = NULL,
  jaccard_threshold = 0.05,
  msig_subset = NULL, 
  metric = "jaccard"
)
```

---

## Example 2: Restrict comparison to a custom subset of MSigDB

```{r, fig.width=6, fig.height=6, out.width="60%"}

geneset_similarity(
  signatures = list(Sig1 = signature1, Sig2 = signature2),
  other_user_signatures = NULL,
  collection = "C2",
  subcollection = "CP:KEGG_LEGACY",
  num_sigs_toplot = NULL,
  jaccard_threshold = 0,
  msig_subset = c("KEGG_MTOR_SIGNALING_PATHWAY", "KEGG_APOPTOSIS", "NON_EXISTENT_PATHWAY"), 
  metric = "jaccard"
)
```

---

# Similarity via Log Odds Ratio

The log odds ratio (logOR) provides a statistically grounded alternative for assessing gene set similarity. It measures **enrichment of one set within another**, relative to a defined background or **gene universe**, using a 2×2 contingency table and a one-sided **Fisher’s exact test**.

- **Log odds ratio (logOR)**:  
  Derived from contingency tables using:
  - Genes in both sets
  - Genes in one but not the other
  - Gene universe as background  
  Log-transformed odds ratios are visualized; statistical significance is assessed via the adjusted p-value.

> **Note**: When using `metric = "odds_ratio"`, the `universe` parameter **must** be supplied.

---

## Example 3:  Compare against user-defined and MSigDB gene sets

```{r, fig.width=6, fig.height=14, out.width="60%"}
# Define gene universe (e.g., genes from HPA or your dataset)
gene_universe <- unique(c(
  signature1, signature2,
  unlist(signature_list),
  msigdbr::msigdbr(species = "Homo sapiens", category = "C2")$gene_symbol
))

geneset_similarity(
  signatures = list(Sig1 = signature1, Sig2 = signature2),
  other_user_signatures = signature_list,
  collection = "C2",
  subcollection = "CP:KEGG_LEGACY",
  metric = "odds_ratio",
  universe = gene_universe,
  or_threshold = 1,
  pval_threshold = 0.05,
  num_sigs_toplot = 10,
  width_text=50
)
```

---

