---
title: "Supplementary Paper Figures"
author: "Rita Martins-Silva"
date: "05/05/2025"
output: html_document
---

#  {.tabset .tabset-pills}

## Set up {.tabset}
 

### Libraries

```{r}
library("ggplot2")
library("colorspace")
library("scales")
library("scater") 
library("reshape2")
library("data.table")
library("edgeR")
library(ggpubr)
library(cowplot)

```


### Functions

```{r}
wrap_title <- function(title, width = 30) {
  if (nchar(title) <= width) {
    return(title)  # No need to wrap if it fits
  }

  wrapped_title <- ""
  while (nchar(title) > width) {
    # Find positions of capital letters and symbols near the wrap point
    capital_pos <- gregexpr("[A-Z]", title)[[1]]
    symbol_pos <- gregexpr("(_|-|:|\\+|\\\\|/|\\*|\\.|,|;|\\?|!)", title)[[1]]

    # Check for symbol breaks within the last few characters (width - 5 to width)
    valid_symbol_breaks <- symbol_pos[symbol_pos >= (width - 5) & symbol_pos <= width]

    if (length(valid_symbol_breaks) > 0) {
      # If a suitable symbol is found, break at the first valid symbol
      break_at <- valid_symbol_breaks[1]
    } else {
      # If no suitable symbol, look for capital letters within the same range
      valid_capital_breaks <- capital_pos[capital_pos >= (width - 5) & capital_pos <= width]

      if (length(valid_capital_breaks) > 0) {
        # If a capital letter is found, break just before the capital letter
        break_at <- valid_capital_breaks[1] - 1
      } else {
        # If no suitable symbol or capital letter, break at width
        break_at <- width
      }
    }

    # Append the wrapped line
    wrapped_title <- paste0(wrapped_title, substr(title, 1, break_at), "\n")

    # Update title with the remaining text after the break
    title <- substr(title, break_at + 1, nchar(title))
  }

  # Add the remaining part of the title
  wrapped_title <- paste0(wrapped_title, title)

  return(wrapped_title)
}
```

```{r}
PlotScores_Categorical_adapted <- function(data, metadata, gene_sets,
                                           method = c("ssGSEA", "logmedian", "ranking"),
                                           ColorVariable = NULL, GroupingVariable = NULL,
                                           ColorValues = NULL, ConnectGroups = FALSE, ncol = NULL, nrow = NULL, title = NULL,
                                           widthTitle = 10, titlesize = 12, limits = NULL, legend_nrow = NULL, pointSize = 2,
                                           xlab = NULL, labsize = 10, compute_cohen=TRUE, cond_cohend = NULL, pvalcalc = FALSE, mode = c("simple","medium","extensive"),
                                           widthlegend=22, cohen_threshold=0.6, colorPalette="Set3") {
  
  method <- match.arg(method)
  
  ResultsList <- CalculateScores(data = data,
                                 metadata = metadata,
                                 gene_sets = gene_sets,
                                 method = method)
  
  # if grouping variable is NULL, then the function displays a density / distribution of scores
  if (is.null(GroupingVariable) | is.null(metadata)) {
    
    plot_list <- list()
    
    for (signature in names(ResultsList)) {
      
      df <- ResultsList[[signature]]
      # Wrap the signature name using the helper function
      wrapped_title <- wrap_title(signature, width = widthTitle)
      
      ColorValues <- if (is.null(ColorValues)) "#ECBD78" else ColorValues
      
      p <- ggplot2::ggplot(df, ggplot2::aes(x = score)) +
        ggplot2::geom_density(fill = ColorValues, alpha = 0.5) +
        ggplot2::labs(title = "Density Plot of Score", x = xlab, y = "Density") +
        # add points below density
        ggplot2::geom_rug(ggplot2::aes(x = score), color=ColorValues, sides = "b",  alpha = 0.8, size = .5, length = grid::unit(0.035, "npc"))
      
      # Customize the plot appearance.
      p <- p + ggplot2::theme_classic() +
        ggplot2::labs(title = wrapped_title, color = "", x = "", y = "") +
        ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 45, hjust = 1, size = labsize - .5),
                       axis.text.y = ggplot2::element_text(  size = labsize - .5),
                       plot.title = ggplot2::element_text(hjust = 0.5, size = titlesize-1),
                       plot.subtitle = ggplot2::element_text(hjust = 0.5, size = titlesize - 1.5, face = "italic"),
                       legend.position="none")
      
      # If limits is specified, crop the plot without adjusting the data (violins).
      if (!is.null(limits)) {
        p <- p + ggplot2::coord_cartesian(xlim = limits)
      }
      
      plot_list[[signature]] <- p
      
    }
    
    n <- length(plot_list)
    
    # Determine grid layout
    if (is.null(ncol) && is.null(nrow)) {
      ncol <- ceiling(sqrt(n))
      nrow <- ceiling(n / ncol)
    } else if (is.null(ncol)) {
      ncol <- ceiling(n / nrow)
    } else if (is.null(nrow)) {
      nrow <- ceiling(n / ncol)
    }
    
    # create label for y axis
    if (method == "ssGSEA") {
      xlab <- "ssGSEA Enrichment Score"
    } else if (method == "logmedian") {
      xlab <- "Normalised Signature Score"
    } else if (method == "ranking") {
      xlab <- "Signature Genes' Ranking"
    }
    
    combined_plot <- ggpubr::ggarrange(plotlist = plot_list, ncol = ncol, nrow = nrow,  align = "h")
    combined_plot <- ggpubr::annotate_figure(combined_plot,
                                             left = grid::textGrob("Density",
                                                                   rot = 90, vjust = 1, gp = grid::gpar(cex = 1.3, fontsize = labsize)),
                                             bottom = grid::textGrob(xlab, gp = grid::gpar(cex = 1.3, fontsize = labsize)),
                                             top = grid::textGrob(title, gp = grid::gpar(cex = 1.3, fontsize = titlesize + 2)))
    return(combined_plot)
  }
  
  if (!(GroupingVariable %in% colnames(metadata)))
    stop(paste0(GroupingVariable, " not in metadata columns. Please check metadata."))
  
  # Initialize an empty list to store individual ggplot objects.
  plot_list <- list()
  
  # Loop over each gene signature in the ResultsList.
  for (signature in names(ResultsList)) {
    # Extract the data frame for the current signature.
    df <- ResultsList[[signature]]
    
    # Using factors so we can retrieve the first condition for Cohen's d if none is specified.
    df[, GroupingVariable] <- factor(df[, GroupingVariable],
                                     levels = sort(unique(as.character(df[, GroupingVariable]))))
    
    # Wrap the signature name using the helper function.
    wrapped_title <- wrap_title(signature, width = widthTitle)
    
    # Create a base ggplot object with the specified grouping on the x-axis and score on the y-axis.
    p <- ggplot2::ggplot(df, ggplot2::aes_string(x = GroupingVariable, y = "score"))
    
    # Add jittered points, optionally colored by ColorVariable.
    if (!is.null(ColorVariable)) {
      p <- p + ggplot2::geom_jitter(ggplot2::aes_string(color = ColorVariable), size = pointSize, alpha = 0.5)
    } else {
      p <- p + ggplot2::geom_jitter(size = pointSize, alpha = 0.5) +
        ggplot2::scale_color_brewer(palette = colorPalette)
    }
    
    # Overlay violin plots.
    p <- p + ggplot2::geom_violin(alpha = 0.5, scale = "width")
    
    # Add median summary crossbar.
    p <- p + ggplot2::stat_summary(fun = median, fun.min = median, fun.max = median,
                                   geom = "crossbar", width = 0.25,
                                   position = ggplot2::position_dodge(width = 0.13))
    
    # Add stats: Compute Cohen's d (and optionally p‑value)
    if(compute_cohen){
      if (!is.null(cond_cohend)){
        # can be of the following form:
        # cond_cohend <- list(A=c("Senescent"),
        #                     B=c("Proliferative","Quiescent"))
        
        if (sum(unlist(cond_cohend) %in% unique(df[, GroupingVariable])) != length(unique(df[, GroupingVariable])))
          warning("Warning: Not all conditions of GroupingVariable were specified for Cohen's d calculation")
        
        x <- df[df[[GroupingVariable]] %in% cond_cohend[[1]], "score", drop = TRUE]
        y <- df[df[[GroupingVariable]]  %in% cond_cohend[[2]], "score", drop = TRUE]
        
        cohen_d_results <- cohen_d(x, y)
        
        # df$cohen <- ifelse(df[, GroupingVariable] %in% cond_cohend[[1]], names(cond_cohend)[1], names(cond_cohend)[2])
        # cohen_d_results <- rstatix::cohens_d(df, formula = score ~ cohen)
        
        if (pvalcalc) {
          df$cohen <- ifelse(df[, GroupingVariable] %in% cond_cohend[[1]], names(cond_cohend)[1], names(cond_cohend)[2])
          ttest_results <- rstatix::t_test(df, formula = score ~ cohen)
          p_val <- ttest_results$p[1]
          line1 <- wrap_title(paste0("Cohen's d = ", round(cohen_d_results, 3)), width = widthTitle)
          line2 <- wrap_title(paste0("p = ", round(p_val, 3)), width = widthTitle)
          subtitle <- paste(line1, line2, sep = "\n")
        } else {
          subtitle <- wrap_title(paste0("Cohen's d = ", round(cohen_d_results, 3)), width = widthTitle)
        }
        
        
      } else {
        
        if(length(unique(df[, GroupingVariable])) < 2){
          
          warning("Not enough conditions available to report Cohen's d.")
          
        } else if(length(unique(df[, GroupingVariable])) == 2) {
          
          # Calculate Cohen's d based on ordering of the x axis
          group1 <- levels(df[, GroupingVariable])[1]
          group2 <- levels(df[, GroupingVariable])[2]
          
          x <- df[df[[GroupingVariable]] == group1, "score", drop = TRUE]
          y <- df[df[[GroupingVariable]] == group2, "score", drop = TRUE]
          
          cohen_d_results <- cohen_d(x, y)
          
          if (pvalcalc) {
            ttest_results <- rstatix::t_test(df, formula = score ~ GroupingVariable)
            p_val <- ttest_results$p[1]
            line1 <- wrap_title(paste0("Cohen's d = ", round(cohen_d_results$effsize, 3)), width = widthTitle)
            line2 <- wrap_title(paste0("p = ", round(p_val, 3)), width = widthTitle)
            subtitle <- paste(line1, line2, sep = "\n")
          } else {
            subtitle <- wrap_title(paste0("Cohen's d = ", round(cohen_d_results$effsize, 3)), width = widthTitle)
          }
          
          
          
        } else if(length(unique(df[, GroupingVariable])) > 2){
          
          # Calculate Cohen's f
          type <- identify_variable_type(df, GroupingVariable)[GroupingVariable]
          #Without scaling, the coefficient represents the change in score per unit increase in the variable (if numeric, the unit of the variable. Makes sense to not scale...)
          model <- lm(score ~ get(GroupingVariable), data = df)
          results_var <- compute_cohens_f_pval(model, type)
          
          
          if (pvalcalc) {
            line1 <- wrap_title(paste0("Cohen's f = ", round(results_var["Cohen_f"], 3)), width = widthTitle)
            line2 <- wrap_title(paste0("p = ", round(results_var["P_Value"], 3)), width = widthTitle)
            subtitle <- paste(line1, line2, sep = "\n")
          } else {
            subtitle <- wrap_title(paste0("Cohen's f = ", round(results_var["Cohen_f"], 3)), width = widthTitle)
          }
          
        }
        
        
      }
      
    } else {
      
      subtitle <- NULL
      
    }
    # If ConnectGroups is TRUE, add a line connecting medians across groups.
    if (ConnectGroups && !is.null(ColorVariable)) {
      p <- p + ggplot2::stat_summary(ggplot2::aes_string(group = ColorVariable, color = ColorVariable),
                                     fun.y = median, geom = "line", size = 1.5, alpha = 0.75,
                                     show.legend = FALSE)
    }
    
    # Customize the plot appearance.
    p <- p + ggplot2::theme_bw() +
      ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 45, hjust = 1, size = labsize),
                     axis.text.y = ggplot2::element_text(  size = labsize),
                     plot.title = ggplot2::element_text(hjust = 0.5, size = titlesize-1),
                     plot.subtitle = ggplot2::element_text(hjust = 0.5, size = titlesize - 1.5, face = "italic")) +
      ggplot2::labs(title = wrapped_title, subtitle = subtitle, color = "", x = "", y = "")
    
    # If ColorValues is provided, use a manual color scale; otherwise, if ColorVariable is provided,
    # use a default brewer palette.
    if (!is.null(ColorValues)) {
      p <- p + ggplot2::scale_color_manual(values = ColorValues)
    } else if (!is.null(ColorVariable)) {
      p <- p + ggplot2::scale_color_brewer(palette = colorPalette)
    }
    
    # If limits is specified, crop the plot without adjusting the data (violins).
    if (!is.null(limits)) {
      p <- p + ggplot2::coord_cartesian(ylim = limits)
    }
    
    # Adjust legend rows if legend_nrow is specified.
    if (!is.null(legend_nrow)) {
      p <- p + ggplot2::guides(color = ggplot2::guide_legend(nrow = legend_nrow))
    }
    
    # Store the plot in the list.
    plot_list[[signature]] <- p + theme(legend.position = "none")
  }
  
  n <- length(plot_list)
  
  # Determine grid layout.
  if (is.null(ncol) && is.null(nrow)) {
    ncol <- ceiling(sqrt(n))
    nrow <- ceiling(n / ncol)
  } else if (is.null(ncol)) {
    ncol <- ceiling(n / nrow)
  } else if (is.null(nrow)) {
    nrow <- ceiling(n / ncol)
  }
  
  # Combine plots.
  combined_plot <- ggpubr::ggarrange(plotlist = plot_list, ncol = ncol, nrow = nrow , align = "h") #, common.legend = TRUE
  
  # Annotate with axis labels.
  if (is.null(xlab)) {
    xlab <- GroupingVariable
  }
  
  if (!is.null(title)) title <- wrap_title(title, width = widthTitle)
  
  # Create label for y axis based on method.
  if (method == "ssGSEA") {
    ylab <- "ssGSEA Enrichment Score"
  } else if (method == "logmedian") {
    ylab <- "Normalised Signature Score"
  } else if (method == "ranking") {
    ylab <- "Signature Genes' Ranking"
  }
  
  combined_plot <- ggpubr::annotate_figure(combined_plot,
                                           bottom = grid::textGrob(xlab, gp = grid::gpar(cex = 1.3, fontsize = labsize)),
                                           top = grid::textGrob(title, gp = grid::gpar(cex = 1.3, fontsize = titlesize)))
  return(combined_plot)
}
```

```{r}
# adapted to not have the color legend

PlotScores_adapted <- function(data, metadata, gene_sets,
                               method = c("ssGSEA", "logmedian", "ranking", "all"),
                               ColorVariable = NULL, Variable = NULL,
                               ColorValues = NULL, ConnectGroups = FALSE, ncol = NULL, nrow = NULL, title = NULL,
                               widthTitle = 20, titlesize = 12, limits = NULL, legend_nrow = NULL, pointSize = 4,
                               xlab = NULL, labsize = 10, compute_cohen=TRUE, cond_cohend = NULL, pvalcalc = FALSE, mode = c("simple","medium","extensive"),
                               widthlegend=22, sig_threshold=0.05, cohen_threshold=0.5, colorPalette="Set3", cor=c("pearson","spearman","kendall")) {
  
  method <- match.arg(method)
  
  type <- identify_variable_type(metadata, Variable)#[Variable]
  
  if (method == "all") { # returns heatmap
    
    if (type =="Numeric"){
      
      cohenlist <- CohenF_allConditions(data = data, metadata = metadata, gene_sets = gene_sets, variable = Variable )
      
    } else {
      
      cohenlist <- CohenD_allConditions(data = data, metadata = metadata, gene_sets = gene_sets, variable = Variable, mode = mode)
      
    }
    
    # if user wants "all" methods, a heatmap of Cohen's d's is returned, for all combination of variables in GroupingVariable
    Heatmap_Final <- Heatmap_Cohen_adapted(cohenlist = cohenlist,
                                   nrow = nrow,
                                   ncol = ncol,
                                   limits = limits,
                                   widthTitle = widthTitle,
                                   titlesize = titlesize,
                                   ColorValues = ColorValues,
                                   title = title )
    
    Volcano_Cohen <- Volcano_Cohen(cohenlist = cohenlist,
                                   titlesize = 12,
                                   ColorValues = ColorValues,
                                   title = title,
                                   widthlegend = widthlegend,
                                   pointSize = pointSize,
                                   sig_threshold = sig_threshold,
                                   cohen_threshold = cohen_threshold,
                                   colorPalette =colorPalette,
                                   ncol = ncol,
                                   nrow = nrow)
    
    return(list(heatmap=Heatmap_Final$plt,
                volcano=Volcano_Cohen$plt))
    
  } else {
    
    
    
    if (type!="Numeric"){
      
      return(
        
        PlotScores_Categorical_adapted(data=data, metadata=metadata, gene_sets=gene_sets,
                                       method = method,
                                       ColorVariable = ColorVariable, GroupingVariable = Variable,
                                       ColorValues = ColorValues, ConnectGroups = ConnectGroups, ncol = ncol, nrow = nrow, title = title,
                                       widthTitle = widthTitle, titlesize = titlesize, limits = limits, legend_nrow = legend_nrow, pointSize = pointSize,
                                       xlab = xlab, labsize = labsize, compute_cohen=compute_cohen, cond_cohend = cond_cohend, pvalcalc = pvalcalc, mode = mode,
                                       widthlegend=widthlegend, cohen_threshold=cohen_threshold, colorPalette=colorPalette)
        
      )
      
    } else {
      
      return(
        
        PlotScores_Numeric(data=data,
                           metadata=metadata,
                           gene_sets=gene_sets,
                           method = method,
                           Variable = Variable,
                           ColorValues = ColorValues,
                           ncol = ncol,
                           nrow = nrow,
                           title = title,
                           widthTitle = widthTitle,
                           titlesize = titlesize,
                           limits = limits,
                           pointSize = pointSize,
                           xlab = xlab,
                           labsize = labsize,
                           compute_cohen = compute_cohen,
                           pvalcalc = pvalcalc,
                           colorPalette = colorPalette,
                           cor=cor)
        
      )
      
    }
    
    
  }
  
  
}
```

```{r}
PlotScores_Numeric_adapted <- function(data,
                                       metadata,
                                       gene_sets,
                                       method = c("ssGSEA", "logmedian", "ranking"),
                                       Variable = NULL,
                                       ColorValues = NULL,
                                       ncol = NULL,
                                       nrow = NULL,
                                       title = NULL,
                                       widthTitle = 10,
                                       titlesize = 12,
                                       limits = NULL,
                                       pointSize = 2,
                                       xlab = NULL,
                                       labsize = 10,
                                       compute_cohen = TRUE,
                                       pvalcalc = FALSE,
                                       colorPalette = "Set3",
                                       cor = c("pearson","spearman","kendall")) {
  
  method <- match.arg(method)
  
  
  ResultsList <- CalculateScores(data = data,
                                 metadata = metadata,
                                 gene_sets = gene_sets,
                                 method = method)
  
  
  # if grouping variable is NULL, then the function displays a density / distribution of scores
  if (is.null(Variable) | is.null(metadata)) {
    
    plot_list <- list()
    
    for (signature in names(ResultsList)) {
      
      df <- ResultsList[[signature]]
      # Wrap the signature name using the helper function
      wrapped_title <- wrap_title(signature, width = widthTitle)
      
      ColorValues <- if (is.null(ColorValues)) "#ECBD78" else ColorValues
      
      p <- ggplot2::ggplot(df, ggplot2::aes(x = score)) +
        ggplot2::geom_density(fill = ColorValues, alpha = 0.5) +
        ggplot2::labs(title = "Density Plot of Score", x = xlab, y = "Density")
      
      # Customize the plot appearance.
      p <- p + ggplot2::theme_classic() +
        ggplot2::labs( color = "", x = "", y = "") +
        ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 45, hjust = 1, size = labsize - .5),
                       axis.text.y = ggplot2::element_text(  size = labsize - .5),
                       plot.title = ggplot2::element_text(hjust = 0.5, size = titlesize-1),
                       plot.subtitle = ggplot2::element_text(hjust = 0.5, size = titlesize - 1.5, face = "italic"))
      
      
      
      # If limits is specified, crop the plot without adjusting the data (violins).
      if (!is.null(limits)) {
        p <- p + ggplot2::coord_cartesian(xlim = limits)
      }
      
      plot_list[[signature]] <- p
      
    }
    
    n <- length(plot_list)
    
    # Determine grid layout
    if (is.null(ncol) && is.null(nrow)) {
      ncol <- ceiling(sqrt(n))
      nrow <- ceiling(n / ncol)
    } else if (is.null(ncol)) {
      ncol <- ceiling(n / nrow)
    } else if (is.null(nrow)) {
      nrow <- ceiling(n / ncol)
    }
    
    # create label for y axis
    if (method == "ssGSEA") {
      xlab <- "ssGSEA Enrichment Score"
    } else if (method == "logmedian") {
      xlab <- "Normalised Signature Score"
    } else if (method == "ranking") {
      xlab <- "Signature Genes' Ranking"
    }
    
    combined_plot <- ggpubr::ggarrange(plotlist = plot_list, ncol = ncol, nrow = nrow, common.legend = TRUE, align = "h")
    combined_plot <- ggpubr::annotate_figure(combined_plot,
                                             left = grid::textGrob("Density",
                                                                   rot = 90, vjust = 1, gp = grid::gpar(cex = 1.3, fontsize = labsize)),
                                             bottom = grid::textGrob(xlab, gp = grid::gpar(cex = 1.3, fontsize = labsize)),
                                             top = grid::textGrob(title, gp = grid::gpar(cex = 1.3, fontsize = titlesize + 2)))
    return(combined_plot)
  }
  
  if (!(Variable %in% colnames(metadata)))
    stop(paste0(Variable, " not in metadata columns. Please check metadata."))
  
  # Initialize an empty list to store individual ggplot objects.
  plot_list <- list()
  
  # Loop over each gene signature in the ResultsList.
  for (signature in names(ResultsList)) {
    # Extract the data frame for the current signature.
    df <- ResultsList[[signature]]
    #
    #       # Using factors so we can retrieve the first condition for Cohen's d if none is specified.
    #       df[, Variable] <- factor(df[, Variable],
    #                                        levels = sort(unique(as.character(df[, Variable]))))
    
    # Wrap the signature name using the helper function.
    wrapped_title <- wrap_title(signature, width = widthTitle)
    
    # Create a base ggplot object with the specified grouping on the x-axis and score on the y-axis.
    p <- ggplot2::ggplot(df, ggplot2::aes_string(x = Variable, y = "score"))
    
    #add points
    # If ColorValues is provided, use a manual color scale;
    # use a default brewer palette.
    ColorValues <- if (is.null(ColorValues)) "#5264B6" else ColorValues
    p <- p + ggplot2::geom_point(size = pointSize, alpha = 0.5, color=ColorValues[1])+
      # add density lines
      geom_density2d( colour="white")
    
    # Add  line
    p <- p + ggplot2::geom_smooth(method = "lm", col = "black", se = FALSE, size=2) + ggpubr::stat_cor(method=cor) # cor in "pearson" (default), "kendall", or "spearman".
    
    # Add stats: Compute Cohen's f (and optionally p‑value)
    if(compute_cohen){
      
      # Calculate Cohen's f
      type <- identify_variable_type(df, Variable)[Variable]
      #Without scaling, the coefficient represents the change in score per unit increase in the variable (if numeric, the unit of the variable. Makes sense to not scale...)
      model <- lm(score ~ get(Variable), data = df)
      results_var <- compute_cohens_f_pval(model, type)
      
      if (pvalcalc) {
        line1 <- wrap_title(paste0("Cohen's f = ", round(results_var["Cohen_f"], 3)), width = widthTitle)
        line2 <- wrap_title(paste0("p = ", round(results_var["P_Value"], 3)), width = widthTitle)
        subtitle <- paste(line1, line2, sep = "; ")
        
      } else {
        subtitle <- wrap_title(paste0("Cohen's f = ", round(results_var["Cohen_f"], 3)), width = widthTitle)
      }
      
    } else {
      subtitle <- NULL
    }
    
    
    # Customize the plot appearance.
    p <- p + ggplot2::theme_bw() +
      ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 45, hjust = 1, size = labsize),
                     axis.text.y = ggplot2::element_text(  size = labsize),
                     plot.title = ggplot2::element_text(hjust = 0.5, size = titlesize-1),
                     plot.subtitle = ggplot2::element_text(hjust = 0.5, size = titlesize - 1.5, face = "italic")) +
      ggplot2::labs(title =  subtitle, color = "", x = "", y = "") 
    
    
    
    # If limits is specified, crop the plot without adjusting the data (violins).
    if (!is.null(limits)) {
      p <- p + ggplot2::coord_cartesian(ylim = limits)
    }
    
    # Store the plot in the list.
    plot_list[[signature]] <- p
  }
  
  n <- length(plot_list)
  
  # Determine grid layout.
  if (is.null(ncol) && is.null(nrow)) {
    ncol <- ceiling(sqrt(n))
    nrow <- ceiling(n / ncol)
  } else if (is.null(ncol)) {
    ncol <- ceiling(n / nrow)
  } else if (is.null(nrow)) {
    nrow <- ceiling(n / ncol)
  }
  
  # Combine plots.
  combined_plot <- ggpubr::ggarrange(plotlist = plot_list, ncol = ncol, nrow = nrow, common.legend = TRUE, align = "h")
  
  # Annotate with axis labels.
  if (is.null(xlab)) {
    xlab <- Variable
  }
  
  if (!is.null(title)) title <- wrap_title(title, width = widthTitle)
  
  # Create label for y axis based on method.
  if (method == "ssGSEA") {
    ylab <- "Score"
  } else if (method == "logmedian") {
    ylab <- "Score"
  } else if (method == "ranking") {
    ylab <- "Score"
  }
  
  combined_plot <- ggpubr::annotate_figure(combined_plot,
                                           left = grid::textGrob(ylab,
                                                                 rot = 90, vjust = 2 , gp = grid::gpar(cex = 1.3, fontsize = labsize)),
                                           bottom = grid::textGrob(xlab, gp = grid::gpar(cex = 1.3, fontsize = labsize),vjust = -2 ),
                                           top = grid::textGrob(title, gp = grid::gpar(cex = 1.3, fontsize = titlesize)))
  return(combined_plot)
  
}
```



```{r}
PlotScores_Categorical_adapted_noX <- function(data, metadata, gene_sets,
                                           method = c("ssGSEA", "logmedian", "ranking"),
                                           ColorVariable = NULL, GroupingVariable = NULL,
                                           ColorValues = NULL, ConnectGroups = FALSE, ncol = NULL, nrow = NULL, title = NULL,
                                           widthTitle = 10, titlesize = 12, limits = NULL, legend_nrow = NULL, pointSize = 2,
                                           xlab = NULL, labsize = 10, compute_cohen=TRUE, cond_cohend = NULL, pvalcalc = FALSE, mode = c("simple","medium","extensive"),
                                           widthlegend=22, cohen_threshold=0.6, colorPalette="Set3") {
  
  method <- match.arg(method)
  
  ResultsList <- CalculateScores(data = data,
                                 metadata = metadata,
                                 gene_sets = gene_sets,
                                 method = method)
  
  # if grouping variable is NULL, then the function displays a density / distribution of scores
  if (is.null(GroupingVariable) | is.null(metadata)) {
    
    plot_list <- list()
    
    for (signature in names(ResultsList)) {
      
      df <- ResultsList[[signature]]
      # Wrap the signature name using the helper function
      wrapped_title <- wrap_title(signature, width = widthTitle)
      
      ColorValues <- if (is.null(ColorValues)) "#ECBD78" else ColorValues
      
      p <- ggplot2::ggplot(df, ggplot2::aes(x = score)) +
        ggplot2::geom_density(fill = ColorValues, alpha = 0.5) +
        ggplot2::labs(title = "Density Plot of Score", x = xlab, y = "Density") +
        # add points below density
        ggplot2::geom_rug(ggplot2::aes(x = score), color=ColorValues, sides = "b",  alpha = 0.8, size = .5, length = grid::unit(0.035, "npc"))
      
      # Customize the plot appearance.
      p <- p + ggplot2::theme_classic() +
        ggplot2::labs(title = wrapped_title, color = "", x = "", y = "") +
        ggplot2::theme(axis.text.x = ggplot2::element_blank(),
                       axis.text.y = ggplot2::element_text(  size = labsize - .5),
                       plot.title = ggplot2::element_text(hjust = 0.5, size = titlesize-1),
                       plot.subtitle = ggplot2::element_text(hjust = 0.5, size = titlesize - 1.5, face = "italic"),
                       legend.position="top")+
         guides(color = guide_legend(override.aes = list(size = 6)))
      
      # If limits is specified, crop the plot without adjusting the data (violins).
      if (!is.null(limits)) {
        p <- p + ggplot2::coord_cartesian(xlim = limits)
      }
      
      plot_list[[signature]] <- p
      
    }
    
    n <- length(plot_list)
    
    # Determine grid layout
    if (is.null(ncol) && is.null(nrow)) {
      ncol <- ceiling(sqrt(n))
      nrow <- ceiling(n / ncol)
    } else if (is.null(ncol)) {
      ncol <- ceiling(n / nrow)
    } else if (is.null(nrow)) {
      nrow <- ceiling(n / ncol)
    }
    
    # create label for y axis
    if (method == "ssGSEA") {
      xlab <- "ssGSEA Enrichment Score"
    } else if (method == "logmedian") {
      xlab <- "Normalised Signature Score"
    } else if (method == "ranking") {
      xlab <- "Signature Genes' Ranking"
    }
    
    combined_plot <- ggpubr::ggarrange(plotlist = plot_list, ncol = ncol, nrow = nrow,  align = "h",common.legend = TRUE)
    combined_plot <- ggpubr::annotate_figure(combined_plot,
                                             left = grid::textGrob("Density",
                                                                   rot = 90, vjust = 1, gp = grid::gpar(cex = 1.3, fontsize = labsize)),
                                             bottom = grid::textGrob(xlab, gp = grid::gpar(cex = 1.3, fontsize = labsize)),
                                             top = grid::textGrob(title, gp = grid::gpar(cex = 1.3, fontsize = titlesize + 2)))
    return(combined_plot)
  }
  
  if (!(GroupingVariable %in% colnames(metadata)))
    stop(paste0(GroupingVariable, " not in metadata columns. Please check metadata."))
  
  # Initialize an empty list to store individual ggplot objects.
  plot_list <- list()
  
  # Loop over each gene signature in the ResultsList.
  for (signature in names(ResultsList)) {
    # Extract the data frame for the current signature.
    df <- ResultsList[[signature]]
    
    # Using factors so we can retrieve the first condition for Cohen's d if none is specified.
    df[, GroupingVariable] <- factor(df[, GroupingVariable],
                                     levels = sort(unique(as.character(df[, GroupingVariable]))))
    
    # Wrap the signature name using the helper function.
    wrapped_title <- wrap_title(signature, width = widthTitle)
    
    # Create a base ggplot object with the specified grouping on the x-axis and score on the y-axis.
    p <- ggplot2::ggplot(df, ggplot2::aes_string(x = GroupingVariable, y = "score"))
    
    # Add jittered points, optionally colored by ColorVariable.
    if (!is.null(ColorVariable)) {
      p <- p + ggplot2::geom_jitter(ggplot2::aes_string(color = ColorVariable), size = pointSize, alpha = 0.5)
    } else {
      p <- p + ggplot2::geom_jitter(size = pointSize, alpha = 0.5) +
        ggplot2::scale_color_brewer(palette = colorPalette)
    }
    
    # Overlay violin plots.
    p <- p + ggplot2::geom_violin(alpha = 0.5, scale = "width")
    
    # Add median summary crossbar.
    p <- p + ggplot2::stat_summary(fun = median, fun.min = median, fun.max = median,
                                   geom = "crossbar", width = 0.25,
                                   position = ggplot2::position_dodge(width = 0.13))
    
    # Add stats: Compute Cohen's d (and optionally p‑value)
    if(compute_cohen){
      if (!is.null(cond_cohend)){
        # can be of the following form:
        # cond_cohend <- list(A=c("Senescent"),
        #                     B=c("Proliferative","Quiescent"))
        
        if (sum(unlist(cond_cohend) %in% unique(df[, GroupingVariable])) != length(unique(df[, GroupingVariable])))
          warning("Warning: Not all conditions of GroupingVariable were specified for Cohen's d calculation")
        
        x <- df[df[[GroupingVariable]] %in% cond_cohend[[1]], "score", drop = TRUE]
        y <- df[df[[GroupingVariable]]  %in% cond_cohend[[2]], "score", drop = TRUE]
        
        cohen_d_results <- cohen_d(x, y)
        
        # df$cohen <- ifelse(df[, GroupingVariable] %in% cond_cohend[[1]], names(cond_cohend)[1], names(cond_cohend)[2])
        # cohen_d_results <- rstatix::cohens_d(df, formula = score ~ cohen)
        
        if (pvalcalc) {
          df$cohen <- ifelse(df[, GroupingVariable] %in% cond_cohend[[1]], names(cond_cohend)[1], names(cond_cohend)[2])
          ttest_results <- rstatix::t_test(df, formula = score ~ cohen)
          p_val <- ttest_results$p[1]
          line1 <- wrap_title(paste0("Cohen's d = ", round(cohen_d_results, 3)), width = widthTitle)
          line2 <- wrap_title(paste0("p = ", round(p_val, 3)), width = widthTitle)
          subtitle <- paste(line1, line2, sep = "\n")
        } else {
          subtitle <- wrap_title(paste0("Cohen's d = ", round(cohen_d_results, 3)), width = widthTitle)
        }
        
        
      } else {
        
        if(length(unique(df[, GroupingVariable])) < 2){
          
          warning("Not enough conditions available to report Cohen's d.")
          
        } else if(length(unique(df[, GroupingVariable])) == 2) {
          
          # Calculate Cohen's d based on ordering of the x axis
          group1 <- levels(df[, GroupingVariable])[1]
          group2 <- levels(df[, GroupingVariable])[2]
          
          x <- df[df[[GroupingVariable]] == group1, "score", drop = TRUE]
          y <- df[df[[GroupingVariable]] == group2, "score", drop = TRUE]
          
          cohen_d_results <- cohen_d(x, y)
          
          if (pvalcalc) {
            ttest_results <- rstatix::t_test(df, formula = score ~ GroupingVariable)
            p_val <- ttest_results$p[1]
            line1 <- wrap_title(paste0("Cohen's d = ", round(cohen_d_results$effsize, 3)), width = widthTitle)
            line2 <- wrap_title(paste0("p = ", round(p_val, 3)), width = widthTitle)
            subtitle <- paste(line1, line2, sep = "\n")
          } else {
            subtitle <- wrap_title(paste0("Cohen's d = ", round(cohen_d_results$effsize, 3)), width = widthTitle)
          }
          
          
          
        } else if(length(unique(df[, GroupingVariable])) > 2){
          
          # Calculate Cohen's f
          type <- identify_variable_type(df, GroupingVariable)[GroupingVariable]
          #Without scaling, the coefficient represents the change in score per unit increase in the variable (if numeric, the unit of the variable. Makes sense to not scale...)
          model <- lm(score ~ get(GroupingVariable), data = df)
          results_var <- compute_cohens_f_pval(model, type)
          
          
          if (pvalcalc) {
            line1 <- wrap_title(paste0("Cohen's f = ", round(results_var["Cohen_f"], 3)), width = widthTitle)
            line2 <- wrap_title(paste0("p = ", round(results_var["P_Value"], 3)), width = widthTitle)
            subtitle <- paste(line1, line2, sep = "\n")
          } else {
            subtitle <- wrap_title(paste0("Cohen's f = ", round(results_var["Cohen_f"], 3)), width = widthTitle)
          }
          
        }
        
        
      }
      
    } else {
      
      subtitle <- NULL
      
    }
    # If ConnectGroups is TRUE, add a line connecting medians across groups.
    if (ConnectGroups && !is.null(ColorVariable)) {
      p <- p + ggplot2::stat_summary(ggplot2::aes_string(group = ColorVariable, color = ColorVariable),
                                     fun.y = median, geom = "line", size = 1.5, alpha = 0.75,
                                     show.legend = FALSE)
    }
    
    # Customize the plot appearance.
    p <- p + ggplot2::theme_bw() +
      ggplot2::theme(axis.text.x = ggplot2::element_blank(),
                     axis.text.y = ggplot2::element_text(  size = labsize),
                     plot.title = ggplot2::element_text(hjust = 0.5, size = titlesize-1),
                     plot.subtitle = ggplot2::element_text(hjust = 0.5, size = titlesize - 1.5, face = "italic")) +
      ggplot2::labs(title = wrapped_title, subtitle = subtitle, color = "", x = "", y = "")  
    
    # If ColorValues is provided, use a manual color scale; otherwise, if ColorVariable is provided,
    # use a default brewer palette.
    if (!is.null(ColorValues)) {
      p <- p + ggplot2::scale_color_manual(values = ColorValues)
    } else if (!is.null(ColorVariable)) {
      p <- p + ggplot2::scale_color_brewer(palette = colorPalette)
    }
    
    # If limits is specified, crop the plot without adjusting the data (violins).
    if (!is.null(limits)) {
      p <- p + ggplot2::coord_cartesian(ylim = limits)
    }
    
    # Adjust legend rows if legend_nrow is specified.
    if (!is.null(legend_nrow)) {
      p <- p + ggplot2::guides(color = ggplot2::guide_legend(nrow = legend_nrow, override.aes = list(size = pointSize + 2)))
    }
    
    # Store the plot in the list.
    plot_list[[signature]] <- p + theme(legend.position = "top")
  }
  
  n <- length(plot_list)
  
  # Determine grid layout.
  if (is.null(ncol) && is.null(nrow)) {
    ncol <- ceiling(sqrt(n))
    nrow <- ceiling(n / ncol)
  } else if (is.null(ncol)) {
    ncol <- ceiling(n / nrow)
  } else if (is.null(nrow)) {
    nrow <- ceiling(n / ncol)
  }
  
  # Combine plots.
  combined_plot <- ggpubr::ggarrange(plotlist = plot_list, ncol = ncol, nrow = nrow , align = "h",common.legend = TRUE) #, common.legend = TRUE
  
  # Annotate with axis labels.
  if (is.null(xlab)) {
    xlab <- GroupingVariable
  }
  
  if (!is.null(title)) title <- wrap_title(title, width = widthTitle)
  
  # Create label for y axis based on method.
  if (method == "ssGSEA") {
    ylab <- "ssGSEA Enrichment Score"
  } else if (method == "logmedian") {
    ylab <- "Normalised Signature Score"
  } else if (method == "ranking") {
    ylab <- "Signature Genes' Ranking"
  }
  
  combined_plot <- ggpubr::annotate_figure(combined_plot, 
                                           top = grid::textGrob(title, gp = grid::gpar(cex = 1.3, fontsize = titlesize)))
  return(combined_plot)
}
```


```{r}
# adapted to not have the x axis 

PlotScores_adapted_noX <- function(data, metadata, gene_sets,
                               method = c("ssGSEA", "logmedian", "ranking", "all"),
                               ColorVariable = NULL, Variable = NULL,
                               ColorValues = NULL, ConnectGroups = FALSE, ncol = NULL, nrow = NULL, title = NULL,
                               widthTitle = 20, titlesize = 12, limits = NULL, legend_nrow = NULL, pointSize = 4,
                               xlab = NULL, labsize = 10, compute_cohen=TRUE, cond_cohend = NULL, pvalcalc = FALSE, mode = c("simple","medium","extensive"),
                               widthlegend=22, sig_threshold=0.05, cohen_threshold=0.5, colorPalette="Set3", cor=c("pearson","spearman","kendall")) {
  
  method <- match.arg(method)
  
  type <- identify_variable_type(metadata, Variable)#[Variable]
  
  if (method == "all") { # returns heatmap
    
    if (type =="Numeric"){
      
      cohenlist <- CohenF_allConditions(data = data, metadata = metadata, gene_sets = gene_sets, variable = Variable )
      
    } else {
      
      cohenlist <- CohenD_allConditions(data = data, metadata = metadata, gene_sets = gene_sets, variable = Variable, mode = mode)
      
    }
    
    # if user wants "all" methods, a heatmap of Cohen's d's is returned, for all combination of variables in GroupingVariable
    Heatmap_Final <- Heatmap_Cohen_noX(cohenlist = cohenlist,
                                   nrow = nrow,
                                   ncol = ncol,
                                   limits = limits,
                                   widthTitle = widthTitle,
                                   titlesize = titlesize,
                                   ColorValues = ColorValues,
                                   title = title )
    
    Volcano_Cohen <- Volcano_Cohen(cohenlist = cohenlist,
                                   titlesize = 12,
                                   ColorValues = ColorValues,
                                   title = title,
                                   widthlegend = widthlegend,
                                   pointSize = pointSize,
                                   sig_threshold = sig_threshold,
                                   cohen_threshold = cohen_threshold,
                                   colorPalette =colorPalette,
                                   ncol = ncol,
                                   nrow = nrow)
    
    return(list(heatmap=Heatmap_Final$plt,
                volcano=Volcano_Cohen$plt))
    
  } else {
    
    
    
    if (type!="Numeric"){
      
      return(
        
        PlotScores_Categorical_adapted_noX(data=data, metadata=metadata, gene_sets=gene_sets,
                                       method = method,
                                       ColorVariable = ColorVariable, GroupingVariable = Variable,
                                       ColorValues = ColorValues, ConnectGroups = ConnectGroups, ncol = ncol, nrow = nrow, title = title,
                                       widthTitle = widthTitle, titlesize = titlesize, limits = limits, legend_nrow = legend_nrow, pointSize = pointSize,
                                       xlab = xlab, labsize = labsize, compute_cohen=compute_cohen, cond_cohend = cond_cohend, pvalcalc = pvalcalc, mode = mode,
                                       widthlegend=widthlegend, cohen_threshold=cohen_threshold, colorPalette=colorPalette)
        
      )
      
    } else {
      
      return(
        
        PlotScores_Numeric(data=data,
                           metadata=metadata,
                           gene_sets=gene_sets,
                           method = method,
                           Variable = Variable,
                           ColorValues = ColorValues,
                           ncol = ncol,
                           nrow = nrow,
                           title = title,
                           widthTitle = widthTitle,
                           titlesize = titlesize,
                           limits = limits,
                           pointSize = pointSize,
                           xlab = xlab,
                           labsize = labsize,
                           compute_cohen = compute_cohen,
                           pvalcalc = pvalcalc,
                           colorPalette = colorPalette,
                           cor=cor)
        
      )
      
    }
    
    
  }
  
  
}
```


```{r}
Heatmap_Cohen_noX <- function(cohenlist, nrow = NULL, ncol = NULL, limits = NULL,
                          widthTitle = 22, titlesize = 12, ColorValues = NULL,
                          title=NULL ) {

  cohentype <- ifelse("CohenD" %in% names(cohenlist[[1]]), "d",
                      ifelse("CohenF" %in% names(cohenlist[[1]]), "f", NULL))

  heatmaps <- list()

  if(!is.null(ColorValues)) {
    if (!is.null(ColorValues[["heatmap"]])) {
      ColorValues <- ColorValues[["heatmap"]]
    } else {
      ColorValues <- ColorValues[[1]]
      }
  }

  if (is.null(ColorValues)) ColorValues <- c("#F9F4AE", "#B44141")



  for (signature in names(cohenlist)) {

    if (cohentype=="d"){
      cohen_mat <- t(as.matrix(cohenlist[[signature]]$CohenD))
    } else if (cohentype=="f"){
      cohen_mat <- t(as.matrix(cohenlist[[signature]]$CohenF))
    } else {
      stop("Error: cohenlist format not valid.")
    }

    p_value_mat <- t(as.matrix(cohenlist[[signature]]$padj)) # adjusted p-value

    # Convert to long format manually
    long_data <- data.frame(
      Var1 = rep(rownames(cohen_mat), times = ncol(cohen_mat)),
      Var2 = rep(colnames(cohen_mat), each = nrow(cohen_mat)),
      Cohen = abs(as.vector(cohen_mat)), # absolute value
      PValue = as.vector(p_value_mat),
      stringsAsFactors = FALSE
    )

    # Generate text labels
    long_data$label <- paste0(sprintf("%.2f", long_data$Cohen), "\n(",
                              format.pval(long_data$PValue, digits = 1), ")")

    # Wrap the signature title using an internal helper function
    signature_title <- wrap_title(signature, widthTitle)

    limits <- if (is.null(limits)) c(0 , max(long_data$Cohen, na.rm = TRUE)) else limits

    # Create heatmap using ggplot2
    p <- ggplot2::ggplot(long_data, ggplot2::aes(x = .data$Var2, y = .data$Var1, fill = .data$Cohen)) +
      ggplot2::geom_tile() +
      ggplot2::geom_text(aes(label = .data$label), color = "black", size = 4) +
      ggplot2::scale_fill_gradientn(colors = ColorValues, limits = limits) +
      ggplot2::labs(title = signature_title, x = NULL, y = NULL, fill =
                      ifelse(cohentype=="d", "|Cohen\'s d|", "|Cohen\'s f|")) +
      ggplot2::theme_bw(base_size = 14) +
      ggplot2::theme(axis.text.x = ggplot2::element_blank(),
                     plot.title = ggplot2::element_text(hjust = 0.5, size = titlesize) )

    heatmaps[[signature]] <- p
  }

  # Determine grid layout if not provided
  num_signatures <- length(heatmaps)
  if (is.null(nrow) & is.null(ncol)) {
    ncol <- ceiling(sqrt(num_signatures))
    nrow <- ceiling(num_signatures / ncol)
  } else if (is.null(nrow)) {
    nrow <- ceiling(num_signatures / ncol)
  } else if (is.null(ncol)) {
    ncol <- ceiling(num_signatures / nrow)
  }

  # Adjust axis labels based on grid position
  for (i in seq_along(heatmaps)) {
    row_idx <- ceiling(i / ncol)  # Current row number
    col_idx <- (i - 1) %% ncol + 1  # Current column number

    p <- heatmaps[[i]] +
      theme(
        axis.text.y = if (col_idx == 1) ggplot2::element_text() else ggplot2::element_blank(),
        axis.ticks.y = if (col_idx == 1) ggplot2::element_line() else ggplot2::element_blank(),
        plot.margin = ggplot2::margin(4, 0, 0, 0)  # Adjust plot margins
      )

    heatmaps[[i]] <- p
  }

  # Compute dynamic column widths: first column is wider
  widths <- c(1.5, rep(1, ncol - 1))

  # Combine heatmaps into a single grid plot using ggpubr
  plt <- ggpubr::ggarrange(
    plotlist = heatmaps,
    ncol = ncol,
    nrow = nrow,
    common.legend = TRUE,
    legend = "right",
    align = "h",
    widths = widths
  )

  plt <- ggpubr::annotate_figure(plt, top = grid::textGrob(title,
                                                           gp = grid::gpar(cex = 1.3, fontsize = titlesize+2)))


  return(list(plt = plt, data = cohenlist))
}

```
 
```{r}
# no color legend
Heatmap_Cohen_adapted <- function(cohenlist, nrow = NULL, ncol = NULL, limits = NULL,
                          widthTitle = 22, titlesize = 12, ColorValues = NULL,
                          title=NULL ) {

  cohentype <- ifelse("CohenD" %in% names(cohenlist[[1]]), "d",
                      ifelse("CohenF" %in% names(cohenlist[[1]]), "f", NULL))

  heatmaps <- list()

  if(!is.null(ColorValues)) {
    if (!is.null(ColorValues[["heatmap"]])) {
      ColorValues <- ColorValues[["heatmap"]]
    } else {
      ColorValues <- ColorValues[[1]]
      }
  }

  if (is.null(ColorValues)) ColorValues <- c("#F9F4AE", "#B44141")



  for (signature in names(cohenlist)) {

    if (cohentype=="d"){
      cohen_mat <- t(as.matrix(cohenlist[[signature]]$CohenD))
    } else if (cohentype=="f"){
      cohen_mat <- t(as.matrix(cohenlist[[signature]]$CohenF))
    } else {
      stop("Error: cohenlist format not valid.")
    }

    p_value_mat <- t(as.matrix(cohenlist[[signature]]$padj)) # adjusted p-value

    # Convert to long format manually
    long_data <- data.frame(
      Var1 = rep(rownames(cohen_mat), times = ncol(cohen_mat)),
      Var2 = rep(colnames(cohen_mat), each = nrow(cohen_mat)),
      Cohen = abs(as.vector(cohen_mat)), # absolute value
      PValue = as.vector(p_value_mat),
      stringsAsFactors = FALSE
    )

    # Generate text labels
    long_data$label <- paste0(sprintf("%.2f", long_data$Cohen), "\n(",
                              format.pval(long_data$PValue, digits = 1), ")")

    # Wrap the signature title using an internal helper function
    signature_title <- wrap_title(signature, widthTitle)

    limits <- if (is.null(limits)) c(0 , max(long_data$Cohen, na.rm = TRUE)) else limits

    # Create heatmap using ggplot2
    p <- ggplot2::ggplot(long_data, ggplot2::aes(x = .data$Var2, y = .data$Var1, fill = .data$Cohen)) +
      ggplot2::geom_tile() +
      ggplot2::geom_text(aes(label = .data$label), color = "black", size = 4) +
      ggplot2::scale_fill_gradientn(colors = ColorValues, limits = limits) +
      ggplot2::labs(title = signature_title, x = NULL, y = NULL, fill =
                      ifelse(cohentype=="d", "|Cohen\'s d|", "|Cohen\'s f|")) +
      ggplot2::theme_bw(base_size = 14) +
      ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 45, hjust = 1),
                     plot.title = ggplot2::element_text(hjust = 0.5, size = titlesize),
                     legend.position = "none")

    heatmaps[[signature]] <- p
  }

  # Determine grid layout if not provided
  num_signatures <- length(heatmaps)
  if (is.null(nrow) & is.null(ncol)) {
    ncol <- ceiling(sqrt(num_signatures))
    nrow <- ceiling(num_signatures / ncol)
  } else if (is.null(nrow)) {
    nrow <- ceiling(num_signatures / ncol)
  } else if (is.null(ncol)) {
    ncol <- ceiling(num_signatures / nrow)
  }

  # Adjust axis labels based on grid position
  for (i in seq_along(heatmaps)) {
    row_idx <- ceiling(i / ncol)  # Current row number
    col_idx <- (i - 1) %% ncol + 1  # Current column number

    p <- heatmaps[[i]] +
      theme(
        axis.text.y = if (col_idx == 1) ggplot2::element_text() else ggplot2::element_blank(),
        axis.ticks.y = if (col_idx == 1) ggplot2::element_line() else ggplot2::element_blank(),
        plot.margin = ggplot2::margin(4, 0, 0, 0)  # Adjust plot margins
      )

    heatmaps[[i]] <- p
  }

  # Compute dynamic column widths: first column is wider
  widths <- c(1.5, rep(1, ncol - 1))

  # Combine heatmaps into a single grid plot using ggpubr
  plt <- ggpubr::ggarrange(
    plotlist = heatmaps,
    ncol = ncol,
    nrow = nrow,
    align = "h",
    widths = widths
  )

  plt <- ggpubr::annotate_figure(plt, top = grid::textGrob(title,
                                                           gp = grid::gpar(cex = 1.3, fontsize = titlesize+2)))


  return(list(plt = plt, data = cohenlist))
}

```





```{r}
AUC_Scores_adapted <- function(data, metadata, gene_sets,
                       method = c("logmedian", "ssGSEA", "ranking", "all"),
                       mode = c("simple","medium","extensive"), variable,
                       nrow = NULL, ncol = NULL, limits = NULL, widthTitle = 22,
                       titlesize = 12, ColorValues = c("#F9F4AE", "#B44141"),
                       title = NULL) {
  data <- as.data.frame(data) # Ensure data is a data frame
  # Calculate the AUC scores using the given method
  auc_list <- ROCAUC_Scores_Calculate(data = data, metadata = metadata,
                                      gene_sets = gene_sets, method = method,
                                      variable = variable, mode = mode)

  heatmaps <- list()

  # Loop over the gene signatures
  for (signature_name in names(auc_list[[1]])) {

    # Initialize a matrix to store AUC values for this signature
    auc_matrix <- matrix(nrow = length(auc_list[[1]][[signature_name]]), ncol = length(auc_list))
    rownames(auc_matrix) <- names(auc_list[[1]][[signature_name]])  # Contrasts
    colnames(auc_matrix) <- names(auc_list)  # Methods

    # Fill the matrix with AUC values
    for (method_name in names(auc_list)) {
      for (contrast_name in names(auc_list[[method_name]][[signature_name]])) {
        auc_matrix[contrast_name, method_name] <- auc_list[[method_name]][[signature_name]][[contrast_name]]$AUC
      }
    }

    # Convert the AUC matrix into long format for ggplot
    long_data <- as.data.frame(as.table(auc_matrix))
    colnames(long_data) <- c("Contrast", "Method", "AUC")

    # Generate text labels showing the AUC value
    long_data$label <- sprintf("%.2f", long_data$AUC)

    # Wrap the signature title using an internal helper function
    signature_title <- wrap_title(signature_name, widthTitle)

    limits <- if (is.null(limits)) c(0.5 , 1) else limits

    # Create heatmap using ggplot2
    p <- ggplot2::ggplot(long_data, ggplot2::aes(x = .data$Method, y = .data$Contrast, fill = .data$AUC)) +
      ggplot2::geom_tile() +
      ggplot2::geom_text(aes(label = .data$label), color = "black", size = 4) +
      ggplot2::scale_fill_gradientn(colors = ColorValues, limits = limits) +
      ggplot2::labs(title = signature_title, x = NULL, y = NULL, fill = "AUC") +
      ggplot2::theme_bw(base_size = 14) +
      ggplot2::theme(
        axis.text.x = ggplot2::element_text(angle = 45, hjust = 1),
        plot.title = ggplot2::element_text(hjust = 0.5, size = titlesize),
        legend.position="none"
      )

    heatmaps[[signature_name]] <- p
  }

  # Determine grid layout if not provided
  num_signatures <- length(heatmaps)
  if (is.null(nrow) & is.null(ncol)) {
    ncol <- ceiling(sqrt(num_signatures))
    nrow <- ceiling(num_signatures / ncol)
  } else if (is.null(nrow)) {
    nrow <- ceiling(num_signatures / ncol)
  } else if (is.null(ncol)) {
    ncol <- ceiling(num_signatures / nrow)
  }


  # Adjust axis labels based on grid position
  for (i in seq_along(heatmaps)) {
    row_idx <- ceiling(i / ncol)  # Current row number
    col_idx <- (i - 1) %% ncol + 1  # Current column number

    p <- heatmaps[[i]] +
      theme(
        axis.text.y = if (col_idx == 1) ggplot2::element_text() else ggplot2::element_blank(),
        axis.ticks.y = if (col_idx == 1) ggplot2::element_line() else ggplot2::element_blank(),
        plot.margin = ggplot2::margin(4, 0, 0, 0)  # Adjust plot margins
      )

    heatmaps[[i]] <- p
  }

  # Compute dynamic column widths: first column is wider
  widths <- c(1.5, rep(1, ncol - 1))


  # Combine heatmaps into a single grid plot using ggpubr
  plt <- ggpubr::ggarrange(
    plotlist = heatmaps,
    ncol = ncol,
    nrow = nrow, 
    align = "h",
    widths = widths
  )

  # Add the title to the grid if provided
  if (!is.null(title)) {
    plt <- ggpubr::annotate_figure(plt, top = grid::textGrob(title, gp = grid::gpar(cex = 1.3, fontsize = titlesize + 2)))
  }

  return(plt)
}


```


```{r}
AUC_Scores_noX  <- function(data, metadata, gene_sets,
                       method = c("logmedian", "ssGSEA", "ranking", "all"),
                       mode = c("simple","medium","extensive"), variable,
                       nrow = NULL, ncol = NULL, limits = NULL, widthTitle = 22,
                       titlesize = 12, ColorValues = c("#F9F4AE", "#B44141"),
                       title = NULL) {
  data <- as.data.frame(data) # Ensure data is a data frame
  # Calculate the AUC scores using the given method
  auc_list <- ROCAUC_Scores_Calculate(data = data, metadata = metadata,
                                      gene_sets = gene_sets, method = method,
                                      variable = variable, mode = mode)

  heatmaps <- list()

  # Loop over the gene signatures
  for (signature_name in names(auc_list[[1]])) {

    # Initialize a matrix to store AUC values for this signature
    auc_matrix <- matrix(nrow = length(auc_list[[1]][[signature_name]]), ncol = length(auc_list))
    rownames(auc_matrix) <- names(auc_list[[1]][[signature_name]])  # Contrasts
    colnames(auc_matrix) <- names(auc_list)  # Methods

    # Fill the matrix with AUC values
    for (method_name in names(auc_list)) {
      for (contrast_name in names(auc_list[[method_name]][[signature_name]])) {
        auc_matrix[contrast_name, method_name] <- auc_list[[method_name]][[signature_name]][[contrast_name]]$AUC
      }
    }

    # Convert the AUC matrix into long format for ggplot
    long_data <- as.data.frame(as.table(auc_matrix))
    colnames(long_data) <- c("Contrast", "Method", "AUC")

    # Generate text labels showing the AUC value
    long_data$label <- sprintf("%.2f", long_data$AUC)

    # Wrap the signature title using an internal helper function
    signature_title <- wrap_title(signature_name, widthTitle)

    limits <- if (is.null(limits)) c(0.5 , 1) else limits

    # Create heatmap using ggplot2
    p <- ggplot2::ggplot(long_data, ggplot2::aes(x = .data$Method, y = .data$Contrast, fill = .data$AUC)) +
      ggplot2::geom_tile() +
      ggplot2::geom_text(aes(label = .data$label), color = "black", size = 4) +
      ggplot2::scale_fill_gradientn(colors = ColorValues, limits = limits) +
      ggplot2::labs(title = signature_title, x = NULL, y = NULL, fill = "AUC") +
      ggplot2::theme_bw(base_size = 14) +
      ggplot2::theme(
        axis.text.x = element_blank(),
        plot.title = ggplot2::element_text(hjust = 0.5, size = titlesize) 
                        
      )

    heatmaps[[signature_name]] <- p
  }

  # Determine grid layout if not provided
  num_signatures <- length(heatmaps)
  if (is.null(nrow) & is.null(ncol)) {
    ncol <- ceiling(sqrt(num_signatures))
    nrow <- ceiling(num_signatures / ncol)
  } else if (is.null(nrow)) {
    nrow <- ceiling(num_signatures / ncol)
  } else if (is.null(ncol)) {
    ncol <- ceiling(num_signatures / nrow)
  }


  # Adjust axis labels based on grid position
  for (i in seq_along(heatmaps)) {
    row_idx <- ceiling(i / ncol)  # Current row number
    col_idx <- (i - 1) %% ncol + 1  # Current column number

    p <- heatmaps[[i]] +
      theme(
        axis.text.y = if (col_idx == 1) ggplot2::element_text() else ggplot2::element_blank(),
        axis.ticks.y = if (col_idx == 1) ggplot2::element_line() else ggplot2::element_blank(),
        plot.margin = ggplot2::margin(4, 0, 0, 0)  # Adjust plot margins
      )

    heatmaps[[i]] <- p
  }

  # Compute dynamic column widths: first column is wider
  widths <- c(1.5, rep(1, ncol - 1))


  # Combine heatmaps into a single grid plot using ggpubr
  plt <- ggpubr::ggarrange(
    plotlist = heatmaps,
    ncol = ncol,
    nrow = nrow,
    common.legend = TRUE,
    legend = "right",
    align = "h",
    widths = widths
  )

  # Add the title to the grid if provided
  if (!is.null(title)) {
    plt <- ggpubr::annotate_figure(plt, top = grid::textGrob(title, gp = grid::gpar(cex = 1.3, fontsize = titlesize + 2)))
  }

  return(plt)
}
```

 

```{r}
PlotScores_Categorical_adapted_noX_noleg <- function(data, metadata, gene_sets,
                                           method = c("ssGSEA", "logmedian", "ranking"),
                                           ColorVariable = NULL, GroupingVariable = NULL,
                                           ColorValues = NULL, ConnectGroups = FALSE, ncol = NULL, nrow = NULL, title = NULL,
                                           widthTitle = 10, titlesize = 12, limits = NULL, legend_nrow = NULL, pointSize = 2,
                                           xlab = NULL, labsize = 10, compute_cohen=TRUE, cond_cohend = NULL, pvalcalc = FALSE, mode = c("simple","medium","extensive"),
                                           widthlegend=22, cohen_threshold=0.6, colorPalette="Set3") {
  
  method <- match.arg(method)
  
  ResultsList <- CalculateScores(data = data,
                                 metadata = metadata,
                                 gene_sets = gene_sets,
                                 method = method)
  
  # if grouping variable is NULL, then the function displays a density / distribution of scores
  if (is.null(GroupingVariable) | is.null(metadata)) {
    
    plot_list <- list()
    
    for (signature in names(ResultsList)) {
      
      df <- ResultsList[[signature]]
      # Wrap the signature name using the helper function
      wrapped_title <- wrap_title(signature, width = widthTitle)
      
      ColorValues <- if (is.null(ColorValues)) "#ECBD78" else ColorValues
      
      p <- ggplot2::ggplot(df, ggplot2::aes(x = score)) +
        ggplot2::geom_density(fill = ColorValues, alpha = 0.5) +
        ggplot2::labs(title = "Density Plot of Score", x = xlab, y = "Density") +
        # add points below density
        ggplot2::geom_rug(ggplot2::aes(x = score), color=ColorValues, sides = "b",  alpha = 0.8, size = .5, length = grid::unit(0.035, "npc"))
      
      # Customize the plot appearance.
      p <- p + ggplot2::theme_classic() +
        ggplot2::labs(title = wrapped_title, color = "", x = "", y = "") +
        ggplot2::theme(axis.text.x = ggplot2::element_blank(),
                       axis.text.y = ggplot2::element_text(  size = labsize - .5),
                       plot.title = ggplot2::element_text(hjust = 0.5, size = titlesize-1),
                       plot.subtitle = ggplot2::element_text(hjust = 0.5, size = titlesize - 1.5, face = "italic"),
                       legend.position="none")
      
      # If limits is specified, crop the plot without adjusting the data (violins).
      if (!is.null(limits)) {
        p <- p + ggplot2::coord_cartesian(xlim = limits)
      }
      
      plot_list[[signature]] <- p 
      
    }
    
    n <- length(plot_list)
    
    # Determine grid layout
    if (is.null(ncol) && is.null(nrow)) {
      ncol <- ceiling(sqrt(n))
      nrow <- ceiling(n / ncol)
    } else if (is.null(ncol)) {
      ncol <- ceiling(n / nrow)
    } else if (is.null(nrow)) {
      nrow <- ceiling(n / ncol)
    }
    
    # create label for y axis
    if (method == "ssGSEA") {
      xlab <- "ssGSEA Enrichment Score"
    } else if (method == "logmedian") {
      xlab <- "Normalised Signature Score"
    } else if (method == "ranking") {
      xlab <- "Signature Genes' Ranking"
    }
    
    combined_plot <- ggpubr::ggarrange(plotlist = plot_list, ncol = ncol, nrow = nrow,  align = "h" )
    combined_plot <- ggpubr::annotate_figure(combined_plot,
                                             left = grid::textGrob("Density",
                                                                   rot = 90, vjust = 1, gp = grid::gpar(cex = 1.3, fontsize = labsize)),
                                             bottom = grid::textGrob(xlab, gp = grid::gpar(cex = 1.3, fontsize = labsize)),
                                             top = grid::textGrob(title, gp = grid::gpar(cex = 1.3, fontsize = titlesize + 2)))
    return(combined_plot)
  }
  
  if (!(GroupingVariable %in% colnames(metadata)))
    stop(paste0(GroupingVariable, " not in metadata columns. Please check metadata."))
  
  # Initialize an empty list to store individual ggplot objects.
  plot_list <- list()
  
  # Loop over each gene signature in the ResultsList.
  for (signature in names(ResultsList)) {
    # Extract the data frame for the current signature.
    df <- ResultsList[[signature]]
    
    # Using factors so we can retrieve the first condition for Cohen's d if none is specified.
    df[, GroupingVariable] <- factor(df[, GroupingVariable],
                                     levels = sort(unique(as.character(df[, GroupingVariable]))))
    
    # Wrap the signature name using the helper function.
    wrapped_title <- wrap_title(signature, width = widthTitle)
    
    # Create a base ggplot object with the specified grouping on the x-axis and score on the y-axis.
    p <- ggplot2::ggplot(df, ggplot2::aes_string(x = GroupingVariable, y = "score"))
    
    # Add jittered points, optionally colored by ColorVariable.
    if (!is.null(ColorVariable)) {
      p <- p + ggplot2::geom_jitter(ggplot2::aes_string(color = ColorVariable), size = pointSize, alpha = 0.5)
    } else {
      p <- p + ggplot2::geom_jitter(size = pointSize, alpha = 0.5) +
        ggplot2::scale_color_brewer(palette = colorPalette)
    }
    
    # Overlay violin plots.
    p <- p + ggplot2::geom_violin(alpha = 0.5, scale = "width")
    
    # Add median summary crossbar.
    p <- p + ggplot2::stat_summary(fun = median, fun.min = median, fun.max = median,
                                   geom = "crossbar", width = 0.25,
                                   position = ggplot2::position_dodge(width = 0.13))
    
    # Add stats: Compute Cohen's d (and optionally p‑value)
    if(compute_cohen){
      if (!is.null(cond_cohend)){
        # can be of the following form:
        # cond_cohend <- list(A=c("Senescent"),
        #                     B=c("Proliferative","Quiescent"))
        
        if (sum(unlist(cond_cohend) %in% unique(df[, GroupingVariable])) != length(unique(df[, GroupingVariable])))
          warning("Warning: Not all conditions of GroupingVariable were specified for Cohen's d calculation")
        
        x <- df[df[[GroupingVariable]] %in% cond_cohend[[1]], "score", drop = TRUE]
        y <- df[df[[GroupingVariable]]  %in% cond_cohend[[2]], "score", drop = TRUE]
        
        cohen_d_results <- cohen_d(x, y)
        
        # df$cohen <- ifelse(df[, GroupingVariable] %in% cond_cohend[[1]], names(cond_cohend)[1], names(cond_cohend)[2])
        # cohen_d_results <- rstatix::cohens_d(df, formula = score ~ cohen)
        
        if (pvalcalc) {
          df$cohen <- ifelse(df[, GroupingVariable] %in% cond_cohend[[1]], names(cond_cohend)[1], names(cond_cohend)[2])
          ttest_results <- rstatix::t_test(df, formula = score ~ cohen)
          p_val <- ttest_results$p[1]
          line1 <- wrap_title(paste0("Cohen's d = ", round(cohen_d_results, 3)), width = widthTitle)
          line2 <- wrap_title(paste0("p = ", round(p_val, 3)), width = widthTitle)
          subtitle <- paste(line1, line2, sep = "\n")
        } else {
          subtitle <- wrap_title(paste0("Cohen's d = ", round(cohen_d_results, 3)), width = widthTitle)
        }
        
        
      } else {
        
        if(length(unique(df[, GroupingVariable])) < 2){
          
          warning("Not enough conditions available to report Cohen's d.")
          
        } else if(length(unique(df[, GroupingVariable])) == 2) {
          
          # Calculate Cohen's d based on ordering of the x axis
          group1 <- levels(df[, GroupingVariable])[1]
          group2 <- levels(df[, GroupingVariable])[2]
          
          x <- df[df[[GroupingVariable]] == group1, "score", drop = TRUE]
          y <- df[df[[GroupingVariable]] == group2, "score", drop = TRUE]
          
          cohen_d_results <- cohen_d(x, y)
          
          if (pvalcalc) {
            ttest_results <- rstatix::t_test(df, formula = score ~ GroupingVariable)
            p_val <- ttest_results$p[1]
            line1 <- wrap_title(paste0("Cohen's d = ", round(cohen_d_results$effsize, 3)), width = widthTitle)
            line2 <- wrap_title(paste0("p = ", round(p_val, 3)), width = widthTitle)
            subtitle <- paste(line1, line2, sep = "\n")
          } else {
            subtitle <- wrap_title(paste0("Cohen's d = ", round(cohen_d_results$effsize, 3)), width = widthTitle)
          }
          
          
          
        } else if(length(unique(df[, GroupingVariable])) > 2){
          
          # Calculate Cohen's f
          type <- identify_variable_type(df, GroupingVariable)[GroupingVariable]
          #Without scaling, the coefficient represents the change in score per unit increase in the variable (if numeric, the unit of the variable. Makes sense to not scale...)
          model <- lm(score ~ get(GroupingVariable), data = df)
          results_var <- compute_cohens_f_pval(model, type)
          
          
          if (pvalcalc) {
            line1 <- wrap_title(paste0("Cohen's f = ", round(results_var["Cohen_f"], 3)), width = widthTitle)
            line2 <- wrap_title(paste0("p = ", round(results_var["P_Value"], 3)), width = widthTitle)
            subtitle <- paste(line1, line2, sep = "\n")
          } else {
            subtitle <- wrap_title(paste0("Cohen's f = ", round(results_var["Cohen_f"], 3)), width = widthTitle)
          }
          
        }
        
        
      }
      
    } else {
      
      subtitle <- NULL
      
    }
    # If ConnectGroups is TRUE, add a line connecting medians across groups.
    if (ConnectGroups && !is.null(ColorVariable)) {
      p <- p + ggplot2::stat_summary(ggplot2::aes_string(group = ColorVariable, color = ColorVariable),
                                     fun.y = median, geom = "line", size = 1.5, alpha = 0.75,
                                     show.legend = FALSE)
    }
    
    # Customize the plot appearance.
    p <- p + ggplot2::theme_bw() +
      ggplot2::theme(axis.text.x = ggplot2::element_blank(),
                     axis.text.y = ggplot2::element_text(  size = labsize),
                     plot.title = ggplot2::element_text(hjust = 0.5, size = titlesize-1),
                     plot.subtitle = ggplot2::element_text(hjust = 0.5, size = titlesize - 1.5, face = "italic"),
                     legend.position = "none") +
      ggplot2::labs(title = wrapped_title, subtitle = subtitle,x = "", y = "")  
    
    # If ColorValues is provided, use a manual color scale; otherwise, if ColorVariable is provided,
    # use a default brewer palette.
    if (!is.null(ColorValues)) {
      p <- p + ggplot2::scale_color_manual(values = ColorValues)
    } else if (!is.null(ColorVariable)) {
      p <- p + ggplot2::scale_color_brewer(palette = colorPalette)
    }
    
    # If limits is specified, crop the plot without adjusting the data (violins).
    if (!is.null(limits)) {
      p <- p + ggplot2::coord_cartesian(ylim = limits)
    }
    
    # Adjust legend rows if legend_nrow is specified.
    if (!is.null(legend_nrow)) {
      p <- p #+ ggplot2::guides(color = ggplot2::guide_legend(nrow = legend_nrow, override.aes = list(size = pointSize + 2)))
    }
    
    # Store the plot in the list.
    plot_list[[signature]] <- p   
  }
  
  n <- length(plot_list)
  
  # Determine grid layout.
  if (is.null(ncol) && is.null(nrow)) {
    ncol <- ceiling(sqrt(n))
    nrow <- ceiling(n / ncol)
  } else if (is.null(ncol)) {
    ncol <- ceiling(n / nrow)
  } else if (is.null(nrow)) {
    nrow <- ceiling(n / ncol)
  }
  
  # Combine plots.
  combined_plot <- ggpubr::ggarrange(plotlist = plot_list, ncol = ncol, nrow = nrow , align = "h") #, common.legend = TRUE
  
  # Annotate with axis labels.
  if (is.null(xlab)) {
    xlab <- GroupingVariable
  }
  
  if (!is.null(title)) title <- wrap_title(title, width = widthTitle)
  
  # Create label for y axis based on method.
  if (method == "ssGSEA") {
    ylab <- "ssGSEA Enrichment Score"
  } else if (method == "logmedian") {
    ylab <- "Normalised Signature Score"
  } else if (method == "ranking") {
    ylab <- "Signature Genes' Ranking"
  }
  
  combined_plot <- ggpubr::annotate_figure(combined_plot, 
                                           top = grid::textGrob(title, gp = grid::gpar(cex = 1.3, fontsize = titlesize)))
  return(combined_plot)
}
```


```{r}
# adapted to not have the color legend

PlotScores_adapted_noX_noleg <- function(data, metadata, gene_sets,
                               method = c("ssGSEA", "logmedian", "ranking", "all"),
                               ColorVariable = NULL, Variable = NULL,
                               ColorValues = NULL, ConnectGroups = FALSE, ncol = NULL, nrow = NULL, title = NULL,
                               widthTitle = 20, titlesize = 12, limits = NULL, legend_nrow = NULL, pointSize = 4,
                               xlab = NULL, labsize = 10, compute_cohen=TRUE, cond_cohend = NULL, pvalcalc = FALSE, mode = c("simple","medium","extensive"),
                               widthlegend=22, sig_threshold=0.05, cohen_threshold=0.5, colorPalette="Set3", cor=c("pearson","spearman","kendall")) {
  
  method <- match.arg(method)
  
  type <- identify_variable_type(metadata, Variable)#[Variable]
  
  if (method == "all") { # returns heatmap
    
    if (type =="Numeric"){
      
      cohenlist <- CohenF_allConditions(data = data, metadata = metadata, gene_sets = gene_sets, variable = Variable )
      
    } else {
      
      cohenlist <- CohenD_allConditions(data = data, metadata = metadata, gene_sets = gene_sets, variable = Variable, mode = mode)
      
    }
    
    # if user wants "all" methods, a heatmap of Cohen's d's is returned, for all combination of variables in GroupingVariable
    Heatmap_Final <- Heatmap_Cohen(cohenlist = cohenlist,
                                   nrow = nrow,
                                   ncol = ncol,
                                   limits = limits,
                                   widthTitle = widthTitle,
                                   titlesize = titlesize,
                                   ColorValues = ColorValues,
                                   title = title )
    
    Volcano_Cohen <- Volcano_Cohen(cohenlist = cohenlist,
                                   titlesize = 12,
                                   ColorValues = ColorValues,
                                   title = title,
                                   widthlegend = widthlegend,
                                   pointSize = pointSize,
                                   sig_threshold = sig_threshold,
                                   cohen_threshold = cohen_threshold,
                                   colorPalette =colorPalette,
                                   ncol = ncol,
                                   nrow = nrow)
    
    return(list(heatmap=Heatmap_Final$plt,
                volcano=Volcano_Cohen$plt))
    
  } else {
    
    
    
    if (type!="Numeric"){
      
      return(
        
        PlotScores_Categorical_adapted_noX_noleg(data=data, metadata=metadata, gene_sets=gene_sets,
                                       method = method,
                                       ColorVariable = ColorVariable, GroupingVariable = Variable,
                                       ColorValues = ColorValues, ConnectGroups = ConnectGroups, ncol = ncol, nrow = nrow, title = title,
                                       widthTitle = widthTitle, titlesize = titlesize, limits = limits, legend_nrow = legend_nrow, pointSize = pointSize,
                                       xlab = xlab, labsize = labsize, compute_cohen=compute_cohen, cond_cohend = cond_cohend, pvalcalc = pvalcalc, mode = mode,
                                       widthlegend=widthlegend, cohen_threshold=cohen_threshold, colorPalette=colorPalette)
        
      )
      
    } else {
      
      return(
        
        PlotScores_Numeric(data=data,
                           metadata=metadata,
                           gene_sets=gene_sets,
                           method = method,
                           Variable = Variable,
                           ColorValues = ColorValues,
                           ncol = ncol,
                           nrow = nrow,
                           title = title,
                           widthTitle = widthTitle,
                           titlesize = titlesize,
                           limits = limits,
                           pointSize = pointSize,
                           xlab = xlab,
                           labsize = labsize,
                           compute_cohen = compute_cohen,
                           pvalcalc = pvalcalc,
                           colorPalette = colorPalette,
                           cor=cor)
        
      )
      
    }
    
    
  }
  
  
}
```

 
### Load relevant data

```{r}
metadata <- readRDS("../data_aux/metadata.rds")
counts <- readRDS("../data_aux/counts.rds")
filtered_counts <- readRDS("../data_aux/filtered_counts.rds")
normalised_counts <- readRDS("../data_aux/normalised_counts.rds")
corrcounts <- readRDS("../data_aux/corrcounts.rds")
signatures_bidirectional <- readRDS("../data_aux/SenescenceSigntures_Bidirectional.rds")


GTEx_alltissues <- readRDS("../data_aux/GTExV8_voyagercorrected.rds") # https://github.com/DiseaseTranscriptomicsLab/voyAGEr/tree/main/Corrected_Counts
metadata_GTEx_alltissues <- readRDS("../data_aux/GTExV8_metadata.rds") # Restricted access
``` 

### Colors

```{r}
Condition_colors <- c(
  "Senescent"     = "#F4A261",  # Soft orange (not goldenrod)
  "Proliferative" = "#A1C298",  # Desaturated sage green
  "Quiescent"     = "#9DB4C0"   # Dusty teal-gray
)

CellTypecols <- c(
  "Fibroblast"   = "#E69F00",  # Orange
  "Keratinocyte" = "#56B4E9",  # Sky Blue
  "Melanocyte"   = "#009E73",  # Bluish Green
  "Endothelial"  = "#F0E442",  # Yellow
  "Neuronal"     = "#0072B2",  # Blue
  "Mesenchymal"  = "#D55E00"   # Vermilion
)

CellTypecols_alt <- c(
  "Fibroblast"   = "#1b9e77",  # Teal green
  "Keratinocyte" = "#d95f02",  # Warm orange-brown
  "Melanocyte"   = "#7570b3",  # Muted purple
  "Endothelial"  = "#e7298a",  # Dark pink
  "Neuronal"     = "#66a61e",  # Olive green
  "Mesenchymal"  = "#e6ab02"   # Mustard
)



dataset_colors <- c(
  "HernandezSegura" = "#8DD3C7",
  "Mangelinck"      = "#FFFFB3",
  "Chan"            = "#BEBADA",
  "Purcell"         = "#FB8072",
  "Marthandan2016"  = "#80B1D3",
  "Marthandan2015"  = "#FDB462",
  "Wang"            = "#B3DE69",
  "Suda"            = "#FCCDE5",
  "McHugh"          = "#D9D9D9",
  "Savic"           = "#BC80BD",
  "Skea"            = "#CCEBC5",
  "Laurent"         = "#FFED6F",
  "Tasdemir"        = "#D0B7E1",
  "Lenain"          = "#FFB3BA",
  "Aarts"           = "#C2C2F0",
  "Casella"         = "#FFDAC1",
  "Numa"            = "#C6E2FF",
  "Admasu"          = "#F5CBA7",
  "Urata"           = "#C7CEEA",
  "Wang2023"        = "#E6B0AA",
  "Fu"              = "#A9DFBF",
  "Tanke"           = "#F9E79F",
  "Bawens"          = "#AED6F1"
)


```

```{r}
show_col(Condition_colors, ncol=3)
show_col(CellTypecols_alt)
show_col(dataset_colors)
```
 
## Supplementary Figure 1 - Senescence Datasets Processing

Filtering lowly expressed genes

 
```{r fig.width=4, fig.height=4}
# Average expression per gene per condition
avg_expr_quiesc <- log10(rowMeans(counts[, metadata$Condition == "Quiescent"] + 1))
avg_expr_prol   <- log10(rowMeans(counts[, metadata$Condition == "Proliferative"] + 1))
avg_expr_sen    <- log10(rowMeans(counts[, metadata$Condition == "Senescent"] + 1))

# Combine into dataframe
df_avg_expr <- data.frame(expression = c(avg_expr_quiesc, avg_expr_prol, avg_expr_sen),
                          Condition = rep(c("Quiescent", "Proliferative", "Senescent"),
                                          each = nrow(counts)))

# Plot
(plt_beforefilt <- ggplot(df_avg_expr, aes(x = expression, fill = Condition)) +
  geom_density(alpha = 0.7) +
  theme_bw(base_size = 14) +
  xlab("Mean log10(counts + 1) per gene") +
  ylab("Density") +
  geom_vline(xintercept = log10(70), linetype = "dotted", color = "black", size = 1) +
  scale_fill_manual(values = Condition_colors) +
  theme(legend.position="bottom", legend.title = element_blank())+ theme( legend.text.align = 0, 
         legend.key.width  = unit(0.8, "cm") ))
 
# Average expression per gene per condition
avg_expr_quiesc <- log10(rowMeans(filtered_counts[, metadata$Condition == "Quiescent"] + 1))
avg_expr_prol   <- log10(rowMeans(filtered_counts[, metadata$Condition == "Proliferative"] + 1))
avg_expr_sen    <- log10(rowMeans(filtered_counts[, metadata$Condition == "Senescent"] + 1))

# Combine into dataframe
df_avg_expr <- data.frame(expression = c(avg_expr_quiesc, avg_expr_prol, avg_expr_sen),
                          Condition = rep(c("Quiescent", "Proliferative", "Senescent"),
                                          each = nrow(filtered_counts)))

# Plot
(plt_afterfilt <- ggplot(df_avg_expr, aes(x = expression, fill = Condition)) +
  geom_density(alpha = 0.7) +
  theme_bw(base_size = 14) +
  xlab("Mean log10(counts + 1) per gene") +
  ylab("Density") +
  geom_vline(xintercept = log10(70), linetype = "dotted", color = "black", size = 1) +
  scale_fill_manual(values = Condition_colors) +
  theme(legend.position="bottom", legend.title = element_blank() )+ theme( legend.text.align = 0, 
         legend.key.width  = unit(0.8, "cm") ))

```


```{r fig.width=4, fig.height=6}
# (filtplot <- ggarrange(plt_beforefilt,plt_afterfilt, ncol=1, common.legend = T))
```

Normalisation
 
```{r fig.width=10, fig.height=4}
samples_ordered <- colnames(filtered_counts)[order(colSums(filtered_counts))]
N <- 30

filtcounts_ggplot <- stack(log2(filtered_counts[,c(samples_ordered[1:N], samples_ordered[(length(samples_ordered)-N +1):length(samples_ordered)])]))
colnames(filtcounts_ggplot) <- c("expression", "sampleID")
filtcounts_ggplot <- merge(filtcounts_ggplot,metadata, by="sampleID")
  

(plt_beforenorm <- ggplot(filtcounts_ggplot, aes(sampleID,expression)) + 
    geom_boxplot(aes(fill=DatasetID))+ 
    xlab("Samples") + 
    ylab("Gene Expression \n(log2(counts))")    +  
    theme(axis.text.x = element_text(angle = 45, hjust=1))+
  theme_bw(base_size = 14)+ 
  theme(plot.title = element_text(hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank(), legend.title = element_blank())  + 
    scale_fill_manual(values = dataset_colors))
 

normcounts_ggplot <- stack(log2(normalised_counts[,c(samples_ordered[1:N], samples_ordered[(length(samples_ordered)-N +1):length(samples_ordered)])]))
colnames(normcounts_ggplot) <- c("expression", "sampleID")
normcounts_ggplot <- merge(normcounts_ggplot,metadata, by="sampleID")
  

(plt_afternorm <- ggplot(normcounts_ggplot, aes(sampleID,expression)) + 
    geom_boxplot(aes(fill=DatasetID))+ 
    xlab("Samples") + 
    ylab("Gene Expression \n(log2(CPM))")   +  
    theme(axis.text.x = element_text(angle = 45, hjust=1))+
  theme_bw(base_size = 14)+ 
  theme(plot.title = element_text(hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank(), legend.title = element_blank()) + 
    scale_fill_manual(values = dataset_colors))

```


```{r fig.width=8, fig.height=5}
#(normplot <- ggarrange(plt_beforenorm,plt_afternorm,ncol=1, common.legend = T, legend="top"))
```


Batch effect correction

(before)


```{r fig.width=6, fig.height=5}
y <- DGEList(log2(normalised_counts+1), samples= metadata)
PCA_beforeBC  <- prcomp(t(y$counts), scale=FALSE, center=TRUE)
PCACounts_beforeBC  <- PCA_beforeBC$x
PCACounts_beforeBC  <- as.data.frame(PCACounts_beforeBC )
PCAData_beforeBC <-  cbind(PCACounts_beforeBC[1:10],y$samples) 

ev = PCA_beforeBC$sdev^2 
pc1= round(100*ev[1]/sum(ev),2) 
pc2 = round(100*ev[2]/sum(ev),2)  

(PCA_beforeBEC_Dataset <- ggplot(PCAData_beforeBC, aes(PC1,PC2))+
  geom_point(aes(fill=DatasetID), size=5, alpha=0.6, shape=21)+ 
  xlab(paste0("PC1: ",pc1,"% variance")) +
  ylab(paste0("PC2: ",pc2,"% variance")) + 
  theme_bw(base_size = 14)+
  theme(legend.position="bottom", legend.box="vertical", legend.margin=margin())  + 
  geom_vline(xintercept=0, linetype="dotted") + 
  geom_hline(yintercept=0, linetype="dotted") + 
    scale_fill_manual(values = dataset_colors) +
  theme( legend.title = element_blank())+ guides(fill=guide_legend(ncol=2,byrow=TRUE))+ theme( legend.text.align = 0, 
         legend.key.width  = unit(0.8, "cm") ))

(PCA_beforeBEC_Conditon <- ggplot(PCAData_beforeBC, aes(PC1,PC2))+
  geom_point(aes(fill=Condition), size=5, alpha=0.6, shape=21)+ 
  xlab(paste0("PC1: ",pc1,"% variance")) +
  ylab(paste0("PC2: ",pc2,"% variance")) + 
  theme_bw(base_size = 14)+
  theme(legend.position="bottom", legend.box="vertical", legend.margin=margin())  + 
  geom_vline(xintercept=0, linetype="dotted") + 
  geom_hline(yintercept=0, linetype="dotted") + 
    scale_fill_manual(values = Condition_colors) +
  theme( legend.title = element_blank())+ theme( legend.text.align = 0, 
         legend.key.width  = unit(0.8, "cm") ))

(PCA_beforeBEC_CellType <- ggplot(PCAData_beforeBC, aes(PC1,PC2))+
  geom_point(aes(fill=CellType), size=5, alpha=0.6, shape=21)+ 
  xlab(paste0("PC1: ",pc1,"% variance")) +
  ylab(paste0("PC2: ",pc2,"% variance")) + 
  theme_bw(base_size = 14)+
  theme(legend.position="bottom", legend.box="vertical", legend.margin=margin())  + 
  geom_vline(xintercept=0, linetype="dotted") + 
  geom_hline(yintercept=0, linetype="dotted") + 
    scale_fill_manual(values = CellTypecols_alt) +
  theme( legend.title = element_blank())+
      theme(
    legend.text.align = 0,                   # left-align text 
    legend.key.width  = unit(0.8, "cm")      # keep box size compact
  ))
 


 
```

Batch effect correction

(after)



```{r fig.width=6, fig.height=5}
y <- DGEList(log2(corrcounts + 1), samples= metadata)
PCA_afterBC  <- prcomp(t(y$counts), scale=FALSE, center=TRUE)
PCACounts_afterBC  <- PCA_afterBC$x
PCACounts_afterBC  <- as.data.frame(PCACounts_afterBC )
PCAData_afterBC <-  cbind(PCACounts_afterBC[1:10],y$samples) 

ev = PCA_afterBC$sdev^2 
pc1= round(100*ev[1]/sum(ev),2) 
pc2 = round(100*ev[2]/sum(ev),2)  

(PCA_afterBEC_Dataset <- ggplot(PCAData_afterBC, aes(PC1,PC2))+
  geom_point(aes(fill=DatasetID), size=5, alpha=0.6, shape=21)+ 
  xlab(paste0("PC1: ",pc1,"% variance")) +
  ylab(paste0("PC2: ",pc2,"% variance")) + 
  theme_bw(base_size = 14)+
  theme(legend.position="bottom", legend.box="vertical", legend.margin=margin())  + 
  geom_vline(xintercept=0, linetype="dotted") + 
  geom_hline(yintercept=0, linetype="dotted") + 
    scale_fill_manual(values = dataset_colors) +
  theme( legend.title = element_blank())+ guides(fill=guide_legend(ncol=3,byrow=TRUE))+ theme( legend.text.align = 0, 
         legend.key.width  = unit(0.8, "cm") ))

(PCA_afterBEC_Conditon <- ggplot(PCAData_afterBC, aes(PC1,PC2))+
  geom_point(aes(fill=Condition), size=5, alpha=0.6, shape=21)+ 
  xlab(paste0("PC1: ",pc1,"% variance")) +
  ylab(paste0("PC2: ",pc2,"% variance")) + 
  theme_bw(base_size = 14)+
  theme(legend.position="bottom", legend.box="vertical", legend.margin=margin())  + 
  geom_vline(xintercept=0, linetype="dotted") + 
  geom_hline(yintercept=0, linetype="dotted") + 
    scale_fill_manual(values = Condition_colors) +
  theme( legend.title = element_blank())+ theme( legend.text.align = 0, 
         legend.key.width  = unit(0.8, "cm") ))

(PCA_afterBEC_CellType <- ggplot(PCAData_afterBC, aes(PC1,PC2))+
  geom_point(aes(fill=CellType), size=5, alpha=0.6, shape=21)+ 
  xlab(paste0("PC1: ",pc1,"% variance")) +
  ylab(paste0("PC2: ",pc2,"% variance")) + 
  theme_bw(base_size = 14)+
  theme(legend.position="bottom", legend.box="vertical", legend.margin=margin())  + 
  geom_vline(xintercept=0, linetype="dotted") + 
  geom_hline(yintercept=0, linetype="dotted") + 
    scale_fill_manual(values = CellTypecols_alt) +
  theme( legend.title = element_blank())+ theme( legend.text.align = 0, 
         legend.key.width  = unit(0.8, "cm") ))
 
 
```


```{r}

corrcounts_ggplot <- stack(log2(corrcounts[,c(samples_ordered[1:N], samples_ordered[(length(samples_ordered)-N +1):length(samples_ordered)])]))
colnames(corrcounts_ggplot) <- c("expression", "sampleID")
corrcounts_ggplot <- merge(corrcounts_ggplot,metadata, by="sampleID")
  

(plt_afterBEC <- ggplot(corrcounts_ggplot, aes(sampleID,expression)) + 
    geom_boxplot(aes(fill=DatasetID))+ 
    xlab("Samples") + 
    ylab("Gene Expression \n(log2(CPM))")   +  
    theme(axis.text.x = element_text(angle = 45, hjust=1))+
  theme_bw(base_size = 14)+ 
  theme(plot.title = element_text(hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank(), legend.title = element_blank(),
        legend.position = "bottom") + 
    scale_fill_manual(values = dataset_colors))
```


```{r}
# Local job (lost in the meantime, this is a good approximation)

#' Estimate variance explained by each metadata variable
#'
#' @param expr numeric matrix or data.frame of normalized expression (genes × samples)
#' @param meta data.frame with sample-level metadata (rows = samples)
#' @param nPC integer; number of top principal components to consider
#' @param scale logical; whether to scale genes before PCA (default FALSE)
#' @return data.frame with variables and % variance explained
#' @examples
#' res <- explained_variance(expr, coldata)
#' print(res)
# explained_variance <- function(expr, meta, nPC = 10, scale = FALSE) {
#   stopifnot(all(colnames(expr) %in% rownames(meta)))
#   meta <- meta[colnames(expr), , drop = FALSE]
#   
#   # PCA
#   pca <- prcomp(t(expr), scale. = scale)
#   pcs <- pca$x[, seq_len(min(nPC, ncol(pca$x))), drop = FALSE]
#   var_exp <- (pca$sdev^2 / sum(pca$sdev^2))[seq_len(ncol(pcs))]
#   
#   # helper to compute R² between numeric vectors
#   r2 <- function(x, y) {
#     if (length(unique(y)) < 2) return(0)
#     mod <- summary(lm(x ~ y))
#     mod$r.squared
#   }
#   
#   res <- sapply(meta, function(v) {
#     if (is.character(v) || is.factor(v)) v <- as.factor(v)
#     if (is.numeric(v)) {
#       mean(sapply(as.data.frame(pcs), function(pc) r2(pc, v) * var_exp))
#     } else {
#       mean(sapply(as.data.frame(pcs), function(pc) r2(pc, model.matrix(~ v)[, -1, drop = FALSE]) * var_exp))
#     }
#   })
#   
#   res <- sort(res, decreasing = TRUE)
#   data.frame(variable = names(res), variance_explained = res, row.names = NULL)
# }
#  

# Another option, I think is not correct...

# 
# explained_variance <- function(expr, meta, nPC = 10, scale = F) {
#   stopifnot(is.matrix(expr) || is.data.frame(expr))
#   stopifnot(is.data.frame(meta))
#   stopifnot(all(colnames(expr) %in% rownames(meta)))
#   meta <- meta[colnames(expr), , drop = FALSE]
# 
#   # PCA on samples
#   pca <- prcomp(t(as.matrix(expr)), scale. = scale)
#   nPCuse <- min(nPC, ncol(pca$x))
#   pcs <- pca$x[, seq_len(nPCuse), drop = FALSE]        # samples × PCs
#   var_exp <- (pca$sdev^2 / sum(pca$sdev^2))[seq_len(nPCuse)]
# 
#   # helper: compute R² of pc ~ predictor
#   r2_single <- function(pc, pred) {
#     # pred can be vector, factor, matrix, or data.frame of predictors
#     if (is.null(ncol(pred))) {
#       # single column predictor (vector or factor)
#       df <- data.frame(.pred = pred)
#       fit <- lm(pc ~ .pred, data = df)
#     } else {
#       # multi-column predictor (matrix/data.frame from model.matrix)
#       df <- as.data.frame(pred)
#       fit <- lm(pc ~ ., data = df)
#     }
#     summary(fit)$r.squared
#   }
# 
#   # iterate over metadata variables
#   res <- sapply(names(meta), function(varname) {
#     v <- meta[[varname]]
#     # prepare predictor: for factors, use model matrix without intercept,
#     # for numeric keep as vector
#     if (is.factor(v) || is.character(v)) {
#       pred <- model.matrix(~ v)[, -1, drop = FALSE]
#       if (ncol(pred) == 0) return(0) # e.g. all NA or single level
#     } else {
#       pred <- v
#     }
#     # compute R2 for each PC, weight by var_exp, and sum
#     r2_per_pc <- vapply(seq_len(ncol(pcs)), function(k) {
#       r2_single(pcs[, k], pred)
#     }, numeric(1))
#     sum(r2_per_pc * var_exp)
#   }, USE.NAMES = TRUE)
# 
#   # return as percentages
#   df <- data.frame(variable = names(res),
#                    perc_variance_explained = unname(res) * 100,
#                    row.names = NULL)
#   df[order(df$perc_variance_explained, decreasing = TRUE), , drop = FALSE]
# }

```


```{r fig.width=4, fig.height=4}

df_variance_corrcounts <- readRDS("/mnt/scratch/home/rsilva/VersionControl/markeR/markeR/inst/Paper/data/df_variance_corrcounts.rds")
df_variance_normcounts <- readRDS("/mnt/scratch/home/rsilva/VersionControl/markeR/markeR/inst/Paper/data/df_variance_normcounts.rds")
vars <- c("DatasetID","Condition","CellType")

(var_beforeBEC <- ggplot(df_variance_normcounts, aes(x = reorder(Variable, -VarianceExplained), y = VarianceExplained, fill = Variable)) +
    geom_bar(stat = "identity") +
    ylab("Mean Variance Explained (%)") +
    xlab("") +
    theme_minimal(base_size = 14) +
    theme(plot.title = element_text(hjust = 0.5),
          legend.position="bottom",
          legend.title=element_blank()) +
  scale_fill_manual(values = c("#F4C430", "#E07A5F", "#3D84A8") )+
    guides(fill = guide_legend(title = "Variable"))+
    ylim(c(0,100)) + 
    theme(legend.position = "none")+
  scale_x_discrete(labels = c(
    "DatasetID" = "Dataset ID",
    "CellType"  = "Cell Type",
    "Condition" = "Cellular State"
  )))


(var_afterBEC <- ggplot(df_variance_corrcounts, aes(x = reorder(Variable, -VarianceExplained), y = VarianceExplained, fill = Variable)) +
    geom_bar(stat = "identity") + 
    ylab("Mean Variance Explained (%)") +
    xlab("") +
    theme_minimal(base_size = 14) +
    theme(plot.title = element_text(hjust = 0.5),
          legend.position="bottom",
          legend.title=element_blank()) +
    scale_fill_manual(values = c("#F4C430", "#E07A5F", "#3D84A8")) +
    guides(fill = guide_legend(title = "Variable"))+
    ylim(c(0,100))+ 
    theme(legend.position = "none")+
  scale_x_discrete(labels = c(
    "DatasetID" = "Dataset ID",
    "CellType"  = "Cell Type",
    "Condition" = "Cellular State"
  )))



```




<!-- ```{r fig.width=16, fig.height=16} -->
<!-- # filtering -->
<!-- (panelA <- ggarrange(plt_beforefilt,plt_afterfilt, ncol=1, common.legend = T, legend="bottom")) -->
<!-- # normalisation -->
<!-- (panelB <- ggarrange(plt_beforenorm,plt_afternorm,ncol=1, common.legend = T, legend="bottom")) -->
<!-- (panelAB <- ggarrange(panelA,panelB,nrow=1, widths=c(0.3,0.7), labels = c("A","B"), -->
<!--                       font.label = list(size = 18, face = "bold"))) -->
<!-- # before BEC -->
<!-- (panelC_left <- ggarrange(PCA_beforeBEC_Dataset+ -->
<!--   theme(plot.margin = margin(t = 5, r = 5, b = 5, l = 30)),PCA_beforeBEC_Conditon, ncol = 1, heights=c(0.7,0.3) )) -->
<!-- (panelC_right <- ggarrange(PCA_beforeBEC_CellType,var_beforeBEC, ncol=1, heights=c(0.45,0.55) )) -->
<!-- (panelC <- ggarrange(panelC_left,panelC_right, ncol=2  )) -->

<!-- # after BEC -->
<!-- (panelD_bottom <- ggarrange(PCA_afterBEC_Dataset,var_afterBEC, nrow=1, widths=c(0.6,0.4))) -->
<!-- (panelD <- ggarrange(plt_afterBEC,panelD_bottom,ncol=1, heights = c(0.4,0.6))) -->
<!-- (panelCD <- ggarrange( -->
<!--   panelC, panelD, -->
<!--   labels = c("C", "D"), -->
<!--   font.label = list(size = 18, face = "bold") -->
<!-- ))   -->
<!-- (plt_sup1 <- ggarrange(panelAB,panelCD, ncol=1, heights=c(0.45,0.65))) -->
<!-- ``` -->


Bigger spacing and labels for each panel
 

```{r fig.width=18, fig.height=18}
 

## Define a blank spacer
spacer <- ggplot() + theme_void()

# filtering
panelA <- ggarrange(
  plt_beforefilt, spacer, plt_afterfilt,
  ncol = 1, heights = c(1, 0.02, 1),
  common.legend = TRUE, legend = "bottom"
)

# normalisation
panelB <- ggarrange(
  plt_beforenorm, spacer, plt_afternorm,
  ncol = 1, heights = c(1, 0.02, 1),
  common.legend = TRUE, legend = "bottom"
)

panelAB <- ggarrange(
  panelA, spacer, panelB,
  nrow = 1, widths = c(0.3, 0.02, 0.7),
  labels = c("A", "", "  B"),
  font.label = list(size = 24, face = "bold"),
  label.x = -0.03,
  label.y = 1.05
)

# # before BEC
# panelC_left <- ggarrange(
#   PCA_beforeBEC_Dataset + theme(plot.margin = margin(t = 20, r = 5, b = 5, l = 40)),
#   spacer,
#   PCA_beforeBEC_Conditon,
#   ncol = 1, heights = c(0.7, 0.02, 0.3)
# )
# 
# panelC_right <- ggarrange(
#   PCA_beforeBEC_CellType, spacer, var_beforeBEC,
#   ncol = 1, heights = c(0.45, 0.02, 0.55)
# )
# 
# panelC <- ggarrange(
#   panelC_left, spacer, panelC_right,
#   ncol = 3, widths = c(1, 0.02, 1)
# )

# Extract legend
legend_big <- cowplot::get_legend(
  PCA_beforeBEC_Dataset + theme(legend.position = "right")
)

# Remove legend from plot
PCA_big_noleg <- PCA_beforeBEC_Dataset + theme(legend.position = "none")

# Top row: PCA (left) + legend (right)
panelC_top <- ggarrange(
  PCA_big_noleg,
  as_ggplot(legend_big),
  ncol = 2,
  widths = c(0.5, 0.5) # adjust as needed
)

# Bottom row: stacked PCAs (left) + barplot (right)
panelC_bottom <- ggarrange(
  ggarrange(
    PCA_beforeBEC_Conditon,
    PCA_beforeBEC_CellType,
    ncol = 1,
    heights = c(0.48, 0.52)
  ),
  var_beforeBEC,
  ncol = 2,
  widths = c(1, 1)
)

# Combine top + bottom
panelC <- ggarrange(
  panelC_top,
  panelC_bottom,
  ncol = 1,
  heights = c(0.3, 0.7) # equal height for top and bottom rows
)


# after BEC
panelD_bottom <- ggarrange(
  PCA_afterBEC_Dataset, spacer, var_afterBEC,
  nrow = 1, widths = c(0.6, 0.02, 0.4)
)

panelD <- ggarrange(
  plt_afterBEC, spacer, panelD_bottom,
  ncol = 1, heights = c(0.4, 0.02, 0.6)
)

panelCD <- ggarrange(
  panelC, spacer, panelD,
  ncol = 3, widths = c(1, 0.02, 1.1),
  labels = c("C", "", " D"),
  font.label = list(size = 24, face = "bold"),
  label.x = -0.03,
  label.y = 1.05
)  

plt_sup1 <- ggarrange(
  panelAB, spacer, panelCD,
  ncol = 1, heights = c(0.45, 0.02, 0.65)
)

# Add extra margins so labels A and C don’t get cropped
plt_sup1 <- annotate_figure(
  plt_sup1,
  fig.lab = NULL,
  fig.lab.face = "bold",
  fig.lab.size = 24
) + theme(plot.margin = margin(t = 20, r = 20, b = 20, l = 40))

plt_sup1
```

```{r}
ggplot2::ggsave("../Figures/Figs/SupFigure1_Preprocessing.png",
                plt_sup1,
                width = 20, height=18, bg = 'white')

ggplot2::ggsave("../Figures/Figs/SupFigure1_Preprocessing.pdf",
                plt_sup1,
                width = 20, height=18, bg = 'white')
```


## Supplementary Figure 2 - Function of Gene Sets

```{r}
# transform list into a list of vectors
gene_sets <- lapply(signatures_bidirectional, function(x) {
  if (is.data.frame(x)) unique(x[[1]])
  else if (is.vector(x)) unique(x)
  else NULL
})

 
```


Crossing information from BioMart (gene characterisation) and COMPARTMENTS (protein localisation)

COMPARTMENTS: https://academic.oup.com/database/article/doi/10.1093/database/bau012/2633793#47519723
Human Protein Atlas: https://www.proteinatlas.org/about/assays+annotation#ih_reliability

```{r}
gene_annotations <- as.data.frame(fread("../data_aux/BioMart_Gene_Characterisation_March2024.txt"))
gene_annotations <- subset(gene_annotations, gene_annotations$`Protein stable ID`!="")


protein_localisation <- as.data.frame(fread("../data_aux/human_compartment_knowledge_full.tsv"))
protein_localisation <- subset(protein_localisation, V5 == "UniProtKB")
protein_localisation <- subset(protein_localisation, V6 %in% c("CURATED")) #c("CURATED", "IDA", "TAS", "NAS")  

protein_localisation_HumanAtlas_all <- as.data.frame(fread("../data_aux/HumanAtlas_subcellular_location.tsv"))
protein_localisation_HumanAtlas <- subset(protein_localisation_HumanAtlas_all, Reliability != "Uncertain")
```

COMPARTMENTS + BioMart

```{r}
gene_annotation_and_localisation <- merge(gene_annotations,protein_localisation, by.x="Protein stable ID", by.y="V1", all.x = TRUE)
colnames(gene_annotation_and_localisation) <- c("ProteinID","GeneName","GeneDescription","GOTermName","GOTermAccession","GOTermDefinition","GODomain","GeneName2","GOTermAccession_Localisation","GOTermName_Localisation","DataBase_Localisation","Evidence","Evidence_Score")
gene_annotation_and_localisation <- subset(gene_annotation_and_localisation, !is.na(GeneName2))
```


```{r}
gene_annotation_and_localisation
protein_localisation_HumanAtlas
```

```{r fig.width=18, fig.height=12}
 
# Assume:
# - `gene_sets`: named list of gene vectors
# - `protein_localisation_HumanAtlas_all`: your dataframe
 

# # Rename columns just in case
# gene_data <- protein_localisation_HumanAtlas_all[, c("Gene name", "Main location", "Extracellular location")]
# colnames(gene_data) <- c("Gene", "Location", "Secretion")
# 
# # Assign extracellular label as either "Predicted to be secreted" or "not reported"
# gene_data$SecretionStatus <- ifelse(grepl("Predicted to be secreted", gene_data$Secretion),
#                                     "Predicted to be secreted", "not reported")
#  
plot_list <- list()

for (sig_name in names(gene_sets)) {
  genes <- gene_sets[[sig_name]]
  
  # Subset the Human Atlas data based on the gene set
  sub_data <- protein_localisation_HumanAtlas_all[protein_localisation_HumanAtlas_all$`Gene name` %in% genes, ]
  
  # Remove duplicates based on the 'Gene name' and 'Main location' columns
  sub_data <- sub_data[!duplicated(sub_data[c("Gene name", "Main location")]), ]

  # Assign secretion status
  sub_data$SecretionStatus <- ifelse(grepl("Predicted to be secreted", sub_data$`Extracellular location`),
                                     "Predicted to be secreted", "Not reported")

  # Initialize empty result data frame
  counts <- data.frame(Location = character(),
                       SecretionStatus = character(),
                       Count = numeric(),
                       stringsAsFactors = FALSE)

  # Count occurrences
  for (i in seq_len(nrow(sub_data))) {
    locs <- unlist(strsplit(sub_data$`Main location`[i], ";"))
    locs <- trimws(locs)
    locs <- locs[locs != ""]
    if (length(locs) == 0) locs <- "Unspecified"
    secretion <- sub_data$SecretionStatus[i]

    for (loc in locs) {
      match_row <- which(counts$Location == loc & counts$SecretionStatus == secretion)
      if (length(match_row) == 0) {
        counts <- rbind(counts, data.frame(Location = loc, SecretionStatus = secretion, Count = 1))
      } else {
        counts$Count[match_row] <- counts$Count[match_row] + 1
      }
    }
  }

  # Reorder Location by total gene count
  total_counts <- tapply(counts$Count, counts$Location, sum)
  counts$Location <- factor(counts$Location, levels = names(sort(total_counts)))

  # Order secretion status so "Predicted" appears on top
  counts$SecretionStatus <- factor(counts$SecretionStatus, levels = c("Predicted to be secreted","Not reported"))

  sig_name <- wrap_title(sig_name,28)
  
  # Plot
  p <- ggplot(counts, aes(x = Count, y = Location)) +
    geom_bar(stat = "identity", aes(fill = SecretionStatus)) +
    labs(title = sig_name, x = "Frequency", y = "Cellular localisation\n") +
    scale_fill_manual(values = c("Predicted to be secreted" = "#3D5D56", 
                                 "Not reported" = "#899C98"),
                      labels = c("Predicted to be secreted", "not reported")) +
    theme_classic() +
    theme(plot.title = element_text(hjust = 0.5),
          axis.text.y = element_text(size = 9),
          legend.title = element_blank())

  plot_list[[sig_name]] <- p
}

# Combine all plots in a grid, with a common legend
final_plot_localisation <- ggarrange(plotlist = plot_list, common.legend = TRUE, heights = c(0.4, 0.3, 0.3))
final_plot_localisation
```


```{r fig.width=18, fig.height=12}
# Load libraries
library(msigdbr)
library(ggplot2)
library(reshape2)
library(ggpubr)

# 1. Load MSigDB Hallmark pathways
pathways_H <- msigdbr(species = "Homo sapiens", category = "H")
pathwaysHset <- split(toupper(pathways_H$gene_symbol), pathways_H$gs_name)

# 2. Create the universe of genes from the Human Protein Atlas
universe_genes <- unique(toupper(protein_localisation_HumanAtlas_all$`Gene name`))

# 3. Initialize result matrices
df_OR <- matrix(nrow = length(pathwaysHset), ncol = length(gene_sets))
df_pval <- matrix(nrow = length(pathwaysHset), ncol = length(gene_sets))

# 4. Fisher's exact test for each pathway × signature
for (i in seq_along(pathwaysHset)) {
  pathway_genes <- pathwaysHset[[i]]
  pathway_name <- names(pathwaysHset)[i]

  for (j in seq_along(gene_sets)) {
    signature_genes <- toupper(gene_sets[[j]])
    signature_name <- names(gene_sets)[j]

    # Define contingency table
    a <- length(intersect(signature_genes, pathway_genes))                           # In both
    b <- length(setdiff(signature_genes, pathway_genes))                             # In signature only
    c <- length(setdiff(pathway_genes, signature_genes))                             # In pathway only
    d <- length(setdiff(universe_genes, union(signature_genes, pathway_genes)))      # In neither

    contingency_table <- matrix(c(a, b, c, d), nrow = 2)

    # Perform Fisher's exact test
    fisher_result <- fisher.test(contingency_table)
    df_OR[i, j] <- fisher_result$estimate
    df_pval[i, j] <- fisher_result$p.value
  }
}

# 5. Set row and column names
rownames(df_pval) <- names(pathwaysHset)
colnames(df_pval) <- names(gene_sets)
rownames(df_OR) <- names(pathwaysHset)
colnames(df_OR) <- names(gene_sets)

# 6. Convert to long format
df_pval_melt <- melt(df_pval)
colnames(df_pval_melt) <- c("Pathway", "Signature", "pval")

df_OR_melt <- melt(df_OR)
colnames(df_OR_melt) <- c("Pathway", "Signature", "OR")

# 7. Adjust p-values
df_pval_melt$padj <- p.adjust(df_pval_melt$pval, method = "BH")

# 8. Merge OR and adjusted p-values
enrichment_results <- merge(df_OR_melt,
                            df_pval_melt[, c("Pathway", "Signature", "padj")],
                            by = c("Pathway", "Signature"))

# 9. Filter significant results (padj ≤ 0.05)
enrichment_results <- enrichment_results[!is.na(enrichment_results$padj) & enrichment_results$padj <= 0.05, ]

# 10. Plot per signature
plot_list <- list()
for (sig in unique(enrichment_results$Signature)) {
  data_sig <- enrichment_results[enrichment_results$Signature == sig, ]
  
    sig <- wrap_title(sig,28)
    
  p <- ggplot(data_sig, aes(x = log10(OR), y = reorder(Pathway, log10(OR)))) +
    geom_bar(stat = "identity", fill = "#3D5D56") +
    labs(title = sig, x = "log10(Fisher's OR)", y = "Enriched Hallmark \nPathways") +
    theme_bw() +
    geom_vline(xintercept = 0, linetype = "dashed") +
    theme(plot.title = element_text(hjust = 0.5, size = 12),
          axis.text.y = element_text(size = 9))
  
  plot_list[[sig]] <- p
}

# 11. Arrange plots in a 3x3 grid with common legend
final_plot_function <- ggarrange(plotlist = plot_list,
                        ncol = 3, nrow = 3,
                        common.legend = TRUE, legend = "bottom", 
                        heights=c(0.45,0.35,0.2))

# 12. Display final plot
print(final_plot_function)

# 13. Optionally save it
# ggsave("GeneSets_HallmarkPathways_Enrichment.png", plot = final_plot,
#        width = 15, height = 18, bg = 'white')


```


```{r fig.width=18, fig.height=24}
(supfig2 <- ggarrange(final_plot_localisation,final_plot_function, labels="AUTO", ncol=1,font.label = list(size = 22)))
```

```{r}
ggplot2::ggsave("Figs/Sup2_GeneSets_Function_Localisation.png",
                supfig2,
                width = 20, height=24, bg = 'white')
```



### Alternative - using Signature Similarity

```{r}
gene_sets <- lapply(signatures_bidirectional, function(x) {
  if (is.data.frame(x)) unique(x[[1]])
  else if (is.vector(x)) unique(x)
  else NULL
})


```
 


```{r fig.width=8, fig.height=16}
# changed internally in the function
#      x = "Senescence Gene Set"
#      y = "Compared Gene Set"
(plt_logodds <- geneset_similarity(
  signatures = gene_sets,
  other_user_signatures = NULL,
  collection = "H",
  subcollection = NULL,
  metric = "odds_ratio",
  unique(toupper(protein_localisation_HumanAtlas_all$`Gene name`)), # universe of protein coding geenes, as they are probably the ones important in this
  or_threshold = 1,
  pval_threshold = 0.05,
  limits = NULL,
  title_size = 12,
  title = "",
  jaccard_threshold = 0.1,
  msig_subset = NULL,
  width_text = 28
)$plot)
```
 

```{r fig.width=10, fig.height=16}
 library(ggplot2)
library(ggpubr)

plot_list <- list()
plot_heights <- numeric()

for (sig_name in names(gene_sets)) {
  genes <- gene_sets[[sig_name]]
  sub_data <- protein_localisation_HumanAtlas_all[protein_localisation_HumanAtlas_all$`Gene name` %in% genes, ]
  sub_data <- sub_data[!duplicated(sub_data[c("Gene name", "Main location")]), ]

  sub_data$SecretionStatus <- ifelse(grepl("Predicted to be secreted", sub_data$`Extracellular location`),
                                     "Predicted to be secreted", "Not reported")

  counts <- data.frame(Location = character(),
                       SecretionStatus = character(),
                       Count = numeric(),
                       stringsAsFactors = FALSE)

  for (i in seq_len(nrow(sub_data))) {
    locs <- unlist(strsplit(sub_data$`Main location`[i], ";"))
    locs <- trimws(locs)
    locs <- locs[locs != ""]
    if (length(locs) == 0) locs <- "Unspecified"
    secretion <- sub_data$SecretionStatus[i]

    for (loc in locs) {
      match_row <- which(counts$Location == loc & counts$SecretionStatus == secretion)
      if (length(match_row) == 0) {
        counts <- rbind(counts, data.frame(Location = loc, SecretionStatus = secretion, Count = 1))
      } else {
        counts$Count[match_row] <- counts$Count[match_row] + 1
      }
    }
  }

  total_counts <- tapply(counts$Count, counts$Location, sum)
  counts$Location <- factor(counts$Location, levels = names(sort(total_counts)))
  counts$SecretionStatus <- factor(counts$SecretionStatus, levels = c("Predicted to be secreted","Not reported"))

  sig_name_wrapped <- wrap_title(sig_name, 28)

  p <- ggplot(counts, aes(x = Count, y = Location)) +
    geom_bar(stat = "identity", aes(fill = SecretionStatus)) +
    labs(title = sig_name_wrapped, x = "Frequency", y = "Cellular localisation\n") +
    scale_fill_manual(values = c("Predicted to be secreted" = "#3D5D56", 
                                 "Not reported" = "#899C98"),
                      labels = c("Predicted to be secreted", "not reported")) +
    theme_classic() +
    theme(plot.title = element_text(hjust = 0.5, size=10),
          axis.text.y = element_text(size = 10),
          
          axis.text.x = element_text(size = 10),
          legend.title = element_blank(), legend.position = "none")

  plot_list[[sig_name]] <- p
  
  # Add number of y-axis rows as a proxy for height
  n_locs <- length(levels(counts$Location))
  plot_heights <- c(plot_heights, n_locs)
}

# Normalize heights so that they sum to 1
plot_heights <- plot_heights / sum(plot_heights)
names(plot_heights) <- names(plot_list)

plot_heights["GOBP_POSITIVE_REGULATION_OF_CELLULAR_SENESCENCE"] <- 0.06
plot_heights["GOBP_NEGATIVE_REGULATION_OF_CELLULAR_SENESCENCE"] <- 0.06

# Combine plots in one column, adjusting heights automatically
final_plot_localisation2_1 <- ggarrange(plotlist = plot_list[c("CellAge","CSgene","REACTOME_CELLULAR_SENESCENCE","SAUL_SEN_MAYO")],
                                      ncol = 1,
                                      heights = plot_heights[c("CellAge","CSgene","REACTOME_CELLULAR_SENESCENCE","SAUL_SEN_MAYO")] )

final_plot_localisation2_2 <- ggarrange(plotlist = plot_list[!names(plot_list) %in% c("CellAge","CSgene","REACTOME_CELLULAR_SENESCENCE","SAUL_SEN_MAYO")],
                                      ncol = 1,
                                      heights = plot_heights[!names(plot_heights) %in% c("CellAge","CSgene","REACTOME_CELLULAR_SENESCENCE","SAUL_SEN_MAYO")],
                                      common.legend = TRUE,
                                      legend = "top")


(final_plot_localisation2 <- ggarrange(final_plot_localisation2_2,final_plot_localisation2_1,
                                      nrow = 1,
                                      heights = plot_heights))
```

```{r fig.width=15, fig.height=14}
(plt_sup2_alternative <- ggarrange(final_plot_localisation2,plt_logodds, nrow=1, labels = "AUTO", font.label = list(size = 20), widths = c(0.5,0.5)))
```


```{r}
ggplot2::ggsave("Figs/SupFigure2_GeneSets_Function_Localisation_alternative.png",
                plt_sup2_alternative,
                width = 15, height=14, bg = 'white')

ggplot2::ggsave("Figs/SupFigure2_GeneSets_Function_Localisation_alternative.pdf",
                plt_sup2_alternative,
                width = 15, height=14, bg = 'white')
```


## Supplementary Figure 3 - Score distribution and correlations




```{r}
 
flat_df_scores <- imap_dfr(scoresList, function(signature_list, method_name) {
  imap_dfr(signature_list, function(signature_df, signature_name) {
    signature_df %>%
      mutate(
        Method = method_name,
        Signature = signature_name
      )
  })
})

flat_df_scores

```
```{r fig.width=16, fig.height=12}
# Get the unique signatures
signatures <- unique(flat_df_scores$Signature)

# Create an empty list to store plots
plot_list <- list()

# Loop over each signature
for (sig in signatures) {
  df_sig <- flat_df_scores[flat_df_scores$Signature == sig, ]

  p <- ggplot(df_sig, aes(x = Condition, y = score)) +
    geom_jitter(size=2, aes(color = CellType), alpha=0.7)+
    geom_violin(trim = FALSE, alpha=0.6, scale = "width") +  
    stat_summary(aes(group = CellType, color = CellType),
                                     fun.y = median, geom = "line", size = 1.5, alpha = 0.75,
                                     show.legend = FALSE) +
    stat_summary(fun = median, fun.min = median, fun.max = median,
                                   geom = "crossbar", width = 0.25,
                                   position = position_dodge(width = 0.13)) +
    facet_wrap(~ Method, scales = "free_y") +
    scale_color_manual(values = CellTypecols_alt) +
    theme_minimal(base_size = 16) +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1),
      strip.text = element_text(face = "bold"),
      legend.position = "right",
      plot.title = ggplot2::element_text(hjust = 0.5),

    ) +
    labs(
      title = wrap_title(sig, width = 28),
      x = "",
      y = "Score",
      color = ""
    ) + 
    guides(color = guide_legend(nrow = 1, override.aes = list(size = 4)))

  plot_list[[sig]] <- p
}


# Arrange all plots in a grid with ggarrange
plt_scores_2 <- ggarrange(plotlist = plot_list, ncol = 3, nrow=3, common.legend = T, legend = "top", align="h")
plt_scores_2

```



```{r}
(scores_logmedian <- CalculateScores(data = corrcounts, 
                                     metadata = metadata, 
                                     method = "logmedian", 
                                     gene_sets = signatures_bidirectional))

(scores_ssGSEA <- CalculateScores(data = corrcounts, 
                                     metadata = metadata, 
                                     method = "ssGSEA", 
                                     gene_sets = signatures_bidirectional))

(scores_ranking <- CalculateScores(data = corrcounts, 
                                     metadata = metadata, 
                                     method = "ranking", 
                                     gene_sets = signatures_bidirectional))

scoresList <- list(logmedian=scores_logmedian,
               ssGSEA=scores_ssGSEA,
               ranking=scores_ranking)
```

```{r fig.width=8, fig.height=4}
library(dplyr)
library(tidyr)
 
 

df_all <- lapply(names(scoresList$logmedian), function(gene_set) {
  
  df_logmedian <- scoresList$logmedian[[gene_set]] %>%
    select(sample, score, SenescentType) %>%
    rename(logmedian = score)
  
  df_ssGSEA <- scoresList$ssGSEA[[gene_set]] %>%
    select(sample, score) %>%
    rename(ssGSEA = score)
  
  df_ranking <- scoresList$ranking[[gene_set]] %>%
    select(sample, score) %>%
    rename(ranking = score)
  
  # merge all three by sampleID
  df <- df_logmedian %>%
    inner_join(df_ssGSEA, by = "sample") %>%
    inner_join(df_ranking, by = "sample") %>%
    mutate(GeneSet = gene_set)
  
  return(df)
})

df_combined <- bind_rows(df_all)

 
library(ggplot2)
library(ggpubr)
library(dplyr)
library(patchwork)
library(scales)  # Make sure this is loaded

# Create darker colors
darker_geneSet_colors <- darken(geneSet_colors, amount = 0.3)
 
df_combined$GeneSet <-  vapply(df_combined$GeneSet, FUN = function(x) wrap_title(x,28),"character")


# Plot
p1 <- ggplot(df_combined, aes(x = logmedian, y = ssGSEA)) +
  geom_point(aes(color = GeneSet), alpha = 0.6) +
  scale_color_manual(values = geneSet_colors) +
 guides(color = guide_legend(nrow = 3, override.aes = list(size = 4))) +
  ggnewscale::new_scale_color() +
  geom_smooth(aes(color = GeneSet), method = "lm", se = FALSE, show.legend = FALSE) +
  scale_color_manual(values = darker_geneSet_colors) +
  theme_classic(base_size=16) +
  labs(title = "logmedian vs ssGSEA", x = "logmedian score", y = "ssGSEA score") +
  theme(
    legend.position = "bottom",
    legend.background = element_rect(fill = "white"),
    legend.title = element_blank()
  ) 


p2 <- ggplot(df_combined, aes(x = logmedian, y = ranking)) +
  geom_point(aes(color = GeneSet), alpha = 0.6) +
  scale_color_manual(values = geneSet_colors) +
  guides(color = guide_legend(nrow = 3, override.aes = list(size = 4)))+
  ggnewscale::new_scale_color() +
  geom_smooth(aes(color = GeneSet), method = "lm", se = FALSE, show.legend = FALSE) +
  scale_color_manual(values = darker_geneSet_colors) +
  theme_classic(base_size=16) +
  labs(title = "logmedian vs ranking", x = "logmedian score", y = "ranking score") +
  theme(
    legend.position = "bottom",
    legend.background = element_rect(fill = "white"),
    legend.title = element_blank())
   

p3 <- ggplot(df_combined, aes(x = ssGSEA, y = ranking)) +
  geom_point(aes(color = GeneSet), alpha = 0.4, size=2) +
  scale_color_manual(values = geneSet_colors) +
  guides(color = guide_legend(nrow = 3, override.aes = list(size = 4))) +
  ggnewscale::new_scale_color() +
  geom_smooth(aes(color = GeneSet), method = "lm", se = FALSE, show.legend = FALSE) +
  scale_color_manual(values = darker_geneSet_colors) +
  theme_classic(base_size=16) +
  labs(title = "ssGSEA vs ranking", x = "ssGSEA score", y = "ranking score") +
  theme(
    legend.position = "bottom",
    legend.background = element_rect(fill = "white"),
    legend.title = element_blank()
  )  




# Arrange the three plots in a single row with common legend
(plt_corr_methods <- ggarrange(p1, p2, p3, nrow = 1, common.legend = TRUE, legend = "none"))



```

```{r fig.width=10, fig.height=5}

library(purrr)

# Compute Spearman correlations for each gene set
cor_table <- df_combined %>%
  group_by(GeneSet) %>%
  summarise(
    "logmedian vs ssGSEA" = cor.test(logmedian, ssGSEA, method = "spearman") %>%
      {sprintf("ρ=%.2f, p=%.2g", .$estimate, .$p.value)},
    "logmedian vs ranking" = cor.test(logmedian, ranking, method = "spearman") %>%
      {sprintf("ρ=%.2f, p=%.2g", .$estimate, .$p.value)},
    "ssGSEA vs ranking" = cor.test(ssGSEA, ranking, method = "spearman") %>%
      {sprintf("ρ=%.2f, p=%.2g", .$estimate, .$p.value)}
  )

cor_table


# Assuming cor_table already exists

# Convert to long format for plotting
cor_long <- cor_table %>%
  pivot_longer(
    cols = -GeneSet,
    names_to = "Comparison",
    values_to = "Correlation"
  )

# Plot
(plt_table_methods <- ggplot(cor_long, aes(x = Comparison, y = GeneSet)) +
  geom_tile(fill = "white", color = "black") +  # background rectangles for readability
  geom_text(aes(label = Correlation), size = 5) +   # display correlation text
  theme_minimal(base_size = 14) +
  labs(title = "Spearman Correlations by Gene Set",
       x = "Comparison",
       y = "Gene Set") +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid = element_blank()
  ))


```


```{r fig.width = 6, fig.height=4}


 
# Create a long-format correlation table
cor_long <- df_combined %>%
  group_by(GeneSet) %>%
  summarise(
    `logmedian \nvs ssGSEA` = cor(logmedian, ssGSEA, method = "spearman"),
    `logmedian \nvs ranking` = cor(logmedian, ranking, method = "spearman"),
    `ssGSEA \nvs ranking` = cor(ssGSEA, ranking, method = "spearman")
  ) %>%
  pivot_longer(-GeneSet, names_to = "Comparison", values_to = "SpearmanRho")


#cor_long$GeneSet <-  vapply(cor_long$GeneSet, FUN = function(x) wrap_title(x,28),"character")


# Violin plot with proper rho symbol
(plt_violin_methods <- ggplot(cor_long, aes(x = Comparison, y = SpearmanRho)) +
  geom_jitter(aes(color = GeneSet), width = 0.2, size = 3) +
  geom_violin(fill = "grey80", color = "grey50", alpha = 0.4) +
  theme_minimal(base_size=16) +
  labs(
    x = "Method Comparison",
    y = expression("Spearman's " * rho)   # <-- rho symbol
  ) +
  theme(
    legend.position = "right",
    legend.title = element_blank(),
    axis.text.x = element_text(angle = 60, vjust = 0.5)
  ) +
  scale_color_manual(values = geneSet_colors) +
  ylim(c(0,1)))

```

```{r fig.width=16, fig.height=4}
(plt_scorecorrelations <- ggarrange(plt_corr_methods, plt_violin_methods, nrow=1, widths = c(0.6,0.4)))
```
```{r fig.width=18, fig.height=16}
(supfig3_new <- ggarrange(plt_scores_2, plt_scorecorrelations, ncol=1, heights=c(0.8,0.25), labels = "AUTO",font.label = list(size = 20)))
```
```{r}
ggplot2::ggsave("Figs/SupFigure3_ScoreDistribution_Correlations.png",
                supfig3_new,
                width = 18, height=15, bg = 'white')

ggplot2::ggsave("Figs/SupFigure3_ScoreDistribution_Correlations.pdf",
                supfig3_new,
                width = 18, height=15, bg = 'white')
```

## Supplementary Figure 4 - Score Distribution on Different Senescence-Inducing Factors
 

 

```{r fig.width=10, fig.height=20}
plt_logmedian_stressors_1 <-  PlotScores_adapted_noX_noleg(data = corrcounts, 
           metadata = metadata, 
           method = "logmedian", 
           gene_sets = signatures_bidirectional[1:8],  
           ColorVariable = "CellType", 
           Variable="SenescentType", 
           ColorValues = CellTypecols_alt, 
           ConnectGroups=F,  
           ncol = 1, 
           widthTitle=30, 
           limits = NULL, 
           legend_nrow = 1, 
           compute_cohen=F,
           pointSize=1,
           titlesize=14, xlab = "Senescence Inducer")

plt_logmedian_stressors_2 <-  PlotScores_adapted(data = corrcounts, 
           metadata = metadata, 
           method = "logmedian", 
           gene_sets = signatures_bidirectional[9],  
           ColorVariable = "CellType", 
           Variable="SenescentType", 
           ColorValues = CellTypecols_alt, 
           ConnectGroups=F,  
           ncol = 1, 
           widthTitle=30, 
           limits = NULL, 
           legend_nrow = 1, 
           compute_cohen=F,
           pointSize=1,
           titlesize=14, xlab = "Senescence Inducer")

(plt_logmedian_stressors <- ggarrange(plt_logmedian_stressors_1,plt_logmedian_stressors_2, ncol=1, heights=c(0.8,0.16) ))
(plt_logmedian_stressors <- ggpubr::annotate_figure(plt_logmedian_stressors, left = grid::textGrob("Gene Set's Score (logmedian)",
                                                                 rot = 90, vjust = 1, gp = grid::gpar(cex = 1.3, fontsize = 10))))

plt_ssGSEA_stressors_1 <-  PlotScores_adapted_noX_noleg(data = corrcounts, 
           metadata = metadata, 
           method = "ssGSEA", 
           gene_sets = signatures_bidirectional[1:8],  
           ColorVariable = "CellType", 
           Variable="SenescentType", 
           ColorValues = CellTypecols_alt, 
           ConnectGroups=F,  
           ncol = 1, 
           widthTitle=30, 
           limits = NULL, 
           legend_nrow = 1, 
           compute_cohen=F,
           pointSize=1,
           titlesize=14, xlab = "Senescence Inducer")

plt_ssGSEA_stressors_2 <-  PlotScores_adapted(data = corrcounts, 
           metadata = metadata, 
           method = "ssGSEA", 
           gene_sets = signatures_bidirectional[9],  
           ColorVariable = "CellType", 
           Variable="SenescentType", 
           ColorValues = CellTypecols_alt, 
           ConnectGroups=F,  
           ncol = 1, 
           widthTitle=30, 
           limits = NULL, 
           legend_nrow = 1, 
           compute_cohen=F,
           pointSize=1,
           titlesize=14, xlab = "Senescence Inducer")

(plt_ssGSEA_stressors <- ggarrange(plt_ssGSEA_stressors_1,plt_ssGSEA_stressors_2, ncol=1, heights=c(0.8,0.16)))
(plt_ssGSEA_stressors <- ggpubr::annotate_figure(plt_ssGSEA_stressors, left = grid::textGrob("Gene Set's Score (ssGSEA)",
                                                                 rot = 90, vjust = 1, gp = grid::gpar(cex = 1.3, fontsize = 10))))

plt_ranking_stressors_1 <-  PlotScores_adapted_noX_noleg(data = corrcounts, 
           metadata = metadata, 
           method = "ranking", 
           gene_sets = signatures_bidirectional[1:8],  
           ColorVariable = "CellType", 
           Variable="SenescentType", 
           ColorValues = CellTypecols_alt, 
           ConnectGroups=F,  
           ncol = 1, 
           widthTitle=30, 
           limits = NULL, 
           legend_nrow = 1, 
           compute_cohen=F,
           pointSize=1,
           titlesize=14, xlab = "Senescence Inducer")

plt_ranking_stressors_2 <-  PlotScores_adapted(data = corrcounts, 
           metadata = metadata, 
           method = "ranking", 
           gene_sets = signatures_bidirectional[9],  
           ColorVariable = "CellType", 
           Variable="SenescentType", 
           ColorValues = CellTypecols_alt, 
           ConnectGroups=F,  
           ncol = 1, 
           widthTitle=30, 
           limits = NULL, 
           legend_nrow = 1, 
           compute_cohen=F,
           pointSize=1,
           titlesize=14, xlab = "Senescence Inducer")

(plt_ranking_stressors <- ggarrange(plt_ranking_stressors_1,plt_ranking_stressors_2, ncol=1, heights=c(0.8,0.16), common.legend = T, legend = "none"))
(plt_ranking_stressors <- ggpubr::annotate_figure(plt_ranking_stressors, left = grid::textGrob("Gene Set's Score (ranking)",
                                                                 rot = 90, vjust = 1, gp = grid::gpar(cex = 1.3, fontsize = 10))))
```



```{r}
library(ggplot2)
library(ggpubr)

# ----------------------------------------------------------
# Colours
# ----------------------------------------------------------
celltype_colors <- c(
  "Fibroblast"   = "#1b9e77",
  "Keratinocyte" = "#d95f02",
  "Melanocyte"   = "#7570b3",
  "Endothelial"  = "#e7298a",
  "Neuronal"     = "#66a61e",
  "Mesenchymal"  = "#e6ab02"
)

# Dummy data (one row per category)
df_dummy <- data.frame(
  celltype = names(celltype_colors),
  x = 1,
  y = 1
)

# ----------------------------------------------------------
# Dummy plot to extract legend
# ----------------------------------------------------------
p_dummy <- ggplot(df_dummy, aes(x, y, colour = celltype)) +
  geom_point(size = 5) +
  scale_colour_manual(
    values = celltype_colors,
    name = "Cell type"
  ) +
  guides(colour = guide_legend(nrow = 1, byrow = TRUE)) +   # max 3 rows
  theme_void() +
  theme(
    legend.position = "bottom",
    legend.title = element_text(size = 12),
    legend.text = element_text(size = 11)
  )

legend_celltypes <- ggpubr::get_legend(p_dummy)

legend_celltypes <- as_ggplot(legend_celltypes)

legend_celltypes 
```


```{r fig.width=12, fig.height=24}
supfigure_3 <- ggarrange(plt_logmedian_stressors,
          plt_ssGSEA_stressors,
          plt_ranking_stressors, 
          common.legend=T, nrow=1, heights=c(0.3,0.3,0.3), labels = "AUTO",font.label = list(size = 20))

(supfigure_3 <-
  ggarrange(
    supfigure_3,          
    legend_celltypes,     # the legend
    ncol = 1,
    heights = c(1, 0.05)  # adjust legend space as needed
  ))
```

```{r}
ggplot2::ggsave("../Figures/Figs/SupFigure4_Scores_SenescenceStressors.png",
                supfigure_3,
                width = 12, height=24, bg = 'white')

ggplot2::ggsave("../Figures/Figs/SupFigure4_Scores_SenescenceStressors.pdf",
                supfigure_3,
                width = 12, height=24, bg = 'white')
```




## Supplementary Figure 5 - Cohen's d and AUC for all signatures (scores)


<!-- ```{r  fig.width=8, fig.height=8} -->


<!-- Scores_Overview_All <- PlotScores(data = corrcounts,  -->
<!--                                   metadata = metadata,   -->
<!--                                   gene_sets=signatures_bidirectional,  -->
<!--                                   Variable="Condition",   -->
<!--                                   method ="all",    -->
<!--                                   ncol = 3,  -->
<!--                                   nrow = 3,  -->
<!--                                   widthTitle=28,  -->
<!--                                   limits = c(0,2),    -->
<!--                                   title=NULL,  -->
<!--                                   titlesize = 10, -->
<!--                                   #ColorValues = list(heatmap=c("#F9F4AE", "#B44141") ), -->
<!--                                   mode="simple"#, -->
<!--                                   #widthlegend=30,  -->
<!--                                   #sig_threshold=0.05,  -->
<!--                                   #cohen_threshold=0.6, -->
<!--                                   #pointSize=6, -->
<!--                                   #colorPalette="Paired" -->
<!-- ) -->

<!-- plt_cohenheatmap <- Scores_Overview_All$heatmap -->
<!-- plt_cohenheatmap -->
<!-- ``` -->

<!-- ```{r fig.width=8, fig.height=8} -->
<!-- plt_auc_heatmap <- AUC_Scores(data = corrcounts,  -->
<!--                               metadata = metadata,  -->
<!--                               gene_sets=signatures_bidirectional,  -->
<!--                               method = "all",  -->
<!--                               mode = "simple",  -->
<!--                               variable="Condition",  -->
<!--                               nrow = 3,  -->
<!--                               ncol = 3,  -->
<!--                               limits = c(0.5,1),  -->
<!--                               widthTitle = 28,  -->
<!--                               titlesize = 10,  -->
<!--                               ColorValues = c("#F9F4AE", "#B44141") )  -->
<!-- plt_auc_heatmap -->

<!-- ``` -->
<!-- ```{r fig.width=11, fig.height=12} -->
<!-- (supfig4 <- ggarrange(plt_cohenheatmap,plt_auc_heatmap,ncol=1, labels="AUTO", heights=c(0.55,0.45) )) -->
<!-- ``` -->

<!-- ```{r} -->
<!-- ggplot2::ggsave("../Figures/Figs/SupFigure5_Scores_AUCCohen.png", -->
<!--                 supfig4, -->
<!--                 width = 11, height=12, bg = 'white') -->
<!-- ``` -->





```{r fig.width=8,fig.height=20}
plt_scores_overview_1 <-  PlotScores_adapted_noX(data = corrcounts, 
                                  metadata = metadata,  
                                  gene_sets=signatures_bidirectional[1:8], 
                                  Variable="Condition",  
                                  method ="all",   
                                  ncol = 1,  
                                  widthTitle=60, 
                                  limits = c(0,2),   
                                  title=NULL, 
                                  titlesize = 10,
                                  labsize = 16, 
                                  mode="simple" 
)

plt_scores_overview_2 <-  PlotScores_adapted(data = corrcounts, 
                                  metadata = metadata,  
                                  gene_sets=signatures_bidirectional[9], 
                                  Variable="Condition",  
                                  method ="all",   
                                  ncol = 1,  
                                  widthTitle=60, 
                                  limits = c(0,2),   
                                  title=NULL, 
                                  titlesize = 10, 
                                  mode="simple",
                                  labsize = 16 
)

 
 
# Laura proporções de largura
w_main  <- 1      # largura do plot principal
w_small <- 0.83   # largura desejada para o segundo plot (ajusta aqui)

(plt_scores_overview <- ggarrange(
  # linha 1: heatmap principal, colunas com larguras definidas
  ggarrange(
    plt_scores_overview_1$heatmap, NULL,
    ncol = 2, widths = c(w_main, 0), align = "h"
  ),
  # linha 2: heatmap pequeno, alinhado à esquerda mas mais curto
  ggarrange(
    plt_scores_overview_2$heatmap, NULL,
    ncol = 2, widths = c(w_small, 1 - w_small), align = "h"
  ),
  ncol = 1,
  heights = c(0.8, 0.14)
))
 
```

```{r fig.width=8,fig.height=20}
plt_auc_heatmap_1 <- AUC_Scores_noX(data = corrcounts, 
                              metadata = metadata, 
                              gene_sets=signatures_bidirectional[1:8], 
                              method = "all", 
                              mode = "simple", 
                              variable="Condition",  
                              ncol = 1, 
                              limits = c(0.5,1), 
                              widthTitle = 60, 
                              titlesize = 10, 
                              ColorValues = c("#F9F4AE", "#B44141") ) 

plt_auc_heatmap_2 <- AUC_Scores_adapted(data = corrcounts, 
                              metadata = metadata, 
                              gene_sets=signatures_bidirectional[9], 
                              method = "all", 
                              mode = "simple", 
                              variable="Condition",  
                              ncol = 1, 
                              limits = c(0.5,1), 
                              widthTitle = 60, 
                              titlesize = 10, 
                              ColorValues = c("#F9F4AE", "#B44141") ) 

 
w_main  <- 1      
w_small <- 0.9    

(plt_auc_overview <- ggarrange( 
  ggarrange(
    plt_auc_heatmap_1, NULL,
    ncol = 2, widths = c(w_main, 0), align = "h"
  ),
 
  ggarrange(
    plt_auc_heatmap_2, NULL,
    ncol = 2, widths = c(w_small, 1 - w_small), align = "h"
  ),
  ncol = 1,
  heights = c(0.8, 0.14)
))
 

```
```{r fig.width=14, fig.height=20}
(supfig5_alt <- ggarrange(plt_scores_overview,plt_auc_overview,nrow=1, labels="AUTO", widths = c(0.53,0.49)))
```


```{r}
ggplot2::ggsave("../Figures/Figs/SupFigure5_Scores_AUCCohen.png",
                supfig5_alt,
                width = 14, height=20, bg = 'white')


ggplot2::ggsave("../Figures/Figs/SupFigure5_Scores_AUCCohen.pdf",
                supfig5_alt,
                width = 14, height=20, bg = 'white')
```

## Supplementary Figure 6 - FPR for all signatures

 
```{r fig.width=12, fig.height=18}
 
# Local job
set.seed("12345678")

# plt_fdrsim_allsignatures_v3 <- FPR_Simulation(data = corrcounts,
#                                          metadata = metadata,
#                                          original_signatures = signatures_bidirectional,
#                                          gene_list = row.names(corrcounts),
#                                          number_of_sims = 100,
#                                          widthTitle = 30,
#                                          Variable = "Condition",
#                                          titlesize = 12,
#                                          pointSize = 3,
#                                          labsize = 10,
#                                          mode = "simple",
#                                          ColorValues=NULL,
#                                          ncol=NULL,
#                                          nrow=3 )
 

plt_fdrsim_allsignatures_v3 <-readRDS("../data_aux/FPRsim_100_data_and_plot_Sept30_2025_allsigs.rds")
(plt_fdrsim_allsignatures_v3 <- plt_fdrsim_allsignatures_v3$plot)
```

```{r}
ggplot2::ggsave("Figs/SupFigure6_FPRsimulations.png",
                plt_fdrsim_allsignatures_v3,
                width = 12, height=18, bg = 'white')

ggplot2::ggsave("Figs/SupFigure6_FPRsimulations.pdf",
                plt_fdrsim_allsignatures_v3,
                width = 12, height=18, bg = 'white')
```


## Supplementary Figure 7 - SenMayo for GTEx
 
 


```{r}
# local job: ~5 days
# 
# 
# methods <- c("logmedian","ranking","ssGSEA")
# gene_set <- list(HernandezSegura=signatures_bidirectional$HernandezSegura,
#                  SAUL_SEN_MAYO=signatures_bidirectional$SAUL_SEN_MAYO)
# tissues <- unique(metadata_GTEx_alltissues$SMTSD)
# 
# results_df_score <- data.frame(NULL)
# results_df_gsea <- data.frame(NULL)
# 
# # Initialize progress bar
# pb <- txtProgressBar(min = 0, max = length(tissues), style = 3)
# 
# for (i in seq_along(tissues)) {
# 
#   tissue <- tissues[i]
# 
#   subset_metadata <- metadata_GTEx_alltissues[metadata_GTEx_alltissues$SMTSD == tissue,]
#   subset_data <- GTEx_alltissues[,subset_metadata$SAMPID]
# 
# 
#   # Update progress bar
#   setTxtProgressBar(pb, i)
# 
#   for (sig in names(gene_set)){
# 
#     signature <- list(gene_set[[sig]])
#     names(signature) <- sig
# 
#     data_varassoc_gsea <- suppressWarnings(suppressMessages(GSEA_VariableAssociation(data=subset_data,
#                                                                                      metadata=subset_metadata,
#                                                                                      cols=c("AGE"),
#                                                                                      mode="simple",
#                                                                                      gene_set=signature)$data))
# 
#     data_varassoc_gsea <- data_varassoc_gsea[,c("NES","pval","Contrast")]
#     data_varassoc_gsea$signature <- sig
#     data_varassoc_gsea$method <- "GSEA"
#     data_varassoc_gsea$tissue <- tissue
# 
#     results_df_gsea <- rbind(results_df_gsea, data_varassoc_gsea)
# 
#     for (method in methods){
# 
#       data_varassoc_score <- suppressWarnings(suppressMessages(Score_VariableAssociation(data=subset_data,
#                                                                                          metadata=subset_metadata,
#                                                                                          cols=c("AGE"), # SMRIN was a variable for batch correction
#                                                                                          method=method,
#                                                                                          gene_set = signature,
#                                                                                          mode="simple", printplt = F)$Overall))
# 
# 
#       df_permutations <- data.frame(NULL)
#       for (j in 1:nperm){
#         set.seed(j)
#         metadata_subset_shuffleAGE <- subset_metadata
#         metadata_subset_shuffleAGE$AGE <- sample(metadata_subset_shuffleAGE$AGE)
# 
#         cohend_shuffle <- suppressWarnings(suppressMessages(Score_VariableAssociation(data=subset_data[,metadata_subset_shuffleAGE$SAMPID],
#                                                                                            metadata=metadata_subset_shuffleAGE,
#                                                                                            cols=c("AGE"), # SMRIN was a variable for batch correction
#                                                                                            method=method,
#                                                                                            gene_set = signature,
#                                                                                            mode="simple", printplt = F)$Overall))
# 
# 
#         cohend_shuffle$signature <- sig
#         cohend_shuffle$method <- method
#         cohend_shuffle$tissue <- tissue
#         df_permutations <- rbind(df_permutations, cohend_shuffle)
# 
#       }
# 
#       # calculateFPR
# 
#       df_permutations$Cohen_f
#       fpr <- sum(df_permutations$Cohen_f > data_varassoc_score$Cohen_f)/length(df_permutations$Cohen_f)
# 
# 
#       data_varassoc_score$signature <- sig
#       data_varassoc_score$method <- method
#       data_varassoc_score$tissue <- tissue
#       data_varassoc_score$fpr <- fpr
#       results_df_score <- rbind(results_df_score, data_varassoc_score)
# 
# 
# 
# 
#     }
# 
#   }
# 
# }
# 
# # Close the progress bar
# close(pb)
# 
# 
# saveRDS( results_df_gsea,"results_df_gsea_GTEx_FPR.rds")
# saveRDS( results_df_score,"results_df_score_GTEx_FPR.rds") # same results for t-test's p value and 1000 perms
```


```{r}
results_df_score <- readRDS("../data_aux/results_df_score_GTEx_FPR.rds")
results_df_gsea <- readRDS("../data_aux/results_df_gsea_GTEx_FPR.rds")


results_df_gsea <- results_df_gsea %>%
  group_by(signature) %>%
  mutate(padj = p.adjust(pval, method = "BH")) %>%
  ungroup()

results_df_score <- results_df_score %>%
  group_by(signature, method) %>%
  mutate(P_adj = p.adjust(fpr, method = "BH")) %>%
  ungroup()


# Alphabetically order tissue factor
tissue_levels <- sort(unique(results_df_score$tissue), decreasing = T)
results_df_score$tissue <- factor(results_df_score$tissue, levels = tissue_levels)
results_df_gsea$tissue  <- factor(results_df_gsea$tissue, levels = tissue_levels)

results_df_gsea
results_df_score
```

 

```{r fig.width=10, fig.height=5.5}
library(ggplot2)
library(dplyr)
library(patchwork)

# Ensure correct factor levels for method and tissue
results_df_score$method <- factor(results_df_score$method, levels = c("logmedian", "ranking", "ssGSEA"))
results_df_gsea$method <- "GSEA"  # single method name

# Add asterisk for significance
results_df_score <- results_df_score %>%
  mutate(signif = ifelse(P_adj < 0.05, "*", ""))

results_df_gsea <- results_df_gsea %>%
  mutate(signif = ifelse(padj < 0.05, "*", ""))

# Loop through each signature
sigs <- unique(results_df_score$signature)

list_plts_sigs <- list()
for (sig in sigs) {
  # Subset for this signature
  score_data <- results_df_score %>% filter(signature == sig)
  gsea_data  <- results_df_gsea %>% filter(signature == sig)

  # Score plot
  p_score <- ggplot(score_data, aes(x = tissue, y = method, fill = Cohen_f)) +
    geom_tile(color = "gray90") +
    geom_text(aes(label = signif), size = 6, vjust = 0.8, hjust = 0.5) +
    scale_fill_gradient(low = "white", high = "#49B0AB", name = "Cohen's f", limits=c(0,0.4) , breaks=c(0,0.2,0.4)) +
    theme_minimal(base_size = 12) +
    theme(
      axis.title = element_blank(),
      axis.text.x = element_text(angle = 60, hjust = 1, size = 12),
      axis.text.y = element_text(size = 12),
      legend.position = "top"
    )

  # GSEA plot
  p_gsea <- ggplot(gsea_data, aes(x = tissue, y = method, fill = abs(NES))) +
    geom_tile(color = "gray90") +
    geom_text(aes(label = signif), size = 6, vjust = 0.8, hjust = 0.5) +
    scale_fill_gradient(low = "white", high = "#B04975", name = "|NES|", limits=c(0,2.5), breaks = c(0, 1, 2),na.value = "#B04975"  ) + 
    theme_minimal(base_size = 12) +
    theme(
      axis.title = element_blank(), 
      axis.text.x = element_blank(),
      axis.ticks.x = element_blank(),
      legend.position = "top"
    )

  # Combine score and GSEA plots
  combined_plot <- p_gsea + p_score  + 
    plot_layout(ncol = 1, guides = "collect", heights = c(0.25, 0.75)) & 
    theme(legend.position = "top")

  list_plts_sigs[[sig]] <- combined_plot
}

 
plt_gtex_scores_SenMayo <- list_plts_sigs[["SAUL_SEN_MAYO"]]
plt_gtex_scores_SenMayo

```

```{r}
tissues_signif_HS <- c("Artery - Aorta","Breast - Mammary Tissue", "Cells - Cultured fibroblasts","Thyroid")
tissues_signif_SenMayo <- c("Artery - Tibial", "Brain - Anterior cingulate cortex (BA24)", "Brain - Hippocampus", "Colon - Sigmoid", "Minor Salivary Gland", "Muscle - Skeletal", "Nerve - Tibial", "Prostate")

```

```{r fig.width=12, fig.height=8}

methods <- c("logmedian","ranking","ssGSEA")
pltlist <- list()

for (tissue in tissues_signif_SenMayo){
  
  subset_metadata <- metadata_GTEx_alltissues[metadata_GTEx_alltissues$SMTSD == tissue,]
  subset_data <- GTEx_alltissues[,subset_metadata$SAMPID]
  
  pltlist_aux <- list()
  
  for (method in methods){ 
    
    scores_df <- CalculateScores(data = subset_data,
                                 metadata = subset_metadata,
                                 gene_sets = list(SAUL_SEN_MAYO=signatures_bidirectional$SAUL_SEN_MAYO), method = method)
     
    
    pltlist_aux[[method]] <- ggplot(scores_df$SAUL_SEN_MAYO, aes(x=AGE, y=score)) +
      geom_jitter(size=2, color="#8F95B1")+ # to preserve the donor's age; correlation was calculated with "real" data
      geom_density2d( colour="white", size=0.3) +
      ggplot2::geom_smooth(method = "lm", col = "black", se = FALSE, size=1.3)+ ggpubr::stat_cor(aes(label = ..r.label..), 
                                                                                                 method = "spearman",
                                                                                                     cor.coef.name="rho",  
                                                                                                 label.x = 20, size=4) +  
      xlab("Age (years)") + ylab("Score") +
      theme_bw() +
      ggtitle(method) +
      #center title 
      theme(plot.title = element_text(hjust = 0.5 ),
            base_size = 16,
            axis.text.x = element_text( size=12),
            axis.text.y = element_text(size=12),
            axis.title.x = element_text(size=14),
            axis.title.y = element_text(size=14)) 
  }
  
  pltlist[[tissue]] <- ggarrange(plotlist = pltlist_aux, ncol = 3, nrow = 1, common.legend = TRUE) +
    ggtitle( tissue)  + theme(plot.title = element_text(hjust = 0.5, size = 14, face = "bold"))
  
  
}

plt_scoredistribution_SenMayo_GTEx <- ggarrange(plotlist = pltlist, ncol = 2, nrow=4, common.legend = TRUE, legend = "bottom")  
plt_scoredistribution_SenMayo_GTEx
```


```{r fig.width=13, fig.height=15}
 # sup figure?

(supfig6 <-  ggarrange(plt_gtex_scores_SenMayo,plt_scoredistribution_SenMayo_GTEx, heights=c(0.35,0.65), labels = c("A","B"), nrow=2))
ggplot2::ggsave("../Figures/Figs/SupFigure7_GTEx_SenMayo.png",
                supfig6,
                width = 13, height=15, bg = 'white')
ggplot2::ggsave("../Figures/Figs/SupFigure7_GTEx_SenMayo.pdf",
                supfig6,
                width = 13, height=15, bg = 'white')

```



## Supplementary Figure 8 - SenMayo with and without inflammation genes


```{r}
# Retrieve all hallmark gene sets (category H)
hallmark_sets <- msigdbr(species = "Homo sapiens", category = "H")

# View unique hallmark gene set names
unique(hallmark_sets$gs_name)

inflammation_sets <- c(
  "HALLMARK_INFLAMMATORY_RESPONSE",
  "HALLMARK_TNFA_SIGNALING_VIA_NFKB",
  "HALLMARK_INTERFERON_GAMMA_RESPONSE",
  "HALLMARK_INTERFERON_ALPHA_RESPONSE",
  "HALLMARK_IL6_JAK_STAT3_SIGNALING",
  #"HALLMARK_COMPLEMENT",
  "HALLMARK_INFLAMMATORY_RESPONSE"
)

inflammation_genes <- msigdbr(species = "Homo sapiens", category = "H") |>
  dplyr::filter(gs_name %in% inflammation_sets) |>
  dplyr::pull(gene_symbol) |>
  unique()

inflammation_genes

overlap <- intersect(signatures_bidirectional$SAUL_SEN_MAYO$gene, inflammation_genes)

overlap

SenMayo_woInflammation <- signatures_bidirectional$SAUL_SEN_MAYO[signatures_bidirectional$SAUL_SEN_MAYO$gene %in% overlap,]
SenMayo_InflammationOnly <- signatures_bidirectional$SAUL_SEN_MAYO[!signatures_bidirectional$SAUL_SEN_MAYO$gene %in% overlap,]

SenMayo_Signatures <- list(SAUL_SEN_MAYO=signatures_bidirectional$SAUL_SEN_MAYO,
                           "Non-inflammatory SAUL_SEN_MAYO"=SenMayo_woInflammation,
                           "Inflammatory SAUL_SEN_MAYO"=SenMayo_InflammationOnly
)

```


```{r fig.width=8, fig.height=4}
(plt_senmayo_scores <- PlotScores(data = corrcounts, 
           metadata = metadata, 
           method = "ranking", 
           gene_sets = SenMayo_Signatures,  
           ColorVariable = "CellType", 
           Variable="Condition", 
           ColorValues = CellTypecols_alt, 
           ConnectGroups=T,  
           ncol = 3, 
           widthTitle=30, 
           limits = NULL, 
           legend_nrow = 1, 
           compute_cohen=F,
           pointSize=3,
           titlesize=14, xlab = "",
            labsize = 12))

```

 
```{r}

degenes_senmayo <- calculateDE(data=corrcounts, 
                       metadata=metadata, 
                       variables="Condition",   
                       modelmat = NULL, 
                       contrasts = c("Senescent - Proliferative",
                                     "Senescent - Quiescent"#,
                                     #"Proliferative - Quiescent"
                       )) 

degenes_senmayo
```


 



```{r}
GSEAresults_senmayo <- runGSEA(degenes_senmayo, SenMayo_Signatures, stat = NULL)
GSEAresults_senmayo
```


 

```{r fig.width=11, fig.height=3}

(plotNESlollipop_senmayo <- plotNESlollipop(GSEA_results=GSEAresults_senmayo, sig_threshold = 0.05,saturation_value=0.00001, nonsignif_color = "white", signif_color = "#934873",
                                    grid = T, nrow = 1, ncol = NULL, widthlabels=16, title=NULL, titlesize=14))

```



```{r} 

methods <- c("logmedian","ranking","ssGSEA")
gene_set <- SenMayo_Signatures[2:3] #wo/ senmayo original, wich we already got
tissues <- unique(metadata_GTEx_alltissues$SMTSD) 

results_df_score_SenMayoCustom <- data.frame(NULL)
results_df_gsea_SenMayoCustom <- data.frame(NULL)

# Initialize progress bar
pb <- txtProgressBar(min = 0, max = length(tissues), style = 3)

for (i in seq_along(tissues)) {
  
  tissue <- tissues[i]
  
  subset_metadata <- metadata_GTEx_alltissues[metadata_GTEx_alltissues$SMTSD == tissue,]
  subset_data <- GTEx_alltissues[,subset_metadata$SAMPID]
  
  
  # Update progress bar
  setTxtProgressBar(pb, i)
  
  for (sig in names(gene_set)){
    
    signature <- list(gene_set[[sig]])
    names(signature) <- sig
    
    data_varassoc_gsea <- suppressWarnings(suppressMessages(GSEA_VariableAssociation(data=subset_data, 
                                                                                     metadata=subset_metadata, 
                                                                                     cols=c("AGE"),  
                                                                                     mode="simple", 
                                                                                     gene_set=signature)$data))
    
    data_varassoc_gsea <- data_varassoc_gsea[,c("NES","pval","Contrast")]
    data_varassoc_gsea$signature <- sig
    data_varassoc_gsea$method <- "GSEA"
    data_varassoc_gsea$tissue <- tissue
    
    results_df_gsea_SenMayoCustom <- rbind(results_df_gsea_SenMayoCustom, data_varassoc_gsea)
    
    for (method in methods){
      
      data_varassoc_score <- suppressWarnings(suppressMessages(Score_VariableAssociation(data=subset_data, 
                                                                                         metadata=subset_metadata, 
                                                                                         cols=c("AGE"), # SMRIN was a variable for batch correction
                                                                                         method=method, 
                                                                                         gene_set = signature, 
                                                                                         mode="simple", printplt = F)$Overall))
      data_varassoc_score$signature <- sig
      data_varassoc_score$method <- method
      data_varassoc_score$tissue <- tissue
      results_df_score_SenMayoCustom <- rbind(results_df_score_SenMayoCustom, data_varassoc_score)
      
      
      
    }
    
  }
  
}

# Close the progress bar
close(pb)


results_df_gsea_SenMayoCustom
results_df_score_SenMayoCustom
```
 
```{r}

results_df_gsea_SenMayoCustom  <- results_df_gsea_SenMayoCustom  %>%
  group_by(signature) %>%
  mutate(padj = p.adjust(pval, method = "BH")) %>%
  ungroup()

results_df_score_SenMayoCustom <- results_df_score_SenMayoCustom  %>%
  group_by(signature, method) %>%
  mutate(P_adj = p.adjust(P_Value, method = "BH")) %>%
  ungroup()


# Alphabetically order tissue factor
tissue_levels <- sort(unique(results_df_score_SenMayoCustom$tissue), decreasing = T)
results_df_score_SenMayoCustom$tissue <- factor(results_df_score_SenMayoCustom$tissue, levels = tissue_levels)
results_df_gsea_SenMayoCustom$tissue  <- factor(results_df_gsea_SenMayoCustom$tissue, levels = tissue_levels)

results_df_gsea_SenMayoCustom
results_df_score_SenMayoCustom
```


```{r}
results_df_score_SenMayoCustom
saveRDS(results_df_score_SenMayoCustom, "../data_aux/results_df_score_SenMayoCustom.rds")
results_df_gsea_SenMayoCustom
saveRDS(results_df_gsea_SenMayoCustom, "../data_aux/results_df_gsea_SenMayoCustom.rds")

```


```{r fig.width=12, fig.height=5}

# Ensure correct factor levels for method and tissue
results_df_score_SenMayoCustom$method <- factor(results_df_score_SenMayoCustom$method, levels = c("logmedian", "ranking", "ssGSEA"))
results_df_gsea_SenMayoCustom$method <- "GSEA"  # single method name

# Add asterisk for significance
results_df_score_SenMayoCustom <- results_df_score_SenMayoCustom %>%
  mutate(signif = ifelse(P_adj < 0.05, "*", ""))

results_df_gsea_SenMayoCustom <- results_df_gsea_SenMayoCustom %>%
  mutate(signif = ifelse(padj < 0.05, "*", ""))

# Loop through each signature
sigs <- unique(results_df_score_SenMayoCustom$signature)

list_plts_sigs <- list()
for (sig in sigs) {
  # Subset for this signature
  score_data <- results_df_score_SenMayoCustom %>% filter(signature == sig)
  gsea_data  <- results_df_gsea_SenMayoCustom %>% filter(signature == sig)

  # Score plot
  p_score <- ggplot(score_data, aes(x = tissue, y = method, fill = Cohen_f)) +
    geom_tile(color = "gray90") +
    geom_text(aes(label = signif), size = 6, vjust = 0.8, hjust = 0.5) +
    scale_fill_gradient(low = "white", high = "#49B0AB", name = "Cohen's f", limits=c(0,0.4) ) +
    theme_minimal(base_size = 12) +
    theme(
      axis.title = element_blank(),
      axis.text.x = element_text(angle = 60, hjust = 1, size = 12),
      axis.text.y = element_text(size = 12),
      legend.position = "none"
    )

  # GSEA plot
  p_gsea <- ggplot(gsea_data, aes(x = tissue, y = method, fill = abs(NES))) +
    geom_tile(color = "gray90") +
    geom_text(aes(label = signif), size = 6, vjust = 0.8, hjust = 0.5) +
    scale_fill_gradient(low = "white", high = "#B04975", name = "|NES|", limits=c(0,2.5) ) + 
    theme_minimal(base_size = 12) +
    theme(
      axis.title = element_blank(), 
      axis.text.x = element_blank(),
      axis.ticks.x = element_blank(),
      legend.position = "none"
    )

  # Combine score and GSEA plots
  combined_plot <- p_gsea + p_score  + 
    plot_layout(ncol = 1, guides = "collect", heights = c(0.25, 0.75))  

  list_plts_sigs[[sig]] <- combined_plot
}

plt_senmayo_original <- plt_gtex_scores_SenMayo + plot_annotation(
  title = "SAUL_SEN_MAYO",
  theme = theme(
    plot.title = element_text(hjust = 0.5, size = 16 )
  )
)
plt_senmayo_woinfl <- list_plts_sigs[["Non-inflammatory SAUL_SEN_MAYO"]] +
  plot_annotation(
  title = "Non-inflammatory SAUL_SEN_MAYO",
  theme = theme(
    plot.title = element_text(hjust = 0.5, size = 16 )
  )
)
plt_senmayo_infl <- list_plts_sigs[["Inflammatory SAUL_SEN_MAYO"]] +
  plot_annotation(
  title = "Inflammatory SAUL_SEN_MAYO",
  theme = theme(
    plot.title = element_text(hjust = 0.5, size = 16 )
  )
)

```
```{r fig.width=12, fig.height=10}
 

# Adjust plot 1: keep legend, hide x-axis
plt_senmayo_original <- plt_senmayo_original + 
  theme(axis.title.x = element_blank(),
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank())

# Adjust plot 2: remove legend and x-axis
plt_senmayo_woinfl <- plt_senmayo_woinfl + 
  theme(axis.title.x = element_blank(),
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank() )
 

# Combine plots vertically
(plt_senmayo_gtex_custom <- ggarrange(
  plt_senmayo_original,
  plt_senmayo_woinfl,
  plt_senmayo_infl,
  ncol = 1,
  align = "v",
  heights=c(0.3,0.2,0.5)
))
 
```
```{r fig.width=12, fig.height=6}
(plt_senmayo_invitro <- ggarrange(plt_senmayo_scores,plotNESlollipop_senmayo, nrow=2, heights=c(0.65,0.35), labels=c("A","B")))
```

```{r fig.width=12, fig.height= 16}
(plt_senmayo_custom <- ggarrange(plt_senmayo_invitro,plt_senmayo_gtex_custom, labels=c("","C"), ncol=1, heights=c(0.4,0.6)))
```
```{r}
ggplot2::ggsave("../Figures/Figs/SupFigure7_GTEx_SenMayo_Custom.png",
                plt_senmayo_custom,
                width = 12, height=16, bg = 'white')
```


Alternative, more quantitative

```{r}
 

# Combine data frames
results_df_gsea <- results_df_gsea_SenMayoCustom %>%
  rename(EffectSize = NES, PValue = padj) %>%
  select(tissue, method, signature, EffectSize, PValue)

results_df_score <- results_df_score_SenMayoCustom %>%
  rename(EffectSize = Cohen_f, PValue = P_adj) %>%
  select(tissue, method, signature, EffectSize, PValue)

results_combined <- bind_rows(results_df_gsea, results_df_score)

# Add inflammation column
results_combined <- results_combined %>%
  mutate(inflammation = ifelse(grepl("Non-inflammatory SAUL_SEN_MAYO", signature), "Without", "With"))
results_combined
 
```
 


<!-- ```{r fig.width=14, fig.height=10} -->
<!-- library(dplyr) -->
<!-- library(tidyr) -->
<!-- library(ggplot2) -->
<!-- library(ggrepel) -->
<!-- library(patchwork) -->

<!-- # ---------------------------------------------------------- -->
<!-- # Tissues to highlight -->
<!-- # ---------------------------------------------------------- -->
<!-- my_tissues <- c( -->
<!--   "Artery - Tibial", -->
<!--   "Brain - Anterior cingulate cortex (BA24)", -->
<!--   "Brain - Hippocampus", -->
<!--   "Colon - Sigmoid", -->
<!--   "Minor Salivary Gland", -->
<!--   "Muscle - Skeletal", -->
<!--   "Nerve - Tibial", -->
<!--   "Prostate", -->
<!--   "Artery - Aorta", -->
<!--   "Breast - Mammary Tissue", -->
<!--   "Cells - Cultured fibroblasts", -->
<!--   "Thyroid" -->
<!-- ) -->

<!-- tissue_colors <- c( -->
<!--   "Artery - Tibial" = "#E69E9E", -->
<!--   "Brain - Anterior cingulate cortex (BA24)" = "#7FBFFF", -->
<!--   "Brain - Hippocampus" = "#4DCC99", -->
<!--   "Colon - Sigmoid" = "#FFFF7F", -->
<!--   "Minor Salivary Gland" = "#FF9F4D", -->
<!--   "Muscle - Skeletal" = "#4A4141", -->
<!--   "Nerve - Tibial" = "#99FFCC", -->
<!--   "Prostate" = "#FF5F5F", -->

<!--   # four shades of the SAME color family (lavender/purple) -->
<!--   "Artery - Aorta" = "#D7B9FF",     -->
<!--   "Breast - Mammary Tissue" = "#B48CFF",  # light lavender -->
<!--   "Cells - Cultured fibroblasts" = "#9660FF",  # medium lavender -->
<!--   "Thyroid" = "#583896"   # deeper lavender -->
<!-- ) -->
<!-- # ---------------------------------------------------------- -->
<!-- # Absolute NES version -->
<!-- # ---------------------------------------------------------- -->
<!-- results_combined_absNES <- results_combined -->
<!-- results_combined_absNES$EffectSize <- abs(results_combined_absNES$EffectSize) -->

<!-- # Assign point color: use tissue color if in my_tissues, otherwise method-based -->
<!-- # Fix point_color as character -->
<!-- results_combined_absNES$point_color <- ifelse( -->
<!--   results_combined_absNES$tissue %in% my_tissues, -->
<!--   as.character(results_combined_absNES$tissue),    # tissue name for color -->
<!--   ifelse(results_combined_absNES$method == "GSEA", "Other_GSEA", "Other_Other") -->
<!-- ) -->

<!-- # Color mapping -->
<!-- all_colors <- c(tissue_colors, -->
<!--                 "Other_GSEA" = "#B04975", -->
<!--                 "Other_Other" = "#49B0AB") -->

<!-- # ---------------------------------------------------------- -->
<!-- # Pivot wider for scatter plots -->
<!-- # ---------------------------------------------------------- -->
<!-- df_wide <- results_combined_absNES %>% -->
<!--   pivot_wider( -->
<!--     id_cols = c(tissue, method, point_color), -->
<!--     names_from = inflammation, -->
<!--     values_from = c(EffectSize, PValue) -->
<!--   ) -->

<!-- # ---------------------------------------------------------- -->
<!-- # Helper: Effect size label -->
<!-- # ---------------------------------------------------------- -->
<!-- get_effect_label <- function(method_name) { -->
<!--   if (method_name == "GSEA") return("|NES|") else return("Cohen's f") -->
<!-- } -->

<!-- # ---------------------------------------------------------- -->
<!-- # Volcano plot -->
<!-- # ---------------------------------------------------------- -->
<!-- make_volcano_plot <- function(method_name, infl_group, threshold_y = 0, threshold_x = 0, show_title = FALSE) { -->

<!--   df_sub <- results_combined_absNES %>% -->
<!--     filter(method == method_name, inflammation == infl_group) -->

<!--   df_lab <- df_sub %>% -->
<!--     filter(tissue %in% my_tissues & (-log10(PValue) >= threshold_y | EffectSize >= threshold_x)) -->

<!--   ggplot(df_sub, aes(x = EffectSize, y = -log10(PValue), color = point_color)) + -->
<!--     geom_point(size = 4) + -->
<!--     # geom_text_repel( -->
<!--     #   data = df_lab, -->
<!--     #   aes(label = tissue), -->
<!--     #   color = "black", -->
<!--     #   size = 3, -->
<!--     #   hjust = 1.1 -->
<!--     # ) + -->
<!--     scale_color_manual( -->
<!--       values = all_colors, -->
<!--       breaks = my_tissues,   # show only tissues of interest in legend -->
<!--       name = NULL -->
<!--     ) + -->
<!--     labs( -->
<!--       x = get_effect_label(method_name), -->
<!--       y = "-log10(adj. p-value)", -->
<!--       title = ifelse(show_title, method_name, "") -->
<!--     ) + -->
<!--     theme_minimal(base_size = 12) + -->
<!--     theme(plot.title = element_text(hjust = 0.5))+ -->
<!--     geom_hline(yintercept = -log10(0.05), linetype="dashed", color="#A6A3A1") -->
<!-- } -->

<!-- # ---------------------------------------------------------- -->
<!-- # Effect-size scatter plot -->
<!-- # ---------------------------------------------------------- -->
<!-- make_effect_plot <- function(df, method_name, show_title = TRUE) { -->
<!--   df_sub <- df %>% filter(method == method_name) -->
<!--   df_lab <- df_sub %>% filter(tissue %in% my_tissues) -->

<!--   ggplot(df_sub, aes(x = EffectSize_Without, y = EffectSize_With, color = point_color)) + -->
<!--     geom_point(size = 4) + -->
<!--     geom_abline(slope = 1, intercept = 0, linetype = "dashed") + -->
<!--     #geom_text_repel(data = df_lab, aes(label = tissue), color = "black", size = 3) + -->
<!--     scale_color_manual( -->
<!--       values = all_colors, -->
<!--       breaks = my_tissues, -->
<!--       name = NULL -->
<!--     ) + -->
<!--     labs( -->
<!--       x = paste0(get_effect_label(method_name), " \nNon-inflammatory SAUL_SEN_MAYO"), -->
<!--       y = paste0(get_effect_label(method_name), " \nInflammatory \nSAUL_SEN_MAYO"), -->
<!--       title = ifelse(show_title, method_name, "") -->
<!--     ) + -->
<!--     theme_minimal(base_size = 12) + -->
<!--     theme(plot.title = element_text(hjust = 0.5)) -->
<!-- } -->

<!-- # ---------------------------------------------------------- -->
<!-- # P-value scatter plot -->
<!-- # ---------------------------------------------------------- -->
<!-- make_pval_plot <- function(df, method_name) { -->
<!--   df_sub <- df %>% filter(method == method_name) -->
<!--   df_lab <- df_sub %>% filter(tissue %in% my_tissues) -->

<!--   ggplot(df_sub, aes(x = -log10(PValue_Without), y = -log10(PValue_With), color = point_color)) + -->
<!--     geom_point(size = 4) + -->
<!--     geom_abline(slope = 1, intercept = 0, linetype = "dashed") + -->
<!--     #geom_text_repel(data = df_lab, aes(label = tissue), color = "black", size = 3) + -->
<!--     scale_color_manual( -->
<!--       values = all_colors, -->
<!--       breaks = my_tissues, -->
<!--       name = NULL -->
<!--     ) + -->
<!--     labs( -->
<!--       x = "-log10(adj. p-value) \nNon-inflammatory SAUL_SEN_MAYO", -->
<!--       y = "-log10(adj. p-value) \nInflammatory \nSAUL_SEN_MAYO" -->
<!--     ) + -->
<!--     theme_minimal(base_size = 12) + -->
<!--     theme(plot.title = element_text(hjust = 0.5)) + -->
<!--     geom_vline(xintercept = -log10(0.05), linetype="dashed", color="#A6A3A1")+ -->
<!--     geom_hline(yintercept = -log10(0.05), linetype="dashed", color="#A6A3A1") -->
<!-- } -->

<!-- # ---------------------------------------------------------- -->
<!-- # Build all 16 plots (with correct labeling) -->
<!-- # ---------------------------------------------------------- -->

<!-- # --- VOLCANO: INFLAMMATORY (Without) --- -->
<!-- v1 <- make_volcano_plot("GSEA", "Without", threshold_y = 1.9, threshold_x = 1.8, show_title = TRUE) + -->
<!--   labs(y = "-log10(adj. p-value)\nInflammatory \nSAUL_SEN_MAYO") + -->
<!--   labs(x = NULL) -->

<!-- v2 <- make_volcano_plot("ssGSEA", "Without", threshold_y = 4, threshold_x = 0.18, show_title = TRUE) + -->
<!--   labs(y = NULL, x = NULL) -->

<!-- v3 <- make_volcano_plot("logmedian", "Without", threshold_y = 3, threshold_x = 0.18, show_title = TRUE) + -->
<!--   labs(y = NULL, x = NULL) -->

<!-- v4 <- make_volcano_plot("ranking", "Without", threshold_y = 3, threshold_x = 0.18, show_title = TRUE) + -->
<!--   labs(y = NULL, x = NULL) -->


<!-- # --- VOLCANO: NON-INFLAMMATORY (With) --- -->
<!-- v5 <- make_volcano_plot("GSEA", "With", threshold_y = 1.5, threshold_x = 1.8) + -->
<!--   labs(y = "-log10(adj. p-value)\nNon-inflammatory \nSAUL_SEN_MAYO") -->

<!-- v6 <- make_volcano_plot("ssGSEA", "With", threshold_y = 3, threshold_x = 0.18) + -->
<!--   labs(y = NULL) -->

<!-- v7 <- make_volcano_plot("logmedian", "With", threshold_y = 2, threshold_x = 0.18) + -->
<!--   labs(y = NULL) -->

<!-- v8 <- make_volcano_plot("ranking", "With", threshold_y = 5, threshold_x = 0.18) + -->
<!--   labs(y = NULL) -->


<!-- # --- EFFECT-SIZE SCATTER --- -->
<!-- p1 <- make_effect_plot(df_wide, "GSEA") -->

<!-- p2 <- make_effect_plot(df_wide, "ssGSEA")  -->

<!-- p3 <- make_effect_plot(df_wide, "logmedian") + -->
<!--   labs(y = NULL) -->

<!-- p4 <- make_effect_plot(df_wide, "ranking") + -->
<!--   labs(y = NULL) -->


<!-- # --- P-VALUE SCATTER --- -->
<!-- p5 <- make_pval_plot(df_wide, "GSEA") -->

<!-- p6 <- make_pval_plot(df_wide, "ssGSEA")   -->

<!-- p7 <- make_pval_plot(df_wide, "logmedian") + -->
<!--   labs(y = NULL) -->

<!-- p8 <- make_pval_plot(df_wide, "ranking") + -->
<!--   labs(y = NULL) -->


<!-- # ---------------------------------------------------------- -->
<!-- # Combine into 4x4 panel with shared legend -->
<!-- # ---------------------------------------------------------- -->
<!-- final_plot <-  ggarrange(v1,v2,v3,v4, -->
<!--                          v5,v6,v7,v8, -->
<!--                          p1,p2,p3,p4, -->
<!--                          p5,p6,p7,p8, -->
<!--                          common.legend = T, legend = "bottom" -->
<!-- ) -->

<!-- final_plot -->
<!-- ``` -->


```{r  fig.width=14, fig.height=4}
library(ggplot2)
library(cowplot)
library(ggpubr)

# --- Dummy data for legends ---
senmayo_tissues <- data.frame(
  tissue = c("Artery - Tibial", "Brain - Anterior cingulate cortex (BA24)", "Brain - Hippocampus",
             "Colon - Sigmoid", "Minor Salivary Gland", "Muscle - Skeletal", 
             "Nerve - Tibial", "Prostate")
)

hernandez_tissues <- data.frame(
  tissue = c("Artery - Aorta", "Breast - Mammary Tissue", "Cells - Cultured fibroblasts", "Thyroid")
)

other_method <- data.frame(
  method = c("GSEA", "Score-based")
)

# --- Colors ---
senmayo_colors <- c("#E69E9E","#7FBFFF","#4DCC99","#FFFF7F","#FF9F4D","#4A4141","#99FFCC","#FF5F5F")
hernandez_colors <- c("#D7B9FF","#B48CFF","#9660FF","#583896")
method_colors <- c("#B04975","#49B0AB")

# --- Dummy plots with 4-row legends and title above ---
p_senmayo <- ggplot(senmayo_tissues, aes(x = tissue, y = 1, color = tissue)) +
  geom_point(show.legend = TRUE, size=6) +
  scale_color_manual(values = setNames(senmayo_colors, senmayo_tissues$tissue),
                    name = "SAUL_SEN_MAYO \nselected tissues") +
  guides(color = guide_legend(nrow = 4, byrow = TRUE, title.position = "left")) +
  theme_minimal() +
  theme(axis.text = element_blank(), axis.title = element_blank(), axis.ticks = element_blank(),
        legend.position = "bottom")

p_hernandez <- ggplot(hernandez_tissues, aes(x = tissue, y = 1, color = tissue)) +
  geom_point(show.legend = TRUE, size=6) +
  scale_color_manual(values = setNames(hernandez_colors, hernandez_tissues$tissue),
                    name = "HernandezSegura \nselected tissues") +
  guides(color = guide_legend(nrow = 4, byrow = TRUE, title.position = "left")) +
  theme_minimal() +
  theme(axis.text = element_blank(), axis.title = element_blank(), axis.ticks = element_blank(),
        legend.position = "bottom")

p_method <- ggplot(other_method, aes(x = method, y = 1, color = method)) +
  geom_point(show.legend = TRUE, size=6) +
  scale_color_manual(values = setNames(method_colors, other_method$method),
                    name = "Other tissues \n(method used)") +
  guides(color = guide_legend(nrow = 2, byrow = TRUE, title.position = "left")) +
  theme_minimal() +
  theme(axis.text = element_blank(), axis.title = element_blank(), axis.ticks = element_blank(),
        legend.position = "bottom")

# --- Extract legends ---
leg_senmayo <- ggpubr::as_ggplot(ggpubr::get_legend(p_senmayo))
leg_hernandez <- ggpubr::as_ggplot(ggpubr::get_legend(p_hernandez))
leg_method <- ggpubr::as_ggplot(ggpubr::get_legend(p_method))

 
# --- Add a small spacer ---
spacer <- ggplot() + theme_void()

# --- Combine legends with a spacer ---
combined_legend <- ggarrange(
  leg_senmayo,
  spacer,                 # adds space between SenMayo and Hernandez-Segura
  leg_hernandez,
  leg_method,
  nrow = 1,
  widths = c(1, 0.05, 1, 0.5)  # adjust the spacer height (0.1) as needed
)

combined_legend
```

```{r fig.width=14, fig.height=10}
library(dplyr)
library(tidyr)
library(ggplot2)
library(ggrepel)
library(patchwork)

# ----------------------------------------------------------
# Tissues to highlight
# ----------------------------------------------------------
my_tissues <- c(
  "Artery - Tibial",
  "Brain - Anterior cingulate cortex (BA24)",
  "Brain - Hippocampus",
  "Colon - Sigmoid",
  "Minor Salivary Gland",
  "Muscle - Skeletal",
  "Nerve - Tibial",
  "Prostate",
  "Artery - Aorta",
  "Breast - Mammary Tissue",
  "Cells - Cultured fibroblasts",
  "Thyroid"
)

tissue_colors <- c(
  "Artery - Tibial" = "#E69E9E",
  "Brain - Anterior cingulate cortex (BA24)" = "#7FBFFF",
  "Brain - Hippocampus" = "#4DCC99",
  "Colon - Sigmoid" = "#FFFF7F",
  "Minor Salivary Gland" = "#FF9F4D",
  "Muscle - Skeletal" = "#4A4141",
  "Nerve - Tibial" = "#99FFCC",
  "Prostate" = "#FF5F5F",

  # four shades of the SAME color family (lavender/purple)
  "Artery - Aorta" = "#D7B9FF",    
  "Breast - Mammary Tissue" = "#B48CFF",  # light lavender
  "Cells - Cultured fibroblasts" = "#9660FF",  # medium lavender
  "Thyroid" = "#583896"   # deeper lavender
)
# ----------------------------------------------------------
# Absolute NES version
# ----------------------------------------------------------
results_combined_absNES <- results_combined
results_combined_absNES$EffectSize <- abs(results_combined_absNES$EffectSize)

# Assign point color: use tissue color if in my_tissues, otherwise method-based
# Fix point_color as character
results_combined_absNES$point_color <- ifelse(
  results_combined_absNES$tissue %in% my_tissues,
  as.character(results_combined_absNES$tissue),    # tissue name for color
  ifelse(results_combined_absNES$method == "GSEA", "Other_GSEA", "Other_Other")
)

# Color mapping
all_colors <- c(tissue_colors,
                "Other_GSEA" = "#B04975",
                "Other_Other" = "#49B0AB")

# ----------------------------------------------------------
# Pivot wider for scatter plots
# ----------------------------------------------------------
df_wide <- results_combined_absNES %>%
  pivot_wider(
    id_cols = c(tissue, method, point_color),
    names_from = inflammation,
    values_from = c(EffectSize, PValue)
  )

# ----------------------------------------------------------
# Helper: Effect size label
# ----------------------------------------------------------
get_effect_label <- function(method_name) {
  if (method_name == "GSEA") return("|NES|") else return("Cohen's f")
}

# ----------------------------------------------------------
# Volcano plot
# ----------------------------------------------------------
make_volcano_plot <- function(method_name, infl_group, threshold_y = 0, threshold_x = 0, show_title = FALSE) {
  
  df_sub <- results_combined_absNES %>%
    filter(method == method_name, inflammation == infl_group)
  
  df_lab <- df_sub %>%
    filter(tissue %in% my_tissues & (-log10(PValue) >= threshold_y | EffectSize >= threshold_x))
  
  ggplot(df_sub, aes(x = EffectSize, y = -log10(PValue), color = point_color)) +
    geom_point(size = 4, alpha=0.7) +
    # geom_text_repel(
    #   data = df_lab,
    #   aes(label = tissue),
    #   color = "black",
    #   size = 3,
    #   hjust = 1.1
    # ) +
    scale_color_manual(
      values = all_colors,
      breaks = my_tissues,   # show only tissues of interest in legend
      name = NULL
    ) +
    labs(
      x = get_effect_label(method_name),
      y = "-log10(adj. p-value)",
      title = ifelse(show_title, method_name, "")
    ) +
    theme_minimal(base_size = 12) +
    theme(plot.title = element_text(hjust = 0.5))+
    geom_hline(yintercept = -log10(0.05), linetype="dashed", color="#A6A3A1")
}

# ----------------------------------------------------------
# Effect-size scatter plot
# ----------------------------------------------------------
make_effect_plot <- function(df, method_name, show_title = TRUE) {
  df_sub <- df %>% filter(method == method_name)
  df_lab <- df_sub %>% filter(tissue %in% my_tissues)
  
  ggplot(df_sub, aes(x = EffectSize_Without, y = EffectSize_With, color = point_color)) +
    geom_point(size = 4, alpha=0.7) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
    #geom_text_repel(data = df_lab, aes(label = tissue), color = "black", size = 3) +
    scale_color_manual(
      values = all_colors,
      breaks = my_tissues,
      name = NULL
    ) +
    labs(
      x = paste0(get_effect_label(method_name), " \nNon-inflammatory \nSAUL_SEN_MAYO"),
      y = paste0(get_effect_label(method_name), " \nInflammatory \nSAUL_SEN_MAYO"),
      title = ifelse(show_title, method_name, "")
    ) +
    theme_minimal(base_size = 12) +
    theme(plot.title = element_text(hjust = 0.5))
}

# ----------------------------------------------------------
# P-value scatter plot
# ----------------------------------------------------------
make_pval_plot <- function(df, method_name) {
  df_sub <- df %>% filter(method == method_name)
  df_lab <- df_sub %>% filter(tissue %in% my_tissues)
  
  ggplot(df_sub, aes(x = -log10(PValue_Without), y = -log10(PValue_With), color = point_color)) +
    geom_point(size = 4, alpha=0.7) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
    #geom_text_repel(data = df_lab, aes(label = tissue), color = "black", size = 3) +
    scale_color_manual(
      values = all_colors,
      breaks = my_tissues,
      name = NULL
    ) +
    labs(
      x = "-log10(adj. p-value) \nNon-inflammatory \nSAUL_SEN_MAYO",
      y = "-log10(adj. p-value) \nInflammatory \nSAUL_SEN_MAYO"
    ) +
    theme_minimal(base_size = 12) +
    theme(plot.title = element_text(hjust = 0.5)) +
    geom_vline(xintercept = -log10(0.05), linetype="dashed", color="#A6A3A1")+
    geom_hline(yintercept = -log10(0.05), linetype="dashed", color="#A6A3A1")
}

# ----------------------------------------------------------
# Build all 16 plots (with correct labeling)
# ----------------------------------------------------------

# --- VOLCANO: INFLAMMATORY (Without) ---
v1 <- make_volcano_plot("GSEA", "Without", threshold_y = 1.9, threshold_x = 1.8, show_title = TRUE) +
  labs(y = "-log10(adj. p-value)\nInflammatory \nSAUL_SEN_MAYO") +
  labs(x = NULL)

v2 <- make_volcano_plot("ssGSEA", "Without", threshold_y = 4, threshold_x = 0.18, show_title = TRUE) +
  labs(y = NULL, x = NULL)

v3 <- make_volcano_plot("logmedian", "Without", threshold_y = 3, threshold_x = 0.18, show_title = TRUE) +
  labs(y = NULL, x = NULL)

v4 <- make_volcano_plot("ranking", "Without", threshold_y = 3, threshold_x = 0.18, show_title = TRUE) +
  labs(y = NULL, x = NULL)


# --- VOLCANO: NON-INFLAMMATORY (With) ---
v5 <- make_volcano_plot("GSEA", "With", threshold_y = 1.5, threshold_x = 1.8) +
  labs(y = "-log10(adj. p-value)\nNon-inflammatory \nSAUL_SEN_MAYO")

v6 <- make_volcano_plot("ssGSEA", "With", threshold_y = 3, threshold_x = 0.18) +
  labs(y = NULL)

v7 <- make_volcano_plot("logmedian", "With", threshold_y = 2, threshold_x = 0.18) +
  labs(y = NULL)

v8 <- make_volcano_plot("ranking", "With", threshold_y = 5, threshold_x = 0.18) +
  labs(y = NULL)


# --- EFFECT-SIZE SCATTER ---
p1 <- make_effect_plot(df_wide, "GSEA")

p2 <- make_effect_plot(df_wide, "ssGSEA") 

p3 <- make_effect_plot(df_wide, "logmedian") +
  labs(y = NULL)

p4 <- make_effect_plot(df_wide, "ranking") +
  labs(y = NULL)


# --- P-VALUE SCATTER ---
p5 <- make_pval_plot(df_wide, "GSEA")

p6 <- make_pval_plot(df_wide, "ssGSEA")  

p7 <- make_pval_plot(df_wide, "logmedian") +
  labs(y = NULL)

p8 <- make_pval_plot(df_wide, "ranking") +
  labs(y = NULL)


# ----------------------------------------------------------
# Combine into 4x4 panel with shared legend
# ----------------------------------------------------------
 

# ----------------------------------------------------------
# Remove individual plot legends
# ----------------------------------------------------------
remove_legend <- function(p) {
  p + theme(legend.position = "none")
}

plots_no_legend <- lapply(list(v1,v2,v3,v4,
                               v5,v6,v7,v8,
                               p1,p2,p3,p4,
                               p5,p6,p7,p8), remove_legend)

# ----------------------------------------------------------
# Combine plots in 4x4 layout
# ----------------------------------------------------------
main_grid <- wrap_plots(plots_no_legend, ncol = 4)

# ----------------------------------------------------------
# Add the combined legend at the bottom
# ----------------------------------------------------------
final_plot_with_legend <- main_grid / combined_legend + 
  plot_layout(heights = c(20, 4))  # main grid taller than legend

final_plot_with_legend
```
 

```{r fig.width=12, fig.height=6}
(plt_senmayo_invitro_alt <- ggarrange(plt_senmayo_scores,plotNESlollipop_senmayo, nrow=2, heights=c(0.65,0.3), labels=c("A","B")))
```



```{r fig.width=13, fig.height= 20}
(plt_senmayo_custom_alt <- ggarrange(plt_senmayo_invitro_alt, final_plot_with_legend, labels=c("","C","D"), ncol=1, heights=c(0.45,0.85)))
```

```{r}
ggplot2::ggsave("../Figures/Figs/SupFigure8_GTEx_SenMayo_Custom_alternative.png",
                plt_senmayo_custom_alt,
                width = 13, height=20, bg = 'white')
ggplot2::ggsave("../Figures/Figs/SupFigure8_GTEx_SenMayo_Custom_alternative.pdf",
                plt_senmayo_custom_alt,
                width = 13, height=20, bg = 'white')
```





