---
title: "Response to Reviewers"
author: "Rita Martins-Silva"
date: "02/03/2026"
output: html_document
---


# Response to Reviewers {.tabset .tabset-pills}

## Set up {.tabset}

```{r}
setwd("../")
```
### Relevant Data

```{r}
signatures_bidirectional <- readRDS("../data/SenescenceSigntures_Bidirectional.rds")
t2gene2 <- fread("../data_aux/RefSeq_Transcripts_To_Genes.csv")
t2gene2 <- t2gene2[,c("name","name2")]
colnames(t2gene2) <- c("transcript","gene")
```

### Libraries

```{r}
library(data.table)
library("ggplot2")
library("colorspace")
library("scales")
#library("scater") 
library("reshape2")
#library("markeR")
#library("ggbreak")
#library("ggnewscale")
library("dplyr")
library("tidyr")  
library(patchwork)
library("ggpubr") 
library(purrr)
library(edgeR)
library(RColorBrewer)
library(pROC) 
#library("ggforce")
library(ggrepel)
#devtools::install_github("DiseaseTranscriptomicsLab/markeR@*release")
library(biomaRt)
library("tximport")
```

### Functions

Load markeR functions directly

```{r}
# Path to your folder with functions
fun_dir <- "functions_Sept18_2025/"

# Get all .R files in the folder
files <- list.files(paste0("../",fun_dir), pattern = "\\.R$", full.names = TRUE)

# Source them one by one
invisible(lapply(files, source))

```

### Colors

```{r}
Condition_colors <- c(
  "Senescent"     = "#F4A261",  # Soft orange (not goldenrod)
  "Proliferative" = "#A1C298",  # Desaturated sage green
  "Quiescent"     = "#9DB4C0"   # Dusty teal-gray
)

CellTypecols <- c(
  "Fibroblast"   = "#E69F00",  # Orange
  "Keratinocyte" = "#56B4E9",  # Sky Blue
  "Melanocyte"   = "#009E73",  # Bluish Green
  "Endothelial"  = "#F0E442",  # Yellow
  "Neuronal"     = "#0072B2",  # Blue
  "Mesenchymal"  = "#D55E00"   # Vermilion
)

CellTypecols_alt <- c(
  "Fibroblast"   = "#1b9e77",  # Teal green
  "Keratinocyte" = "#d95f02",  # Warm orange-brown
  "Melanocyte"   = "#7570b3",  # Muted purple
  "Endothelial"  = "#e7298a",  # Dark pink
  "Neuronal"     = "#66a61e",  # Olive green
  "Mesenchymal"  = "#e6ab02"   # Mustard
)


dataset_colors <- c(
  "HernandezSegura" = "#8DD3C7",
  "Mangelinck"      = "#FFFFB3",
  "Chan"            = "#BEBADA",
  "Purcell"         = "#FB8072",
  "Marthandan2016"  = "#80B1D3",
  "Marthandan2015"  = "#FDB462",
  "Wang"            = "#B3DE69",
  "Suda"            = "#FCCDE5",
  "McHugh"          = "#D9D9D9",
  "Savic"           = "#BC80BD",
  "Skea"            = "#CCEBC5",
  "Laurent"         = "#FFED6F",
  "Tasdemir"        = "#D0B7E1",
  "Lenain"          = "#FFB3BA",
  "Aarts"           = "#C2C2F0",
  "Casella"         = "#FFDAC1",
  "Numa"            = "#C6E2FF",
  "Admasu"          = "#F5CBA7",
  "Urata"           = "#C7CEEA",
  "Wang2023"        = "#E6B0AA",
  "Fu"              = "#A9DFBF",
  "Tanke"           = "#F9E79F",
  "Bawens"          = "#AED6F1"
)

geneSet_colors <- brewer.pal(9, "Set1")  # 9 distinct colors, mostly bright and distinct

```

```{r}
show_col(Condition_colors, ncol=3)
show_col(CellTypecols)
show_col(dataset_colors)
show_col(geneSet_colors)
```


## Comment 3.3 — External validation on an independent dataset

R.: Apply markeR's benchmarking mode to a senescence dataset not included in the original compendium, to evaluate whether HernandezSegura and SAUL_SEN_MAYO retain their discriminatory performance in a fully independent evaluation set. 



### GSE297560 (Space Radiation)
HiSat2 + feature counts, hg38

#### Data

```{r}
data <- as.data.frame(fread("../data_aux/GSE297560_counts.txt")[,-c(2:6)])
row.names(data) <- data$Geneid
data$Geneid <-  NULL
head(data)
# Convert row names from ENSEMBL ID to Gene Name using biomaRt

# Connect to Ensembl
mart <- useMart("ensembl", dataset = "hsapiens_gene_ensembl")

# Query the conversion
gene_conversion <- getBM(
  attributes = c("ensembl_gene_id", "hgnc_symbol"),
  filters    = "ensembl_gene_id",
  values     = rownames(data),
  mart       = mart
)

# Replace rownames — keeping only genes with a match and a non-empty symbol
gene_conversion <- gene_conversion[!duplicated(gene_conversion$hgnc_symbol), ]
data <- data[gene_conversion$ensembl_gene_id, ]
rownames(data) <- gene_conversion$hgnc_symbol
#remove .trimmed from sample names
colnames(data) <- gsub("\\.trimmed", "", colnames(data))

```


- Healthy_CTRL: Low-passage fibroblasts that were never shipped. Represent the true untreated baseline but are not the appropriate matched control for irradiation comparisons, as any differences relative to irradiated samples would conflate shipping stress with radiation effects.

- MOCK_CTRL: Fibroblasts that were shipped to the radiation facility but not irradiated, paired with the charged particle irradiation experiments (Hydrogen, Iron, Silicon). Control for the confounding effect of shipping stress, isolating the effect of radiation.

- MOCKgamma_CTRL: Identical to MOCK_CTRL but paired specifically with the gamma irradiation experiment. Kept separate to maintain matched experimental pairing.

```{r}
# Load metadata
metadata <- read.csv("../data_aux/GSE297560_metadata.csv")
metadata <- metadata[,c("Run","Sample.Name","source_name", "treatment")]
metadata$datasetID <- "GSE297560"

metadata$condition <- ifelse(grepl("control", metadata$treatment, ignore.case = TRUE), "Proliferative", ifelse(grepl("irradiation", metadata$treatment, ignore.case = TRUE), "Senescent", "Proliferative"))


# Attempt to match SRR with colnames
sample_to_srr <- c(
  "Healthy_CTRL_1_HISAT2"   = "SRR33626854",#
  "Healthy_CTRL_2_HISAT2"   = "SRR33626853",#
  "Healthy_CTRL_3_HISAT2"   = "SRR33626852",#
  "4Hydrogen_1_HISAT2"      = "SRR33626842",#
  "4Hydrogen_2_HISAT2"      = "SRR33626841",#
  "4Hydrogen_3_HISAT2"      = "SRR33626840",#
  "4Iron_1_HISAT2"          = "SRR33626845",#
  "4Iron_2_HISAT2"          = "SRR33626844",#
  "4Iron_3_HISAT2"          = "SRR33626843",#
  "4Silicon_1_HISAT2"       = "SRR33626848",#
  "4Silicon_2_HISAT2"       = "SRR33626847",#
  "4Silicon_3_HISAT2"       = "SRR33626846",#
  "MOCK_CTRL_1_HISAT2"      = "SRR33626857",#
  "MOCK_CTRL_2_HISAT2"      = "SRR33626856",#
  "MOCK_CTRL_3_HISAT2"      = "SRR33626855",#
  "4Gamma_1_HISAT2"         = "SRR33626839",#
  "4Gamma_3_HISAT2"         = "SRR33626838",#
  "4Gamma_2_HISAT2"         = "SRR33626837",#
  "MOCKgamma_CTRL_1_HISAT2" = "SRR33626851",
  "MOCKgamma_CTRL_2_HISAT2" = "SRR33626850",
  "MOCKgamma_CTRL_3_HISAT2" = "SRR33626849"
)

# Rename columns of your expression matrix
colnames(data) <- sample_to_srr[colnames(data)]

# Reorder metadata to match column order of your_df
metadata <- metadata[match(colnames(data), metadata$Run), ]

treatment_map <- c(
  "Low passage\\, unshipped"                            = "Healthy_CTRL",
  "Irradiation control shipped not irradiated"          = "MOCK_CTRL",
  "Gamma irradiation control shipped not irradiated"    = "MOCKgamma_CTRL",
  "4Gy Gamma irradiation"                               = "4Gy_Gamma",
  "4Gy Hydrogen irradiation"                            = "4Gy_Hydrogen",
  "4Gy Iron irradiation"                                = "4Gy_Iron",
  "4Gy Silicon irradiation"                             = "4Gy_Silicon"
)

metadata$treatment_simple <- treatment_map[metadata$treatment]
colnames(metadata)[1] <- "sampleID"
```



Remove Gamma and mock gamma, as in the paper the 4 gamma are not senescent...
Remove also Healthy_CTRL, they don't validate those samples in the paper either

```{r}
metadata <- metadata[!grepl("Gamma", metadata$treatment_simple, ignore.case = T), ]
metadata <- metadata[!grepl("Healthy_CTRL", metadata$treatment_simple, ignore.case = T), ]
data <- data[,metadata$sampleID]
```


```{r}
data
metadata
```

#### Preprocessing


##### Filtering

```{r}
# Plot
(plt_beforefilt <- ggplot() + aes(x=log10(rowSums(data))) +
   geom_density(fill="grey", alpha=0.5) + theme_bw() + 
   theme(axis.text.x = element_text(angle = 45, hjust=1),
         plot.title = element_text(hjust = 0.5),
         plot.subtitle = element_text(hjust = 0.5)) +
   ggtitle("Before Filtering") +
   xlab(expression(log[10]~"(counts per gene)")) +
   ylab("Frequency") + 
   geom_vline(xintercept = 2.9, linetype="dotted",color = "black", size=1.5))

```


```{r}
filtdata <- data[rowSums(data) > 10^2.9,]
nrow(data)
nrow(filtdata)


cat("Number of genes before filtering:", nrow(data), "\n")
cat("Number of genes after filtering:", nrow(filtdata), "\n")

```


```{r}

# Plot
(plt_afterfilt <- ggplot() + aes(x=log10(rowSums(filtdata))) +
   geom_density(fill="grey", alpha=0.5) + theme_bw() + 
   theme(axis.text.x = element_text(angle = 45, hjust=1),
         plot.title = element_text(hjust = 0.5),
         plot.subtitle = element_text(hjust = 0.5)) +
   ggtitle("After Filtering") +
   xlab(expression(log[10]~"(counts per gene)")) +
   ylab("Frequency") )

```



##### Normalisation

```{r fig.width=6, fig.height=4}

filtcounts_ggplot <- stack(log2(filtdata))
colnames(filtcounts_ggplot) <- c("expression", "sampleID")
filtcounts_ggplot <- merge(filtcounts_ggplot,metadata, by="sampleID")


(plt_beforenorm <- ggplot(filtcounts_ggplot, aes(sampleID,expression)) + geom_boxplot(alpha=0.4)+ xlab("Samples") + ylab("log2(counts)")   +  theme(axis.text.x = element_text(angle = 45, hjust=1))+
    theme_bw()+ 
    theme(plot.title = element_text(hjust = 0.5),
          plot.subtitle = element_text(hjust = 0.5),
          axis.text.x=element_blank(),
          axis.ticks.x=element_blank()) +
    ggtitle("Before Normalisation") )

```


```{r}
dgelist_normcounts <- DGEList(filtdata) # the matrix with gene names as row names and counts in columns
dgelist_normcounts <- calcNormFactors(dgelist_normcounts, method = "TMM") 
normalised_counts <- as.data.frame(voom(dgelist_normcounts,plot=F)$E) 
#normalised_counts <- as.data.frame(edgeR::cpm(dgelist_normcounts, log = FALSE)) 
normdata <- 2^normalised_counts
lognormdata <- log2(normdata+1)
```


```{r fig.width=6, fig.height=4}

normcounts_ggplot <- stack(log2(normdata))
colnames(normcounts_ggplot) <- c("expression", "sampleID")
normcounts_ggplot <- merge(normcounts_ggplot,metadata, by="sampleID")


(plt_afternorm <- ggplot(normcounts_ggplot, aes(sampleID,expression)) + geom_boxplot(alpha=0.4)+ xlab("Samples") + ylab("log2(counts)")   +  theme(axis.text.x = element_text(angle = 45, hjust=1))+
    theme_bw()+ 
    theme(plot.title = element_text(hjust = 0.5),
          plot.subtitle = element_text(hjust = 0.5),
          axis.text.x=element_blank(),
          axis.ticks.x=element_blank()) +
    ggtitle("After Normalisation") )

```


##### PCA

```{r fig.width=6, fig.height=4}
y <- DGEList(lognormdata, samples= metadata)
PCA  <- prcomp(t(y$counts), scale=FALSE, center=TRUE)
PCACounts  <- PCA$x
PCACounts  <- as.data.frame(PCACounts )
PCAData <-  cbind(PCACounts[1:10],y$samples) 

ev = PCA$sdev^2 
pc1= round(100*ev[1]/sum(ev),2) 
pc2 = round(100*ev[2]/sum(ev),2) 
pc3 = round(100*ev[3]/sum(ev),2) 
pc4 = round(100*ev[4]/sum(ev),2) 
pc5 = round(100*ev[5]/sum(ev),2) 
pc6 = round(100*ev[6]/sum(ev),2)  


(pltpca_1 <- ggplot(PCAData, aes(PC1,PC2))+
    geom_point(aes(color=treatment_simple), size=5, alpha=0.6)+ 
    xlab(paste0("PC1: ",pc1,"% variance")) +
    ylab(paste0("PC2: ",pc2,"% variance")) + 
    theme_bw()+
    guides(color = guide_legend(nrow = 2)) +
    theme(legend.position="top", legend.box="vertical", legend.title = element_blank(), legend.margin=margin())  + 
    geom_vline(xintercept=0, linetype="dotted") + 
    geom_hline(yintercept=0, linetype="dotted") )

(pltpca_2 <- ggplot(PCAData, aes(PC1,PC2))+
    geom_point(aes(color=condition), size=5, alpha=0.6)+ 
    xlab(paste0("PC1: ",pc1,"% variance")) +
    ylab(paste0("PC2: ",pc2,"% variance")) + 
    theme_bw()+
    theme(legend.position="top", legend.box="vertical", legend.title = element_blank(), legend.margin=margin())  + 
    geom_vline(xintercept=0, linetype="dotted") + 
    guides(color = guide_legend(nrow = 2)) +
    geom_hline(yintercept=0, linetype="dotted") +
    scale_color_manual(values=Condition_colors))


```

#### Calculate Scores

```{r}
(pltscores <- PlotScores(data = normdata, 
                         metadata = metadata, 
                         gene_sets = signatures_bidirectional[c("HernandezSegura", "SAUL_SEN_MAYO")],
                         Variable = "condition",
                         ColorVariable = "treatment_simple",
                         method = "logmedian",
                         xlab = "Cellular State", 
                         labsize = 8))
```

```{r fig.width=11, fig.height=5}
plt_filt <- ggarrange(plt_beforefilt,plt_afterfilt, nrow=1)
plt_norm <- ggarrange(plt_beforenorm,plt_afternorm, nrow=1)
plt_pca <- ggarrange(pltpca_1,pltpca_2, nrow=1)

plt1 <- ggarrange(plt_filt, plt_norm, nrow=1)
plt2 <- ggarrange(plt_pca, pltscores, nrow=1, widths=c(0.5,0.5))

plt_GSE297560 <- ggarrange(plt1, plt2, ncol=1, heights = c(0.4,0.7))
plt_GSE297560 <- annotate_figure(plt_GSE297560, top = text_grob("GSE297560: Space Radiation\n", face = "bold", size = 14))
plt_GSE297560


```

### GSE288940 (IMR90 + Eto)

My quantification pipeline 

#### Data

```{r}
# Get a list of the paths to the samples. Each sample is inside its own folder, under the name "quant.sf"
listfolders <- list.dirs("../data_aux/kallisto/Farr/")[-1] # remove "." directory
listfolders <- paste0(listfolders,"/abundance.tsv")
listfolders

# Naming the files (important so that when using tximport each column has the corresponding sample name)
samples <- gsub("../data_aux/kallisto/Farr//","",listfolders)
samples <- gsub("/abundance.tsv","",samples)
names(listfolders) <- samples
listfolders

# Merge all counts in an txi object. The argument txOut=TRUE allows to avoid gene-level summarization (we want the transcripts)
txi.kallisto <- tximport(listfolders, type = "kallisto", txOut=FALSE, tx2gene = t2gene2)

data <- as.data.frame(txi.kallisto$counts) 

```

```{r}
metadata <- as.data.frame(fread("../data_aux/GSE288940_metadata.csv"))
metadata <- metadata[,c("Run","Sample Name","cell_line","treatment")]
colnames(metadata)[1] <- "sampleID"


gsm_to_timepoint <- c(
  "GSM8779535" = "D0",
  "GSM8779536" = "D0",
  "GSM8779537" = "D0",
  "GSM8779538" = "D0",
  "GSM8779539" = "D0",
  "GSM8779540" = "D7",
  "GSM8779541" = "D7",
  "GSM8779542" = "D7",
  "GSM8779543" = "D7",
  "GSM8779544" = "D7",
  "GSM8779545" = "D7",
  "GSM8779546" = "D28",
  "GSM8779547" = "D28",
  "GSM8779548" = "D28",
  "GSM8779549" = "D28",
  "GSM8779550" = "D28",
  "GSM8779551" = "D28"
)

metadata$timepoint <- gsm_to_timepoint[metadata$`Sample Name`]
metadata$timepoint <- factor(metadata$timepoint, levels = c("D0", "D7", "D28"))

metadata$condition2 <- ifelse(grepl("D0", metadata$timepoint), "Proliferative", ifelse(grepl("D7", metadata$timepoint), "Prol > Sen", "Senescent"))

metadata$condition <- ifelse(grepl("D0", metadata$timepoint), "Proliferative",  "Senescent")

data <- data[,metadata$sampleID]
```


```{r}
data
metadata
```

#### Preprocessing


##### Filtering

```{r}
# Plot
(plt_beforefilt <- ggplot() + aes(x=log10(rowSums(data))) +
   geom_density(fill="grey", alpha=0.5) + theme_bw() + 
   theme(axis.text.x = element_text(angle = 45, hjust=1),
         plot.title = element_text(hjust = 0.5),
         plot.subtitle = element_text(hjust = 0.5)) +
   ggtitle("Before Filtering") +
   xlab(expression(log[10]~"(counts per gene)")) +
   ylab("Frequency") + 
   geom_vline(xintercept = 3.5, linetype="dotted",color = "black", size=1.5))

```


```{r}
filtdata <- data[rowSums(data) > 10^3.5,]
nrow(data)
nrow(filtdata)


cat("Number of genes before filtering:", nrow(data), "\n")
cat("Number of genes after filtering:", nrow(filtdata), "\n")

```


```{r}

# Plot
(plt_afterfilt <- ggplot() + aes(x=log10(rowSums(filtdata))) +
   geom_density(fill="grey", alpha=0.5) + theme_bw() + 
   theme(axis.text.x = element_text(angle = 45, hjust=1),
         plot.title = element_text(hjust = 0.5),
         plot.subtitle = element_text(hjust = 0.5)) +
   ggtitle("After Filtering") +
   xlab(expression(log[10]~"(counts per gene)")) +
   ylab("Frequency") )

```



##### Normalisation

```{r fig.width=6, fig.height=4}

filtcounts_ggplot <- stack(log2(filtdata))
colnames(filtcounts_ggplot) <- c("expression", "sampleID")
filtcounts_ggplot <- merge(filtcounts_ggplot,metadata, by="sampleID")


(plt_beforenorm <- ggplot(filtcounts_ggplot, aes(sampleID,expression)) + geom_boxplot(alpha=0.4)+ xlab("Samples") + ylab("log2(counts)")   +  theme(axis.text.x = element_text(angle = 45, hjust=1))+
    theme_bw()+ 
    theme(plot.title = element_text(hjust = 0.5),
          plot.subtitle = element_text(hjust = 0.5),
          axis.text.x=element_blank(),
          axis.ticks.x=element_blank()) +
    ggtitle("Before Normalisation") )

```


```{r}
dgelist_normcounts <- DGEList(filtdata) # the matrix with gene names as row names and counts in columns
dgelist_normcounts <- calcNormFactors(dgelist_normcounts, method = "TMM") 
normalised_counts <- as.data.frame(voom(dgelist_normcounts,plot=F)$E) 
#normalised_counts <- as.data.frame(edgeR::cpm(dgelist_normcounts, log = FALSE)) 
normdata <- 2^normalised_counts
lognormdata <- log2(normdata+1)
```


```{r fig.width=6, fig.height=4}

normcounts_ggplot <- stack(log2(normdata))
colnames(normcounts_ggplot) <- c("expression", "sampleID")
normcounts_ggplot <- merge(normcounts_ggplot,metadata, by="sampleID")


(plt_afternorm <- ggplot(normcounts_ggplot, aes(sampleID,expression)) + geom_boxplot(alpha=0.4)+ xlab("Samples") + ylab("log2(counts)")   +  theme(axis.text.x = element_text(angle = 45, hjust=1))+
    theme_bw()+ 
    theme(plot.title = element_text(hjust = 0.5),
          plot.subtitle = element_text(hjust = 0.5),
          axis.text.x=element_blank(),
          axis.ticks.x=element_blank()) +
    ggtitle("After Normalisation") )

```


##### PCA

```{r fig.width=6, fig.height=4}
y <- DGEList(lognormdata, samples= metadata)
PCA  <- prcomp(t(y$counts), scale=FALSE, center=TRUE)
PCACounts  <- PCA$x
PCACounts  <- as.data.frame(PCACounts )
PCAData <-  cbind(PCACounts[1:10],y$samples) 

ev = PCA$sdev^2 
pc1= round(100*ev[1]/sum(ev),2) 
pc2 = round(100*ev[2]/sum(ev),2) 
pc3 = round(100*ev[3]/sum(ev),2) 
pc4 = round(100*ev[4]/sum(ev),2) 
pc5 = round(100*ev[5]/sum(ev),2) 
pc6 = round(100*ev[6]/sum(ev),2)  

timepoint_colors <- c(
  "D0"  = "#AEC6E8",  # pastel blue
  "D7"  = "#C3A8D1",  # pastel purple
  "D28" = "#F4A6A0"   # pastel red/pink
)

(pltpca_1 <- ggplot(PCAData, aes(PC1,PC2))+
    geom_point(aes(color=timepoint), size=5, alpha=0.6)+ 
    xlab(paste0("PC1: ",pc1,"% variance")) +
    ylab(paste0("PC2: ",pc2,"% variance")) + 
    theme_bw()+
    guides(color = guide_legend(nrow = 1)) +
    theme(legend.position="top", legend.box="vertical", legend.title = element_blank(), legend.margin=margin())  + 
    geom_vline(xintercept=0, linetype="dotted") + 
    geom_hline(yintercept=0, linetype="dotted") +
    scale_color_manual(values=timepoint_colors))

(pltpca_2 <- ggplot(PCAData, aes(PC1,PC2))+
    geom_point(aes(color=condition), size=5, alpha=0.6)+ 
    xlab(paste0("PC1: ",pc1,"% variance")) +
    ylab(paste0("PC2: ",pc2,"% variance")) + 
    theme_bw()+
    theme(legend.position="top", legend.box="vertical", legend.title = element_blank(), legend.margin=margin())  + 
    geom_vline(xintercept=0, linetype="dotted") + 
    guides(color = guide_legend(nrow = 1)) +
    geom_hline(yintercept=0, linetype="dotted") +
    scale_color_manual(values=Condition_colors))


```

#### Calculate Scores

```{r}
(pltscores <- PlotScores(data = normdata, 
                         metadata = metadata, 
                         gene_sets = signatures_bidirectional[c("HernandezSegura", "SAUL_SEN_MAYO")],
                         Variable = "condition",
                         ColorVariable = "timepoint",
                         method = "logmedian",
                         ColorValues = timepoint_colors,
                         xlab = "Cellular State",
                         labsize = 8))
```

```{r fig.width=11, fig.height=5}
plt_filt <- ggarrange(plt_beforefilt,plt_afterfilt, nrow=1)
plt_norm <- ggarrange(plt_beforenorm,plt_afternorm, nrow=1)
plt_pca <- ggarrange(pltpca_1,pltpca_2, nrow=1)

plt1 <- ggarrange(plt_filt, plt_norm, nrow=1)
plt2 <- ggarrange(plt_pca, pltscores, nrow=1, widths=c(0.5,0.5))

plt_GSE288940 <- ggarrange(plt1, plt2, ncol=1, heights = c(0.4,0.7))
plt_GSE288940 <- annotate_figure(plt_GSE288940, top = text_grob("GSE288940: IMR90 + Etoposide\n", face = "bold", size = 14))
plt_GSE288940


```






### GSE287058 (Quiescent Fibroblasts)

counts tables were generated using HTSeq
Assembly: hg38

#### Data

```{r}

# List all files
files <- list.files(path = "../data_aux/GSE287058_RAW/", pattern = "\\.txt\\.gz$", full.names = TRUE)

# Read and combine
data <- lapply(files, function(f) {
  sample_name <- gsub("(GSM[0-9]+).*", "\\1", basename(f))
  df <- read.table(f, header = FALSE, col.names = c("gene", sample_name))
  df
}) %>%
  Reduce(function(x, y) full_join(x, y, by = "gene"), .) %>%
  tibble::column_to_rownames("gene")

data
```

```{r}
metadata <- as.data.frame(fread("../data_aux/GSE287058_metadata.csv"))
metadata <- metadata[,c("Run","Library Name","cell_line","treatment")]
colnames(metadata)[1] <- "sampleID"



# Reorder metadata to match column order of data
metadata <- metadata[match(colnames(data), metadata$`Library Name`), ]

# Verify order is correct before renaming
all(colnames(data) == metadata$`Library Name`)

# Then rename
colnames(data) <- metadata$sampleID

metadata$condition <- ifelse(grepl("Proliferating", metadata$treatment, ignore.case = TRUE), "Proliferative", "Quiescent")
metadata$condition2 <-  ifelse(metadata$condition == "Quiescent",
                               paste0(metadata$condition, "\n(", metadata$treatment, ")"),
                               metadata$condition)
metadata$condition2 <- gsub(" ","\n",metadata$condition2)

```

```{r}
data
metadata
```
Remove LF1 samples as they don't have controls

```{r}
metadata <- metadata[!grepl("LF1", metadata$cell_line), ]
data <- data[,metadata$sampleID]
```


```{r}
data
metadata
```


#### Preprocessing


##### Filtering

```{r}
# Plot
(plt_beforefilt <- ggplot() + aes(x=log10(rowSums(data))) +
   geom_density(fill="grey", alpha=0.5) + theme_bw() + 
   theme(axis.text.x = element_text(angle = 45, hjust=1),
         plot.title = element_text(hjust = 0.5),
         plot.subtitle = element_text(hjust = 0.5)) +
   ggtitle("Before Filtering") +
   xlab(expression(log[10]~"(counts per gene)")) +
   ylab("Frequency") + 
   geom_vline(xintercept = 3.2, linetype="dotted",color = "black", size=1.5))

```


```{r}
filtdata <- data[rowSums(data) > 10^3.2,]
nrow(data)
nrow(filtdata)


cat("Number of genes before filtering:", nrow(data), "\n")
cat("Number of genes after filtering:", nrow(filtdata), "\n")

```


```{r}

# Plot
(plt_afterfilt <- ggplot() + aes(x=log10(rowSums(filtdata))) +
   geom_density(fill="grey", alpha=0.5) + theme_bw() + 
   theme(axis.text.x = element_text(angle = 45, hjust=1),
         plot.title = element_text(hjust = 0.5),
         plot.subtitle = element_text(hjust = 0.5)) +
   ggtitle("After Filtering") +
   xlab(expression(log[10]~"(counts per gene)")) +
   ylab("Frequency") )

```



##### Normalisation

```{r fig.width=6, fig.height=4}

filtcounts_ggplot <- stack(log2(filtdata))
colnames(filtcounts_ggplot) <- c("expression", "sampleID")
filtcounts_ggplot <- merge(filtcounts_ggplot,metadata, by="sampleID")


(plt_beforenorm <- ggplot(filtcounts_ggplot, aes(sampleID,expression)) + geom_boxplot(alpha=0.4, outlier.size = 0.5)+ xlab("Samples") + ylab("log2(counts)")   +  theme(axis.text.x = element_text(angle = 45, hjust=1))+
    theme_bw()+ 
    theme(plot.title = element_text(hjust = 0.5),
          plot.subtitle = element_text(hjust = 0.5),
          axis.text.x=element_blank(),
          axis.ticks.x=element_blank()) +
    ggtitle("Before Normalisation") )

```


```{r}
dgelist_normcounts <- DGEList(filtdata) # the matrix with gene names as row names and counts in columns
dgelist_normcounts <- calcNormFactors(dgelist_normcounts, method = "TMM") 
normalised_counts <- as.data.frame(voom(dgelist_normcounts,plot=F)$E) 
#normalised_counts <- as.data.frame(edgeR::cpm(dgelist_normcounts, log = FALSE)) 
normdata <- 2^normalised_counts
lognormdata <- log2(normdata+1)
```


```{r fig.width=6, fig.height=4}

normcounts_ggplot <- stack(log2(normdata))
colnames(normcounts_ggplot) <- c("expression", "sampleID")
normcounts_ggplot <- merge(normcounts_ggplot,metadata, by="sampleID")


(plt_afternorm <- ggplot(normcounts_ggplot, aes(sampleID,expression)) + geom_boxplot(alpha=0.4, outlier.size = 0.5)+ xlab("Samples") + ylab("log2(counts)")   +  theme(axis.text.x = element_text(angle = 45, hjust=1))+
    theme_bw()+ 
    theme(plot.title = element_text(hjust = 0.5),
          plot.subtitle = element_text(hjust = 0.5),
          axis.text.x=element_blank(),
          axis.ticks.x=element_blank()) +
    ggtitle("After Normalisation") )

```


##### PCA

```{r fig.width=6, fig.height=4}
y <- DGEList(lognormdata, samples= metadata)
PCA  <- prcomp(t(y$counts), scale=FALSE, center=TRUE)
PCACounts  <- PCA$x
PCACounts  <- as.data.frame(PCACounts )
PCAData <-  cbind(PCACounts[1:10],y$samples) 

ev = PCA$sdev^2 
pc1= round(100*ev[1]/sum(ev),2) 
pc2 = round(100*ev[2]/sum(ev),2) 
pc3 = round(100*ev[3]/sum(ev),2) 
pc4 = round(100*ev[4]/sum(ev),2) 
pc5 = round(100*ev[5]/sum(ev),2) 
pc6 = round(100*ev[6]/sum(ev),2)  

cell_line_colors <- c(
  "BJ"                    = "#6DBF8A",  # medium pastel green
  "IMR90"                 = "#F47C7C",  # medium pastel red
  "Primary Preadipocytes" = "#F0C060"   # medium pastel yellow
)

(pltpca_1 <- ggplot(PCAData, aes(PC1,PC2))+
    geom_point(aes(color=cell_line), size=5, alpha=0.6)+ 
    xlab(paste0("PC1: ",pc1,"% variance")) +
    ylab(paste0("PC2: ",pc2,"% variance")) + 
    theme_bw()+
    guides(color = guide_legend(nrow = 3)) +
    theme(legend.position="top", legend.box="vertical", legend.title = element_blank(), legend.margin=margin())  + 
    geom_vline(xintercept=0, linetype="dotted") + 
    geom_hline(yintercept=0, linetype="dotted")+
    scale_color_manual(values=cell_line_colors))

(pltpca_2 <- ggplot(PCAData, aes(PC1,PC2))+
    geom_point(aes(color=condition), size=5, alpha=0.6)+ 
    xlab(paste0("PC1: ",pc1,"% variance")) +
    ylab(paste0("PC2: ",pc2,"% variance")) + 
    theme_bw()+
    theme(legend.position="top", legend.box="vertical", legend.title = element_blank(), legend.margin=margin())  + 
    geom_vline(xintercept=0, linetype="dotted") + 
    guides(color = guide_legend(nrow = 3)) +
    geom_hline(yintercept=0, linetype="dotted") +
    scale_color_manual(values=Condition_colors))


```

#### Calculate Scores

```{r}
(pltscores <- PlotScores(data = normdata, 
                         metadata = metadata, 
                         gene_sets = signatures_bidirectional[c("HernandezSegura", "SAUL_SEN_MAYO")],
                         Variable = "condition2",
                         ColorVariable = "cell_line",
                         method = "logmedian",
                         xlab = "Cellular State",
                         ColorValues = cell_line_colors,
                         labsize = 8))
```


```{r fig.width=11, fig.height=5}
plt_filt <- ggarrange(plt_beforefilt,plt_afterfilt, nrow=1)
plt_norm <- ggarrange(plt_beforenorm,plt_afternorm, nrow=1)
plt_pca <- ggarrange(pltpca_1,pltpca_2, nrow=1)

plt1 <- ggarrange(plt_filt, plt_norm, nrow=1)
plt2 <- ggarrange(plt_pca, pltscores, nrow=1, widths=c(0.5,0.5))

plt_GSE287058 <- ggarrange(plt1, plt2, ncol=1, heights = c(0.4,0.7))
plt_GSE287058 <- annotate_figure(plt_GSE287058, top = text_grob("GSE287058: Quiescent Fibroblasts\n", face = "bold", size = 14))
plt_GSE287058


```

### Compile All

```{r fig.width=11, fig.height=16} 
add_margin <- function(plt) plt + theme(plot.margin = margin(t=15, r=5, b=15, l=5))

(plt_3_3 <- ggarrange(add_margin(plt_GSE297560), 
                      add_margin(plt_GSE288940), 
                      add_margin(plt_GSE287058), 
                      ncol=1, labels = c("A","B","C"), 
                      heights = c(0.35,0.35,0.4)))

```

```{r}
ggplot2::ggsave("../Figures/Figs/R3_3_InVitroValidation.png",
                plt_3_3,
                width = 11, height=16, bg = 'white')

ggplot2::ggsave("../Figures/Figs/R3_3_InVitroValidation.pdf",
                plt_3_3,
                width = 11, height=16, bg = 'white')
```


## Comment 1.6 — Applicability of markeR beyond senescence

R.: Apply markeR in discovery mode to a small set of additional biological phenotypes with well-established gene sets and positive control datasets, to demonstrate that the framework generalises beyond senescence. Phenotypes considered include cell proliferation, epithelial-to-mesenchymal transition, and viral/interferon response.



### Sex Signature | GTEx

```{r}
library(biomaRt)
mart <- useMart(biomart="ensembl", dataset="hsapiens_gene_ensembl")
y_genes <- getBM(attributes = c("chromosome_name", "hgnc_symbol"),
                 filters = "chromosome_name", values = "Y", mart = mart)

x_genes <- getBM(attributes = c("chromosome_name", "hgnc_symbol"),
                 filters = "chromosome_name", values = "X", mart = mart)

y_unique <- y_genes[!y_genes$hgnc_symbol %in% x_genes$hgnc_symbol, ]
y_unique <- y_unique[y_unique$hgnc_symbol != "", ]

geneset_Ychromosome <- list(Y_Chromosome = y_unique$hgnc_symbol)
```


```{r}
#  
# base_url <- "https://raw.githubusercontent.com/DiseaseTranscriptomicsLab/voyAGEr/main/Corrected_Counts/"
# 
tissues <- c(
  "Adipose - Subcutaneous",
  "Adipose - Visceral (Omentum)",
  "Adrenal Gland",
  "Artery - Aorta",
  "Artery - Coronary",
  "Artery - Tibial",
  "Brain - Amygdala",
  "Brain - Anterior cingulate cortex (BA24)",
  "Brain - Caudate (basal ganglia)",
  "Brain - Cerebellar Hemisphere",
  "Brain - Cerebellum",
  "Brain - Cortex",
  "Brain - Frontal Cortex (BA9)",
  "Brain - Hippocampus",
  "Brain - Hypothalamus",
  "Brain - Nucleus accumbens (basal ganglia)",
  "Brain - Putamen (basal ganglia)",
  "Brain - Spinal cord (cervical c-1)",
  "Brain - Substantia nigra",
  "Breast - Mammary Tissue",
  "Cells - Cultured fibroblasts",
  "Cells - EBV-transformed lymphocytes",
  "Colon - Sigmoid",
  "Colon - Transverse",
  "Esophagus - Gastroesophageal Junction",
  "Esophagus - Mucosa",
  "Esophagus - Muscularis",
  "Heart - Atrial Appendage",
  "Heart - Left Ventricle",
  "Kidney - Cortex",
  "Liver",
  "Lung",
  "Minor Salivary Gland",
  "Muscle - Skeletal",
  "Nerve - Tibial",
  "Ovary",
  "Pancreas",
  "Pituitary",
  "Prostate",
  "Skin - Not Sun Exposed (Suprapubic)",
  "Skin - Sun Exposed (Lower leg)",
  "Small Intestine - Terminal Ileum",
  "Spleen",
  "Stomach",
  "Testis",
  "Thyroid",
  "Uterus",
  "Vagina",
  "Whole Blood"
)
# 
# gtex <- lapply(tissues, function(tissue) {
#   url <- paste0(base_url, URLencode(paste0(tissue, "_corrected_logCPM.csv.gz")))
#   message("Loading: ", tissue)
#   tryCatch({
#     df <- as.data.frame(fread(url, data.table = FALSE))
#     rownames(df) <- df[[1]]
#     df[[1]] <- NULL
#     df
#   }, error = function(e) { warning("Failed: ", tissue, " — ", e$message); NULL })
# })
# 
# names(gtex) <- tissues
# saveRDS(gtex, "../data_aux/data_gtex.rds")
```

```{r}
#GTEx_alltissues <- readRDS("../data_aux/GTExV8_voyagercorrected.rds") # https://github.com/DiseaseTranscriptomicsLab/voyAGEr/tree/main/Corrected_Counts
data_gtex <- readRDS("../data_aux/data_gtex.rds")

metadata_GTEx_alltissues <- readRDS("../data_aux/GTExV8_metadata.rds") # Restricted access
metadata_GTEx_alltissues$SEX <- ifelse(metadata_GTEx_alltissues$SEX == 1,"Male","Female")
metadata_GTEx_alltissues
```

```{r}
table(metadata_GTEx_alltissues$SMTSD,
      metadata_GTEx_alltissues$SEX)
```



```{r fig.width=11, fig.height=16}
# Assuming your metadata df is called `metadata` with columns SMTSD, sample_id, SEX (or similar)

# Step 1: Calculate scores for all tissues
scores_list <- lapply(tissues, function(tissue) {
  message("Scoring: ", tissue)
  tryCatch({
    expr       <- 2^data_gtex[[tissue]]
    meta_sub   <- metadata_GTEx_alltissues[metadata_GTEx_alltissues$SMTSD == tissue, ]
    # Keep only samples present in expression matrix
    meta_sub   <- meta_sub[meta_sub[[1]] %in% colnames(expr), ]
    #order, just in case
    expr <- expr[, meta_sub[[1]]]
    if (nrow(meta_sub) == 0) { warning("No matching samples: ", tissue); return(NULL) }
    
    CalculateScores(
      data      = expr,
      metadata  = meta_sub,
      method    = "logmedian",
      gene_sets = geneset_Ychromosome
    )
  }, error = function(e) { warning("Failed: ", tissue, " — ", e$message); NULL })
})
names(scores_list) <- tissues


# Step 2: Plot male vs female per tissue
# Define sex-specific tissues to skip
sex_specific <- c("Ovary", "Uterus", "Vagina", "Prostate", "Testis")

# Assumes scores df has a column with the score values and SEX in metadata
# Adjust score_col and sex_col to match your actual column names
score_col <- "logmedian"   # column name in CalculateScores output
sex_col   <- "SEX"         # column in metadata (e.g. 1=Male, 2=Female or "Male"/"Female")

plots_list <- lapply(tissues, function(tissue) {
  if (tissue %in% sex_specific) return(NULL)
  
  df <- scores_list[[tissue]][["Y_Chromosome"]]
  if (is.null(df)) return(NULL)
  
  # Merge scores with sex info if not already present
  if (!sex_col %in% colnames(df)) {
    meta_sub <- metadata_GTEx_alltissues[metadata_GTEx_alltissues$SMTSD == tissue, c("SAMPID","SMTSD","SEX","AGE")]
    df <- merge(df, meta_sub, by = 1)
  } else {
    df <- df[,c("sample","score","SMTSD","SEX","AGE")]
    
  }
  
  #df[[sex_col]] <- factor(df[[sex_col]], levels = c(1, 2), labels = c("Male", "Female"))
  
  ggplot(df, aes(x = SEX, y = score, fill = SEX)) +
    geom_jitter(color = "#DDDDDD", width = 0.15) +
    geom_violin(alpha = 0.4) +
    stat_summary(fun = median, geom = "crossbar", width = 0.1, color = "black") +
    scale_fill_manual(values = c("Male" = "#4E8ECC", "Female" = "#CC4E7A")) +
    labs(title = tissue, x = NULL,
         y = expression(log[2]*"-median score")) +
    theme_bw() +
    theme(legend.position  = "none",
          plot.title      = element_text(hjust = 0.5, size = 10))
})
names(plots_list) <- tissues
plots_list <- Filter(Negate(is.null), plots_list)

# Combine into one figure (optional)
ggarrange(plotlist = plots_list, ncol = 6, nrow = ceiling(length(plots_list) / 6))
```
```{r fig.width=4, fig.height=3}
df_all_sex <- bind_rows(
  lapply(tissues[!tissues %in% sex_specific], function(tissue) {
    df <- scores_list[[tissue]][["Y_Chromosome"]]
    if (is.null(df)) return(NULL)
    df$tissue <- tissue
    df
  })
)


(plt_ychromosomescores <- ggplot(df_all_sex, aes(x = .data[[sex_col]], y = score, fill = .data[[sex_col]])) + 
    geom_jitter(color = "#DDDDDD", width = 0.3, alpha = 0.3, size = 0.5) +
    geom_violin(alpha = 0.5) +
    stat_summary(fun = median, geom = "crossbar", width = 0.1, color = "black") +
    scale_fill_manual(values = c("Male" = "#4E8ECC", "Female" = "#CC4E7A")) +
    labs(x = NULL, y = expression(log[2]*"-median score")) +
    theme_bw() +
    theme(legend.position = "none",
          plot.title = element_text(hjust = 0.5, size = 14),
          text = element_text(size = 14))  +
    ggtitle("Y Chromosome Gene Set (biomaRt)\nAcross GTEx Tissues"))
```

```{r fig.width=4, fig.height=3}
# compute sample counts per group
n_labels <- df_all_sex %>%
  group_by(.data[[sex_col]]) %>%
  summarise(n = n()) %>%
  mutate(label = paste0(.data[[sex_col]], "\n(n=", n, ")"))

label_vec <- setNames(n_labels$label, n_labels[[sex_col]])

(plt_ychromosomescores <- ggplot(df_all_sex, aes(x = .data[[sex_col]], y = score, fill = .data[[sex_col]])) + 
    geom_jitter(color = "#DDDDDD", width = 0.3, alpha = 0.3, size = 0.5) +
    geom_violin(alpha = 0.5) +
    stat_summary(fun = median, geom = "crossbar", width = 0.1, color = "black") +
    scale_fill_manual(values = c("Male" = "#4E8ECC", "Female" = "#CC4E7A")) +
    scale_x_discrete(labels = label_vec) +
    labs(x = NULL, y = expression(log[2]*"-median score")) +
    theme_bw() +
    theme(legend.position = "none",
          plot.title = element_text(hjust = 0.5, size = 14),
          text = element_text(size = 14)) +
    ggtitle("Y Chromosome Gene Set (biomaRt)\nAcross GTEx Tissues"))
```


### Respiratory Infection Signature | GSE161731 (many viral and bacterial infections)


```{r}
signature_respiratoryviral <- list(Zaas2009_RespiratoryViralResponse = as.vector(fread("../data_aux/RespiratoryViralResponse_Zaas2009_simple.txt",header=F)$V1))
```

```{r}
data <- as.data.frame(fread("../data_aux/GSE161731_counts.csv"))
row.names(data) <- data$V1
data$V1 <- NULL
# Convert row names from ENSEMBL ID to Gene Name using biomaRt

# Connect to Ensembl
mart <- useMart("ensembl", dataset = "hsapiens_gene_ensembl")

# Query the conversion
gene_conversion <- getBM(
  attributes = c("ensembl_gene_id", "hgnc_symbol"),
  filters    = "ensembl_gene_id",
  values     = rownames(data),
  mart       = mart
)

# Replace rownames — keeping only genes with a match and a non-empty symbol
gene_conversion <- gene_conversion[!duplicated(gene_conversion$hgnc_symbol), ]
data <- data[gene_conversion$ensembl_gene_id, ]
rownames(data) <- gene_conversion$hgnc_symbol

data
```

```{r}
metadata <- as.data.frame(fread("../data_aux/GSE161731_counts_key.csv"))
colnames(metadata)[1] <- "sampleID"
metadata <- metadata[metadata$sampleID %in% colnames(data),]
data <- data[,metadata$sampleID]
```

```{r}
data
metadata
```


##### Filtering

```{r}
# Plot
(plt_beforefilt <- ggplot() + aes(x=log10(rowSums(data))) +
   geom_density(fill="grey", alpha=0.5) + theme_bw() + 
   theme(axis.text.x = element_text(angle = 45, hjust=1),
         plot.title = element_text(hjust = 0.5),
         plot.subtitle = element_text(hjust = 0.5)) +
   ggtitle("Before Filtering") +
   xlab(expression(log[10]~"(counts per gene)")) +
   ylab("Frequency") + 
   geom_vline(xintercept = 4.2, linetype="dotted",color = "black", size=1.5))

```


```{r}
filtdata <- data[rowSums(data) > 10^4.2,]
nrow(data)
nrow(filtdata)


cat("Number of genes before filtering:", nrow(data), "\n")
cat("Number of genes after filtering:", nrow(filtdata), "\n")

```


```{r}

# Plot
(plt_afterfilt <- ggplot() + aes(x=log10(rowSums(filtdata))) +
   geom_density(fill="grey", alpha=0.5) + theme_bw() + 
   theme(axis.text.x = element_text(angle = 45, hjust=1),
         plot.title = element_text(hjust = 0.5),
         plot.subtitle = element_text(hjust = 0.5)) +
   ggtitle("After Filtering") +
   xlab(expression(log[10]~"(counts per gene)")) +
   ylab("Frequency") )

```



##### Normalisation

```{r fig.width=6, fig.height=4}

filtcounts_ggplot <- stack(log2(filtdata))
colnames(filtcounts_ggplot) <- c("expression", "sampleID")
filtcounts_ggplot <- merge(filtcounts_ggplot,metadata, by="sampleID")


(plt_beforenorm <- ggplot(filtcounts_ggplot, aes(sampleID,expression)) + geom_boxplot(alpha=0.4, outlier.size = 0.5)+ xlab("Samples") + ylab("log2(counts)")   +  theme(axis.text.x = element_text(angle = 45, hjust=1))+
    theme_bw()+ 
    theme(plot.title = element_text(hjust = 0.5),
          plot.subtitle = element_text(hjust = 0.5),
          axis.text.x=element_blank(),
          axis.ticks.x=element_blank()) +
    ggtitle("Before Normalisation") )

```


```{r}
dgelist_normcounts <- DGEList(filtdata) # the matrix with gene names as row names and counts in columns
dgelist_normcounts <- calcNormFactors(dgelist_normcounts, method = "TMM") 
normalised_counts <- as.data.frame(voom(dgelist_normcounts,plot=F)$E) 
#normalised_counts <- as.data.frame(edgeR::cpm(dgelist_normcounts, log = FALSE)) 
normdata <- 2^normalised_counts
lognormdata <- log2(normdata+1)
```


```{r fig.width=6, fig.height=4}

normcounts_ggplot <- stack(log2(normdata))
colnames(normcounts_ggplot) <- c("expression", "sampleID")
normcounts_ggplot <- merge(normcounts_ggplot,metadata, by="sampleID")


(plt_afternorm <- ggplot(normcounts_ggplot, aes(sampleID,expression)) + geom_boxplot(alpha=0.4, outlier.size = 0.5)+ xlab("Samples") + ylab("log2(counts)")   +  theme(axis.text.x = element_text(angle = 45, hjust=1))+
    theme_bw()+ 
    theme(plot.title = element_text(hjust = 0.5),
          plot.subtitle = element_text(hjust = 0.5),
          axis.text.x=element_blank(),
          axis.ticks.x=element_blank()) +
    ggtitle("After Normalisation") )

```


##### PCA

```{r fig.width=6, fig.height=4}
y <- DGEList(lognormdata, samples= metadata)
PCA  <- prcomp(t(y$counts), scale=FALSE, center=TRUE)
PCACounts  <- PCA$x
PCACounts  <- as.data.frame(PCACounts )
PCAData <-  cbind(PCACounts[1:10],y$samples) 

ev = PCA$sdev^2 
pc1= round(100*ev[1]/sum(ev),2) 
pc2 = round(100*ev[2]/sum(ev),2) 
pc3 = round(100*ev[3]/sum(ev),2) 
pc4 = round(100*ev[4]/sum(ev),2) 
pc5 = round(100*ev[5]/sum(ev),2) 
pc6 = round(100*ev[6]/sum(ev),2)  


(pltpca_1 <- ggplot(PCAData, aes(PC1,PC2))+
    geom_point(aes(color=cohort), size=5, alpha=0.6)+ 
    xlab(paste0("PC1: ",pc1,"% variance")) +
    ylab(paste0("PC2: ",pc2,"% variance")) + 
    theme_bw()+
    guides(color = guide_legend(nrow = 2)) +
    theme(legend.position="top", legend.box="vertical", legend.title = element_blank(), legend.margin=margin())  + 
    geom_vline(xintercept=0, linetype="dotted") + 
    geom_hline(yintercept=0, linetype="dotted") )

(pltpca_2 <- ggplot(PCAData, aes(PC1,PC2))+
    geom_point(aes(color=gender), size=5, alpha=0.6)+ 
    xlab(paste0("PC1: ",pc1,"% variance")) +
    ylab(paste0("PC2: ",pc2,"% variance")) + 
    theme_bw()+
    theme(legend.position="top", legend.box="vertical", legend.title = element_blank(), legend.margin=margin())  + 
    geom_vline(xintercept=0, linetype="dotted") + 
    guides(color = guide_legend(nrow = 2)) +
    geom_hline(yintercept=0, linetype="dotted"))

(pltpca_3 <- ggplot(PCAData, aes(PC1,PC2))+
    geom_point(aes(color=race), size=5, alpha=0.6)+ 
    xlab(paste0("PC1: ",pc1,"% variance")) +
    ylab(paste0("PC2: ",pc2,"% variance")) + 
    theme_bw()+
    theme(legend.position="top", legend.box="vertical", legend.title = element_blank(), legend.margin=margin())  + 
    geom_vline(xintercept=0, linetype="dotted") + 
    guides(color = guide_legend(nrow = 2)) +
    geom_hline(yintercept=0, linetype="dotted"))

(pltpca_4 <- ggplot(PCAData, aes(PC1,PC2))+
    geom_point(aes(color=batch), size=5, alpha=0.6)+ 
    xlab(paste0("PC1: ",pc1,"% variance")) +
    ylab(paste0("PC2: ",pc2,"% variance")) + 
    theme_bw()+
    theme(legend.position="top", legend.box="vertical", legend.title = element_blank(), legend.margin=margin())  + 
    geom_vline(xintercept=0, linetype="dotted") + 
    guides(color = guide_legend(nrow = 2)) +
    geom_hline(yintercept=0, linetype="dotted"))


```

#### Calculate Scores


```{r}
#saveRDS(normdata,"../data_aux/data_RespiratoryInfection.rds")
data_RespiratoryInfection <- readRDS("../data_aux/data_RespiratoryInfection.rds")
#saveRDS(metadata,"../data_aux/metadata_RespiratoryInfection.rds")
metadata_ResporatoryInfection <- readRDS("../data_aux/metadata_RespiratoryInfection.rds")
```


```{r fig.width=4, fig.height=3}
(dfscores <- CalculateScores(data = data_ResporatoryInfection, 
                             metadata = metadata_ResporatoryInfection, 
                             gene_sets = signature_respiratoryviral,
                             method = "logmedian")$Zaas2009_RespiratoryViralResponse)

dfscores$cohort2 <- ifelse(dfscores$cohort %in% c("CoV other","COVID-19"), "CoV", dfscores$cohort)
dfscores$cohort2 <- gsub("healthy","Healthy",dfscores$cohort2)
cohort_colors <- c(
  "CoV"       = "#E07B54",
  "Influenza" = "#5B8DB8",
  "Bacterial" = "#6BAF92",
  "Healthy"   = "#B0B0B0"
)

(plt_viralrespiratory <- ggplot(dfscores, aes(x = cohort2, y = score, fill=cohort2)) + 
    geom_jitter(color = "black", width = 0.3, alpha = 0.4, size = 0.5) +
    geom_violin(alpha = 0.7) +
    stat_summary(fun = median, geom = "crossbar", width = 0.1, color = "black") + 
    labs(x = NULL, y = expression(log[2]*"-median score")) +
    theme_bw() +
    theme(legend.position="none",
          plot.title = element_text(hjust = 0.5, size = 14),
          text = element_text(size = 14)) +
    ggtitle("Zaas 2009 Respiratory Viral Gene Set \nin Whole Blood Across \nAcute Respiratory Infections") + scale_fill_manual(values = cohort_colors))
```
```{r fig.width=4, fig.height=3}
n_labels <- dfscores %>%
  group_by(cohort2) %>%
  summarise(n = n()) %>%
  mutate(label = paste0(cohort2, "\n(n=", n, ")"))

label_vec <- setNames(n_labels$label, n_labels$cohort2)

(plt_viralrespiratory <- ggplot(dfscores, aes(x = cohort2, y = score, fill = cohort2)) + 
    geom_jitter(color = "black", width = 0.3, alpha = 0.4, size = 0.5) +
    geom_violin(alpha = 0.7) +
    stat_summary(fun = median, geom = "crossbar", width = 0.1, color = "black") + 
    labs(x = NULL, y = expression(log[2]*"-median score")) +
    theme_bw() +
    theme(legend.position = "none",
          plot.title = element_text(hjust = 0.5, size = 14),
          text = element_text(size = 14)) +
    ggtitle("Zaas 2009 Respiratory Viral Gene Set \nin Whole Blood Across \nAcute Respiratory Infections") +
    scale_fill_manual(values = cohort_colors) +
    scale_x_discrete(labels = label_vec))
```


### EMT transition | KIRC

```{r}
Ramaswamy_Metastasis <- list(Ramaswamy_Metastasis = as.data.frame(fread("../data_aux/RAMASWAMY_METASTASIS_simple.txt",header=T)))
```

https://xenabrowser.net/datapages/?dataset=TCGA-KIRC.star_counts.tsv&host=https%3A%2F%2Fgdc.xenahubs.net&removeHub=https%3A%2F%2Fxena.treehouse.gi.ucsc.edu%3A443

KIRC was chosen given that it has a relatively well documented tumour grade

```{r}
data <- as.data.frame(fread("../data_aux/TCGA-KIRC.star_counts.tsv"))
data$Ensembl_ID <- sub("\\..*", "", data$Ensembl_ID)

# combine rows with same gene id by summing them
data <- rowsum(data[, -1], group = data$Ensembl_ID)

# Connect to Ensembl
mart <- useMart("ensembl", dataset = "hsapiens_gene_ensembl")

# Query the conversion
gene_conversion <- getBM(
  attributes = c("ensembl_gene_id", "hgnc_symbol"),
  filters    = "ensembl_gene_id",
  values     = rownames(data),
  mart       = mart
)

# Replace rownames — keeping only genes with a match and a non-empty symbol
gene_conversion <- gene_conversion[!duplicated(gene_conversion$hgnc_symbol), ]
data <- data[gene_conversion$ensembl_gene_id, ]
rownames(data) <- gene_conversion$hgnc_symbol

data <- 2^data - 1 # unlog...
data

```

```{r}
metadata <- as.data.frame(fread("../data_aux/TCGA-KIRC.clinical.tsv"))
metadata <- metadata[metadata$sample %in% colnames(data),]
data <- data[,metadata$sample]
```


```{r}
metadata$code <- sub(".*-", "", metadata$sample)
table(metadata$code) # no metastasis
table(metadata$tissue_type.samples) # normal vs tumour
table(metadata$sample_type.samples, metadata$ajcc_pathologic_m.diagnoses)

# subset data to M1 or M0 only
# subset data to primary tumour only
metadata <- metadata[metadata$ajcc_pathologic_m.diagnoses %in% c("M0","M1") & metadata$sample_type.samples != "metadata$sample_type.samples" & metadata$tissue_type.samples == "Tumor",]

table(metadata$code) # no metastasis
table(metadata$tissue_type.samples) # normal vs tumour
table(metadata$sample_type.samples, metadata$ajcc_pathologic_m.diagnoses)

data <- data[,metadata$sample]

colnames(metadata)[1] <- "sampleID"
```


```{r}

data
metadata
```




##### Filtering

```{r}
# Plot
(plt_beforefilt <- ggplot() + aes(x=log10(rowSums(data))) +
   geom_density(fill="grey", alpha=0.5) + theme_bw() + 
   theme(axis.text.x = element_text(angle = 45, hjust=1),
         plot.title = element_text(hjust = 0.5),
         plot.subtitle = element_text(hjust = 0.5)) +
   ggtitle("Before Filtering") +
   xlab(expression(log[10]~"(counts per gene)")) +
   ylab("Frequency") + 
   geom_vline(xintercept = 5.2, linetype="dotted",color = "black", size=1.5))

```


```{r}
filtdata <- data[rowSums(data) > 10^5.2,]
nrow(data)
nrow(filtdata)


cat("Number of genes before filtering:", nrow(data), "\n")
cat("Number of genes after filtering:", nrow(filtdata), "\n")

```


```{r}

# Plot
(plt_afterfilt <- ggplot() + aes(x=log10(rowSums(filtdata))) +
   geom_density(fill="grey", alpha=0.5) + theme_bw() + 
   theme(axis.text.x = element_text(angle = 45, hjust=1),
         plot.title = element_text(hjust = 0.5),
         plot.subtitle = element_text(hjust = 0.5)) +
   ggtitle("After Filtering") +
   xlab(expression(log[10]~"(counts per gene)")) +
   ylab("Frequency") )

```



##### Normalisation

```{r fig.width=6, fig.height=4}

# filtcounts_ggplot <- stack(log2(filtdata))
# colnames(filtcounts_ggplot) <- c("expression", "sampleID")
# filtcounts_ggplot <- merge(filtcounts_ggplot,metadata, by="sampleID")
#   


total_coverage <- colSums(filtdata)
top50 <- names(sort(total_coverage, decreasing = TRUE)[1:50])
bot50 <- names(sort(total_coverage)[1:50])
selected_samples <- c(bot50, top50)

filtdata_sub <- filtdata[, selected_samples]

filtcounts_ggplot <- stack(log2(filtdata_sub))
colnames(filtcounts_ggplot) <- c("expression", "sampleID")
filtcounts_ggplot <- merge(filtcounts_ggplot, metadata, by = "sampleID")

# order by total coverage
filtcounts_ggplot$sampleID <- factor(filtcounts_ggplot$sampleID,
                                     levels = names(sort(total_coverage[selected_samples])))

(plt_beforenorm <- ggplot(filtcounts_ggplot, aes(sampleID,expression)) + geom_boxplot(alpha=0.4, outlier.size = 0.5)+ xlab("Samples") + ylab("log2(counts)")   +  theme(axis.text.x = element_text(angle = 45, hjust=1))+
    theme_bw()+ 
    theme(plot.title = element_text(hjust = 0.5),
          plot.subtitle = element_text(hjust = 0.5),
          axis.text.x=element_blank(),
          axis.ticks.x=element_blank()) +
    ggtitle("Before Normalisation") )

```


```{r}
dgelist_normcounts <- DGEList(filtdata) # the matrix with gene names as row names and counts in columns
dgelist_normcounts <- calcNormFactors(dgelist_normcounts, method = "TMM") 
normalised_counts <- as.data.frame(voom(dgelist_normcounts,plot=F)$E) 
#normalised_counts <- as.data.frame(edgeR::cpm(dgelist_normcounts, log = FALSE)) 
normdata <- 2^normalised_counts
lognormdata <- log2(normdata+1)
```



```{r fig.width=6, fig.height=4}

total_coverage <- colSums(normdata)
top50 <- names(sort(total_coverage, decreasing = TRUE)[1:50])
bot50 <- names(sort(total_coverage)[1:50])
selected_samples <- c(bot50, top50)

normdata_sub <- normdata[, selected_samples]

normcounts_ggplot <- stack(log2(normdata_sub))
colnames(normcounts_ggplot) <- c("expression", "sampleID")
normcounts_ggplot <- merge(normcounts_ggplot, metadata, by = "sampleID")

# order by total coverage
normcounts_ggplot$sampleID <- factor(normcounts_ggplot$sampleID,
                                     levels = names(sort(total_coverage[selected_samples])))

(plt_afternorm <- ggplot(normcounts_ggplot, aes(sampleID,expression)) + geom_boxplot(alpha=0.4, outlier.size = 0.5)+ xlab("Samples") + ylab("log2(counts)")   +  theme(axis.text.x = element_text(angle = 45, hjust=1))+
    theme_bw()+ 
    theme(plot.title = element_text(hjust = 0.5),
          plot.subtitle = element_text(hjust = 0.5),
          axis.text.x=element_blank(),
          axis.ticks.x=element_blank()) +
    ggtitle("After Normalisation") )

```


#### Calculate Scores


```{r fig.width=4, fig.height=3}
(dfscores <- CalculateScores(data = normdata, 
                             metadata = metadata, 
                             gene_sets = Ramaswamy_Metastasis,
                             method = "logmedian")$Ramaswamy_Metastasis)

met_colors <- c(
  "M0"       = "#E07B54",
  "M1" = "#5B8DB8" 
)



(plt_emt <- ggplot(dfscores, aes(x = ajcc_pathologic_m.diagnoses, y = score)) + 
    geom_jitter(aes(color=ajcc_pathologic_m.diagnoses), width = 0.3, alpha = 1, size = 2) +
    geom_violin(alpha = 0.5) +
    stat_summary(fun = median, geom = "crossbar", width = 0.1, color = "black") + 
    labs(x = NULL, y = expression(log[2]*"-median score")) +
    theme_bw() +
    theme(legend.position="none",
          plot.title = element_text(hjust = 0.5, size = 14),
          text = element_text(size = 14)) +
    ggtitle("Ramaswamy EMT transition Gene Set\n in KIRC Primary Tumour Samples") +
    scale_color_manual(values = met_colors))
```


### Proliferation | PANCAN

meta-PCNA (Venet et al. 2011)
```{r}
metaPCNA <- list(metaPCNA = as.vector(fread("../data_aux/proliferationsignature_metaPCNA.txt",header=F))$V1)
```

https://xenabrowser.net/datapages/?dataset=EB%2B%2BAdjustPANCAN_IlluminaHiSeq_RNASeqV2.geneExp.xena&host=https%3A%2F%2Fpancanatlas.xenahubs.net&removeHub=https%3A%2F%2Fxena.treehouse.gi.ucsc.edu%3A443

```{r}
data <- as.data.frame(fread("../data_aux/EB++AdjustPANCAN_IlluminaHiSeq_RNASeqV2.geneExp.xena.csv"))
# combine rows with same gene id by summing them
data <- rowsum(data[, -1], group = data$sample)

data <- 2^data - 1 # unlog...
data

```

```{r}
metadata <- as.data.frame(fread("../data_aux/Survival_SupplementalTable_S1_20171025_xena_sp.tsv"))
metadata <- metadata[metadata$sample %in% colnames(data),]
data <- data[,metadata$sample]
```


```{r}
metadata$code <- sub(".*-", "", metadata$sample) 
colnames(metadata)[1] <- "sampleID"
metadata$sample_type <- ifelse(metadata$code == "11", "Normal", "Tumour")
```


```{r}
data
metadata
```


```{r}
data <- na.omit(data)
data <- data + 1 # some negative values, must be of batch correction
```


##### Filtering

No need for filtering!! Distributions look good, data was processed & batch corrected before

```{r}
# Plot
(plt_beforefilt <- ggplot() + aes(x=log10(rowSums(data))) +
   geom_density(fill="grey", alpha=0.5) + theme_bw() + 
   theme(axis.text.x = element_text(angle = 45, hjust=1),
         plot.title = element_text(hjust = 0.5),
         plot.subtitle = element_text(hjust = 0.5)) +
   ggtitle("Before Filtering") +
   xlab(expression(log[10]~"(counts per gene)")) +
   ylab("Frequency") + 
   geom_vline(xintercept = 5.2, linetype="dotted",color = "black", size=1.5))

```


<!-- ```{r} -->
<!-- #filtdata <- data[rowSums(data) > 10^5.2,] -->
<!-- filtdata <- data -->
<!-- nrow(data) -->
<!-- nrow(filtdata) -->


<!-- cat("Number of genes before filtering:", nrow(data), "\n") -->
<!-- cat("Number of genes after filtering:", nrow(filtdata), "\n") -->

<!-- ``` -->


<!-- ```{r} -->

<!-- # Plot -->
<!-- (plt_afterfilt <- ggplot() + aes(x=log10(rowSums(filtdata))) + -->
<!--   geom_density(fill="grey", alpha=0.5) + theme_bw() +  -->
<!--     theme(axis.text.x = element_text(angle = 45, hjust=1), -->
<!--         plot.title = element_text(hjust = 0.5), -->
<!--         plot.subtitle = element_text(hjust = 0.5)) + -->
<!--   ggtitle("After Filtering") + -->
<!--   xlab(expression(log[10]~"(counts per gene)")) + -->
<!--   ylab("Frequency") ) -->

<!-- ``` -->



<!-- ##### Normalisation -->

<!-- ```{r fig.width=6, fig.height=4} -->

<!-- # filtcounts_ggplot <- stack(log2(filtdata)) -->
<!-- # colnames(filtcounts_ggplot) <- c("expression", "sampleID") -->
<!-- # filtcounts_ggplot <- merge(filtcounts_ggplot,metadata, by="sampleID") -->
<!-- #    -->


<!-- total_coverage <- colSums(filtdata) -->
<!-- top50 <- names(sort(total_coverage, decreasing = TRUE)[1:50]) -->
<!-- bot50 <- names(sort(total_coverage)[1:50]) -->
<!-- selected_samples <- c(bot50, top50) -->

<!-- filtdata_sub <- filtdata[, selected_samples] -->

<!-- filtcounts_ggplot <- stack(log2(filtdata_sub)) -->
<!-- colnames(filtcounts_ggplot) <- c("expression", "sampleID") -->
<!-- filtcounts_ggplot <- merge(filtcounts_ggplot, metadata, by = "sampleID") -->

<!-- # order by total coverage -->
<!-- filtcounts_ggplot$sampleID <- factor(filtcounts_ggplot$sampleID, -->
<!--                                       levels = names(sort(total_coverage[selected_samples]))) -->

<!-- (plt_beforenorm <- ggplot(filtcounts_ggplot, aes(sampleID,expression)) + geom_boxplot(alpha=0.4, outlier.size = 0.5)+ xlab("Samples") + ylab("log2(counts)")   +  theme(axis.text.x = element_text(angle = 45, hjust=1))+ -->
<!--   theme_bw()+  -->
<!--   theme(plot.title = element_text(hjust = 0.5), -->
<!--         plot.subtitle = element_text(hjust = 0.5), -->
<!--         axis.text.x=element_blank(), -->
<!--         axis.ticks.x=element_blank()) + -->
<!--   ggtitle("Before Normalisation") ) -->

<!-- ``` -->


<!-- ```{r} -->
<!-- # dgelist_normcounts <- DGEList(filtdata) # the matrix with gene names as row names and counts in columns -->
<!-- # dgelist_normcounts <- calcNormFactors(dgelist_normcounts, method = "TMM")  -->
<!-- # normalised_counts <- as.data.frame(voom(dgelist_normcounts,plot=F)$E)  -->
<!-- # #normalised_counts <- as.data.frame(edgeR::cpm(dgelist_normcounts, log = FALSE))  -->
<!-- # normdata <- 2^normalised_counts -->
<!-- # lognormdata <- log2(normdata+1) -->
<!-- ``` -->
<!-- ```{r} -->
<!-- normdata <- filtdata -->
<!-- ``` -->



<!-- ```{r fig.width=6, fig.height=4} -->

<!-- total_coverage <- colSums(normdata) -->
<!-- top50 <- names(sort(total_coverage, decreasing = TRUE)[1:50]) -->
<!-- bot50 <- names(sort(total_coverage)[1:50]) -->
<!-- selected_samples <- c(bot50, top50) -->

<!-- normdata_sub <- normdata[, selected_samples] -->

<!-- normcounts_ggplot <- stack(log2(normdata_sub)) -->
<!-- colnames(normcounts_ggplot) <- c("expression", "sampleID") -->
<!-- normcounts_ggplot <- merge(normcounts_ggplot, metadata, by = "sampleID") -->

<!-- # order by total coverage -->
<!-- normcounts_ggplot$sampleID <- factor(normcounts_ggplot$sampleID, -->
<!--                                       levels = names(sort(total_coverage[selected_samples]))) -->

<!-- (plt_afternorm <- ggplot(normcounts_ggplot, aes(sampleID,expression)) + geom_boxplot(alpha=0.4, outlier.size = 0.5)+ xlab("Samples") + ylab("log2(counts)")   +  theme(axis.text.x = element_text(angle = 45, hjust=1))+ -->
<!--   theme_bw()+  -->
<!--   theme(plot.title = element_text(hjust = 0.5), -->
<!--         plot.subtitle = element_text(hjust = 0.5), -->
<!--         axis.text.x=element_blank(), -->
<!--         axis.ticks.x=element_blank()) + -->
<!--   ggtitle("After Normalisation") ) -->

<!-- ``` -->


#### Calculate Scores

```{r}
#saveRDS(data,"../data_aux/data_PANCAN.rds")
data_PANCAN <- readRDS("../data_aux/data_PANCAN.rds")
#saveRDS(metadata,"../data_aux/metadata_PANCAN.rds")
metadata_PANCAN <- readRDS("../data_aux/metadata_PANCAN.rds")
```

```{r fig.width=4, fig.height=3}
(dfscores <- CalculateScores(data = data_PANCAN, 
                             metadata = metadata_PANCAN, 
                             gene_sets = metaPCNA,
                             method = "logmedian")$metaPCNA)

samptype_col <- c(
  "Tumour" = "#C0392B",
  "Normal" = "#2E86AB"
)


(plt_prol <- ggplot(dfscores, aes(x = sample_type, y = score, fill=sample_type)) + 
    geom_jitter(color = "#DDDDDD", width = 0.3, alpha = 0.3, size = 0.5) +
    geom_violin(alpha = 0.5) +
    stat_summary(fun = median, geom = "crossbar", width = 0.1, color = "black") + 
    labs(x = NULL, y = expression(log[2]*"-median score")) +
    theme_bw() +
    theme(legend.position="none",
          plot.title = element_text(hjust = 0.5, size = 14),
          text = element_text(size = 14)) +
    ggtitle("Venet 2011 Proliferation Gene Set\n in TCGA PANCAN (Liu 2018)") +
    scale_fill_manual(values = samptype_col))
```
```{r fig.width=4, fig.height=3}
n_labels <- dfscores %>%
  group_by(sample_type) %>%
  summarise(n = n()) %>%
  mutate(label = paste0(sample_type, "\n(n=", n, ")"))

label_vec <- setNames(n_labels$label, n_labels$sample_type)

(plt_prol <- ggplot(dfscores, aes(x = sample_type, y = score, fill = sample_type)) + 
    geom_jitter(color = "#DDDDDD", width = 0.3, alpha = 0.3, size = 0.5) +
    geom_violin(alpha = 0.5) +
    stat_summary(fun = median, geom = "crossbar", width = 0.1, color = "black") + 
    labs(x = NULL, y = expression(log[2]*"-median score")) +
    theme_bw() +
    theme(legend.position = "none",
          plot.title = element_text(hjust = 0.5, size = 14),
          text = element_text(size = 14)) +
    ggtitle("Venet 2011 Proliferation Gene Set\n in TCGA PANCAN (Liu 2018)") +
    scale_fill_manual(values = samptype_col) +
    scale_x_discrete(labels = label_vec))
```

### Combine all

```{r fig.width=12, fig.height=3.5}
ggarrange(plt_ychromosomescores,
          plt_viralrespiratory,
          plt_prol, nrow=1, align = "hv")
```


## Comment 3.1 — Cell type composition as a confounder in GTEx age analyses

R.: Estimate cell type proportions in the GTEx tissues highlighted in the main text using a reference-based deconvolution tool, and incorporate these estimates as covariates in the age–score association models. Results with and without compositional correction are compared for a subset of tissues, both those with and without significant age associations in the original analysis.



## Comment 1.5 — Additional datasets for age correlation analysis

R.: Apply markeR's discovery mode to one or more datasets independent of GTEx, modelling age as a continuous variable, to assess whether age-associated senescence signals are reproducible across transcriptomic resources and pointed limitations are correct. 

## Session Info
```{r}
sessionInfo()
```
