---
title: "Response to Reviewers"
author: "Rita Martins-Silva"
date: "02/03/2026"
output: html_document
---


# Response to Reviewers {.tabset .tabset-pills}

## Set up {.tabset}

```{r}
setwd("../")
```
### Relevant Data

```{r}
signatures_bidirectional <- readRDS("../data/SenescenceSigntures_Bidirectional.rds")
t2gene2 <- fread("../data_aux/RefSeq_Transcripts_To_Genes.csv")
t2gene2 <- t2gene2[,c("name","name2")]
colnames(t2gene2) <- c("transcript","gene")
```

### Libraries

```{r}
library(data.table)
library("ggplot2")
library("colorspace")
library("scales")
#library("scater") 
library("reshape2")
#library("markeR")
#library("ggbreak")
#library("ggnewscale")
library("dplyr")
library("tidyr")  
library(patchwork)
library("ggpubr") 
library(purrr)
library(edgeR)
library(RColorBrewer)
library(pROC) 
#library("ggforce")
library(ggrepel)
#devtools::install_github("DiseaseTranscriptomicsLab/markeR@*release")
library(biomaRt)
library("tximport")
library(tidyverse)
library(patchwork)
library(GEOquery)
```

### Functions

Load markeR functions directly

```{r}
# Path to your folder with functions
fun_dir <- "functions_Sept18_2025/"

# Get all .R files in the folder
files <- list.files(paste0("../",fun_dir), pattern = "\\.R$", full.names = TRUE)

# Source them one by one
invisible(lapply(files, source))

```

### Colors

```{r}
Condition_colors <- c(
  "Senescent"     = "#F4A261",  # Soft orange (not goldenrod)
  "Proliferative" = "#A1C298",  # Desaturated sage green
  "Quiescent"     = "#9DB4C0"   # Dusty teal-gray
)

CellTypecols <- c(
  "Fibroblast"   = "#E69F00",  # Orange
  "Keratinocyte" = "#56B4E9",  # Sky Blue
  "Melanocyte"   = "#009E73",  # Bluish Green
  "Endothelial"  = "#F0E442",  # Yellow
  "Neuronal"     = "#0072B2",  # Blue
  "Mesenchymal"  = "#D55E00"   # Vermilion
)

CellTypecols_alt <- c(
  "Fibroblast"   = "#1b9e77",  # Teal green
  "Keratinocyte" = "#d95f02",  # Warm orange-brown
  "Melanocyte"   = "#7570b3",  # Muted purple
  "Endothelial"  = "#e7298a",  # Dark pink
  "Neuronal"     = "#66a61e",  # Olive green
  "Mesenchymal"  = "#e6ab02"   # Mustard
)


dataset_colors <- c(
  "HernandezSegura" = "#8DD3C7",
  "Mangelinck"      = "#FFFFB3",
  "Chan"            = "#BEBADA",
  "Purcell"         = "#FB8072",
  "Marthandan2016"  = "#80B1D3",
  "Marthandan2015"  = "#FDB462",
  "Wang"            = "#B3DE69",
  "Suda"            = "#FCCDE5",
  "McHugh"          = "#D9D9D9",
  "Savic"           = "#BC80BD",
  "Skea"            = "#CCEBC5",
  "Laurent"         = "#FFED6F",
  "Tasdemir"        = "#D0B7E1",
  "Lenain"          = "#FFB3BA",
  "Aarts"           = "#C2C2F0",
  "Casella"         = "#FFDAC1",
  "Numa"            = "#C6E2FF",
  "Admasu"          = "#F5CBA7",
  "Urata"           = "#C7CEEA",
  "Wang2023"        = "#E6B0AA",
  "Fu"              = "#A9DFBF",
  "Tanke"           = "#F9E79F",
  "Bawens"          = "#AED6F1"
)

geneSet_colors <- brewer.pal(9, "Set1")  # 9 distinct colors, mostly bright and distinct

```

```{r}
show_col(Condition_colors, ncol=3)
show_col(CellTypecols)
show_col(dataset_colors)
show_col(geneSet_colors)
```


## Comment 3.3 — External validation on an independent dataset

R.: Apply markeR's benchmarking mode to a senescence dataset not included in the original compendium, to evaluate whether HernandezSegura and SAUL_SEN_MAYO retain their discriminatory performance in a fully independent evaluation set. 



### GSE297560 (Space Radiation)
HiSat2 + feature counts, hg38

#### Data

```{r}
data <- as.data.frame(fread("../data_aux/GSE297560_counts.txt")[,-c(2:6)])
row.names(data) <- data$Geneid
data$Geneid <-  NULL
head(data)
# Convert row names from ENSEMBL ID to Gene Name using biomaRt

# Connect to Ensembl
mart <- useMart("ensembl", dataset = "hsapiens_gene_ensembl")

# Query the conversion
gene_conversion <- getBM(
  attributes = c("ensembl_gene_id", "hgnc_symbol"),
  filters    = "ensembl_gene_id",
  values     = rownames(data),
  mart       = mart
)

# Replace rownames — keeping only genes with a match and a non-empty symbol
gene_conversion <- gene_conversion[!duplicated(gene_conversion$hgnc_symbol), ]
data <- data[gene_conversion$ensembl_gene_id, ]
rownames(data) <- gene_conversion$hgnc_symbol
#remove .trimmed from sample names
colnames(data) <- gsub("\\.trimmed", "", colnames(data))

```


- Healthy_CTRL: Low-passage fibroblasts that were never shipped. Represent the true untreated baseline but are not the appropriate matched control for irradiation comparisons, as any differences relative to irradiated samples would conflate shipping stress with radiation effects.

- MOCK_CTRL: Fibroblasts that were shipped to the radiation facility but not irradiated, paired with the charged particle irradiation experiments (Hydrogen, Iron, Silicon). Control for the confounding effect of shipping stress, isolating the effect of radiation.

- MOCKgamma_CTRL: Identical to MOCK_CTRL but paired specifically with the gamma irradiation experiment. Kept separate to maintain matched experimental pairing.

```{r}
# Load metadata
metadata <- read.csv("../data_aux/GSE297560_metadata.csv")
metadata <- metadata[,c("Run","Sample.Name","source_name", "treatment")]
metadata$datasetID <- "GSE297560"

metadata$condition <- ifelse(grepl("control", metadata$treatment, ignore.case = TRUE), "Proliferative", ifelse(grepl("irradiation", metadata$treatment, ignore.case = TRUE), "Senescent", "Proliferative"))


# Attempt to match SRR with colnames
sample_to_srr <- c(
  "Healthy_CTRL_1_HISAT2"   = "SRR33626854",#
  "Healthy_CTRL_2_HISAT2"   = "SRR33626853",#
  "Healthy_CTRL_3_HISAT2"   = "SRR33626852",#
  "4Hydrogen_1_HISAT2"      = "SRR33626842",#
  "4Hydrogen_2_HISAT2"      = "SRR33626841",#
  "4Hydrogen_3_HISAT2"      = "SRR33626840",#
  "4Iron_1_HISAT2"          = "SRR33626845",#
  "4Iron_2_HISAT2"          = "SRR33626844",#
  "4Iron_3_HISAT2"          = "SRR33626843",#
  "4Silicon_1_HISAT2"       = "SRR33626848",#
  "4Silicon_2_HISAT2"       = "SRR33626847",#
  "4Silicon_3_HISAT2"       = "SRR33626846",#
  "MOCK_CTRL_1_HISAT2"      = "SRR33626857",#
  "MOCK_CTRL_2_HISAT2"      = "SRR33626856",#
  "MOCK_CTRL_3_HISAT2"      = "SRR33626855",#
  "4Gamma_1_HISAT2"         = "SRR33626839",#
  "4Gamma_3_HISAT2"         = "SRR33626838",#
  "4Gamma_2_HISAT2"         = "SRR33626837",#
  "MOCKgamma_CTRL_1_HISAT2" = "SRR33626851",
  "MOCKgamma_CTRL_2_HISAT2" = "SRR33626850",
  "MOCKgamma_CTRL_3_HISAT2" = "SRR33626849"
)

# Rename columns of your expression matrix
colnames(data) <- sample_to_srr[colnames(data)]

# Reorder metadata to match column order of your_df
metadata <- metadata[match(colnames(data), metadata$Run), ]

treatment_map <- c(
  "Low passage\\, unshipped"                            = "Healthy_CTRL",
  "Irradiation control shipped not irradiated"          = "MOCK_CTRL",
  "Gamma irradiation control shipped not irradiated"    = "MOCKgamma_CTRL",
  "4Gy Gamma irradiation"                               = "4Gy_Gamma",
  "4Gy Hydrogen irradiation"                            = "4Gy_Hydrogen",
  "4Gy Iron irradiation"                                = "4Gy_Iron",
  "4Gy Silicon irradiation"                             = "4Gy_Silicon"
)

metadata$treatment_simple <- treatment_map[metadata$treatment]
colnames(metadata)[1] <- "sampleID"
```



Remove Gamma and mock gamma, as in the paper the 4 gamma are not senescent...
Remove also Healthy_CTRL, they don't validate those samples in the paper either

```{r}
metadata <- metadata[!grepl("Gamma", metadata$treatment_simple, ignore.case = T), ]
metadata <- metadata[!grepl("Healthy_CTRL", metadata$treatment_simple, ignore.case = T), ]
data <- data[,metadata$sampleID]
```


```{r}
data
metadata
```

#### Preprocessing


##### Filtering

```{r}
# Plot
(plt_beforefilt <- ggplot() + aes(x=log10(rowSums(data))) +
   geom_density(fill="grey", alpha=0.5) + theme_bw() + 
   theme(axis.text.x = element_text(angle = 45, hjust=1),
         plot.title = element_text(hjust = 0.5),
         plot.subtitle = element_text(hjust = 0.5)) +
   ggtitle("Before Filtering") +
   xlab(expression(log[10]~"(counts per gene)")) +
   ylab("Frequency") + 
   geom_vline(xintercept = 2.9, linetype="dotted",color = "black", size=1.5))

```


```{r}
filtdata <- data[rowSums(data) > 10^2.9,]
nrow(data)
nrow(filtdata)


cat("Number of genes before filtering:", nrow(data), "\n")
cat("Number of genes after filtering:", nrow(filtdata), "\n")

```


```{r}

# Plot
(plt_afterfilt <- ggplot() + aes(x=log10(rowSums(filtdata))) +
   geom_density(fill="grey", alpha=0.5) + theme_bw() + 
   theme(axis.text.x = element_text(angle = 45, hjust=1),
         plot.title = element_text(hjust = 0.5),
         plot.subtitle = element_text(hjust = 0.5)) +
   ggtitle("After Filtering") +
   xlab(expression(log[10]~"(counts per gene)")) +
   ylab("Frequency") )

```



##### Normalisation

```{r fig.width=6, fig.height=4}

filtcounts_ggplot <- stack(log2(filtdata))
colnames(filtcounts_ggplot) <- c("expression", "sampleID")
filtcounts_ggplot <- merge(filtcounts_ggplot,metadata, by="sampleID")


(plt_beforenorm <- ggplot(filtcounts_ggplot, aes(sampleID,expression)) + geom_boxplot(alpha=0.4)+ xlab("Samples") + ylab("log2(counts)")   +  theme(axis.text.x = element_text(angle = 45, hjust=1))+
    theme_bw()+ 
    theme(plot.title = element_text(hjust = 0.5),
          plot.subtitle = element_text(hjust = 0.5),
          axis.text.x=element_blank(),
          axis.ticks.x=element_blank()) +
    ggtitle("Before Normalisation") )

```


```{r}
dgelist_normcounts <- DGEList(filtdata) # the matrix with gene names as row names and counts in columns
dgelist_normcounts <- calcNormFactors(dgelist_normcounts, method = "TMM") 
normalised_counts <- as.data.frame(voom(dgelist_normcounts,plot=F)$E) 
#normalised_counts <- as.data.frame(edgeR::cpm(dgelist_normcounts, log = FALSE)) 
normdata <- 2^normalised_counts
lognormdata <- log2(normdata+1)
```


```{r fig.width=6, fig.height=4}

normcounts_ggplot <- stack(log2(normdata))
colnames(normcounts_ggplot) <- c("expression", "sampleID")
normcounts_ggplot <- merge(normcounts_ggplot,metadata, by="sampleID")


(plt_afternorm <- ggplot(normcounts_ggplot, aes(sampleID,expression)) + geom_boxplot(alpha=0.4)+ xlab("Samples") + ylab("log2(counts)")   +  theme(axis.text.x = element_text(angle = 45, hjust=1))+
    theme_bw()+ 
    theme(plot.title = element_text(hjust = 0.5),
          plot.subtitle = element_text(hjust = 0.5),
          axis.text.x=element_blank(),
          axis.ticks.x=element_blank()) +
    ggtitle("After Normalisation") )

```


##### PCA

```{r fig.width=6, fig.height=4}
y <- DGEList(lognormdata, samples= metadata)
PCA  <- prcomp(t(y$counts), scale=FALSE, center=TRUE)
PCACounts  <- PCA$x
PCACounts  <- as.data.frame(PCACounts )
PCAData <-  cbind(PCACounts[1:10],y$samples) 

ev = PCA$sdev^2 
pc1= round(100*ev[1]/sum(ev),2) 
pc2 = round(100*ev[2]/sum(ev),2) 
pc3 = round(100*ev[3]/sum(ev),2) 
pc4 = round(100*ev[4]/sum(ev),2) 
pc5 = round(100*ev[5]/sum(ev),2) 
pc6 = round(100*ev[6]/sum(ev),2)  


(pltpca_1 <- ggplot(PCAData, aes(PC1,PC2))+
    geom_point(aes(color=treatment_simple), size=5, alpha=0.6)+ 
    xlab(paste0("PC1: ",pc1,"% variance")) +
    ylab(paste0("PC2: ",pc2,"% variance")) + 
    theme_bw()+
    guides(color = guide_legend(nrow = 2)) +
    theme(legend.position="top", legend.box="vertical", legend.title = element_blank(), legend.margin=margin())  + 
    geom_vline(xintercept=0, linetype="dotted") + 
    geom_hline(yintercept=0, linetype="dotted") )

(pltpca_2 <- ggplot(PCAData, aes(PC1,PC2))+
    geom_point(aes(color=condition), size=5, alpha=0.6)+ 
    xlab(paste0("PC1: ",pc1,"% variance")) +
    ylab(paste0("PC2: ",pc2,"% variance")) + 
    theme_bw()+
    theme(legend.position="top", legend.box="vertical", legend.title = element_blank(), legend.margin=margin())  + 
    geom_vline(xintercept=0, linetype="dotted") + 
    guides(color = guide_legend(nrow = 2)) +
    geom_hline(yintercept=0, linetype="dotted") +
    scale_color_manual(values=Condition_colors))


```

#### Calculate Scores

```{r}
(pltscores <- PlotScores(data = normdata, 
                         metadata = metadata, 
                         gene_sets = signatures_bidirectional[c("HernandezSegura", "SAUL_SEN_MAYO")],
                         Variable = "condition",
                         ColorVariable = "treatment_simple",
                         method = "logmedian",
                         xlab = "Cellular State", 
                         labsize = 8))
```

```{r fig.width=11, fig.height=5}
plt_filt <- ggarrange(plt_beforefilt,plt_afterfilt, nrow=1)
plt_norm <- ggarrange(plt_beforenorm,plt_afternorm, nrow=1)
plt_pca <- ggarrange(pltpca_1,pltpca_2, nrow=1)

plt1 <- ggarrange(plt_filt, plt_norm, nrow=1)
plt2 <- ggarrange(plt_pca, pltscores, nrow=1, widths=c(0.5,0.5))

plt_GSE297560 <- ggarrange(plt1, plt2, ncol=1, heights = c(0.4,0.7))
plt_GSE297560 <- annotate_figure(plt_GSE297560, top = text_grob("GSE297560: Space Radiation\n", face = "bold", size = 14))
plt_GSE297560


```

### GSE288940 (IMR90 + Eto)

My quantification pipeline 

#### Data

```{r}
# Get a list of the paths to the samples. Each sample is inside its own folder, under the name "quant.sf"
listfolders <- list.dirs("../data_aux/kallisto/Farr/")[-1] # remove "." directory
listfolders <- paste0(listfolders,"/abundance.tsv")
listfolders

# Naming the files (important so that when using tximport each column has the corresponding sample name)
samples <- gsub("../data_aux/kallisto/Farr//","",listfolders)
samples <- gsub("/abundance.tsv","",samples)
names(listfolders) <- samples
listfolders

# Merge all counts in an txi object. The argument txOut=TRUE allows to avoid gene-level summarization (we want the transcripts)
txi.kallisto <- tximport(listfolders, type = "kallisto", txOut=FALSE, tx2gene = t2gene2)

data <- as.data.frame(txi.kallisto$counts) 

```

```{r}
metadata <- as.data.frame(fread("../data_aux/GSE288940_metadata.csv"))
metadata <- metadata[,c("Run","Sample Name","cell_line","treatment")]
colnames(metadata)[1] <- "sampleID"


gsm_to_timepoint <- c(
  "GSM8779535" = "D0",
  "GSM8779536" = "D0",
  "GSM8779537" = "D0",
  "GSM8779538" = "D0",
  "GSM8779539" = "D0",
  "GSM8779540" = "D7",
  "GSM8779541" = "D7",
  "GSM8779542" = "D7",
  "GSM8779543" = "D7",
  "GSM8779544" = "D7",
  "GSM8779545" = "D7",
  "GSM8779546" = "D28",
  "GSM8779547" = "D28",
  "GSM8779548" = "D28",
  "GSM8779549" = "D28",
  "GSM8779550" = "D28",
  "GSM8779551" = "D28"
)

metadata$timepoint <- gsm_to_timepoint[metadata$`Sample Name`]
metadata$timepoint <- factor(metadata$timepoint, levels = c("D0", "D7", "D28"))

metadata$condition2 <- ifelse(grepl("D0", metadata$timepoint), "Proliferative", ifelse(grepl("D7", metadata$timepoint), "Prol > Sen", "Senescent"))

metadata$condition <- ifelse(grepl("D0", metadata$timepoint), "Proliferative",  "Senescent")

data <- data[,metadata$sampleID]
```


```{r}
data
metadata
```

#### Preprocessing


##### Filtering

```{r}
# Plot
(plt_beforefilt <- ggplot() + aes(x=log10(rowSums(data))) +
   geom_density(fill="grey", alpha=0.5) + theme_bw() + 
   theme(axis.text.x = element_text(angle = 45, hjust=1),
         plot.title = element_text(hjust = 0.5),
         plot.subtitle = element_text(hjust = 0.5)) +
   ggtitle("Before Filtering") +
   xlab(expression(log[10]~"(counts per gene)")) +
   ylab("Frequency") + 
   geom_vline(xintercept = 3.5, linetype="dotted",color = "black", size=1.5))

```


```{r}
filtdata <- data[rowSums(data) > 10^3.5,]
nrow(data)
nrow(filtdata)


cat("Number of genes before filtering:", nrow(data), "\n")
cat("Number of genes after filtering:", nrow(filtdata), "\n")

```


```{r}

# Plot
(plt_afterfilt <- ggplot() + aes(x=log10(rowSums(filtdata))) +
   geom_density(fill="grey", alpha=0.5) + theme_bw() + 
   theme(axis.text.x = element_text(angle = 45, hjust=1),
         plot.title = element_text(hjust = 0.5),
         plot.subtitle = element_text(hjust = 0.5)) +
   ggtitle("After Filtering") +
   xlab(expression(log[10]~"(counts per gene)")) +
   ylab("Frequency") )

```



##### Normalisation

```{r fig.width=6, fig.height=4}

filtcounts_ggplot <- stack(log2(filtdata))
colnames(filtcounts_ggplot) <- c("expression", "sampleID")
filtcounts_ggplot <- merge(filtcounts_ggplot,metadata, by="sampleID")


(plt_beforenorm <- ggplot(filtcounts_ggplot, aes(sampleID,expression)) + geom_boxplot(alpha=0.4)+ xlab("Samples") + ylab("log2(counts)")   +  theme(axis.text.x = element_text(angle = 45, hjust=1))+
    theme_bw()+ 
    theme(plot.title = element_text(hjust = 0.5),
          plot.subtitle = element_text(hjust = 0.5),
          axis.text.x=element_blank(),
          axis.ticks.x=element_blank()) +
    ggtitle("Before Normalisation") )

```


```{r}
dgelist_normcounts <- DGEList(filtdata) # the matrix with gene names as row names and counts in columns
dgelist_normcounts <- calcNormFactors(dgelist_normcounts, method = "TMM") 
normalised_counts <- as.data.frame(voom(dgelist_normcounts,plot=F)$E) 
#normalised_counts <- as.data.frame(edgeR::cpm(dgelist_normcounts, log = FALSE)) 
normdata <- 2^normalised_counts
lognormdata <- log2(normdata+1)
```


```{r fig.width=6, fig.height=4}

normcounts_ggplot <- stack(log2(normdata))
colnames(normcounts_ggplot) <- c("expression", "sampleID")
normcounts_ggplot <- merge(normcounts_ggplot,metadata, by="sampleID")


(plt_afternorm <- ggplot(normcounts_ggplot, aes(sampleID,expression)) + geom_boxplot(alpha=0.4)+ xlab("Samples") + ylab("log2(counts)")   +  theme(axis.text.x = element_text(angle = 45, hjust=1))+
    theme_bw()+ 
    theme(plot.title = element_text(hjust = 0.5),
          plot.subtitle = element_text(hjust = 0.5),
          axis.text.x=element_blank(),
          axis.ticks.x=element_blank()) +
    ggtitle("After Normalisation") )

```


##### PCA

```{r fig.width=6, fig.height=4}
y <- DGEList(lognormdata, samples= metadata)
PCA  <- prcomp(t(y$counts), scale=FALSE, center=TRUE)
PCACounts  <- PCA$x
PCACounts  <- as.data.frame(PCACounts )
PCAData <-  cbind(PCACounts[1:10],y$samples) 

ev = PCA$sdev^2 
pc1= round(100*ev[1]/sum(ev),2) 
pc2 = round(100*ev[2]/sum(ev),2) 
pc3 = round(100*ev[3]/sum(ev),2) 
pc4 = round(100*ev[4]/sum(ev),2) 
pc5 = round(100*ev[5]/sum(ev),2) 
pc6 = round(100*ev[6]/sum(ev),2)  

timepoint_colors <- c(
  "D0"  = "#AEC6E8",  # pastel blue
  "D7"  = "#C3A8D1",  # pastel purple
  "D28" = "#F4A6A0"   # pastel red/pink
)

(pltpca_1 <- ggplot(PCAData, aes(PC1,PC2))+
    geom_point(aes(color=timepoint), size=5, alpha=0.6)+ 
    xlab(paste0("PC1: ",pc1,"% variance")) +
    ylab(paste0("PC2: ",pc2,"% variance")) + 
    theme_bw()+
    guides(color = guide_legend(nrow = 1)) +
    theme(legend.position="top", legend.box="vertical", legend.title = element_blank(), legend.margin=margin())  + 
    geom_vline(xintercept=0, linetype="dotted") + 
    geom_hline(yintercept=0, linetype="dotted") +
    scale_color_manual(values=timepoint_colors))

(pltpca_2 <- ggplot(PCAData, aes(PC1,PC2))+
    geom_point(aes(color=condition), size=5, alpha=0.6)+ 
    xlab(paste0("PC1: ",pc1,"% variance")) +
    ylab(paste0("PC2: ",pc2,"% variance")) + 
    theme_bw()+
    theme(legend.position="top", legend.box="vertical", legend.title = element_blank(), legend.margin=margin())  + 
    geom_vline(xintercept=0, linetype="dotted") + 
    guides(color = guide_legend(nrow = 1)) +
    geom_hline(yintercept=0, linetype="dotted") +
    scale_color_manual(values=Condition_colors))


```

#### Calculate Scores

```{r}
(pltscores <- PlotScores(data = normdata, 
                         metadata = metadata, 
                         gene_sets = signatures_bidirectional[c("HernandezSegura", "SAUL_SEN_MAYO")],
                         Variable = "condition",
                         ColorVariable = "timepoint",
                         method = "logmedian",
                         ColorValues = timepoint_colors,
                         xlab = "Cellular State",
                         labsize = 8))
```

```{r fig.width=11, fig.height=5}
plt_filt <- ggarrange(plt_beforefilt,plt_afterfilt, nrow=1)
plt_norm <- ggarrange(plt_beforenorm,plt_afternorm, nrow=1)
plt_pca <- ggarrange(pltpca_1,pltpca_2, nrow=1)

plt1 <- ggarrange(plt_filt, plt_norm, nrow=1)
plt2 <- ggarrange(plt_pca, pltscores, nrow=1, widths=c(0.5,0.5))

plt_GSE288940 <- ggarrange(plt1, plt2, ncol=1, heights = c(0.4,0.7))
plt_GSE288940 <- annotate_figure(plt_GSE288940, top = text_grob("GSE288940: IMR90 + Etoposide\n", face = "bold", size = 14))
plt_GSE288940


```






### GSE287058 (Quiescent Fibroblasts)

counts tables were generated using HTSeq
Assembly: hg38

#### Data

```{r}

# List all files
files <- list.files(path = "../data_aux/GSE287058_RAW/", pattern = "\\.txt\\.gz$", full.names = TRUE)

# Read and combine
data <- lapply(files, function(f) {
  sample_name <- gsub("(GSM[0-9]+).*", "\\1", basename(f))
  df <- read.table(f, header = FALSE, col.names = c("gene", sample_name))
  df
}) %>%
  Reduce(function(x, y) full_join(x, y, by = "gene"), .) %>%
  tibble::column_to_rownames("gene")

data
```

```{r}
metadata <- as.data.frame(fread("../data_aux/GSE287058_metadata.csv"))
metadata <- metadata[,c("Run","Library Name","cell_line","treatment")]
colnames(metadata)[1] <- "sampleID"



# Reorder metadata to match column order of data
metadata <- metadata[match(colnames(data), metadata$`Library Name`), ]

# Verify order is correct before renaming
all(colnames(data) == metadata$`Library Name`)

# Then rename
colnames(data) <- metadata$sampleID

metadata$condition <- ifelse(grepl("Proliferating", metadata$treatment, ignore.case = TRUE), "Proliferative", "Quiescent")
metadata$condition2 <-  ifelse(metadata$condition == "Quiescent",
                               paste0(metadata$condition, "\n(", metadata$treatment, ")"),
                               metadata$condition)
metadata$condition2 <- gsub(" ","\n",metadata$condition2)

```

```{r}
data
metadata
```
Remove LF1 samples as they don't have controls

```{r}
metadata <- metadata[!grepl("LF1", metadata$cell_line), ]
data <- data[,metadata$sampleID]
```


```{r}
data
metadata
```


#### Preprocessing


##### Filtering

```{r}
# Plot
(plt_beforefilt <- ggplot() + aes(x=log10(rowSums(data))) +
   geom_density(fill="grey", alpha=0.5) + theme_bw() + 
   theme(axis.text.x = element_text(angle = 45, hjust=1),
         plot.title = element_text(hjust = 0.5),
         plot.subtitle = element_text(hjust = 0.5)) +
   ggtitle("Before Filtering") +
   xlab(expression(log[10]~"(counts per gene)")) +
   ylab("Frequency") + 
   geom_vline(xintercept = 3.2, linetype="dotted",color = "black", size=1.5))

```


```{r}
filtdata <- data[rowSums(data) > 10^3.2,]
nrow(data)
nrow(filtdata)


cat("Number of genes before filtering:", nrow(data), "\n")
cat("Number of genes after filtering:", nrow(filtdata), "\n")

```


```{r}

# Plot
(plt_afterfilt <- ggplot() + aes(x=log10(rowSums(filtdata))) +
   geom_density(fill="grey", alpha=0.5) + theme_bw() + 
   theme(axis.text.x = element_text(angle = 45, hjust=1),
         plot.title = element_text(hjust = 0.5),
         plot.subtitle = element_text(hjust = 0.5)) +
   ggtitle("After Filtering") +
   xlab(expression(log[10]~"(counts per gene)")) +
   ylab("Frequency") )

```



##### Normalisation

```{r fig.width=6, fig.height=4}

filtcounts_ggplot <- stack(log2(filtdata))
colnames(filtcounts_ggplot) <- c("expression", "sampleID")
filtcounts_ggplot <- merge(filtcounts_ggplot,metadata, by="sampleID")


(plt_beforenorm <- ggplot(filtcounts_ggplot, aes(sampleID,expression)) + geom_boxplot(alpha=0.4, outlier.size = 0.5)+ xlab("Samples") + ylab("log2(counts)")   +  theme(axis.text.x = element_text(angle = 45, hjust=1))+
    theme_bw()+ 
    theme(plot.title = element_text(hjust = 0.5),
          plot.subtitle = element_text(hjust = 0.5),
          axis.text.x=element_blank(),
          axis.ticks.x=element_blank()) +
    ggtitle("Before Normalisation") )

```


```{r}
dgelist_normcounts <- DGEList(filtdata) # the matrix with gene names as row names and counts in columns
dgelist_normcounts <- calcNormFactors(dgelist_normcounts, method = "TMM") 
normalised_counts <- as.data.frame(voom(dgelist_normcounts,plot=F)$E) 
#normalised_counts <- as.data.frame(edgeR::cpm(dgelist_normcounts, log = FALSE)) 
normdata <- 2^normalised_counts
lognormdata <- log2(normdata+1)
```


```{r fig.width=6, fig.height=4}

normcounts_ggplot <- stack(log2(normdata))
colnames(normcounts_ggplot) <- c("expression", "sampleID")
normcounts_ggplot <- merge(normcounts_ggplot,metadata, by="sampleID")


(plt_afternorm <- ggplot(normcounts_ggplot, aes(sampleID,expression)) + geom_boxplot(alpha=0.4, outlier.size = 0.5)+ xlab("Samples") + ylab("log2(counts)")   +  theme(axis.text.x = element_text(angle = 45, hjust=1))+
    theme_bw()+ 
    theme(plot.title = element_text(hjust = 0.5),
          plot.subtitle = element_text(hjust = 0.5),
          axis.text.x=element_blank(),
          axis.ticks.x=element_blank()) +
    ggtitle("After Normalisation") )

```


##### PCA

```{r fig.width=6, fig.height=4}
y <- DGEList(lognormdata, samples= metadata)
PCA  <- prcomp(t(y$counts), scale=FALSE, center=TRUE)
PCACounts  <- PCA$x
PCACounts  <- as.data.frame(PCACounts )
PCAData <-  cbind(PCACounts[1:10],y$samples) 

ev = PCA$sdev^2 
pc1= round(100*ev[1]/sum(ev),2) 
pc2 = round(100*ev[2]/sum(ev),2) 
pc3 = round(100*ev[3]/sum(ev),2) 
pc4 = round(100*ev[4]/sum(ev),2) 
pc5 = round(100*ev[5]/sum(ev),2) 
pc6 = round(100*ev[6]/sum(ev),2)  

cell_line_colors <- c(
  "BJ"                    = "#6DBF8A",  # medium pastel green
  "IMR90"                 = "#F47C7C",  # medium pastel red
  "Primary Preadipocytes" = "#F0C060"   # medium pastel yellow
)

(pltpca_1 <- ggplot(PCAData, aes(PC1,PC2))+
    geom_point(aes(color=cell_line), size=5, alpha=0.6)+ 
    xlab(paste0("PC1: ",pc1,"% variance")) +
    ylab(paste0("PC2: ",pc2,"% variance")) + 
    theme_bw()+
    guides(color = guide_legend(nrow = 3)) +
    theme(legend.position="top", legend.box="vertical", legend.title = element_blank(), legend.margin=margin())  + 
    geom_vline(xintercept=0, linetype="dotted") + 
    geom_hline(yintercept=0, linetype="dotted")+
    scale_color_manual(values=cell_line_colors))

(pltpca_2 <- ggplot(PCAData, aes(PC1,PC2))+
    geom_point(aes(color=condition), size=5, alpha=0.6)+ 
    xlab(paste0("PC1: ",pc1,"% variance")) +
    ylab(paste0("PC2: ",pc2,"% variance")) + 
    theme_bw()+
    theme(legend.position="top", legend.box="vertical", legend.title = element_blank(), legend.margin=margin())  + 
    geom_vline(xintercept=0, linetype="dotted") + 
    guides(color = guide_legend(nrow = 3)) +
    geom_hline(yintercept=0, linetype="dotted") +
    scale_color_manual(values=Condition_colors))


```

#### Calculate Scores

```{r}
(pltscores <- PlotScores(data = normdata, 
                         metadata = metadata, 
                         gene_sets = signatures_bidirectional[c("HernandezSegura", "SAUL_SEN_MAYO")],
                         Variable = "condition2",
                         ColorVariable = "cell_line",
                         method = "logmedian",
                         xlab = "Cellular State",
                         ColorValues = cell_line_colors,
                         labsize = 8))
```


```{r fig.width=11, fig.height=5}
plt_filt <- ggarrange(plt_beforefilt,plt_afterfilt, nrow=1)
plt_norm <- ggarrange(plt_beforenorm,plt_afternorm, nrow=1)
plt_pca <- ggarrange(pltpca_1,pltpca_2, nrow=1)

plt1 <- ggarrange(plt_filt, plt_norm, nrow=1)
plt2 <- ggarrange(plt_pca, pltscores, nrow=1, widths=c(0.5,0.5))

plt_GSE287058 <- ggarrange(plt1, plt2, ncol=1, heights = c(0.4,0.7))
plt_GSE287058 <- annotate_figure(plt_GSE287058, top = text_grob("GSE287058: Quiescent Fibroblasts\n", face = "bold", size = 14))
plt_GSE287058


```

### Compile All

```{r fig.width=11, fig.height=16} 
add_margin <- function(plt) plt + theme(plot.margin = margin(t=15, r=5, b=15, l=5))

(plt_3_3 <- ggarrange(add_margin(plt_GSE297560), 
                      add_margin(plt_GSE288940), 
                      add_margin(plt_GSE287058), 
                      ncol=1, labels = c("A","B","C"), 
                      heights = c(0.35,0.35,0.4)))

```

```{r}
ggplot2::ggsave("../Figures/Figs/R3_3_InVitroValidation.png",
                plt_3_3,
                width = 11, height=16, bg = 'white')

ggplot2::ggsave("../Figures/Figs/R3_3_InVitroValidation.pdf",
                plt_3_3,
                width = 11, height=16, bg = 'white')
```


## Comment 1.6 — Applicability of markeR beyond senescence

R.: Apply markeR in discovery mode to a small set of additional biological phenotypes with well-established gene sets and positive control datasets, to demonstrate that the framework generalises beyond senescence. Phenotypes considered include cell proliferation, epithelial-to-mesenchymal transition, and viral/interferon response.



### Sex Signature | GTEx

```{r}
library(biomaRt)
mart <- useMart(biomart="ensembl", dataset="hsapiens_gene_ensembl")
y_genes <- getBM(attributes = c("chromosome_name", "hgnc_symbol"),
                 filters = "chromosome_name", values = "Y", mart = mart)

x_genes <- getBM(attributes = c("chromosome_name", "hgnc_symbol"),
                 filters = "chromosome_name", values = "X", mart = mart)

y_unique <- y_genes[!y_genes$hgnc_symbol %in% x_genes$hgnc_symbol, ]
y_unique <- y_unique[y_unique$hgnc_symbol != "", ]

geneset_Ychromosome <- list(Y_Chromosome = y_unique$hgnc_symbol)
```


```{r}
#  
# base_url <- "https://raw.githubusercontent.com/DiseaseTranscriptomicsLab/voyAGEr/main/Corrected_Counts/"
# 
tissues <- c(
  "Adipose - Subcutaneous",
  "Adipose - Visceral (Omentum)",
  "Adrenal Gland",
  "Artery - Aorta",
  "Artery - Coronary",
  "Artery - Tibial",
  "Brain - Amygdala",
  "Brain - Anterior cingulate cortex (BA24)",
  "Brain - Caudate (basal ganglia)",
  "Brain - Cerebellar Hemisphere",
  "Brain - Cerebellum",
  "Brain - Cortex",
  "Brain - Frontal Cortex (BA9)",
  "Brain - Hippocampus",
  "Brain - Hypothalamus",
  "Brain - Nucleus accumbens (basal ganglia)",
  "Brain - Putamen (basal ganglia)",
  "Brain - Spinal cord (cervical c-1)",
  "Brain - Substantia nigra",
  "Breast - Mammary Tissue",
  "Cells - Cultured fibroblasts",
  "Cells - EBV-transformed lymphocytes",
  "Colon - Sigmoid",
  "Colon - Transverse",
  "Esophagus - Gastroesophageal Junction",
  "Esophagus - Mucosa",
  "Esophagus - Muscularis",
  "Heart - Atrial Appendage",
  "Heart - Left Ventricle",
  "Kidney - Cortex",
  "Liver",
  "Lung",
  "Minor Salivary Gland",
  "Muscle - Skeletal",
  "Nerve - Tibial",
  "Ovary",
  "Pancreas",
  "Pituitary",
  "Prostate",
  "Skin - Not Sun Exposed (Suprapubic)",
  "Skin - Sun Exposed (Lower leg)",
  "Small Intestine - Terminal Ileum",
  "Spleen",
  "Stomach",
  "Testis",
  "Thyroid",
  "Uterus",
  "Vagina",
  "Whole Blood"
)
# 
# gtex <- lapply(tissues, function(tissue) {
#   url <- paste0(base_url, URLencode(paste0(tissue, "_corrected_logCPM.csv.gz")))
#   message("Loading: ", tissue)
#   tryCatch({
#     df <- as.data.frame(fread(url, data.table = FALSE))
#     rownames(df) <- df[[1]]
#     df[[1]] <- NULL
#     df
#   }, error = function(e) { warning("Failed: ", tissue, " — ", e$message); NULL })
# })
# 
# names(gtex) <- tissues
# saveRDS(gtex, "../data_aux/data_gtex.rds")
```

```{r}
#GTEx_alltissues <- readRDS("../data_aux/GTExV8_voyagercorrected.rds") # https://github.com/DiseaseTranscriptomicsLab/voyAGEr/tree/main/Corrected_Counts
data_gtex <- readRDS("../data_aux/data_gtex.rds")

metadata_GTEx_alltissues <- readRDS("../data_aux/GTExV8_metadata.rds") # Restricted access
metadata_GTEx_alltissues$SEX <- ifelse(metadata_GTEx_alltissues$SEX == 1,"Male","Female")
metadata_GTEx_alltissues
```

```{r}
table(metadata_GTEx_alltissues$SMTSD,
      metadata_GTEx_alltissues$SEX)
```



```{r fig.width=11, fig.height=16}
# Assuming your metadata df is called `metadata` with columns SMTSD, sample_id, SEX (or similar)

# Step 1: Calculate scores for all tissues
scores_list <- lapply(tissues, function(tissue) {
  message("Scoring: ", tissue)
  tryCatch({
    expr       <- 2^data_gtex[[tissue]]
    meta_sub   <- metadata_GTEx_alltissues[metadata_GTEx_alltissues$SMTSD == tissue, ]
    # Keep only samples present in expression matrix
    meta_sub   <- meta_sub[meta_sub[[1]] %in% colnames(expr), ]
    #order, just in case
    expr <- expr[, meta_sub[[1]]]
    if (nrow(meta_sub) == 0) { warning("No matching samples: ", tissue); return(NULL) }
    
    CalculateScores(
      data      = expr,
      metadata  = meta_sub,
      method    = "logmedian",
      gene_sets = geneset_Ychromosome
    )
  }, error = function(e) { warning("Failed: ", tissue, " — ", e$message); NULL })
})
names(scores_list) <- tissues


# Step 2: Plot male vs female per tissue
# Define sex-specific tissues to skip
sex_specific <- c("Ovary", "Uterus", "Vagina", "Prostate", "Testis")

# Assumes scores df has a column with the score values and SEX in metadata
# Adjust score_col and sex_col to match your actual column names
score_col <- "logmedian"   # column name in CalculateScores output
sex_col   <- "SEX"         # column in metadata (e.g. 1=Male, 2=Female or "Male"/"Female")

plots_list <- lapply(tissues, function(tissue) {
  if (tissue %in% sex_specific) return(NULL)
  
  df <- scores_list[[tissue]][["Y_Chromosome"]]
  if (is.null(df)) return(NULL)
  
  # Merge scores with sex info if not already present
  if (!sex_col %in% colnames(df)) {
    meta_sub <- metadata_GTEx_alltissues[metadata_GTEx_alltissues$SMTSD == tissue, c("SAMPID","SMTSD","SEX","AGE")]
    df <- merge(df, meta_sub, by = 1)
  } else {
    df <- df[,c("sample","score","SMTSD","SEX","AGE")]
    
  }
  
  #df[[sex_col]] <- factor(df[[sex_col]], levels = c(1, 2), labels = c("Male", "Female"))
  
  ggplot(df, aes(x = SEX, y = score, fill = SEX)) +
    geom_jitter(color = "#DDDDDD", width = 0.15) +
    geom_violin(alpha = 0.4) +
    stat_summary(fun = median, geom = "crossbar", width = 0.1, color = "black") +
    scale_fill_manual(values = c("Male" = "#4E8ECC", "Female" = "#CC4E7A")) +
    labs(title = tissue, x = NULL,
         y = expression(log[2]*"-median score")) +
    theme_bw() +
    theme(legend.position  = "none",
          plot.title      = element_text(hjust = 0.5, size = 10))
})
names(plots_list) <- tissues
plots_list <- Filter(Negate(is.null), plots_list)

# Combine into one figure (optional)
ggarrange(plotlist = plots_list, ncol = 6, nrow = ceiling(length(plots_list) / 6))
```
```{r fig.width=4, fig.height=3}
df_all_sex <- bind_rows(
  lapply(tissues[!tissues %in% sex_specific], function(tissue) {
    df <- scores_list[[tissue]][["Y_Chromosome"]]
    if (is.null(df)) return(NULL)
    df$tissue <- tissue
    df
  })
)


(plt_ychromosomescores <- ggplot(df_all_sex, aes(x = .data[[sex_col]], y = score, fill = .data[[sex_col]])) + 
    geom_jitter(color = "#DDDDDD", width = 0.3, alpha = 0.3, size = 0.5) +
    geom_violin(alpha = 0.5) +
    stat_summary(fun = median, geom = "crossbar", width = 0.1, color = "black") +
    scale_fill_manual(values = c("Male" = "#4E8ECC", "Female" = "#CC4E7A")) +
    labs(x = NULL, y = expression(log[2]*"-median score")) +
    theme_bw() +
    theme(legend.position = "none",
          plot.title = element_text(hjust = 0.5, size = 14),
          text = element_text(size = 14))  +
    ggtitle("Y Chromosome Gene Set (biomaRt)\nacross GTEx Tissues"))
```

```{r fig.width=4, fig.height=3}
# compute sample counts per group
n_labels <- df_all_sex %>%
  group_by(.data[[sex_col]]) %>%
  summarise(n = n()) %>%
  mutate(label = paste0(.data[[sex_col]], "\n(n=", n, ")"))

label_vec <- setNames(n_labels$label, n_labels[[sex_col]])

(plt_ychromosomescores <- ggplot(df_all_sex, aes(x = .data[[sex_col]], y = score, fill = .data[[sex_col]])) + 
    geom_jitter(color = "#DDDDDD", width = 0.3, alpha = 0.3, size = 0.5) +
    geom_violin(alpha = 0.5) +
    stat_summary(fun = median, geom = "crossbar", width = 0.1, color = "black") +
    scale_fill_manual(values = c("Male" = "#4E8ECC", "Female" = "#CC4E7A")) +
    scale_x_discrete(labels = label_vec) +
    labs(x = NULL, y = expression(log[2]*"-median score")) +
    theme_bw() +
    theme(legend.position = "none",
          plot.title = element_text(hjust = 0.5, size = 14),
          text = element_text(size = 14)) +
    ggtitle("Y Chromosome gene set\nacross GTEx Tissues"))
```


### Respiratory Infection Signature | GSE161731 (many viral and bacterial infections)


```{r}
signature_respiratoryviral <- list(Zaas2009_RespiratoryViralResponse = as.vector(fread("../data_aux/RespiratoryViralResponse_Zaas2009_simple.txt",header=F)$V1))
```

```{r}
data <- as.data.frame(fread("../data_aux/GSE161731_counts.csv"))
row.names(data) <- data$V1
data$V1 <- NULL
# Convert row names from ENSEMBL ID to Gene Name using biomaRt

# Connect to Ensembl
mart <- useMart("ensembl", dataset = "hsapiens_gene_ensembl")

# Query the conversion
gene_conversion <- getBM(
  attributes = c("ensembl_gene_id", "hgnc_symbol"),
  filters    = "ensembl_gene_id",
  values     = rownames(data),
  mart       = mart
)

# Replace rownames — keeping only genes with a match and a non-empty symbol
gene_conversion <- gene_conversion[!duplicated(gene_conversion$hgnc_symbol), ]
data <- data[gene_conversion$ensembl_gene_id, ]
rownames(data) <- gene_conversion$hgnc_symbol

data
```

```{r}
metadata <- as.data.frame(fread("../data_aux/GSE161731_counts_key.csv"))
colnames(metadata)[1] <- "sampleID"
metadata <- metadata[metadata$sampleID %in% colnames(data),]
data <- data[,metadata$sampleID]
```

```{r}
data
metadata
```


##### Filtering

```{r}
# Plot
(plt_beforefilt <- ggplot() + aes(x=log10(rowSums(data))) +
   geom_density(fill="grey", alpha=0.5) + theme_bw() + 
   theme(axis.text.x = element_text(angle = 45, hjust=1),
         plot.title = element_text(hjust = 0.5),
         plot.subtitle = element_text(hjust = 0.5)) +
   ggtitle("Before Filtering") +
   xlab(expression(log[10]~"(counts per gene)")) +
   ylab("Frequency") + 
   geom_vline(xintercept = 4.2, linetype="dotted",color = "black", size=1.5))

```


```{r}
filtdata <- data[rowSums(data) > 10^4.2,]
nrow(data)
nrow(filtdata)


cat("Number of genes before filtering:", nrow(data), "\n")
cat("Number of genes after filtering:", nrow(filtdata), "\n")

```


```{r}

# Plot
(plt_afterfilt <- ggplot() + aes(x=log10(rowSums(filtdata))) +
   geom_density(fill="grey", alpha=0.5) + theme_bw() + 
   theme(axis.text.x = element_text(angle = 45, hjust=1),
         plot.title = element_text(hjust = 0.5),
         plot.subtitle = element_text(hjust = 0.5)) +
   ggtitle("After Filtering") +
   xlab(expression(log[10]~"(counts per gene)")) +
   ylab("Frequency") )

```



##### Normalisation

```{r fig.width=6, fig.height=4}

filtcounts_ggplot <- stack(log2(filtdata))
colnames(filtcounts_ggplot) <- c("expression", "sampleID")
filtcounts_ggplot <- merge(filtcounts_ggplot,metadata, by="sampleID")


(plt_beforenorm <- ggplot(filtcounts_ggplot, aes(sampleID,expression)) + geom_boxplot(alpha=0.4, outlier.size = 0.5)+ xlab("Samples") + ylab("log2(counts)")   +  theme(axis.text.x = element_text(angle = 45, hjust=1))+
    theme_bw()+ 
    theme(plot.title = element_text(hjust = 0.5),
          plot.subtitle = element_text(hjust = 0.5),
          axis.text.x=element_blank(),
          axis.ticks.x=element_blank()) +
    ggtitle("Before Normalisation") )

```


```{r}
dgelist_normcounts <- DGEList(filtdata) # the matrix with gene names as row names and counts in columns
dgelist_normcounts <- calcNormFactors(dgelist_normcounts, method = "TMM") 
normalised_counts <- as.data.frame(voom(dgelist_normcounts,plot=F)$E) 
#normalised_counts <- as.data.frame(edgeR::cpm(dgelist_normcounts, log = FALSE)) 
normdata <- 2^normalised_counts
lognormdata <- log2(normdata+1)
```


```{r fig.width=6, fig.height=4}

normcounts_ggplot <- stack(log2(normdata))
colnames(normcounts_ggplot) <- c("expression", "sampleID")
normcounts_ggplot <- merge(normcounts_ggplot,metadata, by="sampleID")


(plt_afternorm <- ggplot(normcounts_ggplot, aes(sampleID,expression)) + geom_boxplot(alpha=0.4, outlier.size = 0.5)+ xlab("Samples") + ylab("log2(counts)")   +  theme(axis.text.x = element_text(angle = 45, hjust=1))+
    theme_bw()+ 
    theme(plot.title = element_text(hjust = 0.5),
          plot.subtitle = element_text(hjust = 0.5),
          axis.text.x=element_blank(),
          axis.ticks.x=element_blank()) +
    ggtitle("After Normalisation") )

```


##### PCA

```{r fig.width=6, fig.height=4}
y <- DGEList(lognormdata, samples= metadata)
PCA  <- prcomp(t(y$counts), scale=FALSE, center=TRUE)
PCACounts  <- PCA$x
PCACounts  <- as.data.frame(PCACounts )
PCAData <-  cbind(PCACounts[1:10],y$samples) 

ev = PCA$sdev^2 
pc1= round(100*ev[1]/sum(ev),2) 
pc2 = round(100*ev[2]/sum(ev),2) 
pc3 = round(100*ev[3]/sum(ev),2) 
pc4 = round(100*ev[4]/sum(ev),2) 
pc5 = round(100*ev[5]/sum(ev),2) 
pc6 = round(100*ev[6]/sum(ev),2)  


(pltpca_1 <- ggplot(PCAData, aes(PC1,PC2))+
    geom_point(aes(color=cohort), size=5, alpha=0.6)+ 
    xlab(paste0("PC1: ",pc1,"% variance")) +
    ylab(paste0("PC2: ",pc2,"% variance")) + 
    theme_bw()+
    guides(color = guide_legend(nrow = 2)) +
    theme(legend.position="top", legend.box="vertical", legend.title = element_blank(), legend.margin=margin())  + 
    geom_vline(xintercept=0, linetype="dotted") + 
    geom_hline(yintercept=0, linetype="dotted") )

(pltpca_2 <- ggplot(PCAData, aes(PC1,PC2))+
    geom_point(aes(color=gender), size=5, alpha=0.6)+ 
    xlab(paste0("PC1: ",pc1,"% variance")) +
    ylab(paste0("PC2: ",pc2,"% variance")) + 
    theme_bw()+
    theme(legend.position="top", legend.box="vertical", legend.title = element_blank(), legend.margin=margin())  + 
    geom_vline(xintercept=0, linetype="dotted") + 
    guides(color = guide_legend(nrow = 2)) +
    geom_hline(yintercept=0, linetype="dotted"))

(pltpca_3 <- ggplot(PCAData, aes(PC1,PC2))+
    geom_point(aes(color=race), size=5, alpha=0.6)+ 
    xlab(paste0("PC1: ",pc1,"% variance")) +
    ylab(paste0("PC2: ",pc2,"% variance")) + 
    theme_bw()+
    theme(legend.position="top", legend.box="vertical", legend.title = element_blank(), legend.margin=margin())  + 
    geom_vline(xintercept=0, linetype="dotted") + 
    guides(color = guide_legend(nrow = 2)) +
    geom_hline(yintercept=0, linetype="dotted"))

(pltpca_4 <- ggplot(PCAData, aes(PC1,PC2))+
    geom_point(aes(color=batch), size=5, alpha=0.6)+ 
    xlab(paste0("PC1: ",pc1,"% variance")) +
    ylab(paste0("PC2: ",pc2,"% variance")) + 
    theme_bw()+
    theme(legend.position="top", legend.box="vertical", legend.title = element_blank(), legend.margin=margin())  + 
    geom_vline(xintercept=0, linetype="dotted") + 
    guides(color = guide_legend(nrow = 2)) +
    geom_hline(yintercept=0, linetype="dotted"))


```

#### Calculate Scores


```{r}
#saveRDS(normdata,"../data_aux/data_RespiratoryInfection.rds")
data_RespiratoryInfection <- readRDS("../data_aux/data_RespiratoryInfection.rds")
#saveRDS(metadata,"../data_aux/metadata_RespiratoryInfection.rds")
metadata_ResporatoryInfection <- readRDS("../data_aux/metadata_RespiratoryInfection.rds")
```


```{r fig.width=4, fig.height=3}
(dfscores <- CalculateScores(data = data_RespiratoryInfection, 
                             metadata = metadata_ResporatoryInfection, 
                             gene_sets = signature_respiratoryviral,
                             method = "logmedian")$Zaas2009_RespiratoryViralResponse)

dfscores$cohort2 <- ifelse(dfscores$cohort %in% c("CoV other","COVID-19"), "CoV", dfscores$cohort)
dfscores$cohort2 <- gsub("healthy","Healthy",dfscores$cohort2)
cohort_colors <- c(
  "CoV"       = "#E07B54",
  "Influenza" = "#5B8DB8",
  "Bacterial" = "#6BAF92",
  "Healthy"   = "#B0B0B0"
)

(plt_viralrespiratory <- ggplot(dfscores, aes(x = cohort2, y = score, fill=cohort2)) + 
    geom_jitter(color = "black", width = 0.3, alpha = 0.4, size = 0.5) +
    geom_violin(alpha = 0.7) +
    stat_summary(fun = median, geom = "crossbar", width = 0.1, color = "black") + 
    labs(x = NULL, y = expression(log[2]*"-median score")) +
    theme_bw() +
    theme(legend.position="none",
          plot.title = element_text(hjust = 0.5, size = 14),
          text = element_text(size = 14)) +
    ggtitle("Zaas 2009 Respiratory Viral Gene Set \nin Whole Blood \nacross Acute Respiratory Infections") + scale_fill_manual(values = cohort_colors))
```
```{r fig.width=4, fig.height=3}
n_labels <- dfscores %>%
  group_by(cohort2) %>%
  summarise(n = n()) %>%
  mutate(label = paste0(cohort2, "\n(n=", n, ")"))

label_vec <- setNames(n_labels$label, n_labels$cohort2)

(plt_viralrespiratory <- ggplot(dfscores, aes(x = cohort2, y = score, fill = cohort2)) + 
    geom_jitter(color = "black", width = 0.3, alpha = 0.4, size = 0.5) +
    geom_violin(alpha = 0.7) +
    stat_summary(fun = median, geom = "crossbar", width = 0.1, color = "black") + 
    labs(x = NULL, y = expression(log[2]*"-median score")) +
    theme_bw() +
    theme(legend.position = "none",
          plot.title = element_text(hjust = 0.5, size = 14),
          text = element_text(size = 14)) +
    ggtitle("Respiratory Viral Response gene set\nacross Acute Respiratory Infections") +
    scale_fill_manual(values = cohort_colors) +
    scale_x_discrete(labels = label_vec))
```


<!-- ### EMT transition | KIRC -->

<!-- ```{r} -->
<!-- Ramaswamy_Metastasis <- list(Ramaswamy_Metastasis = as.data.frame(fread("../data_aux/RAMASWAMY_METASTASIS_simple.txt",header=T))) -->
<!-- ``` -->

<!-- https://xenabrowser.net/datapages/?dataset=TCGA-KIRC.star_counts.tsv&host=https%3A%2F%2Fgdc.xenahubs.net&removeHub=https%3A%2F%2Fxena.treehouse.gi.ucsc.edu%3A443 -->

<!-- KIRC was chosen given that it has a relatively well documented tumour grade -->

<!-- ```{r} -->
<!-- data <- as.data.frame(fread("../data_aux/TCGA-KIRC.star_counts.tsv")) -->
<!-- data$Ensembl_ID <- sub("\\..*", "", data$Ensembl_ID) -->

<!-- # combine rows with same gene id by summing them -->
<!-- data <- rowsum(data[, -1], group = data$Ensembl_ID) -->

<!-- # Connect to Ensembl -->
<!-- mart <- useMart("ensembl", dataset = "hsapiens_gene_ensembl") -->

<!-- # Query the conversion -->
<!-- gene_conversion <- getBM( -->
<!--   attributes = c("ensembl_gene_id", "hgnc_symbol"), -->
<!--   filters    = "ensembl_gene_id", -->
<!--   values     = rownames(data), -->
<!--   mart       = mart -->
<!-- ) -->

<!-- # Replace rownames — keeping only genes with a match and a non-empty symbol -->
<!-- gene_conversion <- gene_conversion[!duplicated(gene_conversion$hgnc_symbol), ] -->
<!-- data <- data[gene_conversion$ensembl_gene_id, ] -->
<!-- rownames(data) <- gene_conversion$hgnc_symbol -->

<!-- data <- 2^data - 1 # unlog... -->
<!-- data -->

<!-- ``` -->

<!-- ```{r} -->
<!-- metadata <- as.data.frame(fread("../data_aux/TCGA-KIRC.clinical.tsv")) -->
<!-- metadata <- metadata[metadata$sample %in% colnames(data),] -->
<!-- data <- data[,metadata$sample] -->
<!-- ``` -->


<!-- ```{r} -->
<!-- metadata$code <- sub(".*-", "", metadata$sample) -->
<!-- table(metadata$code) # no metastasis -->
<!-- table(metadata$tissue_type.samples) # normal vs tumour -->
<!-- table(metadata$sample_type.samples, metadata$ajcc_pathologic_m.diagnoses) -->

<!-- # subset data to M1 or M0 only -->
<!-- # subset data to primary tumour only -->
<!-- metadata <- metadata[metadata$ajcc_pathologic_m.diagnoses %in% c("M0","M1") & metadata$sample_type.samples != "metadata$sample_type.samples" & metadata$tissue_type.samples == "Tumor",] -->

<!-- table(metadata$code) # no metastasis -->
<!-- table(metadata$tissue_type.samples) # normal vs tumour -->
<!-- table(metadata$sample_type.samples, metadata$ajcc_pathologic_m.diagnoses) -->

<!-- data <- data[,metadata$sample] -->

<!-- colnames(metadata)[1] <- "sampleID" -->
<!-- ``` -->


<!-- ```{r} -->

<!-- data -->
<!-- metadata -->
<!-- ``` -->




<!-- ##### Filtering -->

<!-- ```{r} -->
<!-- # Plot -->
<!-- (plt_beforefilt <- ggplot() + aes(x=log10(rowSums(data))) + -->
<!--    geom_density(fill="grey", alpha=0.5) + theme_bw() +  -->
<!--    theme(axis.text.x = element_text(angle = 45, hjust=1), -->
<!--          plot.title = element_text(hjust = 0.5), -->
<!--          plot.subtitle = element_text(hjust = 0.5)) + -->
<!--    ggtitle("Before Filtering") + -->
<!--    xlab(expression(log[10]~"(counts per gene)")) + -->
<!--    ylab("Frequency") +  -->
<!--    geom_vline(xintercept = 5.2, linetype="dotted",color = "black", size=1.5)) -->

<!-- ``` -->


<!-- ```{r} -->
<!-- filtdata <- data[rowSums(data) > 10^5.2,] -->
<!-- nrow(data) -->
<!-- nrow(filtdata) -->


<!-- cat("Number of genes before filtering:", nrow(data), "\n") -->
<!-- cat("Number of genes after filtering:", nrow(filtdata), "\n") -->

<!-- ``` -->


<!-- ```{r} -->

<!-- # Plot -->
<!-- (plt_afterfilt <- ggplot() + aes(x=log10(rowSums(filtdata))) + -->
<!--    geom_density(fill="grey", alpha=0.5) + theme_bw() +  -->
<!--    theme(axis.text.x = element_text(angle = 45, hjust=1), -->
<!--          plot.title = element_text(hjust = 0.5), -->
<!--          plot.subtitle = element_text(hjust = 0.5)) + -->
<!--    ggtitle("After Filtering") + -->
<!--    xlab(expression(log[10]~"(counts per gene)")) + -->
<!--    ylab("Frequency") ) -->

<!-- ``` -->



<!-- ##### Normalisation -->

<!-- ```{r fig.width=6, fig.height=4} -->

<!-- # filtcounts_ggplot <- stack(log2(filtdata)) -->
<!-- # colnames(filtcounts_ggplot) <- c("expression", "sampleID") -->
<!-- # filtcounts_ggplot <- merge(filtcounts_ggplot,metadata, by="sampleID") -->
<!-- #    -->


<!-- total_coverage <- colSums(filtdata) -->
<!-- top50 <- names(sort(total_coverage, decreasing = TRUE)[1:50]) -->
<!-- bot50 <- names(sort(total_coverage)[1:50]) -->
<!-- selected_samples <- c(bot50, top50) -->

<!-- filtdata_sub <- filtdata[, selected_samples] -->

<!-- filtcounts_ggplot <- stack(log2(filtdata_sub)) -->
<!-- colnames(filtcounts_ggplot) <- c("expression", "sampleID") -->
<!-- filtcounts_ggplot <- merge(filtcounts_ggplot, metadata, by = "sampleID") -->

<!-- # order by total coverage -->
<!-- filtcounts_ggplot$sampleID <- factor(filtcounts_ggplot$sampleID, -->
<!--                                      levels = names(sort(total_coverage[selected_samples]))) -->

<!-- (plt_beforenorm <- ggplot(filtcounts_ggplot, aes(sampleID,expression)) + geom_boxplot(alpha=0.4, outlier.size = 0.5)+ xlab("Samples") + ylab("log2(counts)")   +  theme(axis.text.x = element_text(angle = 45, hjust=1))+ -->
<!--     theme_bw()+  -->
<!--     theme(plot.title = element_text(hjust = 0.5), -->
<!--           plot.subtitle = element_text(hjust = 0.5), -->
<!--           axis.text.x=element_blank(), -->
<!--           axis.ticks.x=element_blank()) + -->
<!--     ggtitle("Before Normalisation") ) -->

<!-- ``` -->


<!-- ```{r} -->
<!-- dgelist_normcounts <- DGEList(filtdata) # the matrix with gene names as row names and counts in columns -->
<!-- dgelist_normcounts <- calcNormFactors(dgelist_normcounts, method = "TMM")  -->
<!-- normalised_counts <- as.data.frame(voom(dgelist_normcounts,plot=F)$E)  -->
<!-- #normalised_counts <- as.data.frame(edgeR::cpm(dgelist_normcounts, log = FALSE))  -->
<!-- normdata <- 2^normalised_counts -->
<!-- lognormdata <- log2(normdata+1) -->
<!-- ``` -->



<!-- ```{r fig.width=6, fig.height=4} -->

<!-- total_coverage <- colSums(normdata) -->
<!-- top50 <- names(sort(total_coverage, decreasing = TRUE)[1:50]) -->
<!-- bot50 <- names(sort(total_coverage)[1:50]) -->
<!-- selected_samples <- c(bot50, top50) -->

<!-- normdata_sub <- normdata[, selected_samples] -->

<!-- normcounts_ggplot <- stack(log2(normdata_sub)) -->
<!-- colnames(normcounts_ggplot) <- c("expression", "sampleID") -->
<!-- normcounts_ggplot <- merge(normcounts_ggplot, metadata, by = "sampleID") -->

<!-- # order by total coverage -->
<!-- normcounts_ggplot$sampleID <- factor(normcounts_ggplot$sampleID, -->
<!--                                      levels = names(sort(total_coverage[selected_samples]))) -->

<!-- (plt_afternorm <- ggplot(normcounts_ggplot, aes(sampleID,expression)) + geom_boxplot(alpha=0.4, outlier.size = 0.5)+ xlab("Samples") + ylab("log2(counts)")   +  theme(axis.text.x = element_text(angle = 45, hjust=1))+ -->
<!--     theme_bw()+  -->
<!--     theme(plot.title = element_text(hjust = 0.5), -->
<!--           plot.subtitle = element_text(hjust = 0.5), -->
<!--           axis.text.x=element_blank(), -->
<!--           axis.ticks.x=element_blank()) + -->
<!--     ggtitle("After Normalisation") ) -->

<!-- ``` -->


<!-- #### Calculate Scores -->


<!-- ```{r fig.width=4, fig.height=3} -->
<!-- (dfscores <- CalculateScores(data = normdata,  -->
<!--                              metadata = metadata,  -->
<!--                              gene_sets = Ramaswamy_Metastasis, -->
<!--                              method = "logmedian")$Ramaswamy_Metastasis) -->

<!-- met_colors <- c( -->
<!--   "M0"       = "#E07B54", -->
<!--   "M1" = "#5B8DB8"  -->
<!-- ) -->



<!-- (plt_emt <- ggplot(dfscores, aes(x = ajcc_pathologic_m.diagnoses, y = score)) +  -->
<!--     geom_jitter(aes(color=ajcc_pathologic_m.diagnoses), width = 0.3, alpha = 1, size = 2) + -->
<!--     geom_violin(alpha = 0.5) + -->
<!--     stat_summary(fun = median, geom = "crossbar", width = 0.1, color = "black") +  -->
<!--     labs(x = NULL, y = expression(log[2]*"-median score")) + -->
<!--     theme_bw() + -->
<!--     theme(legend.position="none", -->
<!--           plot.title = element_text(hjust = 0.5, size = 14), -->
<!--           text = element_text(size = 14)) + -->
<!--     ggtitle("Ramaswamy EMT transition Gene Set\n in KIRC Primary Tumour Samples") + -->
<!--     scale_color_manual(values = met_colors)) -->
<!-- ``` -->


### Proliferation | PANCAN

meta-PCNA (Venet et al. 2011)
```{r}
metaPCNA <- list(metaPCNA = as.vector(fread("../data_aux/proliferationsignature_metaPCNA.txt",header=F))$V1)
```

https://xenabrowser.net/datapages/?dataset=EB%2B%2BAdjustPANCAN_IlluminaHiSeq_RNASeqV2.geneExp.xena&host=https%3A%2F%2Fpancanatlas.xenahubs.net&removeHub=https%3A%2F%2Fxena.treehouse.gi.ucsc.edu%3A443

```{r}
data <- as.data.frame(fread("../data_aux/EB++AdjustPANCAN_IlluminaHiSeq_RNASeqV2.geneExp.xena.csv"))
# combine rows with same gene id by summing them
data <- rowsum(data[, -1], group = data$sample)

data <- 2^data - 1 # unlog...
data

```

```{r}
metadata <- as.data.frame(fread("../data_aux/Survival_SupplementalTable_S1_20171025_xena_sp.tsv"))
metadata <- metadata[metadata$sample %in% colnames(data),]
data <- data[,metadata$sample]
```


```{r}
metadata$code <- sub(".*-", "", metadata$sample) 
colnames(metadata)[1] <- "sampleID"
metadata$sample_type <- ifelse(metadata$code == "11", "Normal", "Tumour")
```


```{r}
data
metadata
```


```{r}
data <- na.omit(data)
data <- data + 1 # some negative values, must be of batch correction
```


##### Filtering

No need for filtering!! Distributions look good, data was processed & batch corrected before

```{r}
# Plot
(plt_beforefilt <- ggplot() + aes(x=log10(rowSums(data))) +
   geom_density(fill="grey", alpha=0.5) + theme_bw() + 
   theme(axis.text.x = element_text(angle = 45, hjust=1),
         plot.title = element_text(hjust = 0.5),
         plot.subtitle = element_text(hjust = 0.5)) +
   ggtitle("Before Filtering") +
   xlab(expression(log[10]~"(counts per gene)")) +
   ylab("Frequency") + 
   geom_vline(xintercept = 5.2, linetype="dotted",color = "black", size=1.5))

```


<!-- ```{r} -->
<!-- #filtdata <- data[rowSums(data) > 10^5.2,] -->
<!-- filtdata <- data -->
<!-- nrow(data) -->
<!-- nrow(filtdata) -->


<!-- cat("Number of genes before filtering:", nrow(data), "\n") -->
<!-- cat("Number of genes after filtering:", nrow(filtdata), "\n") -->

<!-- ``` -->


<!-- ```{r} -->

<!-- # Plot -->
<!-- (plt_afterfilt <- ggplot() + aes(x=log10(rowSums(filtdata))) + -->
<!--   geom_density(fill="grey", alpha=0.5) + theme_bw() +  -->
<!--     theme(axis.text.x = element_text(angle = 45, hjust=1), -->
<!--         plot.title = element_text(hjust = 0.5), -->
<!--         plot.subtitle = element_text(hjust = 0.5)) + -->
<!--   ggtitle("After Filtering") + -->
<!--   xlab(expression(log[10]~"(counts per gene)")) + -->
<!--   ylab("Frequency") ) -->

<!-- ``` -->



<!-- ##### Normalisation -->

<!-- ```{r fig.width=6, fig.height=4} -->

<!-- # filtcounts_ggplot <- stack(log2(filtdata)) -->
<!-- # colnames(filtcounts_ggplot) <- c("expression", "sampleID") -->
<!-- # filtcounts_ggplot <- merge(filtcounts_ggplot,metadata, by="sampleID") -->
<!-- #    -->


<!-- total_coverage <- colSums(filtdata) -->
<!-- top50 <- names(sort(total_coverage, decreasing = TRUE)[1:50]) -->
<!-- bot50 <- names(sort(total_coverage)[1:50]) -->
<!-- selected_samples <- c(bot50, top50) -->

<!-- filtdata_sub <- filtdata[, selected_samples] -->

<!-- filtcounts_ggplot <- stack(log2(filtdata_sub)) -->
<!-- colnames(filtcounts_ggplot) <- c("expression", "sampleID") -->
<!-- filtcounts_ggplot <- merge(filtcounts_ggplot, metadata, by = "sampleID") -->

<!-- # order by total coverage -->
<!-- filtcounts_ggplot$sampleID <- factor(filtcounts_ggplot$sampleID, -->
<!--                                       levels = names(sort(total_coverage[selected_samples]))) -->

<!-- (plt_beforenorm <- ggplot(filtcounts_ggplot, aes(sampleID,expression)) + geom_boxplot(alpha=0.4, outlier.size = 0.5)+ xlab("Samples") + ylab("log2(counts)")   +  theme(axis.text.x = element_text(angle = 45, hjust=1))+ -->
<!--   theme_bw()+  -->
<!--   theme(plot.title = element_text(hjust = 0.5), -->
<!--         plot.subtitle = element_text(hjust = 0.5), -->
<!--         axis.text.x=element_blank(), -->
<!--         axis.ticks.x=element_blank()) + -->
<!--   ggtitle("Before Normalisation") ) -->

<!-- ``` -->


<!-- ```{r} -->
<!-- # dgelist_normcounts <- DGEList(filtdata) # the matrix with gene names as row names and counts in columns -->
<!-- # dgelist_normcounts <- calcNormFactors(dgelist_normcounts, method = "TMM")  -->
<!-- # normalised_counts <- as.data.frame(voom(dgelist_normcounts,plot=F)$E)  -->
<!-- # #normalised_counts <- as.data.frame(edgeR::cpm(dgelist_normcounts, log = FALSE))  -->
<!-- # normdata <- 2^normalised_counts -->
<!-- # lognormdata <- log2(normdata+1) -->
<!-- ``` -->
<!-- ```{r} -->
<!-- normdata <- filtdata -->
<!-- ``` -->



<!-- ```{r fig.width=6, fig.height=4} -->

<!-- total_coverage <- colSums(normdata) -->
<!-- top50 <- names(sort(total_coverage, decreasing = TRUE)[1:50]) -->
<!-- bot50 <- names(sort(total_coverage)[1:50]) -->
<!-- selected_samples <- c(bot50, top50) -->

<!-- normdata_sub <- normdata[, selected_samples] -->

<!-- normcounts_ggplot <- stack(log2(normdata_sub)) -->
<!-- colnames(normcounts_ggplot) <- c("expression", "sampleID") -->
<!-- normcounts_ggplot <- merge(normcounts_ggplot, metadata, by = "sampleID") -->

<!-- # order by total coverage -->
<!-- normcounts_ggplot$sampleID <- factor(normcounts_ggplot$sampleID, -->
<!--                                       levels = names(sort(total_coverage[selected_samples]))) -->

<!-- (plt_afternorm <- ggplot(normcounts_ggplot, aes(sampleID,expression)) + geom_boxplot(alpha=0.4, outlier.size = 0.5)+ xlab("Samples") + ylab("log2(counts)")   +  theme(axis.text.x = element_text(angle = 45, hjust=1))+ -->
<!--   theme_bw()+  -->
<!--   theme(plot.title = element_text(hjust = 0.5), -->
<!--         plot.subtitle = element_text(hjust = 0.5), -->
<!--         axis.text.x=element_blank(), -->
<!--         axis.ticks.x=element_blank()) + -->
<!--   ggtitle("After Normalisation") ) -->

<!-- ``` -->


#### Calculate Scores

```{r}
#saveRDS(data,"../data_aux/data_PANCAN.rds")
data_PANCAN <- readRDS("../data_aux/data_PANCAN.rds")
#saveRDS(metadata,"../data_aux/metadata_PANCAN.rds")
metadata_PANCAN <- readRDS("../data_aux/metadata_PANCAN.rds")
```

```{r fig.width=4, fig.height=3}
(dfscores <- CalculateScores(data = data_PANCAN, 
                             metadata = metadata_PANCAN, 
                             gene_sets = metaPCNA,
                             method = "logmedian")$metaPCNA)

samptype_col <- c(
  "Tumour" = "#C0392B",
  "Normal" = "#2E86AB"
)


(plt_prol <- ggplot(dfscores, aes(x = sample_type, y = score, fill=sample_type)) + 
    geom_jitter(color = "#DDDDDD", width = 0.3, alpha = 0.3, size = 0.5) +
    geom_violin(alpha = 0.5) +
    stat_summary(fun = median, geom = "crossbar", width = 0.1, color = "black") + 
    labs(x = NULL, y = expression(log[2]*"-median score")) +
    theme_bw() +
    theme(legend.position="none",
          plot.title = element_text(hjust = 0.5, size = 14),
          text = element_text(size = 14)) +
    ggtitle("Venet 2011 Proliferation Gene Set\n in TCGA PANCAN (Liu 2018)") +
    scale_fill_manual(values = samptype_col))
```
```{r fig.width=4, fig.height=3}
n_labels <- dfscores %>%
  group_by(sample_type) %>%
  summarise(n = n()) %>%
  mutate(label = paste0(sample_type, "\n(n=", n, ")"))

label_vec <- setNames(n_labels$label, n_labels$sample_type)

(plt_prol <- ggplot(dfscores, aes(x = sample_type, y = score, fill = sample_type)) + 
    geom_jitter(color = "#DDDDDD", width = 0.3, alpha = 0.3, size = 0.5) +
    geom_violin(alpha = 0.5) +
    stat_summary(fun = median, geom = "crossbar", width = 0.1, color = "black") + 
    labs(x = NULL, y = expression(log[2]*"-median score")) +
    theme_bw() +
    theme(legend.position = "none",
          plot.title = element_text(hjust = 0.5, size = 14),
          text = element_text(size = 14)) +
    ggtitle("meta-PCNA proliferation gene set\n in TCGA PANCAN") +
    scale_fill_manual(values = samptype_col) +
    scale_x_discrete(labels = label_vec))
```
### Yeast Experiments

```{r}
library(neat)
data(yeast)

# Inspect
length(yeast$esr1)  # rESR (repressed)
length(yeast$esr2)  # iESR (induced)

GeneSet_YeastESR <- list(Gasch2000_ESR = data.frame(gene=c(yeast$esr1,yeast$esr2),
                                                    direction=c(rep(-1,length(yeast$esr1)), rep(1,length(yeast$esr2)))))

```

```{r}
data_yeast <- as.data.frame(fread("../data_aux/GSE135430_Yeast_HeatShock.csv"))
row.names(data_yeast) <- data_yeast$gene_id
data_yeast_o <- data_yeast[,-c(1,2,3)]

data_yeast <- apply(data_yeast_o, 2, function(x) as.numeric(gsub(",", ".", x)))
rownames(data_yeast) <- rownames(data_yeast_o)
data_yeast <- as.data.frame(data_yeast)
data_yeast
```

```{r}
metadata_yeast <- as.data.frame(fread("../data_aux/GSE135430_metadata_yeast.csv"))

colname_to_gsm <- c(
  "Phenol t0 Rep1"  = "GSM4008796",
  "Phenol t0 Rep2"  = "GSM4008797",
  "Phenol t0 Rep3"  = "GSM4008798",
  "Phenol t0 Rep4"  = "GSM4008799",
  "Phenol t20 Rep1" = "GSM4008800",
  "Phenol t20 Rep2" = "GSM4008801",
  "Phenol t20 Rep3" = "GSM4008802",
  "Phenol t20 Rep4" = "GSM4008803",
  "RNeasy t0 Rep1"  = "GSM4008804",
  "RNeasy t0 Rep2"  = "GSM4008805",
  "RNeasy t0 Rep3"  = "GSM4008806",
  "RNeasy t0 Rep4"  = "GSM4008807",
  "RNeasy t20 Rep1" = "GSM4008808",
  "RNeasy t20 Rep2" = "GSM4008809",
  "RNeasy t20 Rep3" = "GSM4008810",
  "RNeasy t20 Rep4" = "GSM4008811",
  "Trizol t0 Rep1"  = "GSM4008812",
  "Trizol t0 Rep2"  = "GSM4008813",
  "Trizol t0 Rep3"  = "GSM4008814",
  "Trizol t0 Rep4"  = "GSM4008815",
  "Trizol t20 Rep1" = "GSM4008816",
  "Trizol t20 Rep2" = "GSM4008817",
  "Trizol t20 Rep3" = "GSM4008818",
  "Trizol t20 Rep4" = "GSM4008819"
)

metadata_yeast_2 <- data.frame(
  sample     = names(colname_to_gsm),
  gsm = colname_to_gsm,
  condition  = ifelse(grepl("t20", names(colname_to_gsm)), "HeatShock", "Unstressed"),
  rna_method = sub(" t[0-9]+ Rep[0-9]+", "", names(colname_to_gsm)),
  row.names  = names(colname_to_gsm)
)

metadata_yeast <- merge(metadata_yeast_2,metadata_yeast, by.y="GEO_Accession (exp)", by.x="gsm")
metadata_yeast <- metadata_yeast[,c("sample",colnames(metadata_yeast))]
metadata_yeast$sample.1 <- NULL
metadata_yeast$condition <- gsub("HeatShock","Heat Shock", metadata_yeast$condition)
```


```{r}
metadata_yeast <- metadata_yeast[metadata_yeast$sample %in% colnames(data_yeast)]
data_yeast <- data_yeast[,metadata_yeast$sample]

```

```{r}
data_yeast
metadata_yeast
```

##### Filtering


```{r}
# Plot
(plt_beforefilt <- ggplot() + aes(x=log10(rowSums(data_yeast))) +
   geom_density(fill="grey", alpha=0.5) + theme_bw() + 
   theme(axis.text.x = element_text(angle = 45, hjust=1),
         plot.title = element_text(hjust = 0.5),
         plot.subtitle = element_text(hjust = 0.5)) +
   ggtitle("Before Filtering") +
   xlab(expression(log[10]~"(counts per gene)")) +
   ylab("Frequency") + 
   geom_vline(xintercept = 3.4, linetype="dotted",color = "black", size=1.5))

```


```{r}
filtdata <- data_yeast[rowSums(data_yeast) > 10^3.4,]
nrow(data_yeast)
nrow(filtdata)


cat("Number of genes before filtering:", nrow(data_yeast), "\n")
cat("Number of genes after filtering:", nrow(filtdata), "\n")

```


```{r}

# Plot
(plt_afterfilt <- ggplot() + aes(x=log10(rowSums(filtdata))) +
   geom_density(fill="grey", alpha=0.5) + theme_bw() + 
   theme(axis.text.x = element_text(angle = 45, hjust=1),
         plot.title = element_text(hjust = 0.5),
         plot.subtitle = element_text(hjust = 0.5)) +
   ggtitle("After Filtering") +
   xlab(expression(log[10]~"(counts per gene)")) +
   ylab("Frequency") )

```


##### Normalisation

```{r fig.width=6, fig.height=4}

filtcounts_ggplot <- stack(log2(filtdata))
colnames(filtcounts_ggplot) <- c("expression", "sample")
filtcounts_ggplot <- merge(filtcounts_ggplot,metadata_yeast, by="sample")


(plt_beforenorm <- ggplot(filtcounts_ggplot, aes(sample,expression)) + geom_boxplot(alpha=0.4, outlier.size = 0.5)+ xlab("Samples") + ylab("log2(counts)")   +  theme(axis.text.x = element_text(angle = 45, hjust=1))+
    theme_bw()+ 
    theme(plot.title = element_text(hjust = 0.5),
          plot.subtitle = element_text(hjust = 0.5),
          axis.text.x=element_blank(),
          axis.ticks.x=element_blank()) +
    ggtitle("Before Normalisation") )

```


```{r}
dgelist_normcounts <- DGEList(filtdata) # the matrix with gene names as row names and counts in columns
dgelist_normcounts <- calcNormFactors(dgelist_normcounts, method = "TMM") 
normalised_counts <- as.data.frame(voom(dgelist_normcounts,plot=F)$E) 
#normalised_counts <- as.data.frame(edgeR::cpm(dgelist_normcounts, log = FALSE)) 
normdata <- 2^normalised_counts
lognormdata <- log2(normdata+1)
```


```{r fig.width=6, fig.height=4}

normcounts_ggplot <- stack(log2(normdata))
colnames(normcounts_ggplot) <- c("expression", "sample")
normcounts_ggplot <- merge(normcounts_ggplot,metadata_yeast, by="sample")


(plt_afternorm <- ggplot(normcounts_ggplot, aes(sample,expression)) + geom_boxplot(alpha=0.4, outlier.size = 0.5)+ xlab("Samples") + ylab("log2(counts)")   +  theme(axis.text.x = element_text(angle = 45, hjust=1))+
    theme_bw()+ 
    theme(plot.title = element_text(hjust = 0.5),
          plot.subtitle = element_text(hjust = 0.5),
          axis.text.x=element_blank(),
          axis.ticks.x=element_blank()) +
    ggtitle("After Normalisation") )

```


##### PCA

```{r fig.width=6, fig.height=4}
y <- DGEList(lognormdata, samples= metadata_yeast)
PCA  <- prcomp(t(y$counts), scale=FALSE, center=TRUE)
PCACounts  <- PCA$x
PCACounts  <- as.data.frame(PCACounts )
PCAData <-  cbind(PCACounts[1:10],y$samples) 

ev = PCA$sdev^2 
pc1= round(100*ev[1]/sum(ev),2) 
pc2 = round(100*ev[2]/sum(ev),2) 
pc3 = round(100*ev[3]/sum(ev),2) 
pc4 = round(100*ev[4]/sum(ev),2) 
pc5 = round(100*ev[5]/sum(ev),2) 
pc6 = round(100*ev[6]/sum(ev),2)  


(pltpca_1 <- ggplot(PCAData, aes(PC1,PC2))+
    geom_point(aes(color=condition), size=5, alpha=0.6)+ 
    xlab(paste0("PC1: ",pc1,"% variance")) +
    ylab(paste0("PC2: ",pc2,"% variance")) + 
    theme_bw()+
    guides(color = guide_legend(nrow = 2)) +
    theme(legend.position="top", legend.box="vertical", legend.title = element_blank(), legend.margin=margin())  + 
    geom_vline(xintercept=0, linetype="dotted") + 
    geom_hline(yintercept=0, linetype="dotted") )

(pltpca_2 <- ggplot(PCAData, aes(PC1,PC2))+
    geom_point(aes(color=rna_method), size=5, alpha=0.6)+ 
    xlab(paste0("PC1: ",pc1,"% variance")) +
    ylab(paste0("PC2: ",pc2,"% variance")) + 
    theme_bw()+
    theme(legend.position="top", legend.box="vertical", legend.title = element_blank(), legend.margin=margin())  + 
    geom_vline(xintercept=0, linetype="dotted") + 
    guides(color = guide_legend(nrow = 2)) +
    geom_hline(yintercept=0, linetype="dotted"))


```

#### Calculate Scores

```{r fig.width=4, fig.height=3}
(dfscores <- CalculateScores(data = normdata, 
                             metadata = metadata_yeast, 
                             gene_sets = GeneSet_YeastESR,
                             method = "logmedian")$Gasch2000_ESR)

condition_colors <- c(
  "Unstressed" = "#F5C842",  # warm yellow
  "Heat Shock"  = "#C0392B"   # red
)

(plt_yeast_heatshock <- ggplot(dfscores, aes(x = condition, y = score, fill=condition)) + 
    geom_jitter(color = "black", width = 0.3, alpha = 0.4, size = 0.5) +    geom_violin(alpha = 0.7) +
    stat_summary(fun = median, geom = "crossbar", width = 0.1, color = "black") + 
    labs(x = NULL, y = expression(log[2]*"-median score")) +
    theme_bw() +
    theme(legend.position="none",
          plot.title = element_text(hjust = 0.5, size = 14),
          text = element_text(size = 14)) +
    ggtitle("Environmental Stress Response (ESR) \ngene set (Gasch et al. 2000) \nin S. cerevisiae Heat Shock") + scale_fill_manual(values = condition_colors))
```
```{r}
# Compute label vector with counts
label_vec <- setNames(
  paste0(levels(factor(dfscores$condition)), "\n(n=", table(dfscores$condition)[levels(factor(dfscores$condition))], ")"),
  levels(factor(dfscores$condition))
)

(plt_yeast_heatshock <- ggplot(dfscores, aes(x = condition, y = score, fill = condition)) + 
    geom_jitter(color = "black", width = 0.3, alpha = 0.4, size = 0.5) +
    geom_violin(alpha = 0.7) +
    stat_summary(fun = median, geom = "crossbar", width = 0.1, color = "black") + 
    labs(x = NULL, y = expression(log[2]*"-median score")) +
    scale_x_discrete(labels = label_vec) +
    theme_bw() +
    theme(legend.position = "none",
          plot.title = element_text(hjust = 0.5, size = 14),
          text = element_text(size = 14)) +
    ggtitle("Environmental Stress Response (ESR) \ngene set in S. cerevisiae Heat Shock")  +
    scale_fill_manual(values = condition_colors))
```



### Combine all

```{r fig.width=8, fig.height=7}
(plt_comment1_6 <- ggarrange(plt_ychromosomescores,
                             plt_viralrespiratory,
                             plt_prol,
                             plt_yeast_heatshock,
                             nrow=2, align = "hv", ncol=2, labels="AUTO"))
```

```{r}
ggplot2::ggsave("../Figures/Figs/R1_6_OtherData_v2.png",
                plt_comment1_6,
                width = 8, height=7, bg = 'white')

ggplot2::ggsave("../Figures/Figs/R1_6_OtherData_v2.pdf",
                plt_comment1_6,
                width = 8, height=7, bg = 'white')
```



## Comment 3.1 — Cell type composition as a confounder in GTEx age analyses

R.: Estimate cell type proportions in the GTEx tissues highlighted in the main text using a reference-based deconvolution tool, and incorporate these estimates as covariates in the age–score association models. Results with and without compositional correction are compared for a subset of tissues, both those with and without significant age associations in the original analysis.

Cell composition correction for age-senescence associations in GTEx

Strategy: GSEA only (enrichment-based), comparing:
- Baseline:  design ~ AGE  (as in the paper)
- Corrected: design ~ AGE + xCell_ct1 + xCell_ct2 + ...

Tissues:
WITH signal    : Artery - Aorta, Breast - Mammary Tissue
WITHOUT signal : Skin - Sun Exposed (Lower leg), Lung

Signatures: HernandezSegura 

```{r}
library(xCell2)
library(limma)
library(dplyr)
library(tidyr)
library(ggplot2)
library(cowplot)
library(beepr)
```

```{r}
# -----------------------------------------------------------------------------
# HOW xCell2 WORKS
# -----------------------------------------------------------------------------
# xCell2 estimates cell type enrichment scores from bulk RNA-seq data using a
# reference panel of known pure cell type expression profiles. It operates in
# two stages:
#
# (1) TRAINING: For each cell type, xCell2 identifies genes that are
#     specifically and consistently highly expressed in that cell type relative
#     to all others in the reference panel (not just highly expressed, but
#     *relatively* highly expressed). Multiple signatures per cell type are
#     built using different parameter combinations (stringency thresholds,
#     reference sample subsets), producing an ensemble that is more robust than
#     a single signature. This is why internal signature names contain suffixes
#     like "#_0.333_1_37_0.97" encoding those parameters.
#
# (2) SCORING: For each sample, xCell2 ranks all genes by expression and asks
#     whether the signature genes for a given cell type are collectively ranked
#     higher than expected by chance — conceptually similar to ssGSEA. A high
#     score means that cell type's transcriptional signal is enriched in the
#     sample; a score near zero means no detectable enrichment. A spillover
#     correction is then applied to reduce cross-talk between related cell
#     types that share marker genes.
#
# Unlike CIBERSORTx, which deconvolves bulk samples into cell type proportions
# that sum to 1 by solving a mixture model, xCell2 scores each cell type
# independently. Scores are not proportions, not bounded, and not comparable
# across cell types in absolute terms. Within a cell type however, higher
# scores consistently reflect greater presence of that cell type's
# transcriptional signal across samples — which is sufficient for our purpose
# of using them as covariates to control for age-related compositional
# variation in GTEx tissues.
# -----------------------------------------------------------------------------
```


```{r}
# =============================================================================
# Cell composition correction for age-senescence associations in GTEx
# Addresses reviewer comment 3.1
#
# Workflow per tissue:
#   1. Run xCell2 to estimate cell type enrichment scores
#   2. Select informative cell type covariates
#   3. Build baseline (~ AGE) and corrected (~ AGE + cell types) design matrices
#   4. Run calculateDE() with each design -> ranked t-statistics for AGE
#   5. Run fgsea on both ranked lists
#   6. Compare NES and significance with and without correction
#
# Tissues:
#   WITH signal    : Artery - Aorta, Breast - Mammary Tissue
#   WITHOUT signal : Skin - Sun Exposed (Lower leg), Lung
#
# Signatures: HernandezSegura, SAUL_SEN_MAYO
#
# NOTE ON xCell2 SCORES:
# xCell2 scores are enrichment scores (conceptually similar to ssGSEA), NOT
# cell type proportions or percentages. Each value reflects how enriched the
# transcriptomic signature of a given cell type is in a sample relative to the
# reference signatures. Values are not constrained to sum to 1 across cell
# types and are not directly comparable across cell types.
#
# For this analysis, scores are used purely as covariates in the limma linear
# model, not for cell type quantification per se. The rationale is: if cell
# type composition varies with age in a given tissue, the xCell2 scores for
# the relevant cell types will correlate with age, and including them as
# covariates will absorb that compositional variance from the model before
# estimating the age effect on the senescence signature. This allows us to
# assess whether the age-senescence associations reported in the paper are
# robust to, or confounded by, age-related shifts in cell type composition.
#
# ON MULTIPLE SIGNATURES PER CELL TYPE:
# Internally, xCell2 stores multiple gene set signatures per cell type,
# identified by names such as:
#   "CD4-positive, alpha-beta T cell#_0.333_1_37_0.97"
# The suffix encodes training parameters (quantile threshold, number of genes,
# performance score). This ensemble approach makes the enrichment scores more
# robust. xCell2Analysis() aggregates across all signatures per cell type
# and returns a single score per cell type per sample — the xcell_scores matrix
# contains one row per cell type (43 total). The internal naming is irrelevant
# for downstream use.
# =============================================================================

```


```{r}
# local job
composition_correction_results <- readRDS("../data_aux/composition_correction_results_noscale_fullGTExdata.rds")

```


```{r fig.width=7, fig.height=11}

results_df <- composition_correction_results
# -----------------------------------------------------------------------------
# 7. Visualisation — all 49 tissues
# -----------------------------------------------------------------------------
# With 49 tissues, a standard dot/line plot becomes unreadable on a single
# panel. We use a heatmap-style layout showing NES for baseline and corrected
# side by side, with significance overlaid as asterisks. Tissues are grouped
# by whether they showed signal in the paper. This allows the reviewer to see
# at a glance: (a) which tissues retain signal after correction, (b) whether
# any new signal emerges, and (c) how much the NES changes overall.

model_colours <- c("Baseline"  = "#2171b5",
                   "Corrected" = "#cb181d")
# rename model levels for clarity
plot_df <- results_df %>%
  mutate(
    # tissue_group = ifelse(tissue %in% TISSUE_SIGNAL,
    #                       "Signal \nin paper", "No signal \nin paper"),
    # tissue_group = factor(tissue_group,
    #                       levels = c("Signal \nin paper", "No signal \nin paper")),
    model        = factor(model,
                          levels   = c("Baseline", "Corrected"),
                          labels   = c("Non-corrected", "Corrected"))
  )

tissue_order <- plot_df %>%
  filter(model == "Non-corrected") %>%
  arrange(desc(tissue)) %>%   # alphabetical, desc so A appears at top
  pull(tissue) %>%
  unique()

plot_df <- plot_df %>%
  mutate(tissue = factor(tissue, levels = tissue_order))

model_colours <- c("Non-corrected" = "#2171b5",
                   "Corrected"     = "#cb181d")

plot_all <- ggplot(plot_df,
                   aes(x = NES, y = tissue, colour = model)) +
  geom_vline(xintercept = 0, linetype = "dashed",
             colour = "grey50", linewidth = 0.4) +
  geom_line(aes(group = tissue), colour = "grey75", linewidth = 0.5) +
  geom_point(aes(shape = sig), size = 3, stroke = 1.2, fill = "white") +
scale_shape_manual(
    values = c("FALSE" = 21, "TRUE"  = 19),
    labels = c("FALSE" = "FDR > 0.05", "TRUE" = "FDR \u2264 0.05"),
    name   = NULL
  ) +
  scale_colour_manual(values = model_colours, name = "Model") +
  #facet_grid(tissue_group ~ ., scales = "free_y", space = "free_y") +
  labs(
    x        = "Normalised Enrichment Score (NES)",
    y        = NULL,
    title    = "HernandezSegura Age-Enrichment \nacross GTEx tissues",
    subtitle = "Non-corrected vs. cell composition-corrected"
  ) +
  theme_bw(13) +
  theme(
    strip.background  = element_rect(fill = "grey92"),
    strip.text.y      = element_text(angle = 90, vjust = 0.5),
    legend.position   = "bottom",
    axis.text.y       = element_text(size = 11),
    plot.caption      = element_text(size = 8, colour = "grey40",
                                     hjust = 0, margin = margin(t = 6)), 
                                     legend.title = element_blank(),
    legend.box        = "vertical",
    legend.spacing.y  = unit(0, "pt")
  )
 

plot_all


```
 



<!-- ```{r} -->
<!-- ggplot2::ggsave("../Figures/Figs/R3_1_CellTypeComposition.png", -->
<!--                 plot_all, -->
<!--                 width = 8, height=11, bg = 'white') -->

<!-- ggplot2::ggsave("../Figures/Figs/R3_1_CellTypeComposition.pdf", -->
<!--                 plot_all, -->
<!--                 width = 7, height=11, bg = 'white') -->
<!-- ``` -->

```{r}
df_null <- readRDS("../data_aux/null_NES_permutations_simpler_1000perms_fullGTExdata.rds")
```

 
```{r fig.width=7, fig.height=11}
# -----------------------------------------------------------------------------
# Compute percentiles of observed NES within the null distribution
# -----------------------------------------------------------------------------

null_plot_df <- df_null %>%
  mutate(tissue = factor(tissue, levels = tissue_order))

percentile_df <- plot_df %>%
  left_join(
    null_plot_df %>%
      group_by(tissue) %>%
      summarise(null_vals = list(NES), .groups = "drop"),
    by = "tissue"
  ) %>%
  rowwise() %>%
  mutate(
    percentile       = round(mean(unlist(null_vals) < NES) * 100, 1),  # signed
    percentile_label = paste0(percentile, "%")
  ) %>%
  ungroup()


# determine label side per tissue based on relative NES positions
# determine which model is left/right per tissue
label_sides <- percentile_df %>%
  dplyr::select(tissue, model, NES) %>%
  pivot_wider(names_from = model, values_from = NES) %>%
  mutate(
    nudge_noncorrected = ifelse(`Non-corrected` <= Corrected, -0.15,  0.15),
    hjust_noncorrected = ifelse(`Non-corrected` <= Corrected,  1,     0),
    nudge_corrected    = ifelse(`Non-corrected` <= Corrected,  0.15, -0.15),
    hjust_corrected    = ifelse(`Non-corrected` <= Corrected,  0,     1)
  ) %>%
  dplyr::select(tissue, nudge_noncorrected, hjust_noncorrected,
                nudge_corrected, hjust_corrected)

percentile_df <- percentile_df %>%
  left_join(label_sides, by = "tissue") %>%
  mutate(
    percentile_label = format(round(percentile, 1), nsmall = 1),
    nudge_x = ifelse(model == "Non-corrected", nudge_noncorrected, nudge_corrected),
    hjust   = ifelse(model == "Non-corrected", hjust_noncorrected, hjust_corrected)
  )

# -----------------------------------------------------------------------------
# Plot
# -----------------------------------------------------------------------------
plot_all <- ggplot() +
  geom_jitter(data  = null_plot_df,
              aes(x = NES, y = tissue, group = tissue),
              fill  = "grey85", colour = "grey70", size = 0.1) +
  geom_violin(data      = null_plot_df,
              aes(x     = NES, y = tissue, group = tissue),
              fill      = "grey85", colour = "grey70",
              alpha     = 0.7, linewidth = 0.3,
              scale     = "width", width = 0.7) +
  geom_vline(xintercept = 0, linetype = "dashed",
             colour = "grey50", linewidth = 0.4) +
  # geom_line(data = plot_df,
  #           aes(x = NES, y = tissue, group = tissue),
  #           colour = "grey75", linewidth = 0.5) +
  geom_point(data = plot_df,
             aes(x = NES, y = tissue, colour = model, shape = sig),
             size = 3, stroke = 1.2, fill = "white") +
geom_text(data     = percentile_df %>% filter(model == "Non-corrected"),
          aes(x    = NES + nudge_x, y = tissue, label = percentile_label,
              hjust = hjust),
          size     = 2.5,
          fontface = "bold",
          colour   = "#2171b5") +
geom_text(data     = percentile_df %>% filter(model == "Corrected"),
          aes(x    = NES + nudge_x, y = tissue, label = percentile_label,
              hjust = hjust),
          size     = 2.5,
          fontface = "bold",
          colour   = "#cb181d") +
  scale_shape_manual(
    values = c("FALSE" = 21, "TRUE"  = 19),
    labels = c("FALSE" = "FDR > 0.05", "TRUE" = "FDR \u2264 0.05"),
    name   = NULL
  ) +
  scale_colour_manual(values = model_colours, name = "Model") +
  #scale_x_continuous(limits = c(-3.8, 3.8)) +
  labs(
    x        = "Normalised Enrichment Score (NES)",
    y        = NULL,
    title    = "HernandezSegura Age-Enrichment \nacross GTEx tissues",
    subtitle = "Non-corrected vs. cell composition-corrected\nPercentiles relative to null distribution shown"
  ) +
  theme_bw(13) +
  theme(
    strip.background  = element_rect(fill = "grey92"),
    strip.text.y      = element_text(angle = 90, vjust = 0.5),
    legend.position   = "bottom",
    axis.text.y       = element_text(size = 11),
    plot.caption      = element_text(size = 8, colour = "grey40",
                                     hjust = 0, margin = margin(t = 6)),
    legend.title      = element_blank(),
    legend.box        = "vertical",
    legend.spacing.y  = unit(0, "pt")
  )

plot_all
```


```{r}
ggplot2::ggsave("../Figures/Figs/R3_1_CellTypeComposition_correctedAllgenes.png",
                plot_all,
                width = 8, height=11, bg = 'white')

ggplot2::ggsave("../Figures/Figs/R3_1_CellTypeComposition_correctedAllgenes.pdf",
                plot_all,
                width = 7, height=11, bg = 'white')
```

### Comparison All Genes vs Common Genes

```{r}
# ── 1. Load data ──────────────────────────────────────────────────────────────
score_common <- readRDS("data_aux/results_df_score_GTEx_FPR.rds")
gsea_common  <- readRDS("data_aux/results_df_gsea_GTEx_FPR.rds")
score_all    <- readRDS("data_aux/results_df_score_GTEx_FPR_agecentered_extendedGTExdata.rds")
gsea_all     <- readRDS("data_aux/results_df_gsea_GTEx_FPR_agecentered_extendedGTExdata.rds")
```

```{r}
# ── 2. Preprocess ─────────────────────────────────────────────────────────────
prep_score <- function(df, gene_set_label) {
  df %>%
    group_by(signature, method) %>%
    mutate(P_adj = p.adjust(fpr, method = "BH")) %>%
    ungroup() %>%
    mutate(
      method  = factor(method, levels = c("logmedian", "ranking", "ssGSEA")),
      signif  = ifelse(P_adj < 0.05, "*", ""),
      geneset = gene_set_label
    )
}

prep_gsea <- function(df, gene_set_label) {
  df %>%
    group_by(signature) %>%
    mutate(padj = p.adjust(pval, method = "BH")) %>%
    ungroup() %>%
    mutate(
      method  = "GSEA",
      signif  = ifelse(padj < 0.05, "*", ""),
      geneset = gene_set_label
    )
}

score_common <- prep_score(score_common, "Common genes \nacross tissues")
score_all    <- prep_score(score_all,    "All genes \nper tissue")
gsea_common  <- prep_gsea(gsea_common,   "Common genes \nacross tissues")
gsea_all     <- prep_gsea(gsea_all,      "All genes \nper tissue")


```

```{r}
# ── 3. Unified tissue order ───────────────────────────────────────────────────
all_tissues <- sort(unique(c(
  score_common$tissue, score_all$tissue,
  gsea_common$tissue,  gsea_all$tissue
)), decreasing = T)

set_levels <- function(df) {
  df %>% mutate(tissue = factor(tissue, levels = all_tissues))
}

score_common <- set_levels(score_common)
score_all    <- set_levels(score_all)
gsea_common  <- set_levels(gsea_common)
gsea_all     <- set_levels(gsea_all)
```

```{r}
# ── 4. Heatmap builders ───────────────────────────────────────────────────────
# All legends removed from individual plots — extracted separately at the end

make_score_heatmap <- function(score_data,
                               show_y = TRUE,
                               title  = NULL) {
  p <- ggplot(score_data,
              aes(x = method, y = tissue, fill = Cohen_f)) +
    geom_tile(color = "gray90") +
    geom_text(aes(label = signif), size = 4.5, vjust = 0.8, hjust = 0.5) +
    scale_fill_gradient(
      low      = "white",
      high     = "#49B0AB",
      name     = "Cohen's f",
      limits   = c(0, 0.4),
      na.value = "#49B0AB",
    ) +
    scale_x_discrete(labels = c(
      logmedian = expression(log[2]*"-median"),
      ranking   = "ranking",
      ssGSEA    = "ssGSEA"
    )) +
    theme_minimal(base_size = 10) +
    theme(
      axis.title      = element_blank(),
      axis.text.x     = element_text(angle = 45, hjust = 1, size = 9),
      axis.text.y     = if (show_y) element_text(size = 7.5) else element_blank(),
      axis.ticks.y    = if (show_y) element_line() else element_blank(),
      legend.position = "none",
      panel.grid      = element_blank(),
      plot.title      = element_text(hjust = 0.5, face = "bold", size = 10,
                                     margin = margin(b = 4))
    )

  if (!is.null(title)) p <- p + labs(title = title)
  p
}

make_gsea_heatmap <- function(gsea_data,
                              title = NULL) {
  p <- ggplot(gsea_data,
              aes(x = method, y = tissue, fill = abs(NES))) +
    geom_tile(color = "gray90") +
    geom_text(aes(label = signif), size = 4.5, vjust = 0.8, hjust = 0.5) +
    scale_fill_gradient(
      low      = "white",
      high     = "#B04975",
      name     = "|NES|",
      limits   = c(0, 2.5),
      na.value = "#B04975"
    ) +
    theme_minimal(base_size = 10) +
    theme(
      axis.title      = element_blank(),
      axis.text.y     = element_blank(),
      axis.ticks.y    = element_blank(),
      axis.text.x     = element_text(angle = 45, hjust = 1, size = 9),
      legend.position = "none",
      panel.grid      = element_blank(),
      plot.title      = element_text(hjust = 0.5, face = "bold", size = 10,
                                     margin = margin(b = 4))
    )

  if (!is.null(title)) p <- p + labs(title = title)
  p
}


```

```{r}
# ── 5. Build heatmap plots ────────────────────────────────────────────────────
# Titles: only on top row (Common genes block), one title per score+gsea pair
# Signature label: added as plot subtitle on the score_common plot only

sig_titles <- list(
  HernandezSegura = "HernandezSegura",
  SAUL_SEN_MAYO   = "SAUL_SEN_MAYO"
)

heatmap_plts <- list()
for (sig in c("HernandezSegura", "SAUL_SEN_MAYO")) {
  for (gs in c("Common genes \nacross tissues", "All genes \nper tissue")) {

    sd <- filter(if (gs == "Common genes \nacross tissues") score_common else score_all,
                 signature == sig)
    gd <- filter(if (gs == "Common genes \nacross tissues") gsea_common  else gsea_all,
                 signature == sig)

    show_y <- (gs == "Common genes \nacross tissues")

    # Block title only on top of score plot, labelled by gene set
    score_title <- gs  # "Common genes" or "All genes"
    gsea_title  <- ""  # empty — one title covers score+gsea block

    # Signature label as subtitle on the leftmost (score common) plot only
    p_score <- make_score_heatmap(sd, show_y = show_y, title = score_title)
    p_gsea  <- make_gsea_heatmap(gd, title = gsea_title)

    # Add signature name as y-axis title on score_common only
    if (gs == "Common genes \nacross tissues") {
      p_score <- p_score +
        #labs(subtitle = sig_titles[[sig]]) +
        theme(plot.subtitle = element_text(
          hjust = 0, face = "bold.italic", size = 9,
          colour = "grey30", margin = margin(b = 2)
        ))
    }

    heatmap_plts[[paste(sig, "score", gs, sep = "||")]] <- p_score
    heatmap_plts[[paste(sig, "gsea",  gs, sep = "||")]] <- p_gsea
  }
}


```


```{r}
# ── 6. Extract legends from dummy plots ───────────────────────────────────────
# Build two dummy plots with legends, extract them cleanly

dummy_score <- ggplot(
  data.frame(x = 1, y = 1, z = 0.2),
  aes(x = x, y = y, fill = z)
) +
  geom_tile() +
  scale_fill_gradient(
    low    = "white",
    high   = "#49B0AB",
    name   = "Cohen's f",
    limits = c(0, 0.4),
    breaks = c(0, 0.1, 0.2, 0.3, 0.4)
  ) +
  theme_void() +
  theme(
    legend.position     = "right",
    legend.title        = element_text(size = 9, face = "bold"),
    legend.text         = element_text(size = 8),
    legend.key.height   = unit(0.8, "cm"),
    legend.key.width    = unit(0.35, "cm")
  )

dummy_gsea <- ggplot(
  data.frame(x = 1, y = 1, z = 1.5),
  aes(x = x, y = y, fill = z)
) +
  geom_tile() +
  scale_fill_gradient(
    low    = "white",
    high   = "#B04975",
    name   = "|NES|",
    limits = c(0, 2.5),
    breaks = c(0, 0.5, 1.0, 1.5, 2.0, 2.5)
  ) +
  theme_void() +
  theme(
    legend.position     = "right",
    legend.title        = element_text(size = 9, face = "bold"),
    legend.text         = element_text(size = 8),
    legend.key.height   = unit(0.8, "cm"),
    legend.key.width    = unit(0.35, "cm")
  )

leg_score <- get_legend(dummy_score)
leg_gsea  <- get_legend(dummy_gsea)

# Stack the two legends vertically with a spacer between them
legend_block <- ggarrange(
  as_ggplot(leg_score),
  as_ggplot(leg_gsea),
  ncol    = 1,
  nrow    = 2,
  heights = c(1, 1)
)

```


```{r fig.width=10, fig.height=8}
# ── 7. Heatmap row ────────────────────────────────────────────────────────────
# 8 heatmap plots + 1 legend block
# widths: score (3 cols) gets ~3x the width of gsea (1 col)
# 
# heatmap_row <- ggarrange(
#   heatmap_plts[["HernandezSegura||score||Common genes"]],
#   heatmap_plts[["HernandezSegura||gsea||Common genes"]] + theme(axis.text.y = element_blank()),
#   heatmap_plts[["HernandezSegura||score||All genes"]] + theme(axis.text.y = element_blank()),
#   heatmap_plts[["HernandezSegura||gsea||All genes"]] + theme(axis.text.y = element_blank()),
#   NULL,
#   heatmap_plts[["SAUL_SEN_MAYO||score||Common genes"]] + theme(axis.text.y = element_blank()),
#   heatmap_plts[["SAUL_SEN_MAYO||gsea||Common genes"]] + theme(axis.text.y = element_blank()),
#   heatmap_plts[["SAUL_SEN_MAYO||score||All genes"]] + theme(axis.text.y = element_blank()),
#   heatmap_plts[["SAUL_SEN_MAYO||gsea||All genes"]] + theme(axis.text.y = element_blank()),
#   legend_block,
#   nrow   = 1,
#   ncol   = 10,
#   widths = c(0.45, 0.05, 0.16, 0.05, 0.05,   # HS block
#              0.16, 0.055, 0.14, 0.055,   # SM block
#              0.10))


heatmap_row1 <- ggarrange(
  heatmap_plts[["HernandezSegura||score||Common genes \nacross tissues"]],
  heatmap_plts[["HernandezSegura||gsea||Common genes \nacross tissues"]] + theme(axis.text.y = element_blank()),
  heatmap_plts[["HernandezSegura||score||All genes \nper tissue"]] + theme(axis.text.y = element_blank()),
  heatmap_plts[["HernandezSegura||gsea||All genes \nper tissue"]] + theme(axis.text.y = element_blank()),
  nrow   = 1, 
  widths = c(0.55, 0.07, 0.16, 0.07),
  align="h")

heatmap_row1 <- annotate_figure(heatmap_row1, top = text_grob("                                                   HernandezSegura", face = "bold", size = 12))


heatmap_row2 <- ggarrange( 
  heatmap_plts[["SAUL_SEN_MAYO||score||Common genes \nacross tissues"]] + theme(axis.text.y = element_blank()),
  heatmap_plts[["SAUL_SEN_MAYO||gsea||Common genes \nacross tissues"]] + theme(axis.text.y = element_blank()),
  heatmap_plts[["SAUL_SEN_MAYO||score||All genes \nper tissue"]] + theme(axis.text.y = element_blank()),
  heatmap_plts[["SAUL_SEN_MAYO||gsea||All genes \nper tissue"]] + theme(axis.text.y = element_blank()),
  nrow   = 1, 
  widths = c(0.16, 0.055, 0.14, 0.055 ),
  align="h")

heatmap_row2 <- annotate_figure(heatmap_row2, top = text_grob("SAUL_SEN_MAYO", face = "bold", size = 12))


heatmap_row3 <- ggarrange( 
  legend_block,
  nrow   = 1, 
  widths = c( 0.10))


heatmap_row <- ggarrange(heatmap_row1,
                         NULL,
                         heatmap_row2,
                         heatmap_row3, 
                         widths=c(0.7,0.05,0.4,0.2), nrow=1)


heatmap_row

```

```{r}
# ── 8. Scatter plot ───────────────────────────────────────────────────────────
score_joined <- inner_join(
  score_common %>% select(signature, tissue, method, Cohen_f, P_adj, signif),
  score_all    %>% select(signature, tissue, method, Cohen_f, P_adj, signif),
  by     = c("signature", "tissue", "method"),
  suffix = c("_common", "_all")
) %>%
  mutate(
    sig_category = case_when(
      signif_common == "*" & signif_all == "*" ~ "Both",
      signif_common == "*" & signif_all != "*" ~ "Common Genes Only",
      signif_common != "*" & signif_all == "*" ~ "All Genes Only",
      TRUE                                     ~ "Neither"
    ),
    sig_category = factor(sig_category,
                          levels = c("Both", "Common Genes Only", "All Genes Only", "Neither")),
    plot_x    = Cohen_f_common,
    plot_y    = Cohen_f_all,
    facet_lab = case_when(
      method == "logmedian" ~ "log[2]-median",
      method == "ranking"   ~ "Ranking",
      method == "ssGSEA"    ~ "ssGSEA"
    ),
    facet_lab = factor(facet_lab, levels = c("log[2]-median", "Ranking", "ssGSEA"))
  )

gsea_joined <- inner_join(
  gsea_common %>% select(signature, tissue, NES, padj, signif),
  gsea_all    %>% select(signature, tissue, NES, padj, signif),
  by     = c("signature", "tissue"),
  suffix = c("_common", "_all")
) %>%
  mutate(
    sig_category = case_when(
      signif_common == "*" & signif_all == "*" ~ "Both",
      signif_common == "*" & signif_all != "*" ~ "Common Genes Only",
      signif_common != "*" & signif_all == "*" ~ "All Genes Only",
      TRUE                                     ~ "Neither"
    ),
    sig_category = factor(sig_category,
                          levels = c("Both", "Common Genes Only", "All Genes Only", "Neither")),
    plot_x    = NES_common,
    plot_y    = NES_all,
    method    = "GSEA",
    facet_lab = factor("GSEA", levels = "GSEA")
  )

combined_scatter_data <- bind_rows(
  score_joined %>% select(signature, tissue, plot_x, plot_y, sig_category, facet_lab),
  gsea_joined  %>% select(signature, tissue, plot_x, plot_y, sig_category, facet_lab)
) %>%
  mutate(
    facet_lab = factor(facet_lab,
                       levels = c("log[2]-median", "Ranking", "ssGSEA", "GSEA"))
  )

sig_colours <- c(
  "Both"        = "#333333",
  "Common Genes Only" = "#49B0AB",
  "All Genes Only"    = "#B04975",
  "Neither"     = "grey80"
)

scatter_plt <- ggplot(combined_scatter_data,
                      aes(x      = plot_x,
                          y      = plot_y,
                          colour = sig_category,
                          shape  = signature)) +
  facet_wrap(
    ~ facet_lab,
    ncol     = 4,
    scales   = "free",
    labeller = labeller(facet_lab = label_parsed)
  ) +
  geom_vline(xintercept = 0, linetype = "dotted",
             colour = "grey70", linewidth = 0.4) +
  geom_hline(yintercept = 0, linetype = "dotted",
             colour = "grey70", linewidth = 0.4) +
  geom_abline(slope = 1, intercept = 0,
              linetype = "dashed", colour = "grey50", linewidth = 0.5) +
  geom_point(size = 2, alpha = 0.85) +
  scale_colour_manual(
    values = sig_colours,
    name   = "FDR \u2264 0.05",
    guide  = guide_legend(order = 1, override.aes = list(size = 3))
  ) +
  scale_shape_manual(
    values = c(HernandezSegura = 16, SAUL_SEN_MAYO = 17),
    name   = "Gene Set",
    guide  = guide_legend(order = 2, override.aes = list(size = 3))
  ) +
  labs(
    x = "Cohen's f \n Common genes",
    y = "Cohen's f \n All genes"
  ) +
  theme_bw(base_size = 10) +
  theme(
    strip.background = element_rect(fill = "grey95", colour = NA),
    strip.text       = element_text(size = 10, face = "bold"),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(colour = "grey93"),
    legend.position  = "right",
    legend.box       = "vertical",
    axis.title       = element_text(size = 9)
  )

```


```{r fig.width=10, fig.height=10}
 # ── 9. Final assembly ─────────────────────────────────────────────────────────
fig_combined <- ggarrange(
  heatmap_row,
  scatter_plt,
  ncol       = 1,
  nrow       = 2,
  heights    = c(2, 0.7),
  labels     = c("A", "B"),
  font.label = list(size = 13, face = "bold")
)

fig_combined
```

```{r}
ggplot2::ggsave("Figures/Figs/R3_1_ImpactSubsetGenes.png",
                fig_combined,
                width = 10, height=10, bg = 'white')

ggplot2::ggsave("Figures/Figs/R3_1_ImpactSubsetGenes.pdf",
                fig_combined,
                width = 10, height=10, bg = 'white')
```

## Comment 1.5 — Additional datasets for age correlation analysis

R.: Apply markeR's discovery mode to one or more datasets independent of GTEx, modelling age as a continuous variable, to assess whether age-associated senescence signals are reproducible across transcriptomic resources and pointed limitations are correct. 

### Skin (microarray) - GSE85358

```{r}
## ============================================================
## Parse Agilent microarray txt.gz files -> expression matrix
## Platform: Agilent-039494, hg19
## Signal:   gProcessedSignal
## Genes:    GeneName (mean across probes)
## Controls: removed (ControlType != 0)
## ============================================================

library(data.table)
library(dplyr)

# ── User settings ─────────────────────────────────────────────
INPUT_DIR  <- "../data_aux/GSE85358_RAW/"   # <-- change this
OUTPUT_CSV <- "../data_aux/GSE85358_expression_matrix.csv"        # output file name
# ──────────────────────────────────────────────────────────────


#' Parse a single Agilent Feature Extraction txt.gz file.
#' Returns a two-column data.frame: GeneName, gProcessedSignal.
parse_agilent_file <- function(filepath) {
  
  # Read all lines (gz is handled transparently by readLines)
  lines <- readLines(filepath, warn = FALSE)
  
  # The file has three blocks separated by "*":
  # FEPARAMS, STATS, FEATURES
  # We need the FEATURES block, which starts with "FEATURES\t..."
  # The actual column header is the line starting with "FEATURES"
  # and data lines start with "DATA"
  
  feat_header_idx <- which(startsWith(lines, "FEATURES\t"))
  
  if (length(feat_header_idx) == 0) {
    warning(paste("No FEATURES block found in:", filepath))
    return(NULL)
  }
  
  # Extract column names from the FEATURES header line
  col_names <- strsplit(lines[feat_header_idx], "\t")[[1]]
  # Replace the first element "FEATURES" with a placeholder
  col_names[1] <- "block_type"
  
  # Find all DATA lines that belong to the FEATURES block
  # (i.e., DATA lines after the FEATURES header)
  data_lines <- lines[(feat_header_idx + 1):length(lines)]
  data_lines <- data_lines[startsWith(data_lines, "DATA\t")]
  
  if (length(data_lines) == 0) {
    warning(paste("No DATA lines found in:", filepath))
    return(NULL)
  }
  
  # Parse into a data.table
  dt <- fread(
    text      = c(paste(col_names, collapse = "\t"), data_lines),
    sep       = "\t",
    header    = TRUE,
    fill      = TRUE,
    showProgress = FALSE
  )
  
  # ── Filter controls ───────────────────────────────────────
  # ControlType == 0 means non-control probes
  dt <- dt[ControlType == 0]
  
  # ── Keep only relevant columns ────────────────────────────
  needed <- c("GeneName", "gProcessedSignal")
  missing_cols <- setdiff(needed, names(dt))
  if (length(missing_cols) > 0) {
    warning(paste("Missing columns", paste(missing_cols, collapse = ", "),
                  "in:", filepath))
    return(NULL)
  }
  
  dt <- dt[, ..needed]
  
  # ── Remove probes with empty/missing/ambiguous gene names ─
  dt <- dt[!is.na(GeneName) & GeneName != "" & GeneName != "---"]
  
  # ── Coerce signal to numeric ──────────────────────────────
  dt[, gProcessedSignal := as.numeric(gProcessedSignal)]
  
  # ── Summarise: mean across probes per gene ────────────────
  dt_mean <- dt[, .(signal = mean(gProcessedSignal, na.rm = TRUE)),
                by = GeneName]
  
  return(dt_mean)
}


# ── Discover files ─────────────────────────────────────────────
files <- list.files(INPUT_DIR, pattern = "\\.txt\\.gz$",
                    full.names = TRUE, recursive = FALSE)


# ── Parse all files ────────────────────────────────────────────
sample_list <- lapply(files, function(f) {
  message("  Processing: ", basename(f))
  result <- tryCatch(
    parse_agilent_file(f),
    error = function(e) {
      warning(paste("Error parsing", basename(f), ":", e$message))
      return(NULL)
    }
  )
  if (!is.null(result)) {
    # Use file basename (strip .txt.gz) as sample name
    sample_name <- sub("\\.txt\\.gz$", "", basename(f))
    setnames(result, "signal", sample_name)
  }
  return(result)
})

# ── Merge all samples by GeneName (full outer join) ────────────
message("Merging samples...")

expr_matrix <- Reduce(function(a, b) {
  merge(a, b, by = "GeneName", all = TRUE)
}, sample_list)

# Genes present in all samples (inner join alternative — optional)
# Uncomment if you prefer to keep only genes present in every sample:
# expr_matrix <- Reduce(function(a, b) merge(a, b, by = "GeneName"), sample_list)


# ── Final formatting ───────────────────────────────────────────
# Set GeneName as row names
expr_df <- as.data.frame(expr_matrix)
rownames(expr_df) <- expr_df$GeneName
expr_df$GeneName  <- NULL

message(sprintf("Final matrix: %d genes x %d samples",
                nrow(expr_df), ncol(expr_df)))


# ── Save ───────────────────────────────────────────────────────
write.csv(expr_df, file = OUTPUT_CSV, quote = FALSE)
message("Saved: ", OUTPUT_CSV)

#saveRDS(expr_df, "../data_aux/GSE85358_expression_matrix.rds")
 expr_df <- readRDS("../data_aux/GSE85358_expression_matrix.rds")
```


```{r}
## ============================================================
## Extract sample metadata from GEO SOFT file
## and align with expression matrix sample names
## ============================================================
 

# ── User settings ─────────────────────────────────────────────
SOFT_FILE  <- "../data_aux/GSE85358_family.soft.gz"  # <-- change this
OUTPUT_CSV <- "../data_aux/GSE85358_metadata.csv"

# Optional: if you already have your expression matrix loaded,
# set this to align metadata rows to your sample columns.
# Otherwise leave as NULL.
EXPR_SAMPLE_NAMES <- colnames(expr_df)
# ──────────────────────────────────────────────────────────────


# ── Parse SOFT file ────────────────────────────────────────────
message("Parsing SOFT file: ", SOFT_FILE)

gse <- getGEO(filename = SOFT_FILE, GSEMatrix = FALSE)

# Extract all GSM (sample) objects
gsm_list <- GSMList(gse)
message(sprintf("Found %d samples.", length(gsm_list)))


# ── Extract metadata from each sample ─────────────────────────
# Each GSM has a Meta slot with all characteristics
extract_gsm_metadata <- function(gsm) {
  meta <- Meta(gsm)
  
  # Core fields always present in GEO
  core <- data.frame(
    sample_id    = meta$geo_accession %||% NA,
    sample_title = meta$title         %||% NA,
    stringsAsFactors = FALSE
  )
  
  # Characteristics lines: format is usually "key: value"
  # e.g. "age: 45", "Sex: Male", "tissue: skin"
  chars <- meta$characteristics_ch1
  
  if (!is.null(chars) && length(chars) > 0) {
    char_pairs <- lapply(chars, function(x) {
      parts <- strsplit(x, ":\\s*", perl = TRUE)[[1]]
      if (length(parts) == 2) {
        key <- trimws(gsub("[^a-zA-Z0-9_]", "_", tolower(parts[1])))
        val <- trimws(parts[2])
        return(setNames(data.frame(val, stringsAsFactors = FALSE), key))
      }
      return(NULL)
    })
    char_df <- bind_cols(Filter(Negate(is.null), char_pairs))
    core <- bind_cols(core, char_df)
  }
  
  return(core)
}

# Fallback for NULL values (like %||% in rlang)
`%||%` <- function(a, b) if (!is.null(a) && length(a) > 0) a else b

metadata <- bind_rows(lapply(gsm_list, extract_gsm_metadata))

message(sprintf("Metadata columns found: %s",
                paste(names(metadata), collapse = ", ")))


# ── Derive sample name matching your expression matrix ────────
# GEO supplementary files are often named after the GSM accession
# OR after the sample title — we keep both so you can match either.
# The expression matrix script named samples after the filename
# (minus .txt.gz), so you may need to adjust the join key below.

message("\nPreview of sample_id and sample_title:")
print(metadata[, c("sample_id", "sample_title")])


# ── Optional: align with expression matrix columns ────────────
if (!is.null(EXPR_SAMPLE_NAMES)) {
  
  # Try matching on sample_id (GSM accession) first
  metadata$matched_expr_col <- NA_character_
  
  for (i in seq_len(nrow(metadata))) {
    gsm_id    <- metadata$sample_id[i]
    gsm_title <- metadata$sample_title[i]
    
    # Match if expr column contains the GSM accession
    hit_id    <- EXPR_SAMPLE_NAMES[grepl(gsm_id, EXPR_SAMPLE_NAMES, fixed = TRUE)]
    # Fallback: match on title
    hit_title <- EXPR_SAMPLE_NAMES[grepl(gsm_title, EXPR_SAMPLE_NAMES, fixed = TRUE)]
    
    if (length(hit_id) == 1) {
      metadata$matched_expr_col[i] <- hit_id
    } else if (length(hit_title) == 1) {
      metadata$matched_expr_col[i] <- hit_title
    } else {
      warning(sprintf("Could not match sample: %s (%s)", gsm_id, gsm_title))
    }
  }
  
  unmatched <- sum(is.na(metadata$matched_expr_col))
  if (unmatched > 0) {
    warning(sprintf("%d sample(s) could not be matched to expression matrix columns.",
                    unmatched))
  } else {
    message("All samples matched successfully.")
  }
  
  # Reorder metadata to match expression matrix column order
  metadata <- metadata[match(EXPR_SAMPLE_NAMES,
                             metadata$matched_expr_col), ]
  rownames(metadata) <- EXPR_SAMPLE_NAMES
}

metadata$sampleID <- row.names(metadata)
metadata <- metadata[,c(8,c(1:7))]

metadata$group <- ifelse(metadata$group == "young", "Young \n (\u2264 25 years)", "Old \n (\u2265 60 years)")
metadata$group <- factor(metadata$group, levels=c("Young \n (\u2264 25 years)","Old \n (\u2265 60 years)"))

# ── Save ───────────────────────────────────────────────────────
write.csv(metadata, file = OUTPUT_CSV, quote = TRUE, row.names = TRUE)
message("Saved: ", OUTPUT_CSV)

#saveRDS(metadata,"../data_aux/metadata_GSE85358.rds")
# metadata is your final sample x variable metadata table

metadata <- readRDS("../data_aux/metadata_GSE85358.rds")
```


##### Filtering

```{r}
data <- expr_df
```

```{r}
# Plot
(plt_beforefilt <- ggplot() + aes(x=log10(rowSums(data))) +
   geom_density(fill="grey", alpha=0.5) + theme_bw() + 
   theme(axis.text.x = element_text(angle = 45, hjust=1),
         plot.title = element_text(hjust = 0.5),
         plot.subtitle = element_text(hjust = 0.5)) +
   ggtitle("Before Filtering") +
   xlab(expression(log[10]~"(counts per gene)")) +
   ylab("Frequency") + 
   geom_vline(xintercept = 3.2, linetype="dotted",color = "black", size=1.5))

```


```{r}
filtdata <- data[rowSums(data) > 10^3.2,]
nrow(data)
nrow(filtdata)


cat("Number of genes before filtering:", nrow(data), "\n")
cat("Number of genes after filtering:", nrow(filtdata), "\n")

```


```{r}

# Plot
(plt_afterfilt <- ggplot() + aes(x=log10(rowSums(filtdata))) +
   geom_density(fill="grey", alpha=0.5) + theme_bw() + 
   theme(axis.text.x = element_text(angle = 45, hjust=1),
         plot.title = element_text(hjust = 0.5),
         plot.subtitle = element_text(hjust = 0.5)) +
   ggtitle("After Filtering") +
   xlab(expression(log[10]~"(counts per gene)")) +
   ylab("Frequency") )

```



##### Normalisation

```{r fig.width=6, fig.height=4}

filtcounts_ggplot <- stack(log2(filtdata))
colnames(filtcounts_ggplot) <- c("expression", "sampleID")
filtcounts_ggplot <- merge(filtcounts_ggplot,metadata, by="sampleID")


(plt_beforenorm <- ggplot(filtcounts_ggplot, aes(sampleID,expression)) + geom_boxplot(alpha=0.4, outlier.size = 0.5)+ xlab("Samples") + ylab("log2(counts)")   +  theme(axis.text.x = element_text(angle = 45, hjust=1))+
    theme_bw()+ 
    theme(plot.title = element_text(hjust = 0.5),
          plot.subtitle = element_text(hjust = 0.5),
          axis.text.x=element_blank(),
          axis.ticks.x=element_blank()) +
    ggtitle("Before Normalisation") )

```


```{r}
dgelist_normcounts <- DGEList(filtdata) # the matrix with gene names as row names and counts in columns
dgelist_normcounts <- calcNormFactors(dgelist_normcounts, method = "TMM") 
normalised_counts <- as.data.frame(voom(dgelist_normcounts,plot=F)$E) 
#normalised_counts <- as.data.frame(edgeR::cpm(dgelist_normcounts, log = FALSE)) 
normdata <- 2^normalised_counts
lognormdata <- log2(normdata+1)
```


```{r fig.width=6, fig.height=4}

normcounts_ggplot <- stack(log2(normdata))
colnames(normcounts_ggplot) <- c("expression", "sampleID")
normcounts_ggplot <- merge(normcounts_ggplot,metadata, by="sampleID")


(plt_afternorm <- ggplot(normcounts_ggplot, aes(sampleID,expression)) + geom_boxplot(alpha=0.4, outlier.size = 0.5)+ xlab("Samples") + ylab("log2(counts)")   +  theme(axis.text.x = element_text(angle = 45, hjust=1))+
    theme_bw()+ 
    theme(plot.title = element_text(hjust = 0.5),
          plot.subtitle = element_text(hjust = 0.5),
          axis.text.x=element_blank(),
          axis.ticks.x=element_blank()) +
    ggtitle("After Normalisation") )

```


##### PCA

```{r fig.width=6, fig.height=4}
y <- DGEList(lognormdata, samples= metadata)
PCA  <- prcomp(t(y$counts), scale=FALSE, center=TRUE)
PCACounts  <- PCA$x
PCACounts  <- as.data.frame(PCACounts )
PCAData <-  cbind(PCACounts[1:10],y$samples) 

ev = PCA$sdev^2 
pc1= round(100*ev[1]/sum(ev),2) 
pc2 = round(100*ev[2]/sum(ev),2) 
pc3 = round(100*ev[3]/sum(ev),2) 
pc4 = round(100*ev[4]/sum(ev),2) 
pc5 = round(100*ev[5]/sum(ev),2) 
pc6 = round(100*ev[6]/sum(ev),2)  

(pltpca_1 <- ggplot(PCAData, aes(PC1,PC2))+
    geom_point(aes(color=group), size=5, alpha=0.6)+ 
    xlab(paste0("PC1: ",pc1,"% variance")) +
    ylab(paste0("PC2: ",pc2,"% variance")) + 
    theme_bw()+
    guides(color = guide_legend(nrow = 3)) +
    theme(legend.position="top", legend.box="vertical", legend.title = element_blank(), legend.margin=margin())  + 
    geom_vline(xintercept=0, linetype="dotted") + 
    geom_hline(yintercept=0, linetype="dotted") )

(pltpca_2 <- ggplot(PCAData, aes(PC1,PC2))+
    geom_point(aes(color=as.numeric(age)), size=5, alpha=0.6)+ 
    xlab(paste0("PC1: ",pc1,"% variance")) +
    ylab(paste0("PC2: ",pc2,"% variance")) + 
    theme_bw()+
    theme(legend.position="top", legend.box="vertical", legend.title = element_blank(), legend.margin=margin())  + 
    geom_vline(xintercept=0, linetype="dotted") +  
    geom_hline(yintercept=0, linetype="dotted"))


```

####Calculate Scores

```{r fig.width=10, fig.height=4}

cols_agegroup <- c(
  "Young \n (\u2264 25 years)"    = "#3A798F",   
  "Old \n (\u2265 60 years)" = "#3C3A8F"    
)

metadata$age <- as.numeric(metadata$age)

(plt_scores_age <- PlotScores(data=normdata, metadata=metadata, gene_sets = signatures_bidirectional[c("HernandezSegura","SAUL_SEN_MAYO")], Variable = "age", method="logmedian", xlab = "Age (years)", pvalcalc = T,ColorValues = "#3A508F85") )

(plt_scores_group <- PlotScores(data=normdata, metadata=metadata, gene_sets = signatures_bidirectional[c("HernandezSegura","SAUL_SEN_MAYO")], Variable = "group", ColorVariable = "group", method="logmedian", xlab = "Age Group", pvalcalc = T, ColorValues = cols_agegroup) )

```


```{r fig.width=10, fig.height=4}

# Reordering so Young appears first

cols_agegroup <- c(
  "Young \n (\u2264 25 years)"    = "#3A798F",   
  "Old \n (\u2265 60 years)" = "#3C3A8F"    
)
 
 metadata$group <- factor(metadata$group, levels=c("Young \n (\u2264 25 years)","Old \n (\u2265 60 years)"))
 
(calcscores_group <- CalculateScores(data=normdata, metadata=metadata, gene_sets = signatures_bidirectional[c("HernandezSegura","SAUL_SEN_MAYO")], method = "logmedian") )

 pltlist <- list()
 
 for (sig in names(calcscores_group)) {
   
   x <- calcscores_group[[sig]][calcscores_group[[sig]][["group"]] %in% levels(metadata$group)[2], "score", drop = TRUE]
   y <- calcscores_group[[sig]][calcscores_group[[sig]][["group"]] %in% levels(metadata$group)[1], "score", drop = TRUE]
   cohen_d_results <- cohen_d(x, y)
   calcscores_group[[sig]]$cohen <- ifelse(calcscores_group[[sig]][, "group"] %in% levels(metadata$group)[1],
                                           levels(metadata$group)[1], levels(metadata$group)[2])
   ttest_results <- rstatix::t_test(calcscores_group[[sig]], formula = score ~ cohen)
   p_val <- ttest_results$p[1]
   
   line1 <- wrap_title(paste0("Cohen's d = ",
                              format(signif(cohen_d_results, digits=3),
                                     scientific = FALSE)),
                       width = 20)
   line2 <- wrap_title(paste0("p = ", format(signif(p_val, digits=3),
                                             scientific = TRUE)),
                       width = 20)
   subtitle <- paste(line1, line2, sep = "\n")
   
   
   pltlist[[sig]] <-  ggplot2::ggplot(calcscores_group[[sig]], ggplot2::aes_string(x = "group", y = "score"))+ 
     ggplot2::geom_jitter(ggplot2::aes_string(color = "group"),
                          size = 3, alpha = 0.5)+
     ggplot2::geom_violin(alpha = 0.5, scale = "width")  + 
     ggplot2::stat_summary(fun = stats::median, fun.min = stats::median, fun.max = stats::median,
                           geom = "crossbar", width = 0.25,
                           position = ggplot2::position_dodge(width = 0.13))+
     ggplot2::theme_bw() +
     ggplot2::theme(legend.position="top",
                    axis.text.x = ggplot2::element_text(angle = 45, hjust = 1,
                                                        size = 10),
                    axis.text.y = ggplot2::element_text(  size = 10),
                    plot.title = ggplot2::element_text(hjust = 0.5,
                                                       size = 14-1),
                    plot.subtitle = ggplot2::element_text(hjust = 0.5,
                                                          size = 14 - 1.5,
                                                          face = "italic")) +
     ggplot2::labs(title = sig, subtitle = subtitle, color = "", x = "", y = "")+ 
     ggplot2::scale_color_manual(values = cols_agegroup)+ 
     ggplot2::guides(color = ggplot2::guide_legend(nrow = 1,
                                                   override.aes = list(size = 3 + 2)))
   
 }
 
 plt_scores_group <- ggarrange(plotlist = pltlist, ncol=length(pltlist), nrow=1, widths = c(1,1), common.legend = T)
   plt_scores_group <- ggpubr::annotate_figure(plt_scores_group,
                                           left = grid::textGrob(expression(log[2]*"-median score"),
                                                                 rot = 90, vjust = 1,
                                                                 gp = grid::gpar(cex = 1.3,
                                                                                 fontsize = 10)),
                                           bottom = grid::textGrob("Age Group",
                                                                   gp = grid::gpar(cex = 1.3,
                                                                                   fontsize = 10)))
   
   
   plt_scores_group
```

```{r}
degenes <- calculateDE(normdata,metadata,variables=c("age"))
gseadf <- runGSEA(degenes,signatures_bidirectional[c("HernandezSegura","SAUL_SEN_MAYO")])
gseadf
plotNESlollipop(gseadf)

(plt_enrich_skin <- plotGSEAenrichment(GSEA_results=gseadf, 
                   DEGList=degenes, 
                   gene_sets=signatures_bidirectional[c("HernandezSegura","SAUL_SEN_MAYO")], 
                   widthTitle=40, grid = TRUE, titlesize = 10, nrow=1  ))

``` 


```{r}

metadata$simpler <- ifelse(metadata$group == "Young \n (\u2264 25 years)", "Young", "Old")
degenes <- calculateDE(normdata,metadata,variables=c("simpler"), contrasts = "Old-Young")
gseadf <- runGSEA(degenes,signatures_bidirectional[c("HernandezSegura","SAUL_SEN_MAYO")])
gseadf
plotNESlollipop(gseadf)

(plt_enrich_skin <- plotGSEAenrichment(GSEA_results=gseadf, 
                   DEGList=degenes, 
                   gene_sets=signatures_bidirectional[c("HernandezSegura","SAUL_SEN_MAYO")], 
                   widthTitle=40, grid = TRUE, titlesize = 10, nrow=1  ))

``` 
```{r fig.width=6, fig.height=12}

plt_scores_skin <- ggarrange(plt_scores_age, plt_scores_group, NULL,plt_enrich_skin, ncol=1, nrow=4,heights=c(1,1.3,0.1,1))
(plt_scores_skin <- annotate_figure(plt_scores_skin, top = text_grob("GSE85358: \nInner Forearm Epidermis\n", face = "bold", size = 14 )))

```

```{r fig.width=6, fig.height=10}

plt_scores_skin <- ggarrange(plt_scores_group, NULL,plt_enrich_skin, ncol=1, nrow=3,heights=c(1.2,0.1,1))
(plt_scores_skin <- annotate_figure(plt_scores_skin, top = text_grob("GSE85358: \nInner Forearm Epidermis\n", face = "bold", size = 14 )))

```

<!-- ```{r} -->
<!-- data_gtex <- readRDS("../data_aux/data_gtex.rds") -->
<!-- GTEx_alltissues <- readRDS("../data_aux/GTExV8_voyagercorrected.rds") -->
<!-- metadata_GTEx_alltissues <- readRDS("../data_aux/GTExV8_metadata.rds") # Restricted access -->
<!-- metadata_GTEx_alltissues$SEX <- ifelse(metadata_GTEx_alltissues$SEX == 1,"Male","Female") -->
<!-- metadata_GTEx_alltissues -->
<!-- ``` -->

<!-- ```{r} -->
<!-- unique(metadata_GTEx_alltissues$SMTSD) -->
<!-- ``` -->

<!-- ```{r} -->
<!-- meta_skin <- metadata_GTEx_alltissues[metadata_GTEx_alltissues$SMTSD == "Skin - Sun Exposed (Lower leg)",] -->
<!-- gtexxskin <- GTEx_alltissues[,meta_skin$SAMPID] -->

<!-- youngoldsamps <- meta_skin$SAMPID[meta_skin$AGE <= 25 | meta_skin$AGE >=60] -->
<!-- meta_skin_youngold <- meta_skin[meta_skin$SAMPID %in% youngoldsamps,] -->
<!-- gtexxskin_youngold <- gtexxskin[,meta_skin_youngold$SAMPID] -->
<!-- meta_skin_youngold$Group <- ifelse(meta_skin_youngold$AGE <= 25, "Young \n (\u2264 25 years)", "Old \n (\u2265 60 years)") -->
<!-- table(meta_skin_youngold$Group) -->
<!-- table(metadata$group) -->
<!-- set.seed("12345678") -->
<!-- youngoldsampssubset <- meta_skin_youngold$SAMPID[meta_skin_youngold$Group == "Young \n (\u2264 25 years)"][sample(1:24)] -->
<!-- youngoldsampssubset <- c(youngoldsampssubset, meta_skin_youngold$SAMPID[meta_skin_youngold$Group == "Old \n (\u2265 60 years)"][sample(1:24)]) -->

<!-- youngoldsampssubset -->
<!-- meta_skin_youngold <- meta_skin_youngold[meta_skin_youngold$SAMPID %in% youngoldsampssubset,] -->
<!-- gtexxskin_youngold <- gtexxskin_youngold[,meta_skin_youngold$SAMPID] -->
<!-- ``` -->

<!-- ```{r} -->
<!-- cols_agegroup <- c( -->
<!--   "Young \n (\u2264 25 years)"    = "#3A798F",    -->
<!--   "Old \n (\u2265 60 years)" = "#3C3A8F"     -->
<!-- ) -->

<!-- PlotScores(data=gtexxskin_youngold, metadata=meta_skin_youngold, gene_sets = signatures_bidirectional[c("HernandezSegura","SAUL_SEN_MAYO")], Variable = "AGE", method="logmedian", xlab = "Age (years)", ColorValues = "#3A508F85")  -->

<!-- PlotScores(data=gtexxskin_youngold, metadata=meta_skin_youngold, gene_sets = signatures_bidirectional[c("HernandezSegura","SAUL_SEN_MAYO")], Variable = "Group", ColorVariable = "Group", method="logmedian", xlab = "Age Group", pvalcalc = T, ColorValues = cols_agegroup)  -->
<!-- ``` -->




### Lung (microarray) - GSE47460

only control samples

```{r}
## ============================================================
## Parse Agilent microarray txt.gz files -> expression matrix
## GSE47460 format: no GeneName column, uses SystematicName
## Maps SystematicName -> HGNC symbol via org.Hs.eg.db
## Signal:   gProcessedSignal
## Controls: removed (ControlType != 0)
## Samples:  CTRL only (filename contains "_CTRL")
## ============================================================

library(data.table)
library(dplyr)
library(org.Hs.eg.db)
library(AnnotationDbi)

# ── User settings ─────────────────────────────────────────────
INPUT_DIR  <- "../data_aux/GSE47460_RAW/"
OUTPUT_CSV <- "../data_aux/GSE47460_expression_matrix.csv"
OUTPUT_RDS <- "../data_aux/GSE47460_expression_matrix.rds"
# ──────────────────────────────────────────────────────────────


# ── Discover CTRL files only ───────────────────────────────────
files <- list.files(INPUT_DIR, pattern = "_CTRL\\.txt\\.gz$",
                    full.names = TRUE, recursive = FALSE)

message(sprintf("Found %d CTRL files.", length(files)))


#' Parse a single Agilent file (GSE47460 format, no GeneName column).
#' Returns a two-column data.table: SystematicName, gProcessedSignal.
parse_agilent_file <- function(filepath) {
  
  lines <- readLines(filepath, warn = FALSE)
  
  # Strip Windows carriage returns if present
  lines <- gsub("\r", "", lines)
  
  # Find FEATURES header line
  feat_header_idx <- which(startsWith(lines, "FEATURES\t"))
  
  if (length(feat_header_idx) == 0) {
    warning(paste("No FEATURES block found in:", filepath))
    return(NULL)
  }
  
  col_names <- strsplit(lines[feat_header_idx], "\t")[[1]]
  col_names[1] <- "block_type"
  
  # DATA lines after the FEATURES header
  data_lines <- lines[(feat_header_idx + 1):length(lines)]
  data_lines <- data_lines[startsWith(data_lines, "DATA\t")]
  
  if (length(data_lines) == 0) {
    warning(paste("No DATA lines found in:", filepath))
    return(NULL)
  }
  
  dt <- fread(
    text         = c(paste(col_names, collapse = "\t"), data_lines),
    sep          = "\t",
    header       = TRUE,
    fill         = TRUE,
    showProgress = FALSE
  )
  
  # ── Filter controls ───────────────────────────────────────
  dt <- dt[ControlType == 0]
  
  # ── Keep relevant columns ─────────────────────────────────
  needed <- c("SystematicName", "gProcessedSignal")
  missing_cols <- setdiff(needed, names(dt))
  if (length(missing_cols) > 0) {
    warning(paste("Missing columns:", paste(missing_cols, collapse = ", "),
                  "in:", filepath))
    return(NULL)
  }
  
  dt <- dt[, ..needed]
  
  # ── Remove empty/ambiguous probe annotations ──────────────
  dt <- dt[!is.na(SystematicName) & SystematicName != "" & SystematicName != "---"]
  
  # ── Coerce signal to numeric ──────────────────────────────
  dt[, gProcessedSignal := as.numeric(gProcessedSignal)]
  
  return(dt)
}


# ── Parse all CTRL files ───────────────────────────────────────
message("Parsing files...")

sample_list <- lapply(files, function(f) {
  message("  Processing: ", basename(f))
  tryCatch(
    parse_agilent_file(f),
    error = function(e) {
      warning(paste("Error parsing", basename(f), ":", e$message))
      return(NULL)
    }
  )
})

names(sample_list) <- sub("\\.txt\\.gz$", "", basename(files))

# Remove failed samples
failed <- which(sapply(sample_list, is.null))
if (length(failed) > 0) {
  warning(sprintf("%d file(s) failed to parse.", length(failed)))
  sample_list <- sample_list[-failed]
}


# ── Build probe-level matrix ───────────────────────────────────
# Collapse duplicate SystematicName entries within each sample (mean),
# then merge across samples.
message("Merging at probe level...")

probe_matrix <- Reduce(function(a, b) {
  merge(a, b, by = "SystematicName", all = TRUE)
}, lapply(names(sample_list), function(nm) {
  dt <- sample_list[[nm]]
  dt <- dt[, .(gProcessedSignal = mean(gProcessedSignal, na.rm = TRUE)),
           by = SystematicName]
  setnames(dt, "gProcessedSignal", nm)
  dt
}))


# ── Annotate: SystematicName -> HGNC symbol via org.Hs.eg.db ──
message("Building annotation lookup from org.Hs.eg.db...")

# RefSeq (NM_, NR_, XM_, XR_) -> gene symbol
refseq_lookup <- AnnotationDbi::select(
  org.Hs.eg.db,
  keys    = AnnotationDbi::keys(org.Hs.eg.db, keytype = "REFSEQ"),
  columns = c("REFSEQ", "SYMBOL"),
  keytype = "REFSEQ"
) %>%
  dplyr::filter(!is.na(SYMBOL), !is.na(REFSEQ), SYMBOL != "", REFSEQ != "") %>%
  dplyr::distinct(REFSEQ, .keep_all = TRUE) %>%
  dplyr::rename(id = REFSEQ, hgnc_symbol = SYMBOL)

# Ensembl transcript (ENST) -> gene symbol
enst_lookup <- AnnotationDbi::select(
  org.Hs.eg.db,
  keys    = AnnotationDbi::keys(org.Hs.eg.db, keytype = "ENSEMBLTRANS"),
  columns = c("ENSEMBLTRANS", "SYMBOL"),
  keytype = "ENSEMBLTRANS"
) %>%
  dplyr::filter(!is.na(SYMBOL), !is.na(ENSEMBLTRANS), SYMBOL != "", ENSEMBLTRANS != "") %>%
  dplyr::distinct(ENSEMBLTRANS, .keep_all = TRUE) %>%
  dplyr::rename(id = ENSEMBLTRANS, hgnc_symbol = SYMBOL)

# Combine both lookups
full_lookup <- dplyr::bind_rows(refseq_lookup, enst_lookup) %>%
  dplyr::distinct(id, .keep_all = TRUE)

message(sprintf("  Lookup table: %d unique IDs mapped to HGNC symbols.",
                nrow(full_lookup)))


# ── Join annotation to probe matrix ───────────────────────────
# Strip version numbers from SystematicName before joining
# e.g. NM_001234.3 -> NM_001234
probe_matrix[, id_clean := sub("\\.\\d+$", "", SystematicName)]

probe_annotated <- merge(
  probe_matrix,
  as.data.table(full_lookup),
  by.x  = "id_clean",
  by.y  = "id",
  all.x = FALSE   # drop probes with no HGNC symbol
)

n_unmapped <- nrow(probe_matrix) - nrow(probe_annotated)
message(sprintf("  %d probes dropped (no HGNC symbol found).", n_unmapped))

# Remove helper columns
probe_annotated[, c("SystematicName", "id_clean") := NULL]


# ── Collapse to gene level: mean across probes ─────────────────
message("Collapsing to gene level (mean across probes)...")

sample_cols <- names(sample_list)

expr_gene <- probe_annotated[,
  lapply(.SD, function(x) mean(as.numeric(x), na.rm = TRUE)),
  by      = hgnc_symbol,
  .SDcols = sample_cols
]


# ── Final formatting ───────────────────────────────────────────
expr_df             <- as.data.frame(expr_gene)
rownames(expr_df)   <- expr_df$hgnc_symbol
expr_df$hgnc_symbol <- NULL

message(sprintf("Final matrix: %d genes x %d samples",
                nrow(expr_df), ncol(expr_df)))

na_count <- sum(is.na(expr_df))
if (na_count > 0) message(sprintf("Note: %d NA values present.", na_count))

expr_df_complete <- expr_df[complete.cases(expr_df), ]
cat(sprintf("Genes retained after removing NAs: %d\n", nrow(expr_df_complete)))
                                                                                                                                                                                                                                                                                

# ── Save ───────────────────────────────────────────────────────
write.csv(expr_df_complete, file = OUTPUT_CSV, quote = FALSE)
saveRDS(expr_df_complete,   file = OUTPUT_RDS)
message("Saved: ", OUTPUT_CSV)
message("Saved: ", OUTPUT_RDS)
```



```{r}
## ============================================================
## Extract sample metadata from GEO SOFT file
## and align with expression matrix sample names
## ============================================================
 

# ── User settings ─────────────────────────────────────────────
SOFT_FILE  <- "../data_aux/GSE47460_family.soft"  # <-- change this
OUTPUT_CSV <- "../data_aux/GSE47460_metadata.csv"

# Optional: if you already have your expression matrix loaded,
# set this to align metadata rows to your sample columns.
# Otherwise leave as NULL.
EXPR_SAMPLE_NAMES <- colnames(expr_df_complete)
# ──────────────────────────────────────────────────────────────


# ── Parse SOFT file ────────────────────────────────────────────
message("Parsing SOFT file: ", SOFT_FILE)

gse <- getGEO(filename = SOFT_FILE, GSEMatrix = FALSE)

# Extract all GSM (sample) objects
gsm_list <- GSMList(gse)
message(sprintf("Found %d samples.", length(gsm_list)))


# ── Extract metadata from each sample ─────────────────────────
# Each GSM has a Meta slot with all characteristics
extract_gsm_metadata <- function(gsm) {
  meta <- Meta(gsm)
  
  # Core fields always present in GEO
  core <- data.frame(
    sample_id    = meta$geo_accession %||% NA,
    sample_title = meta$title         %||% NA,
    stringsAsFactors = FALSE
  )
  
  # Characteristics lines: format is usually "key: value"
  # e.g. "age: 45", "Sex: Male", "tissue: skin"
  chars <- meta$characteristics_ch1
  
  if (!is.null(chars) && length(chars) > 0) {
    char_pairs <- lapply(chars, function(x) {
      parts <- strsplit(x, ":\\s*", perl = TRUE)[[1]]
      if (length(parts) == 2) {
        key <- trimws(gsub("[^a-zA-Z0-9_]", "_", tolower(parts[1])))
        val <- trimws(parts[2])
        return(setNames(data.frame(val, stringsAsFactors = FALSE), key))
      }
      return(NULL)
    })
    char_df <- bind_cols(Filter(Negate(is.null), char_pairs))
    core <- bind_cols(core, char_df)
  }
  
  return(core)
}

# Fallback for NULL values (like %||% in rlang)
`%||%` <- function(a, b) if (!is.null(a) && length(a) > 0) a else b

metadata <- bind_rows(lapply(gsm_list, extract_gsm_metadata))

message(sprintf("Metadata columns found: %s",
                paste(names(metadata), collapse = ", ")))


# ── Derive sample name matching your expression matrix ────────
# GEO supplementary files are often named after the GSM accession
# OR after the sample title — we keep both so you can match either.
# The expression matrix script named samples after the filename
# (minus .txt.gz), so you may need to adjust the join key below.

message("\nPreview of sample_id and sample_title:")
print(metadata[, c("sample_id", "sample_title")])


# ── Optional: align with expression matrix columns ────────────
if (!is.null(EXPR_SAMPLE_NAMES)) {
  
  # Try matching on sample_id (GSM accession) first
  metadata$matched_expr_col <- NA_character_
  
  for (i in seq_len(nrow(metadata))) {
    gsm_id    <- metadata$sample_id[i]
    gsm_title <- metadata$sample_title[i]
    
    # Match if expr column contains the GSM accession
    hit_id    <- EXPR_SAMPLE_NAMES[grepl(gsm_id, EXPR_SAMPLE_NAMES, fixed = TRUE)]
    # Fallback: match on title
    hit_title <- EXPR_SAMPLE_NAMES[grepl(gsm_title, EXPR_SAMPLE_NAMES, fixed = TRUE)]
    
    if (length(hit_id) == 1) {
      metadata$matched_expr_col[i] <- hit_id
    } else if (length(hit_title) == 1) {
      metadata$matched_expr_col[i] <- hit_title
    } else {
      warning(sprintf("Could not match sample: %s (%s)", gsm_id, gsm_title))
    }
  }
  
  unmatched <- sum(is.na(metadata$matched_expr_col))
  if (unmatched > 0) {
    warning(sprintf("%d sample(s) could not be matched to expression matrix columns.",
                    unmatched))
  } else {
    message("All samples matched successfully.")
  }
  
  # Reorder metadata to match expression matrix column order
  metadata <- metadata[match(EXPR_SAMPLE_NAMES,
                             metadata$matched_expr_col), ]
  rownames(metadata) <- EXPR_SAMPLE_NAMES
}

metadata$sampleID <- row.names(metadata)
metadata <- metadata[,c(17,c(1:16))]
 
# ── Save ───────────────────────────────────────────────────────
write.csv(metadata, file = OUTPUT_CSV, quote = TRUE, row.names = TRUE)
message("Saved: ", OUTPUT_CSV)

saveRDS(metadata,"../data_aux/metadata_GSE47460.rds")
# metadata is your final sample x variable metadata table
```


```{r}
expr_df_complete <- readRDS("../data_aux/GSE47460_expression_matrix.rds")
metadata <- readRDS("../data_aux/metadata_GSE47460.rds")
```

##### Filtering

```{r}
data <- expr_df_complete
```

```{r}
# Plot
(plt_beforefilt <- ggplot() + aes(x=log10(rowSums(data))) +
   geom_density(fill="grey", alpha=0.5) + theme_bw() + 
   theme(axis.text.x = element_text(angle = 45, hjust=1),
         plot.title = element_text(hjust = 0.5),
         plot.subtitle = element_text(hjust = 0.5)) +
   ggtitle("Before Filtering") +
   xlab(expression(log[10]~"(counts per gene)")) +
   ylab("Frequency") + 
   geom_vline(xintercept = 3.7, linetype="dotted",color = "black", size=1.5))

```


```{r}
filtdata <- data[rowSums(data) > 10^3.7,]
nrow(data)
nrow(filtdata)


cat("Number of genes before filtering:", nrow(data), "\n")
cat("Number of genes after filtering:", nrow(filtdata), "\n")

```


```{r}

# Plot
(plt_afterfilt <- ggplot() + aes(x=log10(rowSums(filtdata))) +
   geom_density(fill="grey", alpha=0.5) + theme_bw() + 
   theme(axis.text.x = element_text(angle = 45, hjust=1),
         plot.title = element_text(hjust = 0.5),
         plot.subtitle = element_text(hjust = 0.5)) +
   ggtitle("After Filtering") +
   xlab(expression(log[10]~"(counts per gene)")) +
   ylab("Frequency") )

```



##### Normalisation

```{r fig.width=6, fig.height=4}

filtcounts_ggplot <- stack(log2(filtdata))
colnames(filtcounts_ggplot) <- c("expression", "sampleID")
filtcounts_ggplot <- merge(filtcounts_ggplot,metadata, by="sampleID")


(plt_beforenorm <- ggplot(filtcounts_ggplot, aes(sampleID,expression)) + geom_boxplot(alpha=0.4, outlier.size = 0.5)+ xlab("Samples") + ylab("log2(counts)")   +  theme(axis.text.x = element_text(angle = 45, hjust=1))+
    theme_bw()+ 
    theme(plot.title = element_text(hjust = 0.5),
          plot.subtitle = element_text(hjust = 0.5),
          axis.text.x=element_blank(),
          axis.ticks.x=element_blank()) +
    ggtitle("Before Normalisation") )

```


```{r}
dgelist_normcounts <- DGEList(filtdata) # the matrix with gene names as row names and counts in columns
dgelist_normcounts <- calcNormFactors(dgelist_normcounts, method = "TMM") 
normalised_counts <- as.data.frame(voom(dgelist_normcounts,plot=F)$E) 
#normalised_counts <- as.data.frame(edgeR::cpm(dgelist_normcounts, log = FALSE)) 
normdata <- 2^normalised_counts
lognormdata <- log2(normdata+1)
```


```{r fig.width=6, fig.height=4}

normcounts_ggplot <- stack(log2(normdata))
colnames(normcounts_ggplot) <- c("expression", "sampleID")
normcounts_ggplot <- merge(normcounts_ggplot,metadata, by="sampleID")


(plt_afternorm <- ggplot(normcounts_ggplot, aes(sampleID,expression)) + geom_boxplot(alpha=0.4, outlier.size = 0.5)+ xlab("Samples") + ylab("log2(counts)")   +  theme(axis.text.x = element_text(angle = 45, hjust=1))+
    theme_bw()+ 
    theme(plot.title = element_text(hjust = 0.5),
          plot.subtitle = element_text(hjust = 0.5),
          axis.text.x=element_blank(),
          axis.ticks.x=element_blank()) +
    ggtitle("After Normalisation") )

```


##### PCA

```{r fig.width=6, fig.height=4}
y <- DGEList(lognormdata, samples= metadata)
PCA  <- prcomp(t(y$counts), scale=FALSE, center=TRUE)
PCACounts  <- PCA$x
PCACounts  <- as.data.frame(PCACounts )
PCAData <-  cbind(PCACounts[1:10],y$samples) 

ev = PCA$sdev^2 
pc1= round(100*ev[1]/sum(ev),2) 
pc2 = round(100*ev[2]/sum(ev),2) 
pc3 = round(100*ev[3]/sum(ev),2) 
pc4 = round(100*ev[4]/sum(ev),2) 
pc5 = round(100*ev[5]/sum(ev),2) 
pc6 = round(100*ev[6]/sum(ev),2)  

(pltpca_1 <- ggplot(PCAData, aes(PC1,PC2))+
    geom_point(aes(color=sex), size=5, alpha=0.6)+ 
    xlab(paste0("PC1: ",pc1,"% variance")) +
    ylab(paste0("PC2: ",pc2,"% variance")) + 
    theme_bw()+
    guides(color = guide_legend(nrow = 3)) +
    theme(legend.position="top", legend.box="vertical", legend.title = element_blank(), legend.margin=margin())  + 
    geom_vline(xintercept=0, linetype="dotted") + 
    geom_hline(yintercept=0, linetype="dotted") )

(pltpca_2 <- ggplot(PCAData, aes(PC1,PC2))+
    geom_point(aes(color=as.numeric(age)), size=5, alpha=0.6)+ 
    xlab(paste0("PC1: ",pc1,"% variance")) +
    ylab(paste0("PC2: ",pc2,"% variance")) + 
    theme_bw()+
    theme(legend.position="top", legend.box="vertical", legend.title = element_blank(), legend.margin=margin())  + 
    geom_vline(xintercept=0, linetype="dotted") +  
    geom_hline(yintercept=0, linetype="dotted"))
 


```

####Calculate Scores

```{r fig.width=10, fig.height=4}
  
(plt_scores_age <- PlotScores(data=normdata, metadata=metadata, gene_sets = signatures_bidirectional[c("HernandezSegura","SAUL_SEN_MAYO")], Variable = "age", method="logmedian", xlab = "Age (years)", pvalcalc = T,ColorValues = "#3A508F85") )

metadata$age_group <- cut(
  metadata$age,
  breaks = c(-Inf, 49, 59, 69, 79, Inf),
  labels = c("<50", "50-59", "60-69", "70-79", "80+"),
  right  = TRUE
)

metadata$age_group <- factor(
  metadata$age_group,
  levels  = c("<50", "50-59", "60-69", "70-79", "80+"),
  ordered = TRUE
)
 
age_colors <- c(
  "<50"   = "#D0E8F5",
  "50-59" = "#89C4E1",
  "60-69" = "#4A90C4",
  "70-79" = "#1F5C8B",
  "80+"   = "#0A2F4E"
)

(plt_scores_group <- PlotScores(data=normdata, metadata=metadata, gene_sets = signatures_bidirectional[c("HernandezSegura","SAUL_SEN_MAYO")], Variable = "age_group", ColorVariable = "age_group", method="logmedian", xlab = "Age Group", pvalcalc = T, ColorValues = age_colors) )

```







```{r fig.width=10, fig.height=4}

# (plt_scores_age <- PlotScores(data=normdata, metadata=metadata, gene_sets = signatures_bidirectional[c("HernandezSegura","SAUL_SEN_MAYO")], Variable = "age", method="logmedian", xlab = "Age (years)", pvalcalc = T,ColorValues = "#3A508F85") )

  
(calcscores_group <- CalculateScores(data=normdata, metadata=metadata, gene_sets = signatures_bidirectional[c("HernandezSegura","SAUL_SEN_MAYO")], method = "logmedian") )

pltlist <- list()

for (sig in names(calcscores_group)) {
  model <- lm(score ~ get("age"), data = calcscores_group[[sig]])
  results_var <- compute_cohens_f_pval(model, "Numeric")
  
  line1 <- wrap_title(paste0("Cohen's f = ",
                             format(signif(results_var["Cohen_f"], digits=3),
                                    scientific = FALSE)), width = 20)
  line2 <- wrap_title(paste0("p = ",
                             format(signif(results_var["P_Value"], digits=3),
                                    scientific = TRUE)), width = 20)
  subtitle <- paste(line1, line2, sep = "\n")
  
  
  pltlist[[sig]] <- ggplot2::ggplot(calcscores_group[[sig]], ggplot2::aes_string(x = "age", y = "score")) + 
    ggplot2::geom_point(size = 3, alpha = 0.5, color="#3A508F85")+ 
      geom_smooth(method = "lm", col = "black", se = FALSE, size=1.3) + 
      ggpubr::stat_cor(aes(label = ..r.label..), method = "spearman",
                       cor.coef.name="rho", label.x = 30, size=4) +  
 ggplot2::theme_bw() +
      ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 45, hjust = 1,
                                                         size = 10),
                     axis.text.y = ggplot2::element_text(  size = 10),
                     plot.title = ggplot2::element_text(hjust = 0.5,
                                                        size = 14-1),
                     plot.subtitle = ggplot2::element_text(hjust = 0.5,
                                                           size = 14 - 1.5,
                                                           face = "italic")) +
      ggplot2::labs(title = sig, subtitle = subtitle,
                    color = "", x = "", y = "")
    
 }
 
 plt_scores_age <- ggarrange(plotlist = pltlist, ncol=length(pltlist), nrow=1, widths = c(1,1), common.legend = T)
   plt_scores_age <- ggpubr::annotate_figure(plt_scores_age,
                                           left = grid::textGrob(expression(log[2]*"-median score"),
                                                                 rot = 90, vjust = 1,
                                                                 gp = grid::gpar(cex = 1.3,
                                                                                 fontsize = 10)),
                                           bottom = grid::textGrob("Age (years)",
                                                                   gp = grid::gpar(cex = 1.3,
                                                                                   fontsize = 10)))
   
   
   plt_scores_age
```


```{r}
degenes <- calculateDE(normdata,metadata,variables=c("age"))
gseadf <- runGSEA(degenes,signatures_bidirectional[c("HernandezSegura","SAUL_SEN_MAYO")])
gseadf
plotNESlollipop(gseadf)

(plt_enrich_lung <- plotGSEAenrichment(GSEA_results=gseadf, 
                   DEGList=degenes, 
                   gene_sets=signatures_bidirectional[c("HernandezSegura","SAUL_SEN_MAYO")], 
                   widthTitle=40, grid = TRUE, titlesize = 10, nrow=1  ))

``` 
```{r fig.width=6, fig.height=12}

plt_scores_lung <- ggarrange(plt_scores_age, plt_scores_group, NULL, plt_enrich_skin, ncol=1, nrow=4,heights=c(1,1.15,0.1,1))
(plt_scores_lung <- annotate_figure(plt_scores_lung, top = text_grob("GSE47460: \nWhole Lung Homogenate\n", face = "bold", size = 14 )))

```



```{r fig.width=6, fig.height=8}

plt_scores_lung <- ggarrange(plt_scores_age, NULL, plt_enrich_lung, ncol=1, nrow=3,heights=c(1.2,0.1,1))
(plt_scores_lung <- annotate_figure(plt_scores_lung, top = text_grob("GSE47460: \nWhole Lung Homogenate\n", face = "bold", size = 14 )))

```


### Combine

```{r fig.width=12, fig.height=8}
(fig_combined <- ggarrange(plt_scores_skin,NULL, plt_scores_lung,nrow=1, labels=c("A","","B"), widths=c(1,0.1,1)))
```

```{r}
ggplot2::ggsave("../Figures/Figs/R1_5_OtherAgeingDatasets_v2.png",
                fig_combined,
                width = 12, height=8, bg = 'white')

ggplot2::ggsave("../Figures/Figs/R1_5_OtherAgeingDatasets_v2.pdf",
                fig_combined,
                width = 12, height=8, bg = 'white')
```


## Session Info
```{r}
sessionInfo()
```
