---
title: "Response to Reviewers"
author: "Rita Martins-Silva"
date: "02/03/2026"
output: html_document
---


# Response to Reviewers {.tabset .tabset-pills}

## Set up {.tabset}

```{r}
setwd("../")
```
### Relevant Data

```{r}
signatures_bidirectional <- readRDS("../data/SenescenceSigntures_Bidirectional.rds")
t2gene2 <- fread("../data_aux/RefSeq_Transcripts_To_Genes.csv")
t2gene2 <- t2gene2[,c("name","name2")]
colnames(t2gene2) <- c("transcript","gene")
```

### Libraries

```{r}
library(data.table)
library("ggplot2")
library("colorspace")
library("scales")
#library("scater") 
library("reshape2")
#library("markeR")
#library("ggbreak")
#library("ggnewscale")
library("dplyr")
library("tidyr")  
library(patchwork)
library("ggpubr") 
library(purrr)
library(edgeR)
library(RColorBrewer)
library(pROC) 
#library("ggforce")
library(ggrepel)
#devtools::install_github("DiseaseTranscriptomicsLab/markeR@*release")
library(biomaRt)
library("tximport")
```

### Functions

Load markeR functions directly

```{r}
# Path to your folder with functions
fun_dir <- "functions_Sept18_2025/"

# Get all .R files in the folder
files <- list.files(paste0("../",fun_dir), pattern = "\\.R$", full.names = TRUE)

# Source them one by one
invisible(lapply(files, source))

```

### Colors

```{r}
Condition_colors <- c(
  "Senescent"     = "#F4A261",  # Soft orange (not goldenrod)
  "Proliferative" = "#A1C298",  # Desaturated sage green
  "Quiescent"     = "#9DB4C0"   # Dusty teal-gray
)

CellTypecols <- c(
  "Fibroblast"   = "#E69F00",  # Orange
  "Keratinocyte" = "#56B4E9",  # Sky Blue
  "Melanocyte"   = "#009E73",  # Bluish Green
  "Endothelial"  = "#F0E442",  # Yellow
  "Neuronal"     = "#0072B2",  # Blue
  "Mesenchymal"  = "#D55E00"   # Vermilion
)

CellTypecols_alt <- c(
  "Fibroblast"   = "#1b9e77",  # Teal green
  "Keratinocyte" = "#d95f02",  # Warm orange-brown
  "Melanocyte"   = "#7570b3",  # Muted purple
  "Endothelial"  = "#e7298a",  # Dark pink
  "Neuronal"     = "#66a61e",  # Olive green
  "Mesenchymal"  = "#e6ab02"   # Mustard
)


dataset_colors <- c(
  "HernandezSegura" = "#8DD3C7",
  "Mangelinck"      = "#FFFFB3",
  "Chan"            = "#BEBADA",
  "Purcell"         = "#FB8072",
  "Marthandan2016"  = "#80B1D3",
  "Marthandan2015"  = "#FDB462",
  "Wang"            = "#B3DE69",
  "Suda"            = "#FCCDE5",
  "McHugh"          = "#D9D9D9",
  "Savic"           = "#BC80BD",
  "Skea"            = "#CCEBC5",
  "Laurent"         = "#FFED6F",
  "Tasdemir"        = "#D0B7E1",
  "Lenain"          = "#FFB3BA",
  "Aarts"           = "#C2C2F0",
  "Casella"         = "#FFDAC1",
  "Numa"            = "#C6E2FF",
  "Admasu"          = "#F5CBA7",
  "Urata"           = "#C7CEEA",
  "Wang2023"        = "#E6B0AA",
  "Fu"              = "#A9DFBF",
  "Tanke"           = "#F9E79F",
  "Bawens"          = "#AED6F1"
)

geneSet_colors <- brewer.pal(9, "Set1")  # 9 distinct colors, mostly bright and distinct

```

```{r}
show_col(Condition_colors, ncol=3)
show_col(CellTypecols)
show_col(dataset_colors)
show_col(geneSet_colors)
```


## Comment 3.3 — External validation on an independent dataset

R.: Apply markeR's benchmarking mode to a senescence dataset not included in the original compendium, to evaluate whether HernandezSegura and SAUL_SEN_MAYO retain their discriminatory performance in a fully independent evaluation set. 



### GSE297560 (Space Radiation)
HiSat2 + feature counts, hg38

#### Data

```{r}
data <- as.data.frame(fread("../data_aux/GSE297560_counts.txt")[,-c(2:6)])
row.names(data) <- data$Geneid
data$Geneid <-  NULL
head(data)
# Convert row names from ENSEMBL ID to Gene Name using biomaRt

# Connect to Ensembl
mart <- useMart("ensembl", dataset = "hsapiens_gene_ensembl")

# Query the conversion
gene_conversion <- getBM(
  attributes = c("ensembl_gene_id", "hgnc_symbol"),
  filters    = "ensembl_gene_id",
  values     = rownames(data),
  mart       = mart
)

# Replace rownames — keeping only genes with a match and a non-empty symbol
gene_conversion <- gene_conversion[!duplicated(gene_conversion$hgnc_symbol), ]
data <- data[gene_conversion$ensembl_gene_id, ]
rownames(data) <- gene_conversion$hgnc_symbol
#remove .trimmed from sample names
colnames(data) <- gsub("\\.trimmed", "", colnames(data))

```


- Healthy_CTRL: Low-passage fibroblasts that were never shipped. Represent the true untreated baseline but are not the appropriate matched control for irradiation comparisons, as any differences relative to irradiated samples would conflate shipping stress with radiation effects.

- MOCK_CTRL: Fibroblasts that were shipped to the radiation facility but not irradiated, paired with the charged particle irradiation experiments (Hydrogen, Iron, Silicon). Control for the confounding effect of shipping stress, isolating the effect of radiation.

- MOCKgamma_CTRL: Identical to MOCK_CTRL but paired specifically with the gamma irradiation experiment. Kept separate to maintain matched experimental pairing.

```{r}
# Load metadata
metadata <- read.csv("../data_aux/GSE297560_metadata.csv")
metadata <- metadata[,c("Run","Sample.Name","source_name", "treatment")]
metadata$datasetID <- "GSE297560"

metadata$condition <- ifelse(grepl("control", metadata$treatment, ignore.case = TRUE), "Proliferative", ifelse(grepl("irradiation", metadata$treatment, ignore.case = TRUE), "Senescent", "Proliferative"))


# Attempt to match SRR with colnames
sample_to_srr <- c(
  "Healthy_CTRL_1_HISAT2"   = "SRR33626854",#
  "Healthy_CTRL_2_HISAT2"   = "SRR33626853",#
  "Healthy_CTRL_3_HISAT2"   = "SRR33626852",#
  "4Hydrogen_1_HISAT2"      = "SRR33626842",#
  "4Hydrogen_2_HISAT2"      = "SRR33626841",#
  "4Hydrogen_3_HISAT2"      = "SRR33626840",#
  "4Iron_1_HISAT2"          = "SRR33626845",#
  "4Iron_2_HISAT2"          = "SRR33626844",#
  "4Iron_3_HISAT2"          = "SRR33626843",#
  "4Silicon_1_HISAT2"       = "SRR33626848",#
  "4Silicon_2_HISAT2"       = "SRR33626847",#
  "4Silicon_3_HISAT2"       = "SRR33626846",#
  "MOCK_CTRL_1_HISAT2"      = "SRR33626857",#
  "MOCK_CTRL_2_HISAT2"      = "SRR33626856",#
  "MOCK_CTRL_3_HISAT2"      = "SRR33626855",#
  "4Gamma_1_HISAT2"         = "SRR33626839",#
  "4Gamma_3_HISAT2"         = "SRR33626838",#
  "4Gamma_2_HISAT2"         = "SRR33626837",#
  "MOCKgamma_CTRL_1_HISAT2" = "SRR33626851",
  "MOCKgamma_CTRL_2_HISAT2" = "SRR33626850",
  "MOCKgamma_CTRL_3_HISAT2" = "SRR33626849"
)

# Rename columns of your expression matrix
colnames(data) <- sample_to_srr[colnames(data)]

# Reorder metadata to match column order of your_df
metadata <- metadata[match(colnames(data), metadata$Run), ]

treatment_map <- c(
  "Low passage\\, unshipped"                            = "Healthy_CTRL",
  "Irradiation control shipped not irradiated"          = "MOCK_CTRL",
  "Gamma irradiation control shipped not irradiated"    = "MOCKgamma_CTRL",
  "4Gy Gamma irradiation"                               = "4Gy_Gamma",
  "4Gy Hydrogen irradiation"                            = "4Gy_Hydrogen",
  "4Gy Iron irradiation"                                = "4Gy_Iron",
  "4Gy Silicon irradiation"                             = "4Gy_Silicon"
)

metadata$treatment_simple <- treatment_map[metadata$treatment]
colnames(metadata)[1] <- "sampleID"
```



Remove Gamma and mock gamma, as in the paper the 4 gamma are not senescent...
Remove also Healthy_CTRL, they don't validate those samples in the paper either

```{r}
metadata <- metadata[!grepl("Gamma", metadata$treatment_simple, ignore.case = T), ]
metadata <- metadata[!grepl("Healthy_CTRL", metadata$treatment_simple, ignore.case = T), ]
data <- data[,metadata$sampleID]
```


```{r}
data
metadata
```

#### Preprocessing


##### Filtering

```{r}
# Plot
(plt_beforefilt <- ggplot() + aes(x=log10(rowSums(data))) +
   geom_density(fill="grey", alpha=0.5) + theme_bw() + 
   theme(axis.text.x = element_text(angle = 45, hjust=1),
         plot.title = element_text(hjust = 0.5),
         plot.subtitle = element_text(hjust = 0.5)) +
   ggtitle("Before Filtering") +
   xlab(expression(log[10]~"(counts per gene)")) +
   ylab("Frequency") + 
   geom_vline(xintercept = 2.9, linetype="dotted",color = "black", size=1.5))

```


```{r}
filtdata <- data[rowSums(data) > 10^2.9,]
nrow(data)
nrow(filtdata)


cat("Number of genes before filtering:", nrow(data), "\n")
cat("Number of genes after filtering:", nrow(filtdata), "\n")

```


```{r}

# Plot
(plt_afterfilt <- ggplot() + aes(x=log10(rowSums(filtdata))) +
   geom_density(fill="grey", alpha=0.5) + theme_bw() + 
   theme(axis.text.x = element_text(angle = 45, hjust=1),
         plot.title = element_text(hjust = 0.5),
         plot.subtitle = element_text(hjust = 0.5)) +
   ggtitle("After Filtering") +
   xlab(expression(log[10]~"(counts per gene)")) +
   ylab("Frequency") )

```



##### Normalisation

```{r fig.width=6, fig.height=4}

filtcounts_ggplot <- stack(log2(filtdata))
colnames(filtcounts_ggplot) <- c("expression", "sampleID")
filtcounts_ggplot <- merge(filtcounts_ggplot,metadata, by="sampleID")


(plt_beforenorm <- ggplot(filtcounts_ggplot, aes(sampleID,expression)) + geom_boxplot(alpha=0.4)+ xlab("Samples") + ylab("log2(counts)")   +  theme(axis.text.x = element_text(angle = 45, hjust=1))+
    theme_bw()+ 
    theme(plot.title = element_text(hjust = 0.5),
          plot.subtitle = element_text(hjust = 0.5),
          axis.text.x=element_blank(),
          axis.ticks.x=element_blank()) +
    ggtitle("Before Normalisation") )

```


```{r}
dgelist_normcounts <- DGEList(filtdata) # the matrix with gene names as row names and counts in columns
dgelist_normcounts <- calcNormFactors(dgelist_normcounts, method = "TMM") 
normalised_counts <- as.data.frame(voom(dgelist_normcounts,plot=F)$E) 
#normalised_counts <- as.data.frame(edgeR::cpm(dgelist_normcounts, log = FALSE)) 
normdata <- 2^normalised_counts
lognormdata <- log2(normdata+1)
```


```{r fig.width=6, fig.height=4}

normcounts_ggplot <- stack(log2(normdata))
colnames(normcounts_ggplot) <- c("expression", "sampleID")
normcounts_ggplot <- merge(normcounts_ggplot,metadata, by="sampleID")


(plt_afternorm <- ggplot(normcounts_ggplot, aes(sampleID,expression)) + geom_boxplot(alpha=0.4)+ xlab("Samples") + ylab("log2(counts)")   +  theme(axis.text.x = element_text(angle = 45, hjust=1))+
    theme_bw()+ 
    theme(plot.title = element_text(hjust = 0.5),
          plot.subtitle = element_text(hjust = 0.5),
          axis.text.x=element_blank(),
          axis.ticks.x=element_blank()) +
    ggtitle("After Normalisation") )

```


##### PCA

```{r fig.width=6, fig.height=4}
y <- DGEList(lognormdata, samples= metadata)
PCA  <- prcomp(t(y$counts), scale=FALSE, center=TRUE)
PCACounts  <- PCA$x
PCACounts  <- as.data.frame(PCACounts )
PCAData <-  cbind(PCACounts[1:10],y$samples) 

ev = PCA$sdev^2 
pc1= round(100*ev[1]/sum(ev),2) 
pc2 = round(100*ev[2]/sum(ev),2) 
pc3 = round(100*ev[3]/sum(ev),2) 
pc4 = round(100*ev[4]/sum(ev),2) 
pc5 = round(100*ev[5]/sum(ev),2) 
pc6 = round(100*ev[6]/sum(ev),2)  


(pltpca_1 <- ggplot(PCAData, aes(PC1,PC2))+
    geom_point(aes(color=treatment_simple), size=5, alpha=0.6)+ 
    xlab(paste0("PC1: ",pc1,"% variance")) +
    ylab(paste0("PC2: ",pc2,"% variance")) + 
    theme_bw()+
    guides(color = guide_legend(nrow = 2)) +
    theme(legend.position="top", legend.box="vertical", legend.title = element_blank(), legend.margin=margin())  + 
    geom_vline(xintercept=0, linetype="dotted") + 
    geom_hline(yintercept=0, linetype="dotted") )

(pltpca_2 <- ggplot(PCAData, aes(PC1,PC2))+
    geom_point(aes(color=condition), size=5, alpha=0.6)+ 
    xlab(paste0("PC1: ",pc1,"% variance")) +
    ylab(paste0("PC2: ",pc2,"% variance")) + 
    theme_bw()+
    theme(legend.position="top", legend.box="vertical", legend.title = element_blank(), legend.margin=margin())  + 
    geom_vline(xintercept=0, linetype="dotted") + 
    guides(color = guide_legend(nrow = 2)) +
    geom_hline(yintercept=0, linetype="dotted") +
    scale_color_manual(values=Condition_colors))


```

#### Calculate Scores

```{r}
(pltscores <- PlotScores(data = normdata, 
                         metadata = metadata, 
                         gene_sets = signatures_bidirectional[c("HernandezSegura", "SAUL_SEN_MAYO")],
                         Variable = "condition",
                         ColorVariable = "treatment_simple",
                         method = "logmedian",
                         xlab = "Cellular State", 
                         labsize = 8))
```

```{r fig.width=11, fig.height=5}
plt_filt <- ggarrange(plt_beforefilt,plt_afterfilt, nrow=1)
plt_norm <- ggarrange(plt_beforenorm,plt_afternorm, nrow=1)
plt_pca <- ggarrange(pltpca_1,pltpca_2, nrow=1)

plt1 <- ggarrange(plt_filt, plt_norm, nrow=1)
plt2 <- ggarrange(plt_pca, pltscores, nrow=1, widths=c(0.5,0.5))

plt_GSE297560 <- ggarrange(plt1, plt2, ncol=1, heights = c(0.4,0.7))
plt_GSE297560 <- annotate_figure(plt_GSE297560, top = text_grob("GSE297560: Space Radiation\n", face = "bold", size = 14))
plt_GSE297560


```

### GSE288940 (IMR90 + Eto)

My quantification pipeline 

#### Data

```{r}
# Get a list of the paths to the samples. Each sample is inside its own folder, under the name "quant.sf"
listfolders <- list.dirs("../data_aux/kallisto/Farr/")[-1] # remove "." directory
listfolders <- paste0(listfolders,"/abundance.tsv")
listfolders

# Naming the files (important so that when using tximport each column has the corresponding sample name)
samples <- gsub("../data_aux/kallisto/Farr//","",listfolders)
samples <- gsub("/abundance.tsv","",samples)
names(listfolders) <- samples
listfolders

# Merge all counts in an txi object. The argument txOut=TRUE allows to avoid gene-level summarization (we want the transcripts)
txi.kallisto <- tximport(listfolders, type = "kallisto", txOut=FALSE, tx2gene = t2gene2)

data <- as.data.frame(txi.kallisto$counts) 

```

```{r}
metadata <- as.data.frame(fread("../data_aux/GSE288940_metadata.csv"))
metadata <- metadata[,c("Run","Sample Name","cell_line","treatment")]
colnames(metadata)[1] <- "sampleID"


gsm_to_timepoint <- c(
  "GSM8779535" = "D0",
  "GSM8779536" = "D0",
  "GSM8779537" = "D0",
  "GSM8779538" = "D0",
  "GSM8779539" = "D0",
  "GSM8779540" = "D7",
  "GSM8779541" = "D7",
  "GSM8779542" = "D7",
  "GSM8779543" = "D7",
  "GSM8779544" = "D7",
  "GSM8779545" = "D7",
  "GSM8779546" = "D28",
  "GSM8779547" = "D28",
  "GSM8779548" = "D28",
  "GSM8779549" = "D28",
  "GSM8779550" = "D28",
  "GSM8779551" = "D28"
)

metadata$timepoint <- gsm_to_timepoint[metadata$`Sample Name`]
metadata$timepoint <- factor(metadata$timepoint, levels = c("D0", "D7", "D28"))

metadata$condition2 <- ifelse(grepl("D0", metadata$timepoint), "Proliferative", ifelse(grepl("D7", metadata$timepoint), "Prol > Sen", "Senescent"))

metadata$condition <- ifelse(grepl("D0", metadata$timepoint), "Proliferative",  "Senescent")

data <- data[,metadata$sampleID]
```


```{r}
data
metadata
```

#### Preprocessing


##### Filtering

```{r}
# Plot
(plt_beforefilt <- ggplot() + aes(x=log10(rowSums(data))) +
   geom_density(fill="grey", alpha=0.5) + theme_bw() + 
   theme(axis.text.x = element_text(angle = 45, hjust=1),
         plot.title = element_text(hjust = 0.5),
         plot.subtitle = element_text(hjust = 0.5)) +
   ggtitle("Before Filtering") +
   xlab(expression(log[10]~"(counts per gene)")) +
   ylab("Frequency") + 
   geom_vline(xintercept = 3.5, linetype="dotted",color = "black", size=1.5))

```


```{r}
filtdata <- data[rowSums(data) > 10^3.5,]
nrow(data)
nrow(filtdata)


cat("Number of genes before filtering:", nrow(data), "\n")
cat("Number of genes after filtering:", nrow(filtdata), "\n")

```


```{r}

# Plot
(plt_afterfilt <- ggplot() + aes(x=log10(rowSums(filtdata))) +
   geom_density(fill="grey", alpha=0.5) + theme_bw() + 
   theme(axis.text.x = element_text(angle = 45, hjust=1),
         plot.title = element_text(hjust = 0.5),
         plot.subtitle = element_text(hjust = 0.5)) +
   ggtitle("After Filtering") +
   xlab(expression(log[10]~"(counts per gene)")) +
   ylab("Frequency") )

```



##### Normalisation

```{r fig.width=6, fig.height=4}

filtcounts_ggplot <- stack(log2(filtdata))
colnames(filtcounts_ggplot) <- c("expression", "sampleID")
filtcounts_ggplot <- merge(filtcounts_ggplot,metadata, by="sampleID")


(plt_beforenorm <- ggplot(filtcounts_ggplot, aes(sampleID,expression)) + geom_boxplot(alpha=0.4)+ xlab("Samples") + ylab("log2(counts)")   +  theme(axis.text.x = element_text(angle = 45, hjust=1))+
    theme_bw()+ 
    theme(plot.title = element_text(hjust = 0.5),
          plot.subtitle = element_text(hjust = 0.5),
          axis.text.x=element_blank(),
          axis.ticks.x=element_blank()) +
    ggtitle("Before Normalisation") )

```


```{r}
dgelist_normcounts <- DGEList(filtdata) # the matrix with gene names as row names and counts in columns
dgelist_normcounts <- calcNormFactors(dgelist_normcounts, method = "TMM") 
normalised_counts <- as.data.frame(voom(dgelist_normcounts,plot=F)$E) 
#normalised_counts <- as.data.frame(edgeR::cpm(dgelist_normcounts, log = FALSE)) 
normdata <- 2^normalised_counts
lognormdata <- log2(normdata+1)
```


```{r fig.width=6, fig.height=4}

normcounts_ggplot <- stack(log2(normdata))
colnames(normcounts_ggplot) <- c("expression", "sampleID")
normcounts_ggplot <- merge(normcounts_ggplot,metadata, by="sampleID")


(plt_afternorm <- ggplot(normcounts_ggplot, aes(sampleID,expression)) + geom_boxplot(alpha=0.4)+ xlab("Samples") + ylab("log2(counts)")   +  theme(axis.text.x = element_text(angle = 45, hjust=1))+
    theme_bw()+ 
    theme(plot.title = element_text(hjust = 0.5),
          plot.subtitle = element_text(hjust = 0.5),
          axis.text.x=element_blank(),
          axis.ticks.x=element_blank()) +
    ggtitle("After Normalisation") )

```


##### PCA

```{r fig.width=6, fig.height=4}
y <- DGEList(lognormdata, samples= metadata)
PCA  <- prcomp(t(y$counts), scale=FALSE, center=TRUE)
PCACounts  <- PCA$x
PCACounts  <- as.data.frame(PCACounts )
PCAData <-  cbind(PCACounts[1:10],y$samples) 

ev = PCA$sdev^2 
pc1= round(100*ev[1]/sum(ev),2) 
pc2 = round(100*ev[2]/sum(ev),2) 
pc3 = round(100*ev[3]/sum(ev),2) 
pc4 = round(100*ev[4]/sum(ev),2) 
pc5 = round(100*ev[5]/sum(ev),2) 
pc6 = round(100*ev[6]/sum(ev),2)  

timepoint_colors <- c(
  "D0"  = "#AEC6E8",  # pastel blue
  "D7"  = "#C3A8D1",  # pastel purple
  "D28" = "#F4A6A0"   # pastel red/pink
)

(pltpca_1 <- ggplot(PCAData, aes(PC1,PC2))+
    geom_point(aes(color=timepoint), size=5, alpha=0.6)+ 
    xlab(paste0("PC1: ",pc1,"% variance")) +
    ylab(paste0("PC2: ",pc2,"% variance")) + 
    theme_bw()+
    guides(color = guide_legend(nrow = 1)) +
    theme(legend.position="top", legend.box="vertical", legend.title = element_blank(), legend.margin=margin())  + 
    geom_vline(xintercept=0, linetype="dotted") + 
    geom_hline(yintercept=0, linetype="dotted") +
    scale_color_manual(values=timepoint_colors))

(pltpca_2 <- ggplot(PCAData, aes(PC1,PC2))+
    geom_point(aes(color=condition), size=5, alpha=0.6)+ 
    xlab(paste0("PC1: ",pc1,"% variance")) +
    ylab(paste0("PC2: ",pc2,"% variance")) + 
    theme_bw()+
    theme(legend.position="top", legend.box="vertical", legend.title = element_blank(), legend.margin=margin())  + 
    geom_vline(xintercept=0, linetype="dotted") + 
    guides(color = guide_legend(nrow = 1)) +
    geom_hline(yintercept=0, linetype="dotted") +
    scale_color_manual(values=Condition_colors))


```

#### Calculate Scores

```{r}
(pltscores <- PlotScores(data = normdata, 
                         metadata = metadata, 
                         gene_sets = signatures_bidirectional[c("HernandezSegura", "SAUL_SEN_MAYO")],
                         Variable = "condition",
                         ColorVariable = "timepoint",
                         method = "logmedian",
                         ColorValues = timepoint_colors,
                         xlab = "Cellular State",
                         labsize = 8))
```

```{r fig.width=11, fig.height=5}
plt_filt <- ggarrange(plt_beforefilt,plt_afterfilt, nrow=1)
plt_norm <- ggarrange(plt_beforenorm,plt_afternorm, nrow=1)
plt_pca <- ggarrange(pltpca_1,pltpca_2, nrow=1)

plt1 <- ggarrange(plt_filt, plt_norm, nrow=1)
plt2 <- ggarrange(plt_pca, pltscores, nrow=1, widths=c(0.5,0.5))

plt_GSE288940 <- ggarrange(plt1, plt2, ncol=1, heights = c(0.4,0.7))
plt_GSE288940 <- annotate_figure(plt_GSE288940, top = text_grob("GSE288940: IMR90 + Etoposide\n", face = "bold", size = 14))
plt_GSE288940


```






### GSE287058 (Quiescent Fibroblasts)

counts tables were generated using HTSeq
Assembly: hg38

#### Data

```{r}

# List all files
files <- list.files(path = "../data_aux/GSE287058_RAW/", pattern = "\\.txt\\.gz$", full.names = TRUE)

# Read and combine
data <- lapply(files, function(f) {
  sample_name <- gsub("(GSM[0-9]+).*", "\\1", basename(f))
  df <- read.table(f, header = FALSE, col.names = c("gene", sample_name))
  df
}) %>%
  Reduce(function(x, y) full_join(x, y, by = "gene"), .) %>%
  tibble::column_to_rownames("gene")

data
```

```{r}
metadata <- as.data.frame(fread("../data_aux/GSE287058_metadata.csv"))
metadata <- metadata[,c("Run","Library Name","cell_line","treatment")]
colnames(metadata)[1] <- "sampleID"



# Reorder metadata to match column order of data
metadata <- metadata[match(colnames(data), metadata$`Library Name`), ]

# Verify order is correct before renaming
all(colnames(data) == metadata$`Library Name`)

# Then rename
colnames(data) <- metadata$sampleID

metadata$condition <- ifelse(grepl("Proliferating", metadata$treatment, ignore.case = TRUE), "Proliferative", "Quiescent")
metadata$condition2 <-  ifelse(metadata$condition == "Quiescent",
                               paste0(metadata$condition, "\n(", metadata$treatment, ")"),
                               metadata$condition)
metadata$condition2 <- gsub(" ","\n",metadata$condition2)

```

```{r}
data
metadata
```
Remove LF1 samples as they don't have controls

```{r}
metadata <- metadata[!grepl("LF1", metadata$cell_line), ]
data <- data[,metadata$sampleID]
```


```{r}
data
metadata
```


#### Preprocessing


##### Filtering

```{r}
# Plot
(plt_beforefilt <- ggplot() + aes(x=log10(rowSums(data))) +
   geom_density(fill="grey", alpha=0.5) + theme_bw() + 
   theme(axis.text.x = element_text(angle = 45, hjust=1),
         plot.title = element_text(hjust = 0.5),
         plot.subtitle = element_text(hjust = 0.5)) +
   ggtitle("Before Filtering") +
   xlab(expression(log[10]~"(counts per gene)")) +
   ylab("Frequency") + 
   geom_vline(xintercept = 3.2, linetype="dotted",color = "black", size=1.5))

```


```{r}
filtdata <- data[rowSums(data) > 10^3.2,]
nrow(data)
nrow(filtdata)


cat("Number of genes before filtering:", nrow(data), "\n")
cat("Number of genes after filtering:", nrow(filtdata), "\n")

```


```{r}

# Plot
(plt_afterfilt <- ggplot() + aes(x=log10(rowSums(filtdata))) +
   geom_density(fill="grey", alpha=0.5) + theme_bw() + 
   theme(axis.text.x = element_text(angle = 45, hjust=1),
         plot.title = element_text(hjust = 0.5),
         plot.subtitle = element_text(hjust = 0.5)) +
   ggtitle("After Filtering") +
   xlab(expression(log[10]~"(counts per gene)")) +
   ylab("Frequency") )

```



##### Normalisation

```{r fig.width=6, fig.height=4}

filtcounts_ggplot <- stack(log2(filtdata))
colnames(filtcounts_ggplot) <- c("expression", "sampleID")
filtcounts_ggplot <- merge(filtcounts_ggplot,metadata, by="sampleID")


(plt_beforenorm <- ggplot(filtcounts_ggplot, aes(sampleID,expression)) + geom_boxplot(alpha=0.4, outlier.size = 0.5)+ xlab("Samples") + ylab("log2(counts)")   +  theme(axis.text.x = element_text(angle = 45, hjust=1))+
    theme_bw()+ 
    theme(plot.title = element_text(hjust = 0.5),
          plot.subtitle = element_text(hjust = 0.5),
          axis.text.x=element_blank(),
          axis.ticks.x=element_blank()) +
    ggtitle("Before Normalisation") )

```


```{r}
dgelist_normcounts <- DGEList(filtdata) # the matrix with gene names as row names and counts in columns
dgelist_normcounts <- calcNormFactors(dgelist_normcounts, method = "TMM") 
normalised_counts <- as.data.frame(voom(dgelist_normcounts,plot=F)$E) 
#normalised_counts <- as.data.frame(edgeR::cpm(dgelist_normcounts, log = FALSE)) 
normdata <- 2^normalised_counts
lognormdata <- log2(normdata+1)
```


```{r fig.width=6, fig.height=4}

normcounts_ggplot <- stack(log2(normdata))
colnames(normcounts_ggplot) <- c("expression", "sampleID")
normcounts_ggplot <- merge(normcounts_ggplot,metadata, by="sampleID")


(plt_afternorm <- ggplot(normcounts_ggplot, aes(sampleID,expression)) + geom_boxplot(alpha=0.4, outlier.size = 0.5)+ xlab("Samples") + ylab("log2(counts)")   +  theme(axis.text.x = element_text(angle = 45, hjust=1))+
    theme_bw()+ 
    theme(plot.title = element_text(hjust = 0.5),
          plot.subtitle = element_text(hjust = 0.5),
          axis.text.x=element_blank(),
          axis.ticks.x=element_blank()) +
    ggtitle("After Normalisation") )

```


##### PCA

```{r fig.width=6, fig.height=4}
y <- DGEList(lognormdata, samples= metadata)
PCA  <- prcomp(t(y$counts), scale=FALSE, center=TRUE)
PCACounts  <- PCA$x
PCACounts  <- as.data.frame(PCACounts )
PCAData <-  cbind(PCACounts[1:10],y$samples) 

ev = PCA$sdev^2 
pc1= round(100*ev[1]/sum(ev),2) 
pc2 = round(100*ev[2]/sum(ev),2) 
pc3 = round(100*ev[3]/sum(ev),2) 
pc4 = round(100*ev[4]/sum(ev),2) 
pc5 = round(100*ev[5]/sum(ev),2) 
pc6 = round(100*ev[6]/sum(ev),2)  

cell_line_colors <- c(
  "BJ"                    = "#6DBF8A",  # medium pastel green
  "IMR90"                 = "#F47C7C",  # medium pastel red
  "Primary Preadipocytes" = "#F0C060"   # medium pastel yellow
)

(pltpca_1 <- ggplot(PCAData, aes(PC1,PC2))+
    geom_point(aes(color=cell_line), size=5, alpha=0.6)+ 
    xlab(paste0("PC1: ",pc1,"% variance")) +
    ylab(paste0("PC2: ",pc2,"% variance")) + 
    theme_bw()+
    guides(color = guide_legend(nrow = 3)) +
    theme(legend.position="top", legend.box="vertical", legend.title = element_blank(), legend.margin=margin())  + 
    geom_vline(xintercept=0, linetype="dotted") + 
    geom_hline(yintercept=0, linetype="dotted")+
    scale_color_manual(values=cell_line_colors))

(pltpca_2 <- ggplot(PCAData, aes(PC1,PC2))+
    geom_point(aes(color=condition), size=5, alpha=0.6)+ 
    xlab(paste0("PC1: ",pc1,"% variance")) +
    ylab(paste0("PC2: ",pc2,"% variance")) + 
    theme_bw()+
    theme(legend.position="top", legend.box="vertical", legend.title = element_blank(), legend.margin=margin())  + 
    geom_vline(xintercept=0, linetype="dotted") + 
    guides(color = guide_legend(nrow = 3)) +
    geom_hline(yintercept=0, linetype="dotted") +
    scale_color_manual(values=Condition_colors))


```

#### Calculate Scores

```{r}
(pltscores <- PlotScores(data = normdata, 
                         metadata = metadata, 
                         gene_sets = signatures_bidirectional[c("HernandezSegura", "SAUL_SEN_MAYO")],
                         Variable = "condition2",
                         ColorVariable = "cell_line",
                         method = "logmedian",
                         xlab = "Cellular State",
                         ColorValues = cell_line_colors,
                         labsize = 8))
```


```{r fig.width=11, fig.height=5}
plt_filt <- ggarrange(plt_beforefilt,plt_afterfilt, nrow=1)
plt_norm <- ggarrange(plt_beforenorm,plt_afternorm, nrow=1)
plt_pca <- ggarrange(pltpca_1,pltpca_2, nrow=1)

plt1 <- ggarrange(plt_filt, plt_norm, nrow=1)
plt2 <- ggarrange(plt_pca, pltscores, nrow=1, widths=c(0.5,0.5))

plt_GSE287058 <- ggarrange(plt1, plt2, ncol=1, heights = c(0.4,0.7))
plt_GSE287058 <- annotate_figure(plt_GSE287058, top = text_grob("GSE287058: Quiescent Fibroblasts\n", face = "bold", size = 14))
plt_GSE287058


```

### Compile All

```{r fig.width=11, fig.height=16} 
add_margin <- function(plt) plt + theme(plot.margin = margin(t=15, r=5, b=15, l=5))

(plt_3_3 <- ggarrange(add_margin(plt_GSE297560), 
                      add_margin(plt_GSE288940), 
                      add_margin(plt_GSE287058), 
                      ncol=1, labels = c("A","B","C"), 
                      heights = c(0.35,0.35,0.4)))

```

```{r}
ggplot2::ggsave("../Figures/Figs/R3_3_InVitroValidation.png",
                plt_3_3,
                width = 11, height=16, bg = 'white')

ggplot2::ggsave("../Figures/Figs/R3_3_InVitroValidation.pdf",
                plt_3_3,
                width = 11, height=16, bg = 'white')
```


## Comment 1.6 — Applicability of markeR beyond senescence

R.: Apply markeR in discovery mode to a small set of additional biological phenotypes with well-established gene sets and positive control datasets, to demonstrate that the framework generalises beyond senescence. Phenotypes considered include cell proliferation, epithelial-to-mesenchymal transition, and viral/interferon response.



### Sex Signature | GTEx

```{r}
library(biomaRt)
mart <- useMart(biomart="ensembl", dataset="hsapiens_gene_ensembl")
y_genes <- getBM(attributes = c("chromosome_name", "hgnc_symbol"),
                 filters = "chromosome_name", values = "Y", mart = mart)

x_genes <- getBM(attributes = c("chromosome_name", "hgnc_symbol"),
                 filters = "chromosome_name", values = "X", mart = mart)

y_unique <- y_genes[!y_genes$hgnc_symbol %in% x_genes$hgnc_symbol, ]
y_unique <- y_unique[y_unique$hgnc_symbol != "", ]

geneset_Ychromosome <- list(Y_Chromosome = y_unique$hgnc_symbol)
```


```{r}
#  
# base_url <- "https://raw.githubusercontent.com/DiseaseTranscriptomicsLab/voyAGEr/main/Corrected_Counts/"
# 
tissues <- c(
  "Adipose - Subcutaneous",
  "Adipose - Visceral (Omentum)",
  "Adrenal Gland",
  "Artery - Aorta",
  "Artery - Coronary",
  "Artery - Tibial",
  "Brain - Amygdala",
  "Brain - Anterior cingulate cortex (BA24)",
  "Brain - Caudate (basal ganglia)",
  "Brain - Cerebellar Hemisphere",
  "Brain - Cerebellum",
  "Brain - Cortex",
  "Brain - Frontal Cortex (BA9)",
  "Brain - Hippocampus",
  "Brain - Hypothalamus",
  "Brain - Nucleus accumbens (basal ganglia)",
  "Brain - Putamen (basal ganglia)",
  "Brain - Spinal cord (cervical c-1)",
  "Brain - Substantia nigra",
  "Breast - Mammary Tissue",
  "Cells - Cultured fibroblasts",
  "Cells - EBV-transformed lymphocytes",
  "Colon - Sigmoid",
  "Colon - Transverse",
  "Esophagus - Gastroesophageal Junction",
  "Esophagus - Mucosa",
  "Esophagus - Muscularis",
  "Heart - Atrial Appendage",
  "Heart - Left Ventricle",
  "Kidney - Cortex",
  "Liver",
  "Lung",
  "Minor Salivary Gland",
  "Muscle - Skeletal",
  "Nerve - Tibial",
  "Ovary",
  "Pancreas",
  "Pituitary",
  "Prostate",
  "Skin - Not Sun Exposed (Suprapubic)",
  "Skin - Sun Exposed (Lower leg)",
  "Small Intestine - Terminal Ileum",
  "Spleen",
  "Stomach",
  "Testis",
  "Thyroid",
  "Uterus",
  "Vagina",
  "Whole Blood"
)
# 
# gtex <- lapply(tissues, function(tissue) {
#   url <- paste0(base_url, URLencode(paste0(tissue, "_corrected_logCPM.csv.gz")))
#   message("Loading: ", tissue)
#   tryCatch({
#     df <- as.data.frame(fread(url, data.table = FALSE))
#     rownames(df) <- df[[1]]
#     df[[1]] <- NULL
#     df
#   }, error = function(e) { warning("Failed: ", tissue, " — ", e$message); NULL })
# })
# 
# names(gtex) <- tissues
# saveRDS(gtex, "../data_aux/data_gtex.rds")
```

```{r}
#GTEx_alltissues <- readRDS("../data_aux/GTExV8_voyagercorrected.rds") # https://github.com/DiseaseTranscriptomicsLab/voyAGEr/tree/main/Corrected_Counts
data_gtex <- readRDS("../data_aux/data_gtex.rds")

metadata_GTEx_alltissues <- readRDS("../data_aux/GTExV8_metadata.rds") # Restricted access
metadata_GTEx_alltissues$SEX <- ifelse(metadata_GTEx_alltissues$SEX == 1,"Male","Female")
metadata_GTEx_alltissues
```

```{r}
table(metadata_GTEx_alltissues$SMTSD,
      metadata_GTEx_alltissues$SEX)
```



```{r fig.width=11, fig.height=16}
# Assuming your metadata df is called `metadata` with columns SMTSD, sample_id, SEX (or similar)

# Step 1: Calculate scores for all tissues
scores_list <- lapply(tissues, function(tissue) {
  message("Scoring: ", tissue)
  tryCatch({
    expr       <- 2^data_gtex[[tissue]]
    meta_sub   <- metadata_GTEx_alltissues[metadata_GTEx_alltissues$SMTSD == tissue, ]
    # Keep only samples present in expression matrix
    meta_sub   <- meta_sub[meta_sub[[1]] %in% colnames(expr), ]
    #order, just in case
    expr <- expr[, meta_sub[[1]]]
    if (nrow(meta_sub) == 0) { warning("No matching samples: ", tissue); return(NULL) }
    
    CalculateScores(
      data      = expr,
      metadata  = meta_sub,
      method    = "logmedian",
      gene_sets = geneset_Ychromosome
    )
  }, error = function(e) { warning("Failed: ", tissue, " — ", e$message); NULL })
})
names(scores_list) <- tissues


# Step 2: Plot male vs female per tissue
# Define sex-specific tissues to skip
sex_specific <- c("Ovary", "Uterus", "Vagina", "Prostate", "Testis")

# Assumes scores df has a column with the score values and SEX in metadata
# Adjust score_col and sex_col to match your actual column names
score_col <- "logmedian"   # column name in CalculateScores output
sex_col   <- "SEX"         # column in metadata (e.g. 1=Male, 2=Female or "Male"/"Female")

plots_list <- lapply(tissues, function(tissue) {
  if (tissue %in% sex_specific) return(NULL)
  
  df <- scores_list[[tissue]][["Y_Chromosome"]]
  if (is.null(df)) return(NULL)
  
  # Merge scores with sex info if not already present
  if (!sex_col %in% colnames(df)) {
    meta_sub <- metadata_GTEx_alltissues[metadata_GTEx_alltissues$SMTSD == tissue, c("SAMPID","SMTSD","SEX","AGE")]
    df <- merge(df, meta_sub, by = 1)
  } else {
    df <- df[,c("sample","score","SMTSD","SEX","AGE")]
    
  }
  
  #df[[sex_col]] <- factor(df[[sex_col]], levels = c(1, 2), labels = c("Male", "Female"))
  
  ggplot(df, aes(x = SEX, y = score, fill = SEX)) +
    geom_jitter(color = "#DDDDDD", width = 0.15) +
    geom_violin(alpha = 0.4) +
    stat_summary(fun = median, geom = "crossbar", width = 0.1, color = "black") +
    scale_fill_manual(values = c("Male" = "#4E8ECC", "Female" = "#CC4E7A")) +
    labs(title = tissue, x = NULL,
         y = expression(log[2]*"-median score")) +
    theme_bw() +
    theme(legend.position  = "none",
          plot.title      = element_text(hjust = 0.5, size = 10))
})
names(plots_list) <- tissues
plots_list <- Filter(Negate(is.null), plots_list)

# Combine into one figure (optional)
ggarrange(plotlist = plots_list, ncol = 6, nrow = ceiling(length(plots_list) / 6))
```
```{r fig.width=4, fig.height=3}
df_all_sex <- bind_rows(
  lapply(tissues[!tissues %in% sex_specific], function(tissue) {
    df <- scores_list[[tissue]][["Y_Chromosome"]]
    if (is.null(df)) return(NULL)
    df$tissue <- tissue
    df
  })
)


(plt_ychromosomescores <- ggplot(df_all_sex, aes(x = .data[[sex_col]], y = score, fill = .data[[sex_col]])) + 
    geom_jitter(color = "#DDDDDD", width = 0.3, alpha = 0.3, size = 0.5) +
    geom_violin(alpha = 0.5) +
    stat_summary(fun = median, geom = "crossbar", width = 0.1, color = "black") +
    scale_fill_manual(values = c("Male" = "#4E8ECC", "Female" = "#CC4E7A")) +
    labs(x = NULL, y = expression(log[2]*"-median score")) +
    theme_bw() +
    theme(legend.position = "none",
          plot.title = element_text(hjust = 0.5, size = 14),
          text = element_text(size = 14))  +
    ggtitle("Y Chromosome Gene Set (biomaRt)\nAcross GTEx Tissues"))
```

```{r fig.width=4, fig.height=3}
# compute sample counts per group
n_labels <- df_all_sex %>%
  group_by(.data[[sex_col]]) %>%
  summarise(n = n()) %>%
  mutate(label = paste0(.data[[sex_col]], "\n(n=", n, ")"))

label_vec <- setNames(n_labels$label, n_labels[[sex_col]])

(plt_ychromosomescores <- ggplot(df_all_sex, aes(x = .data[[sex_col]], y = score, fill = .data[[sex_col]])) + 
    geom_jitter(color = "#DDDDDD", width = 0.3, alpha = 0.3, size = 0.5) +
    geom_violin(alpha = 0.5) +
    stat_summary(fun = median, geom = "crossbar", width = 0.1, color = "black") +
    scale_fill_manual(values = c("Male" = "#4E8ECC", "Female" = "#CC4E7A")) +
    scale_x_discrete(labels = label_vec) +
    labs(x = NULL, y = expression(log[2]*"-median score")) +
    theme_bw() +
    theme(legend.position = "none",
          plot.title = element_text(hjust = 0.5, size = 14),
          text = element_text(size = 14)) +
    ggtitle("Y Chromosome gene set (biomaRt)\nAcross GTEx Tissues"))
```


### Respiratory Infection Signature | GSE161731 (many viral and bacterial infections)


```{r}
signature_respiratoryviral <- list(Zaas2009_RespiratoryViralResponse = as.vector(fread("../data_aux/RespiratoryViralResponse_Zaas2009_simple.txt",header=F)$V1))
```

```{r}
data <- as.data.frame(fread("../data_aux/GSE161731_counts.csv"))
row.names(data) <- data$V1
data$V1 <- NULL
# Convert row names from ENSEMBL ID to Gene Name using biomaRt

# Connect to Ensembl
mart <- useMart("ensembl", dataset = "hsapiens_gene_ensembl")

# Query the conversion
gene_conversion <- getBM(
  attributes = c("ensembl_gene_id", "hgnc_symbol"),
  filters    = "ensembl_gene_id",
  values     = rownames(data),
  mart       = mart
)

# Replace rownames — keeping only genes with a match and a non-empty symbol
gene_conversion <- gene_conversion[!duplicated(gene_conversion$hgnc_symbol), ]
data <- data[gene_conversion$ensembl_gene_id, ]
rownames(data) <- gene_conversion$hgnc_symbol

data
```

```{r}
metadata <- as.data.frame(fread("../data_aux/GSE161731_counts_key.csv"))
colnames(metadata)[1] <- "sampleID"
metadata <- metadata[metadata$sampleID %in% colnames(data),]
data <- data[,metadata$sampleID]
```

```{r}
data
metadata
```


##### Filtering

```{r}
# Plot
(plt_beforefilt <- ggplot() + aes(x=log10(rowSums(data))) +
   geom_density(fill="grey", alpha=0.5) + theme_bw() + 
   theme(axis.text.x = element_text(angle = 45, hjust=1),
         plot.title = element_text(hjust = 0.5),
         plot.subtitle = element_text(hjust = 0.5)) +
   ggtitle("Before Filtering") +
   xlab(expression(log[10]~"(counts per gene)")) +
   ylab("Frequency") + 
   geom_vline(xintercept = 4.2, linetype="dotted",color = "black", size=1.5))

```


```{r}
filtdata <- data[rowSums(data) > 10^4.2,]
nrow(data)
nrow(filtdata)


cat("Number of genes before filtering:", nrow(data), "\n")
cat("Number of genes after filtering:", nrow(filtdata), "\n")

```


```{r}

# Plot
(plt_afterfilt <- ggplot() + aes(x=log10(rowSums(filtdata))) +
   geom_density(fill="grey", alpha=0.5) + theme_bw() + 
   theme(axis.text.x = element_text(angle = 45, hjust=1),
         plot.title = element_text(hjust = 0.5),
         plot.subtitle = element_text(hjust = 0.5)) +
   ggtitle("After Filtering") +
   xlab(expression(log[10]~"(counts per gene)")) +
   ylab("Frequency") )

```



##### Normalisation

```{r fig.width=6, fig.height=4}

filtcounts_ggplot <- stack(log2(filtdata))
colnames(filtcounts_ggplot) <- c("expression", "sampleID")
filtcounts_ggplot <- merge(filtcounts_ggplot,metadata, by="sampleID")


(plt_beforenorm <- ggplot(filtcounts_ggplot, aes(sampleID,expression)) + geom_boxplot(alpha=0.4, outlier.size = 0.5)+ xlab("Samples") + ylab("log2(counts)")   +  theme(axis.text.x = element_text(angle = 45, hjust=1))+
    theme_bw()+ 
    theme(plot.title = element_text(hjust = 0.5),
          plot.subtitle = element_text(hjust = 0.5),
          axis.text.x=element_blank(),
          axis.ticks.x=element_blank()) +
    ggtitle("Before Normalisation") )

```


```{r}
dgelist_normcounts <- DGEList(filtdata) # the matrix with gene names as row names and counts in columns
dgelist_normcounts <- calcNormFactors(dgelist_normcounts, method = "TMM") 
normalised_counts <- as.data.frame(voom(dgelist_normcounts,plot=F)$E) 
#normalised_counts <- as.data.frame(edgeR::cpm(dgelist_normcounts, log = FALSE)) 
normdata <- 2^normalised_counts
lognormdata <- log2(normdata+1)
```


```{r fig.width=6, fig.height=4}

normcounts_ggplot <- stack(log2(normdata))
colnames(normcounts_ggplot) <- c("expression", "sampleID")
normcounts_ggplot <- merge(normcounts_ggplot,metadata, by="sampleID")


(plt_afternorm <- ggplot(normcounts_ggplot, aes(sampleID,expression)) + geom_boxplot(alpha=0.4, outlier.size = 0.5)+ xlab("Samples") + ylab("log2(counts)")   +  theme(axis.text.x = element_text(angle = 45, hjust=1))+
    theme_bw()+ 
    theme(plot.title = element_text(hjust = 0.5),
          plot.subtitle = element_text(hjust = 0.5),
          axis.text.x=element_blank(),
          axis.ticks.x=element_blank()) +
    ggtitle("After Normalisation") )

```


##### PCA

```{r fig.width=6, fig.height=4}
y <- DGEList(lognormdata, samples= metadata)
PCA  <- prcomp(t(y$counts), scale=FALSE, center=TRUE)
PCACounts  <- PCA$x
PCACounts  <- as.data.frame(PCACounts )
PCAData <-  cbind(PCACounts[1:10],y$samples) 

ev = PCA$sdev^2 
pc1= round(100*ev[1]/sum(ev),2) 
pc2 = round(100*ev[2]/sum(ev),2) 
pc3 = round(100*ev[3]/sum(ev),2) 
pc4 = round(100*ev[4]/sum(ev),2) 
pc5 = round(100*ev[5]/sum(ev),2) 
pc6 = round(100*ev[6]/sum(ev),2)  


(pltpca_1 <- ggplot(PCAData, aes(PC1,PC2))+
    geom_point(aes(color=cohort), size=5, alpha=0.6)+ 
    xlab(paste0("PC1: ",pc1,"% variance")) +
    ylab(paste0("PC2: ",pc2,"% variance")) + 
    theme_bw()+
    guides(color = guide_legend(nrow = 2)) +
    theme(legend.position="top", legend.box="vertical", legend.title = element_blank(), legend.margin=margin())  + 
    geom_vline(xintercept=0, linetype="dotted") + 
    geom_hline(yintercept=0, linetype="dotted") )

(pltpca_2 <- ggplot(PCAData, aes(PC1,PC2))+
    geom_point(aes(color=gender), size=5, alpha=0.6)+ 
    xlab(paste0("PC1: ",pc1,"% variance")) +
    ylab(paste0("PC2: ",pc2,"% variance")) + 
    theme_bw()+
    theme(legend.position="top", legend.box="vertical", legend.title = element_blank(), legend.margin=margin())  + 
    geom_vline(xintercept=0, linetype="dotted") + 
    guides(color = guide_legend(nrow = 2)) +
    geom_hline(yintercept=0, linetype="dotted"))

(pltpca_3 <- ggplot(PCAData, aes(PC1,PC2))+
    geom_point(aes(color=race), size=5, alpha=0.6)+ 
    xlab(paste0("PC1: ",pc1,"% variance")) +
    ylab(paste0("PC2: ",pc2,"% variance")) + 
    theme_bw()+
    theme(legend.position="top", legend.box="vertical", legend.title = element_blank(), legend.margin=margin())  + 
    geom_vline(xintercept=0, linetype="dotted") + 
    guides(color = guide_legend(nrow = 2)) +
    geom_hline(yintercept=0, linetype="dotted"))

(pltpca_4 <- ggplot(PCAData, aes(PC1,PC2))+
    geom_point(aes(color=batch), size=5, alpha=0.6)+ 
    xlab(paste0("PC1: ",pc1,"% variance")) +
    ylab(paste0("PC2: ",pc2,"% variance")) + 
    theme_bw()+
    theme(legend.position="top", legend.box="vertical", legend.title = element_blank(), legend.margin=margin())  + 
    geom_vline(xintercept=0, linetype="dotted") + 
    guides(color = guide_legend(nrow = 2)) +
    geom_hline(yintercept=0, linetype="dotted"))


```

#### Calculate Scores


```{r}
#saveRDS(normdata,"../data_aux/data_RespiratoryInfection.rds")
data_RespiratoryInfection <- readRDS("../data_aux/data_RespiratoryInfection.rds")
#saveRDS(metadata,"../data_aux/metadata_RespiratoryInfection.rds")
metadata_ResporatoryInfection <- readRDS("../data_aux/metadata_RespiratoryInfection.rds")
```


```{r fig.width=4, fig.height=3}
(dfscores <- CalculateScores(data = data_ResporatoryInfection, 
                             metadata = metadata_ResporatoryInfection, 
                             gene_sets = signature_respiratoryviral,
                             method = "logmedian")$Zaas2009_RespiratoryViralResponse)

dfscores$cohort2 <- ifelse(dfscores$cohort %in% c("CoV other","COVID-19"), "CoV", dfscores$cohort)
dfscores$cohort2 <- gsub("healthy","Healthy",dfscores$cohort2)
cohort_colors <- c(
  "CoV"       = "#E07B54",
  "Influenza" = "#5B8DB8",
  "Bacterial" = "#6BAF92",
  "Healthy"   = "#B0B0B0"
)

(plt_viralrespiratory <- ggplot(dfscores, aes(x = cohort2, y = score, fill=cohort2)) + 
    geom_jitter(color = "black", width = 0.3, alpha = 0.4, size = 0.5) +
    geom_violin(alpha = 0.7) +
    stat_summary(fun = median, geom = "crossbar", width = 0.1, color = "black") + 
    labs(x = NULL, y = expression(log[2]*"-median score")) +
    theme_bw() +
    theme(legend.position="none",
          plot.title = element_text(hjust = 0.5, size = 14),
          text = element_text(size = 14)) +
    ggtitle("Zaas 2009 Respiratory Viral Gene Set \nin Whole Blood Across \nAcute Respiratory Infections") + scale_fill_manual(values = cohort_colors))
```
```{r fig.width=4, fig.height=3}
n_labels <- dfscores %>%
  group_by(cohort2) %>%
  summarise(n = n()) %>%
  mutate(label = paste0(cohort2, "\n(n=", n, ")"))

label_vec <- setNames(n_labels$label, n_labels$cohort2)

(plt_viralrespiratory <- ggplot(dfscores, aes(x = cohort2, y = score, fill = cohort2)) + 
    geom_jitter(color = "black", width = 0.3, alpha = 0.4, size = 0.5) +
    geom_violin(alpha = 0.7) +
    stat_summary(fun = median, geom = "crossbar", width = 0.1, color = "black") + 
    labs(x = NULL, y = expression(log[2]*"-median score")) +
    theme_bw() +
    theme(legend.position = "none",
          plot.title = element_text(hjust = 0.5, size = 14),
          text = element_text(size = 14)) +
    ggtitle("Respiratory Viral Response gene set \n(Zaas et al. 2009) in Whole Blood \nAcross Acute Respiratory Infections") +
    scale_fill_manual(values = cohort_colors) +
    scale_x_discrete(labels = label_vec))
```


<!-- ### EMT transition | KIRC -->

<!-- ```{r} -->
<!-- Ramaswamy_Metastasis <- list(Ramaswamy_Metastasis = as.data.frame(fread("../data_aux/RAMASWAMY_METASTASIS_simple.txt",header=T))) -->
<!-- ``` -->

<!-- https://xenabrowser.net/datapages/?dataset=TCGA-KIRC.star_counts.tsv&host=https%3A%2F%2Fgdc.xenahubs.net&removeHub=https%3A%2F%2Fxena.treehouse.gi.ucsc.edu%3A443 -->

<!-- KIRC was chosen given that it has a relatively well documented tumour grade -->

<!-- ```{r} -->
<!-- data <- as.data.frame(fread("../data_aux/TCGA-KIRC.star_counts.tsv")) -->
<!-- data$Ensembl_ID <- sub("\\..*", "", data$Ensembl_ID) -->

<!-- # combine rows with same gene id by summing them -->
<!-- data <- rowsum(data[, -1], group = data$Ensembl_ID) -->

<!-- # Connect to Ensembl -->
<!-- mart <- useMart("ensembl", dataset = "hsapiens_gene_ensembl") -->

<!-- # Query the conversion -->
<!-- gene_conversion <- getBM( -->
<!--   attributes = c("ensembl_gene_id", "hgnc_symbol"), -->
<!--   filters    = "ensembl_gene_id", -->
<!--   values     = rownames(data), -->
<!--   mart       = mart -->
<!-- ) -->

<!-- # Replace rownames — keeping only genes with a match and a non-empty symbol -->
<!-- gene_conversion <- gene_conversion[!duplicated(gene_conversion$hgnc_symbol), ] -->
<!-- data <- data[gene_conversion$ensembl_gene_id, ] -->
<!-- rownames(data) <- gene_conversion$hgnc_symbol -->

<!-- data <- 2^data - 1 # unlog... -->
<!-- data -->

<!-- ``` -->

<!-- ```{r} -->
<!-- metadata <- as.data.frame(fread("../data_aux/TCGA-KIRC.clinical.tsv")) -->
<!-- metadata <- metadata[metadata$sample %in% colnames(data),] -->
<!-- data <- data[,metadata$sample] -->
<!-- ``` -->


<!-- ```{r} -->
<!-- metadata$code <- sub(".*-", "", metadata$sample) -->
<!-- table(metadata$code) # no metastasis -->
<!-- table(metadata$tissue_type.samples) # normal vs tumour -->
<!-- table(metadata$sample_type.samples, metadata$ajcc_pathologic_m.diagnoses) -->

<!-- # subset data to M1 or M0 only -->
<!-- # subset data to primary tumour only -->
<!-- metadata <- metadata[metadata$ajcc_pathologic_m.diagnoses %in% c("M0","M1") & metadata$sample_type.samples != "metadata$sample_type.samples" & metadata$tissue_type.samples == "Tumor",] -->

<!-- table(metadata$code) # no metastasis -->
<!-- table(metadata$tissue_type.samples) # normal vs tumour -->
<!-- table(metadata$sample_type.samples, metadata$ajcc_pathologic_m.diagnoses) -->

<!-- data <- data[,metadata$sample] -->

<!-- colnames(metadata)[1] <- "sampleID" -->
<!-- ``` -->


<!-- ```{r} -->

<!-- data -->
<!-- metadata -->
<!-- ``` -->




<!-- ##### Filtering -->

<!-- ```{r} -->
<!-- # Plot -->
<!-- (plt_beforefilt <- ggplot() + aes(x=log10(rowSums(data))) + -->
<!--    geom_density(fill="grey", alpha=0.5) + theme_bw() +  -->
<!--    theme(axis.text.x = element_text(angle = 45, hjust=1), -->
<!--          plot.title = element_text(hjust = 0.5), -->
<!--          plot.subtitle = element_text(hjust = 0.5)) + -->
<!--    ggtitle("Before Filtering") + -->
<!--    xlab(expression(log[10]~"(counts per gene)")) + -->
<!--    ylab("Frequency") +  -->
<!--    geom_vline(xintercept = 5.2, linetype="dotted",color = "black", size=1.5)) -->

<!-- ``` -->


<!-- ```{r} -->
<!-- filtdata <- data[rowSums(data) > 10^5.2,] -->
<!-- nrow(data) -->
<!-- nrow(filtdata) -->


<!-- cat("Number of genes before filtering:", nrow(data), "\n") -->
<!-- cat("Number of genes after filtering:", nrow(filtdata), "\n") -->

<!-- ``` -->


<!-- ```{r} -->

<!-- # Plot -->
<!-- (plt_afterfilt <- ggplot() + aes(x=log10(rowSums(filtdata))) + -->
<!--    geom_density(fill="grey", alpha=0.5) + theme_bw() +  -->
<!--    theme(axis.text.x = element_text(angle = 45, hjust=1), -->
<!--          plot.title = element_text(hjust = 0.5), -->
<!--          plot.subtitle = element_text(hjust = 0.5)) + -->
<!--    ggtitle("After Filtering") + -->
<!--    xlab(expression(log[10]~"(counts per gene)")) + -->
<!--    ylab("Frequency") ) -->

<!-- ``` -->



<!-- ##### Normalisation -->

<!-- ```{r fig.width=6, fig.height=4} -->

<!-- # filtcounts_ggplot <- stack(log2(filtdata)) -->
<!-- # colnames(filtcounts_ggplot) <- c("expression", "sampleID") -->
<!-- # filtcounts_ggplot <- merge(filtcounts_ggplot,metadata, by="sampleID") -->
<!-- #    -->


<!-- total_coverage <- colSums(filtdata) -->
<!-- top50 <- names(sort(total_coverage, decreasing = TRUE)[1:50]) -->
<!-- bot50 <- names(sort(total_coverage)[1:50]) -->
<!-- selected_samples <- c(bot50, top50) -->

<!-- filtdata_sub <- filtdata[, selected_samples] -->

<!-- filtcounts_ggplot <- stack(log2(filtdata_sub)) -->
<!-- colnames(filtcounts_ggplot) <- c("expression", "sampleID") -->
<!-- filtcounts_ggplot <- merge(filtcounts_ggplot, metadata, by = "sampleID") -->

<!-- # order by total coverage -->
<!-- filtcounts_ggplot$sampleID <- factor(filtcounts_ggplot$sampleID, -->
<!--                                      levels = names(sort(total_coverage[selected_samples]))) -->

<!-- (plt_beforenorm <- ggplot(filtcounts_ggplot, aes(sampleID,expression)) + geom_boxplot(alpha=0.4, outlier.size = 0.5)+ xlab("Samples") + ylab("log2(counts)")   +  theme(axis.text.x = element_text(angle = 45, hjust=1))+ -->
<!--     theme_bw()+  -->
<!--     theme(plot.title = element_text(hjust = 0.5), -->
<!--           plot.subtitle = element_text(hjust = 0.5), -->
<!--           axis.text.x=element_blank(), -->
<!--           axis.ticks.x=element_blank()) + -->
<!--     ggtitle("Before Normalisation") ) -->

<!-- ``` -->


<!-- ```{r} -->
<!-- dgelist_normcounts <- DGEList(filtdata) # the matrix with gene names as row names and counts in columns -->
<!-- dgelist_normcounts <- calcNormFactors(dgelist_normcounts, method = "TMM")  -->
<!-- normalised_counts <- as.data.frame(voom(dgelist_normcounts,plot=F)$E)  -->
<!-- #normalised_counts <- as.data.frame(edgeR::cpm(dgelist_normcounts, log = FALSE))  -->
<!-- normdata <- 2^normalised_counts -->
<!-- lognormdata <- log2(normdata+1) -->
<!-- ``` -->



<!-- ```{r fig.width=6, fig.height=4} -->

<!-- total_coverage <- colSums(normdata) -->
<!-- top50 <- names(sort(total_coverage, decreasing = TRUE)[1:50]) -->
<!-- bot50 <- names(sort(total_coverage)[1:50]) -->
<!-- selected_samples <- c(bot50, top50) -->

<!-- normdata_sub <- normdata[, selected_samples] -->

<!-- normcounts_ggplot <- stack(log2(normdata_sub)) -->
<!-- colnames(normcounts_ggplot) <- c("expression", "sampleID") -->
<!-- normcounts_ggplot <- merge(normcounts_ggplot, metadata, by = "sampleID") -->

<!-- # order by total coverage -->
<!-- normcounts_ggplot$sampleID <- factor(normcounts_ggplot$sampleID, -->
<!--                                      levels = names(sort(total_coverage[selected_samples]))) -->

<!-- (plt_afternorm <- ggplot(normcounts_ggplot, aes(sampleID,expression)) + geom_boxplot(alpha=0.4, outlier.size = 0.5)+ xlab("Samples") + ylab("log2(counts)")   +  theme(axis.text.x = element_text(angle = 45, hjust=1))+ -->
<!--     theme_bw()+  -->
<!--     theme(plot.title = element_text(hjust = 0.5), -->
<!--           plot.subtitle = element_text(hjust = 0.5), -->
<!--           axis.text.x=element_blank(), -->
<!--           axis.ticks.x=element_blank()) + -->
<!--     ggtitle("After Normalisation") ) -->

<!-- ``` -->


<!-- #### Calculate Scores -->


<!-- ```{r fig.width=4, fig.height=3} -->
<!-- (dfscores <- CalculateScores(data = normdata,  -->
<!--                              metadata = metadata,  -->
<!--                              gene_sets = Ramaswamy_Metastasis, -->
<!--                              method = "logmedian")$Ramaswamy_Metastasis) -->

<!-- met_colors <- c( -->
<!--   "M0"       = "#E07B54", -->
<!--   "M1" = "#5B8DB8"  -->
<!-- ) -->



<!-- (plt_emt <- ggplot(dfscores, aes(x = ajcc_pathologic_m.diagnoses, y = score)) +  -->
<!--     geom_jitter(aes(color=ajcc_pathologic_m.diagnoses), width = 0.3, alpha = 1, size = 2) + -->
<!--     geom_violin(alpha = 0.5) + -->
<!--     stat_summary(fun = median, geom = "crossbar", width = 0.1, color = "black") +  -->
<!--     labs(x = NULL, y = expression(log[2]*"-median score")) + -->
<!--     theme_bw() + -->
<!--     theme(legend.position="none", -->
<!--           plot.title = element_text(hjust = 0.5, size = 14), -->
<!--           text = element_text(size = 14)) + -->
<!--     ggtitle("Ramaswamy EMT transition Gene Set\n in KIRC Primary Tumour Samples") + -->
<!--     scale_color_manual(values = met_colors)) -->
<!-- ``` -->


### Proliferation | PANCAN

meta-PCNA (Venet et al. 2011)
```{r}
metaPCNA <- list(metaPCNA = as.vector(fread("../data_aux/proliferationsignature_metaPCNA.txt",header=F))$V1)
```

https://xenabrowser.net/datapages/?dataset=EB%2B%2BAdjustPANCAN_IlluminaHiSeq_RNASeqV2.geneExp.xena&host=https%3A%2F%2Fpancanatlas.xenahubs.net&removeHub=https%3A%2F%2Fxena.treehouse.gi.ucsc.edu%3A443

```{r}
data <- as.data.frame(fread("../data_aux/EB++AdjustPANCAN_IlluminaHiSeq_RNASeqV2.geneExp.xena.csv"))
# combine rows with same gene id by summing them
data <- rowsum(data[, -1], group = data$sample)

data <- 2^data - 1 # unlog...
data

```

```{r}
metadata <- as.data.frame(fread("../data_aux/Survival_SupplementalTable_S1_20171025_xena_sp.tsv"))
metadata <- metadata[metadata$sample %in% colnames(data),]
data <- data[,metadata$sample]
```


```{r}
metadata$code <- sub(".*-", "", metadata$sample) 
colnames(metadata)[1] <- "sampleID"
metadata$sample_type <- ifelse(metadata$code == "11", "Normal", "Tumour")
```


```{r}
data
metadata
```


```{r}
data <- na.omit(data)
data <- data + 1 # some negative values, must be of batch correction
```


##### Filtering

No need for filtering!! Distributions look good, data was processed & batch corrected before

```{r}
# Plot
(plt_beforefilt <- ggplot() + aes(x=log10(rowSums(data))) +
   geom_density(fill="grey", alpha=0.5) + theme_bw() + 
   theme(axis.text.x = element_text(angle = 45, hjust=1),
         plot.title = element_text(hjust = 0.5),
         plot.subtitle = element_text(hjust = 0.5)) +
   ggtitle("Before Filtering") +
   xlab(expression(log[10]~"(counts per gene)")) +
   ylab("Frequency") + 
   geom_vline(xintercept = 5.2, linetype="dotted",color = "black", size=1.5))

```


<!-- ```{r} -->
<!-- #filtdata <- data[rowSums(data) > 10^5.2,] -->
<!-- filtdata <- data -->
<!-- nrow(data) -->
<!-- nrow(filtdata) -->


<!-- cat("Number of genes before filtering:", nrow(data), "\n") -->
<!-- cat("Number of genes after filtering:", nrow(filtdata), "\n") -->

<!-- ``` -->


<!-- ```{r} -->

<!-- # Plot -->
<!-- (plt_afterfilt <- ggplot() + aes(x=log10(rowSums(filtdata))) + -->
<!--   geom_density(fill="grey", alpha=0.5) + theme_bw() +  -->
<!--     theme(axis.text.x = element_text(angle = 45, hjust=1), -->
<!--         plot.title = element_text(hjust = 0.5), -->
<!--         plot.subtitle = element_text(hjust = 0.5)) + -->
<!--   ggtitle("After Filtering") + -->
<!--   xlab(expression(log[10]~"(counts per gene)")) + -->
<!--   ylab("Frequency") ) -->

<!-- ``` -->



<!-- ##### Normalisation -->

<!-- ```{r fig.width=6, fig.height=4} -->

<!-- # filtcounts_ggplot <- stack(log2(filtdata)) -->
<!-- # colnames(filtcounts_ggplot) <- c("expression", "sampleID") -->
<!-- # filtcounts_ggplot <- merge(filtcounts_ggplot,metadata, by="sampleID") -->
<!-- #    -->


<!-- total_coverage <- colSums(filtdata) -->
<!-- top50 <- names(sort(total_coverage, decreasing = TRUE)[1:50]) -->
<!-- bot50 <- names(sort(total_coverage)[1:50]) -->
<!-- selected_samples <- c(bot50, top50) -->

<!-- filtdata_sub <- filtdata[, selected_samples] -->

<!-- filtcounts_ggplot <- stack(log2(filtdata_sub)) -->
<!-- colnames(filtcounts_ggplot) <- c("expression", "sampleID") -->
<!-- filtcounts_ggplot <- merge(filtcounts_ggplot, metadata, by = "sampleID") -->

<!-- # order by total coverage -->
<!-- filtcounts_ggplot$sampleID <- factor(filtcounts_ggplot$sampleID, -->
<!--                                       levels = names(sort(total_coverage[selected_samples]))) -->

<!-- (plt_beforenorm <- ggplot(filtcounts_ggplot, aes(sampleID,expression)) + geom_boxplot(alpha=0.4, outlier.size = 0.5)+ xlab("Samples") + ylab("log2(counts)")   +  theme(axis.text.x = element_text(angle = 45, hjust=1))+ -->
<!--   theme_bw()+  -->
<!--   theme(plot.title = element_text(hjust = 0.5), -->
<!--         plot.subtitle = element_text(hjust = 0.5), -->
<!--         axis.text.x=element_blank(), -->
<!--         axis.ticks.x=element_blank()) + -->
<!--   ggtitle("Before Normalisation") ) -->

<!-- ``` -->


<!-- ```{r} -->
<!-- # dgelist_normcounts <- DGEList(filtdata) # the matrix with gene names as row names and counts in columns -->
<!-- # dgelist_normcounts <- calcNormFactors(dgelist_normcounts, method = "TMM")  -->
<!-- # normalised_counts <- as.data.frame(voom(dgelist_normcounts,plot=F)$E)  -->
<!-- # #normalised_counts <- as.data.frame(edgeR::cpm(dgelist_normcounts, log = FALSE))  -->
<!-- # normdata <- 2^normalised_counts -->
<!-- # lognormdata <- log2(normdata+1) -->
<!-- ``` -->
<!-- ```{r} -->
<!-- normdata <- filtdata -->
<!-- ``` -->



<!-- ```{r fig.width=6, fig.height=4} -->

<!-- total_coverage <- colSums(normdata) -->
<!-- top50 <- names(sort(total_coverage, decreasing = TRUE)[1:50]) -->
<!-- bot50 <- names(sort(total_coverage)[1:50]) -->
<!-- selected_samples <- c(bot50, top50) -->

<!-- normdata_sub <- normdata[, selected_samples] -->

<!-- normcounts_ggplot <- stack(log2(normdata_sub)) -->
<!-- colnames(normcounts_ggplot) <- c("expression", "sampleID") -->
<!-- normcounts_ggplot <- merge(normcounts_ggplot, metadata, by = "sampleID") -->

<!-- # order by total coverage -->
<!-- normcounts_ggplot$sampleID <- factor(normcounts_ggplot$sampleID, -->
<!--                                       levels = names(sort(total_coverage[selected_samples]))) -->

<!-- (plt_afternorm <- ggplot(normcounts_ggplot, aes(sampleID,expression)) + geom_boxplot(alpha=0.4, outlier.size = 0.5)+ xlab("Samples") + ylab("log2(counts)")   +  theme(axis.text.x = element_text(angle = 45, hjust=1))+ -->
<!--   theme_bw()+  -->
<!--   theme(plot.title = element_text(hjust = 0.5), -->
<!--         plot.subtitle = element_text(hjust = 0.5), -->
<!--         axis.text.x=element_blank(), -->
<!--         axis.ticks.x=element_blank()) + -->
<!--   ggtitle("After Normalisation") ) -->

<!-- ``` -->


#### Calculate Scores

```{r}
#saveRDS(data,"../data_aux/data_PANCAN.rds")
data_PANCAN <- readRDS("../data_aux/data_PANCAN.rds")
#saveRDS(metadata,"../data_aux/metadata_PANCAN.rds")
metadata_PANCAN <- readRDS("../data_aux/metadata_PANCAN.rds")
```

```{r fig.width=4, fig.height=3}
(dfscores <- CalculateScores(data = data_PANCAN, 
                             metadata = metadata_PANCAN, 
                             gene_sets = metaPCNA,
                             method = "logmedian")$metaPCNA)

samptype_col <- c(
  "Tumour" = "#C0392B",
  "Normal" = "#2E86AB"
)


(plt_prol <- ggplot(dfscores, aes(x = sample_type, y = score, fill=sample_type)) + 
    geom_jitter(color = "#DDDDDD", width = 0.3, alpha = 0.3, size = 0.5) +
    geom_violin(alpha = 0.5) +
    stat_summary(fun = median, geom = "crossbar", width = 0.1, color = "black") + 
    labs(x = NULL, y = expression(log[2]*"-median score")) +
    theme_bw() +
    theme(legend.position="none",
          plot.title = element_text(hjust = 0.5, size = 14),
          text = element_text(size = 14)) +
    ggtitle("Venet 2011 Proliferation Gene Set\n in TCGA PANCAN (Liu 2018)") +
    scale_fill_manual(values = samptype_col))
```
```{r fig.width=4, fig.height=3}
n_labels <- dfscores %>%
  group_by(sample_type) %>%
  summarise(n = n()) %>%
  mutate(label = paste0(sample_type, "\n(n=", n, ")"))

label_vec <- setNames(n_labels$label, n_labels$sample_type)

(plt_prol <- ggplot(dfscores, aes(x = sample_type, y = score, fill = sample_type)) + 
    geom_jitter(color = "#DDDDDD", width = 0.3, alpha = 0.3, size = 0.5) +
    geom_violin(alpha = 0.5) +
    stat_summary(fun = median, geom = "crossbar", width = 0.1, color = "black") + 
    labs(x = NULL, y = expression(log[2]*"-median score")) +
    theme_bw() +
    theme(legend.position = "none",
          plot.title = element_text(hjust = 0.5, size = 14),
          text = element_text(size = 14)) +
    ggtitle("meta-PCNA proliferation \ngene set (Venet et al. 2011)\n in TCGA PANCAN (Liu 2018)") +
    scale_fill_manual(values = samptype_col) +
    scale_x_discrete(labels = label_vec))
```
### Yeast Experiments

```{r}
library(neat)
data(yeast)

# Inspect
length(yeast$esr1)  # rESR (repressed)
length(yeast$esr2)  # iESR (induced)

GeneSet_YeastESR <- list(Gasch2000_ESR = data.frame(gene=c(yeast$esr1,yeast$esr2),
                                                    direction=c(rep(-1,length(yeast$esr1)), rep(1,length(yeast$esr2)))))

```

```{r}
data_yeast <- as.data.frame(fread("../data_aux/GSE135430_Yeast_HeatShock.csv"))
row.names(data_yeast) <- data_yeast$gene_id
data_yeast_o <- data_yeast[,-c(1,2,3)]

data_yeast <- apply(data_yeast_o, 2, function(x) as.numeric(gsub(",", ".", x)))
rownames(data_yeast) <- rownames(data_yeast_o)
data_yeast <- as.data.frame(data_yeast)
data_yeast
```

```{r}
metadata_yeast <- as.data.frame(fread("../data_aux/GSE135430_metadata_yeast.csv"))

colname_to_gsm <- c(
  "Phenol t0 Rep1"  = "GSM4008796",
  "Phenol t0 Rep2"  = "GSM4008797",
  "Phenol t0 Rep3"  = "GSM4008798",
  "Phenol t0 Rep4"  = "GSM4008799",
  "Phenol t20 Rep1" = "GSM4008800",
  "Phenol t20 Rep2" = "GSM4008801",
  "Phenol t20 Rep3" = "GSM4008802",
  "Phenol t20 Rep4" = "GSM4008803",
  "RNeasy t0 Rep1"  = "GSM4008804",
  "RNeasy t0 Rep2"  = "GSM4008805",
  "RNeasy t0 Rep3"  = "GSM4008806",
  "RNeasy t0 Rep4"  = "GSM4008807",
  "RNeasy t20 Rep1" = "GSM4008808",
  "RNeasy t20 Rep2" = "GSM4008809",
  "RNeasy t20 Rep3" = "GSM4008810",
  "RNeasy t20 Rep4" = "GSM4008811",
  "Trizol t0 Rep1"  = "GSM4008812",
  "Trizol t0 Rep2"  = "GSM4008813",
  "Trizol t0 Rep3"  = "GSM4008814",
  "Trizol t0 Rep4"  = "GSM4008815",
  "Trizol t20 Rep1" = "GSM4008816",
  "Trizol t20 Rep2" = "GSM4008817",
  "Trizol t20 Rep3" = "GSM4008818",
  "Trizol t20 Rep4" = "GSM4008819"
)

metadata_yeast_2 <- data.frame(
  sample     = names(colname_to_gsm),
  gsm = colname_to_gsm,
  condition  = ifelse(grepl("t20", names(colname_to_gsm)), "HeatShock", "Unstressed"),
  rna_method = sub(" t[0-9]+ Rep[0-9]+", "", names(colname_to_gsm)),
  row.names  = names(colname_to_gsm)
)

metadata_yeast <- merge(metadata_yeast_2,metadata_yeast, by.y="GEO_Accession (exp)", by.x="gsm")
metadata_yeast <- metadata_yeast[,c("sample",colnames(metadata_yeast))]
metadata_yeast$sample.1 <- NULL
metadata_yeast$condition <- gsub("HeatShock","Heat Shock", metadata_yeast$condition)
```


```{r}
metadata_yeast <- metadata_yeast[metadata_yeast$sample %in% colnames(data_yeast)]
data_yeast <- data_yeast[,metadata_yeast$sample]

```

```{r}
data_yeast
metadata_yeast
```

##### Filtering


```{r}
# Plot
(plt_beforefilt <- ggplot() + aes(x=log10(rowSums(data_yeast))) +
   geom_density(fill="grey", alpha=0.5) + theme_bw() + 
   theme(axis.text.x = element_text(angle = 45, hjust=1),
         plot.title = element_text(hjust = 0.5),
         plot.subtitle = element_text(hjust = 0.5)) +
   ggtitle("Before Filtering") +
   xlab(expression(log[10]~"(counts per gene)")) +
   ylab("Frequency") + 
   geom_vline(xintercept = 3.4, linetype="dotted",color = "black", size=1.5))

```


```{r}
filtdata <- data_yeast[rowSums(data_yeast) > 10^3.4,]
nrow(data_yeast)
nrow(filtdata)


cat("Number of genes before filtering:", nrow(data_yeast), "\n")
cat("Number of genes after filtering:", nrow(filtdata), "\n")

```


```{r}

# Plot
(plt_afterfilt <- ggplot() + aes(x=log10(rowSums(filtdata))) +
   geom_density(fill="grey", alpha=0.5) + theme_bw() + 
   theme(axis.text.x = element_text(angle = 45, hjust=1),
         plot.title = element_text(hjust = 0.5),
         plot.subtitle = element_text(hjust = 0.5)) +
   ggtitle("After Filtering") +
   xlab(expression(log[10]~"(counts per gene)")) +
   ylab("Frequency") )

```


##### Normalisation

```{r fig.width=6, fig.height=4}

filtcounts_ggplot <- stack(log2(filtdata))
colnames(filtcounts_ggplot) <- c("expression", "sample")
filtcounts_ggplot <- merge(filtcounts_ggplot,metadata_yeast, by="sample")


(plt_beforenorm <- ggplot(filtcounts_ggplot, aes(sample,expression)) + geom_boxplot(alpha=0.4, outlier.size = 0.5)+ xlab("Samples") + ylab("log2(counts)")   +  theme(axis.text.x = element_text(angle = 45, hjust=1))+
    theme_bw()+ 
    theme(plot.title = element_text(hjust = 0.5),
          plot.subtitle = element_text(hjust = 0.5),
          axis.text.x=element_blank(),
          axis.ticks.x=element_blank()) +
    ggtitle("Before Normalisation") )

```


```{r}
dgelist_normcounts <- DGEList(filtdata) # the matrix with gene names as row names and counts in columns
dgelist_normcounts <- calcNormFactors(dgelist_normcounts, method = "TMM") 
normalised_counts <- as.data.frame(voom(dgelist_normcounts,plot=F)$E) 
#normalised_counts <- as.data.frame(edgeR::cpm(dgelist_normcounts, log = FALSE)) 
normdata <- 2^normalised_counts
lognormdata <- log2(normdata+1)
```


```{r fig.width=6, fig.height=4}

normcounts_ggplot <- stack(log2(normdata))
colnames(normcounts_ggplot) <- c("expression", "sample")
normcounts_ggplot <- merge(normcounts_ggplot,metadata_yeast, by="sample")


(plt_afternorm <- ggplot(normcounts_ggplot, aes(sample,expression)) + geom_boxplot(alpha=0.4, outlier.size = 0.5)+ xlab("Samples") + ylab("log2(counts)")   +  theme(axis.text.x = element_text(angle = 45, hjust=1))+
    theme_bw()+ 
    theme(plot.title = element_text(hjust = 0.5),
          plot.subtitle = element_text(hjust = 0.5),
          axis.text.x=element_blank(),
          axis.ticks.x=element_blank()) +
    ggtitle("After Normalisation") )

```


##### PCA

```{r fig.width=6, fig.height=4}
y <- DGEList(lognormdata, samples= metadata_yeast)
PCA  <- prcomp(t(y$counts), scale=FALSE, center=TRUE)
PCACounts  <- PCA$x
PCACounts  <- as.data.frame(PCACounts )
PCAData <-  cbind(PCACounts[1:10],y$samples) 

ev = PCA$sdev^2 
pc1= round(100*ev[1]/sum(ev),2) 
pc2 = round(100*ev[2]/sum(ev),2) 
pc3 = round(100*ev[3]/sum(ev),2) 
pc4 = round(100*ev[4]/sum(ev),2) 
pc5 = round(100*ev[5]/sum(ev),2) 
pc6 = round(100*ev[6]/sum(ev),2)  


(pltpca_1 <- ggplot(PCAData, aes(PC1,PC2))+
    geom_point(aes(color=condition), size=5, alpha=0.6)+ 
    xlab(paste0("PC1: ",pc1,"% variance")) +
    ylab(paste0("PC2: ",pc2,"% variance")) + 
    theme_bw()+
    guides(color = guide_legend(nrow = 2)) +
    theme(legend.position="top", legend.box="vertical", legend.title = element_blank(), legend.margin=margin())  + 
    geom_vline(xintercept=0, linetype="dotted") + 
    geom_hline(yintercept=0, linetype="dotted") )

(pltpca_2 <- ggplot(PCAData, aes(PC1,PC2))+
    geom_point(aes(color=rna_method), size=5, alpha=0.6)+ 
    xlab(paste0("PC1: ",pc1,"% variance")) +
    ylab(paste0("PC2: ",pc2,"% variance")) + 
    theme_bw()+
    theme(legend.position="top", legend.box="vertical", legend.title = element_blank(), legend.margin=margin())  + 
    geom_vline(xintercept=0, linetype="dotted") + 
    guides(color = guide_legend(nrow = 2)) +
    geom_hline(yintercept=0, linetype="dotted"))


```

#### Calculate Scores

```{r fig.width=4, fig.height=3}
(dfscores <- CalculateScores(data = normdata, 
                             metadata = metadata_yeast, 
                             gene_sets = GeneSet_YeastESR,
                             method = "logmedian")$Gasch2000_ESR)

condition_colors <- c(
  "Unstressed" = "#F5C842",  # warm yellow
  "Heat Shock"  = "#C0392B"   # red
)

(plt_yeast_heatshock <- ggplot(dfscores, aes(x = condition, y = score, fill=condition)) + 
    geom_jitter(color = "black", width = 0.3, alpha = 0.4, size = 0.5) +    geom_violin(alpha = 0.7) +
    stat_summary(fun = median, geom = "crossbar", width = 0.1, color = "black") + 
    labs(x = NULL, y = expression(log[2]*"-median score")) +
    theme_bw() +
    theme(legend.position="none",
          plot.title = element_text(hjust = 0.5, size = 14),
          text = element_text(size = 14)) +
    ggtitle("Environmental Stress Response (ESR) \ngene set (Gasch et al. 2000) \nin S. cerevisiae Heat Shock") + scale_fill_manual(values = condition_colors))
```
```{r}
# Compute label vector with counts
label_vec <- setNames(
  paste0(levels(factor(dfscores$condition)), "\n(n=", table(dfscores$condition)[levels(factor(dfscores$condition))], ")"),
  levels(factor(dfscores$condition))
)

(plt_yeast_heatshock <- ggplot(dfscores, aes(x = condition, y = score, fill = condition)) + 
    geom_jitter(color = "black", width = 0.3, alpha = 0.4, size = 0.5) +
    geom_violin(alpha = 0.7) +
    stat_summary(fun = median, geom = "crossbar", width = 0.1, color = "black") + 
    labs(x = NULL, y = expression(log[2]*"-median score")) +
    scale_x_discrete(labels = label_vec) +
    theme_bw() +
    theme(legend.position = "none",
          plot.title = element_text(hjust = 0.5, size = 14),
          text = element_text(size = 14)) +
    ggtitle("Environmental Stress Response (ESR) \ngene set (Gasch et al. 2000) \nin S. cerevisiae Heat Shock")  +
    scale_fill_manual(values = condition_colors))
```



### Combine all

```{r fig.width=8, fig.height=7}
(plt_comment1_6 <- ggarrange(plt_ychromosomescores,
                             plt_viralrespiratory,
                             plt_prol,
                             plt_yeast_heatshock,
                             nrow=2, align = "hv", ncol=2, labels="AUTO"))
```

```{r}
ggplot2::ggsave("../Figures/Figs/R1_6_OtherData.png",
                plt_comment1_6,
                width = 8, height=7, bg = 'white')

ggplot2::ggsave("../Figures/Figs/R1_6_OtherData.pdf",
                plt_comment1_6,
                width = 8, height=7, bg = 'white')
```



## Comment 3.1 — Cell type composition as a confounder in GTEx age analyses

R.: Estimate cell type proportions in the GTEx tissues highlighted in the main text using a reference-based deconvolution tool, and incorporate these estimates as covariates in the age–score association models. Results with and without compositional correction are compared for a subset of tissues, both those with and without significant age associations in the original analysis.

Cell composition correction for age-senescence associations in GTEx

Strategy: GSEA only (enrichment-based), comparing:
- Baseline:  design ~ AGE  (as in the paper)
- Corrected: design ~ AGE + xCell_ct1 + xCell_ct2 + ...

Tissues:
WITH signal    : Artery - Aorta, Breast - Mammary Tissue
WITHOUT signal : Skin - Sun Exposed (Lower leg), Lung

Signatures: HernandezSegura 

```{r}
library(xCell2)
library(limma)
library(dplyr)
library(tidyr)
library(ggplot2)
library(cowplot)
library(beepr)
```

```{r}
# -----------------------------------------------------------------------------
# HOW xCell2 WORKS
# -----------------------------------------------------------------------------
# xCell2 estimates cell type enrichment scores from bulk RNA-seq data using a
# reference panel of known pure cell type expression profiles. It operates in
# two stages:
#
# (1) TRAINING: For each cell type, xCell2 identifies genes that are
#     specifically and consistently highly expressed in that cell type relative
#     to all others in the reference panel (not just highly expressed, but
#     *relatively* highly expressed). Multiple signatures per cell type are
#     built using different parameter combinations (stringency thresholds,
#     reference sample subsets), producing an ensemble that is more robust than
#     a single signature. This is why internal signature names contain suffixes
#     like "#_0.333_1_37_0.97" encoding those parameters.
#
# (2) SCORING: For each sample, xCell2 ranks all genes by expression and asks
#     whether the signature genes for a given cell type are collectively ranked
#     higher than expected by chance — conceptually similar to ssGSEA. A high
#     score means that cell type's transcriptional signal is enriched in the
#     sample; a score near zero means no detectable enrichment. A spillover
#     correction is then applied to reduce cross-talk between related cell
#     types that share marker genes.
#
# Unlike CIBERSORTx, which deconvolves bulk samples into cell type proportions
# that sum to 1 by solving a mixture model, xCell2 scores each cell type
# independently. Scores are not proportions, not bounded, and not comparable
# across cell types in absolute terms. Within a cell type however, higher
# scores consistently reflect greater presence of that cell type's
# transcriptional signal across samples — which is sufficient for our purpose
# of using them as covariates to control for age-related compositional
# variation in GTEx tissues.
# -----------------------------------------------------------------------------
```


```{r}
# =============================================================================
# Cell composition correction for age-senescence associations in GTEx
# Addresses reviewer comment 3.1
#
# Workflow per tissue:
#   1. Run xCell2 to estimate cell type enrichment scores
#   2. Select informative cell type covariates
#   3. Build baseline (~ AGE) and corrected (~ AGE + cell types) design matrices
#   4. Run calculateDE() with each design -> ranked t-statistics for AGE
#   5. Run fgsea on both ranked lists
#   6. Compare NES and significance with and without correction
#
# Tissues:
#   WITH signal    : Artery - Aorta, Breast - Mammary Tissue
#   WITHOUT signal : Skin - Sun Exposed (Lower leg), Lung
#
# Signatures: HernandezSegura, SAUL_SEN_MAYO
#
# NOTE ON xCell2 SCORES:
# xCell2 scores are enrichment scores (conceptually similar to ssGSEA), NOT
# cell type proportions or percentages. Each value reflects how enriched the
# transcriptomic signature of a given cell type is in a sample relative to the
# reference signatures. Values are not constrained to sum to 1 across cell
# types and are not directly comparable across cell types.
#
# For this analysis, scores are used purely as covariates in the limma linear
# model, not for cell type quantification per se. The rationale is: if cell
# type composition varies with age in a given tissue, the xCell2 scores for
# the relevant cell types will correlate with age, and including them as
# covariates will absorb that compositional variance from the model before
# estimating the age effect on the senescence signature. This allows us to
# assess whether the age-senescence associations reported in the paper are
# robust to, or confounded by, age-related shifts in cell type composition.
#
# ON MULTIPLE SIGNATURES PER CELL TYPE:
# Internally, xCell2 stores multiple gene set signatures per cell type,
# identified by names such as:
#   "CD4-positive, alpha-beta T cell#_0.333_1_37_0.97"
# The suffix encodes training parameters (quantile threshold, number of genes,
# performance score). This ensemble approach makes the enrichment scores more
# robust. xCell2Analysis() aggregates across all signatures per cell type
# and returns a single score per cell type per sample — the xcell_scores matrix
# contains one row per cell type (43 total). The internal naming is irrelevant
# for downstream use.
# =============================================================================

```


```{r}
# local job
composition_correction_results <- readRDS("../Figures/Comment3_1/composition_correction_results_noscale.rds")

```

```{r}
TISSUE_SIGNAL <- c(
  "Artery - Aorta",
  "Breast - Mammary Tissue",
  "Cells - Cultured fibroblasts",
  "Thyroid"
)
```


```{r fig.width=7, fig.height=11}

results_df <- composition_correction_results
# -----------------------------------------------------------------------------
# 7. Visualisation — all 49 tissues
# -----------------------------------------------------------------------------
# With 49 tissues, a standard dot/line plot becomes unreadable on a single
# panel. We use a heatmap-style layout showing NES for baseline and corrected
# side by side, with significance overlaid as asterisks. Tissues are grouped
# by whether they showed signal in the paper. This allows the reviewer to see
# at a glance: (a) which tissues retain signal after correction, (b) whether
# any new signal emerges, and (c) how much the NES changes overall.

model_colours <- c("Baseline"  = "#2171b5",
                   "Corrected" = "#cb181d")
# rename model levels for clarity
plot_df <- results_df %>%
  mutate(
    tissue_group = ifelse(tissue %in% TISSUE_SIGNAL,
                          "Signal \nin paper", "No signal \nin paper"),
    tissue_group = factor(tissue_group,
                          levels = c("Signal \nin paper", "No signal \nin paper")),
    model        = factor(model,
                          levels   = c("Baseline", "Corrected"),
                          labels   = c("Non-corrected", "Corrected"))
  )

tissue_order <- plot_df %>%
  filter(model == "Non-corrected") %>%
  arrange(desc(tissue)) %>%   # alphabetical, desc so A appears at top
  pull(tissue) %>%
  unique()

plot_df <- plot_df %>%
  mutate(tissue = factor(tissue, levels = tissue_order))

model_colours <- c("Non-corrected" = "#2171b5",
                   "Corrected"     = "#cb181d")

plot_all <- ggplot(plot_df,
                   aes(x = NES, y = tissue, colour = model)) +
  geom_vline(xintercept = 0, linetype = "dashed",
             colour = "grey50", linewidth = 0.4) +
  geom_line(aes(group = tissue), colour = "grey75", linewidth = 0.5) +
  geom_point(aes(shape = sig), size = 3, stroke = 1.2, fill = "white") +
scale_shape_manual(
    values = c("FALSE" = 21, "TRUE"  = 19),
    labels = c("FALSE" = "FDR > 0.05", "TRUE" = "FDR \u2264 0.05"),
    name   = NULL
  ) +
  scale_colour_manual(values = model_colours, name = "Model") +
  #facet_grid(tissue_group ~ ., scales = "free_y", space = "free_y") +
  labs(
    x        = "Normalised Enrichment Score (NES)",
    y        = NULL,
    title    = "HernandezSegura Age-Enrichment \nacross GTEx tissues",
    subtitle = "Non-corrected vs. cell composition-corrected"
  ) +
  theme_bw(13) +
  theme(
    strip.background  = element_rect(fill = "grey92"),
    strip.text.y      = element_text(angle = 90, vjust = 0.5),
    legend.position   = "bottom",
    axis.text.y       = element_text(size = 11),
    plot.caption      = element_text(size = 8, colour = "grey40",
                                     hjust = 0, margin = margin(t = 6)), 
                                     legend.title = element_blank(),
    legend.box        = "vertical",
    legend.spacing.y  = unit(0, "pt")
  )
 

plot_all


```
 



```{r}
ggplot2::ggsave("Figures/Figs/R3_1_CellTypeComposition.png",
                plot_all,
                width = 8, height=11, bg = 'white')

ggplot2::ggsave("Figures/Figs/R3_1_CellTypeComposition.pdf",
                plot_all,
                width = 7, height=11, bg = 'white')
```
## Comment 1.5 — Additional datasets for age correlation analysis

R.: Apply markeR's discovery mode to one or more datasets independent of GTEx, modelling age as a continuous variable, to assess whether age-associated senescence signals are reproducible across transcriptomic resources and pointed limitations are correct. 

## Session Info
```{r}
sessionInfo()
```
