---
title: "Discovery Mode Tutorial"
author: "Rita M. Silva"
date: "`r Sys.Date()`"
---


```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette provides a comprehensive introduction to the **`markeR`** package, focusing on its **Discovery Mode**. The discovery mode was designed for users who are interested in examining the relationship between a gene set and one or more metadata variables of interest, being suitable for exploratory or hypothesis-generating analyses.   

# Case-study: Senescence

In this vignette, we use a pre‑processed RNA-seq dataset from Marthandan et al. (2016, GSE63577), with normalised read counts for human fibroblasts under **replicative senescence** and **proliferative control**. See `?counts_example` for preprocessing details and structure. `markeR` requires as input a filtered and normalised, non log-transformed, gene expression matrix (genes × samples). Row names must be gene identifiers; column names must match sample IDs in the metadata. 

We use the accompanying metadata from the Marthandan et al. (2016) study (see `?metadata_example`).

This dataset serves as a working example to demonstrate the main functionalities of the `markeR` package. In particular, it will be used to showcase the two primary modules of `markeR` for quantifying phenotypes using gene sets:

- **Score-based methods**:  log2-median expression, ranking approaches, and single-sample gene set enrichment analysis (ssGSEA) to quantify coordinated expression within a gene set.
- **Enrichment-based methods**: GSEA using moderated t- or B-statistics.

To illustrate the usage of `markeR`, we use the **HernandezSegura** gene set: A transcriptomic gene set identified by Hernandez-Segura et al. (2017) as consistently altered across multiple senescence models. This set includes information on the direction of change in expression of genes upon phenotype.

```{r example}
library(markeR)
```

```{r loadsig} 
# Load example gene sets
data("genesets_example")
HernandezSegura_GeneSet <- list(HernandezSegura=genesets_example$HernandezSegura)
head(HernandezSegura_GeneSet) 
```

```{r loaddata}
data(counts_example)
# Load example data
counts_example[1:5,1:5]
```

For illustration purposes, two synthetic variables were added to the data:

* `DaysToSequencing`, the number of days between sample preparation and sequencing;
* `Researcher`, identifying the person who processed each sample. 

This enables exploration of associations between gene set activity and both categorical and continuous variables.

```{r load_metadata}
data(metadata_example)

set.seed("123456")

metadata_example$Researcher <- sample(c("John","Ana","Francisca"),39, replace = TRUE)
metadata_example$DaysToSequencing <- sample(c(1:20),39, replace = TRUE)
 
head(metadata_example)
```

# Score-Based approaches 
  
A score summarising the collective expression of a gene set is assigned **to each sample**. Scores can be visualised using built-in functions, or used directly in downstream analyses (*e.g.*, comparisons between phenotypic groups of samples, correlations with numerical phenotypes). 
 
## Outputs

If `method = "logmedian"` (or `ssGSEA`, `ranking`), the main function `VariableAssociation()` returns a structured list with:

- **`overall`**: Effect sizes (*Cohen’s f*) and p-values for each variable.
- **`contrasts`**: For categorical variables, pairwise or grouped comparisons using *Cohen’s d* with BH-adjusted p-values.
- **`plot`**: A combined visualization showing:
    1. Lollipop plot of effect sizes (i.e., *Cohen’s f* per variable)
    2. Distribution plots of the score by variable (density or scatter)
    3. Lollipop plots of contrasts (*Cohen’s d*) for categorical variables, if applicable
- **`plot_overall`**, **`plot_contrasts`**, **`plot_distributions`**: Individual components of the combined plot.

## Controlling Contrast Resolution

The `mode` parameter determines the level of contrast detail for categorical variables:

- **`"simple"`**: All pairwise contrasts between group levels (e.g., A vs B, A vs C, B vs C).
- **`"medium"`**: One group vs. the union of all others (e.g., A vs B+C+D).
- **`"extensive"`**: All algebraic combinations of group levels (e.g., A+B vs C+D).

## Results

For this example, we use:

- The **HernandezSegura** gene signature
- The **`logmedian`** scoring method
- **`mode = "extensive"`** for thorough contrast analysis

Though artificial, `DaysToSequencing` and `Researcher` mimic potential **technical covariates**. Strong associations between these and the score could indicate **batch effects**, where technical variation may confound biological interpretation.


In this example, the `Condition` variable shows a large effect size (Cohen’s f and Cohen's d), confirming strong discrimination between Senescent and Proliferative samples. The remaining variables don't show significant associations, suggesting no major batch effects that might be reflected in the computed scores.

```{r variableassoc_score_sen, fig.width=7, fig.height=7, out.width="100%", warning=FALSE, message=FALSE} 
results_scoreassoc_bidirect <- VariableAssociation(data = counts_example, 
                          metadata = metadata_example, 
                          method = "logmedian",
                          cols = c("Condition","Researcher","DaysToSequencing"),  
                          gene_set = HernandezSegura_GeneSet,
                          mode="extensive",
                          nonsignif_color = "white", signif_color = "red", saturation_value=NULL,sig_threshold = 0.05,
                          widthlabels=30, labsize=10, titlesize=14, pointSize=5, discrete_colors=NULL,
                          continuous_color = "#8C6D03", color_palette = "Set2")

results_scoreassoc_bidirect$Overall
results_scoreassoc_bidirect$Contrasts
```



 
# Enrichment-based approaches

Enrichment-based methods implement **Gene Set Enrichment Analysis (GSEA)**. Genes are ranked according to differential expression statistics, and a Normalised Enrichment Score (NES) per variable of interest is computed, accompanied by a p-value adjusted for multiple hypothesis testing.


## Outputs

If `method = "GSEA"`, the main function `VariableAssociation()` returns a structured list with:

- **`data`**: A tidy data frame of GSEA results. For each variable contrast, this includes:
    - **Contrast**: The comparison performed (e.g., A - B, A - (B+C))
    - **Statistic**: The metric used for gene ranking (either *t* or *B*)
    - **NES**: Normalized Enrichment Score
    - **Adjusted p-value**: Multiple-testing corrected p-value (Benjamini–Hochberg)
    - **Gene Set Name**: The gene set being tested

- **`plot`**: A `ggplot2` object showing the NES and significance of each contrast as a **lollipop plot**:
    - **Point position** reflects NES (directional enrichment)
    - **Point color** reflects adjusted p-value significance
    - **Dashed lines** represent contrasts with negative NES under the **B statistic**, indicating likely non-alteration

## Output Interpretation

Depending on the statistic used (`t` or `B`), the interpretation of results varies:

- **Negative NES (Normalized Enrichment Score):**
    - **t statistic**: A negative NES implies the gene set is **depleted** in over-expressed genes — i.e., genes in the set are under-expressed relative to others.
    - **B statistic**: A negative NES suggests the gene set is **less altered** than most other genes — indicating a lack of strong expression change.

- **Dashed Lines in the Plot:**
    - Dashed lines denote contrasts with **negative NES** using the **B statistic**, indicating gene sets that are putatively *not* altered.

- **Subtitle Differences in the Plot:**
    - When using the **B statistic**, the plot subtitle is **"Altered Contrasts"**.
    - When using the **t statistic**, the subtitle becomes **"Enriched/Depleted Contrasts"**.

## Example: Exploring Gene Set Enrichment by Variable

The following code evaluates how three phenotypic variables (`Condition`, `Researcher`, and `DaysToSequencing`) are associated with the **HernandezSegura** gene set. 

 
 In this example, the HernandezSegura gene set shows significant enrichment in samples sequenced by Francisca relative to those processed by Ana or John, which would suggest a differentiated researcher-associated technical impact on the samples' biological phenotype. This gene set shows also a strong depletion in proliferative samples, which is expected given its annotation as senescence-associated and results from the score-based approach.  
 
 
 
```{r GSEA_varassoc, fig.width=6, fig.height=6, out.width="60%", warning=FALSE, message=FALSE}
varassoc_gsea <- VariableAssociation(
  data = counts_example,
  metadata = metadata_example,
  method = "GSEA",
  cols = c("Condition", "Researcher", "DaysToSequencing"),
  mode = "simple",
  gene_set = HernandezSegura_GeneSet,
  saturation_value = NULL,
  nonsignif_color = "white",
  signif_color = "red",
  sig_threshold = 0.05,
  widthlabels = 30,
  labsize = 10,
  titlesize = 14,
  pointSize = 5
)

varassoc_gsea$data
```
 

 
# Session Information

```{r}
sessionInfo()
```
