---
title: "Benchmarking Mode Tutorial"
author: "Rita M. Silva"
date: "`r Sys.Date()`"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


This vignette provides a comprehensive introduction to the **markeR** package, focusing on its **Benchmarking Mode**. This mode is designed to evaluate gene sets' performance in marking a metadata variable, *i.e.*, a phenotype such as disease or cellular condition, returning comparative visualisations across scoring and enrichment methods. It allows users to assess the robustness and reliability of gene sets across various conditions, providing a standardized framework for benchmarking.
 
# Case-study: Senescence

In this vignette, we use a pre‑processed RNA-seq dataset from Marthandan et al. (2016, GSE63577), with normalised read counts for human fibroblasts under **replicative senescence** and **proliferative control**. See `?counts_example` for preprocessing details and structure. `markeR` requires as input a filtered and normalised, non log-transformed, gene expression matrix (genes × samples). Row names must be gene identifiers; column names must match sample IDs in the metadata. 

We use the accompanying metadata from the Marthandan et al. (2016) study (see `?metadata_example`).

This dataset serves as a working example to demonstrate the main functionalities of the `markeR` package. In particular, it will be used to showcase the two primary modules of `markeR` for quantifying phenotypes using gene sets:

- **Score-based methods**:  log2-median expression, ranking approaches, and single-sample gene set enrichment analysis (ssGSEA) to quantify coordinated expression within a gene set.
- **Enrichment-based methods**: GSEA using moderated t- or B-statistics.

To illustrate the usage of `markeR`, we use three gene sets commonly associated with cellular senescence:

-   **LiteratureMarkers**: A small, curated set of well-established senescence-associated genes repeatedly reported in the literature. This concise gene set includes key markers often used for validating senescence phenotypes. This set includes information on the expected direction of change in expression of genes upon phenotype (i.e., whether genes are expected to be up- or down-regulated in senescence).
-   **REACTOME_CELLULAR_SENESCENCE**: A comprehensive gene set from the MSigDB REACTOME collection, representing known molecular pathways involved in cellular senescence and commonly used in enrichment-based analyses. This is also treated as an undirected gene set without information on the up- or down-regulation of individual genes.
-   **HernandezSegura**: A transcriptomic gene set identified by Hernandez-Segura et al. (2017) as consistently altered across multiple senescence models. This set includes information on the expected direction of change in expression of genes upon phenotype. 

```{r example}
library(markeR)
```

```{r loadsig} 
# Load example gene sets
data("genesets_example")
lapply(genesets_example, head, n = 10)
```

```{r loaddata}
data(counts_example)
# Load example data
counts_example[1:5,1:5]
```

For illustration purposes, two synthetic variables were added to the data:

* `DaysToSequencing`, the number of days between sample preparation and sequencing;
* `Researcher`, identifying the person who processed each sample.

This enables exploration of associations between gene set activity and both categorical and continuous variables.
 

```{r load_metadata}
data(metadata_example)

set.seed("123456")

metadata_example$Researcher <- sample(c("John","Ana","Francisca"),39, replace = TRUE)
metadata_example$DaysToSequencing <- sample(c(1:20),39, replace = TRUE)
 
head(metadata_example)
```

# Calculate Senescence Scores 

The `CalculateScores` function computes the gene set scores for each sample based on predefined gene sets, such as a senescence gene set. It returns a named list where each entry corresponds to a specific gene set and includes the calculated scores, along with metadata (if available). When setting `method = "all"`, the function returns a list, where each element corresponds to a scoring method and contains the respective data frame of scores, allowing comparison between methods. The function allows users to select from three different scoring methods:

* **logmedian**: mean of the across-sample normalised log2 median-centred expression levels of the genes in the set; for bidirectional gene sets, the sample score is the partial score for the subset of putatively upregulated genes minus that of the downregulated subset.

* **Ranking**: mean expression rank of gene set members in each sample; for bidirectional gene sets, the sample score is the partial score for the subset of putatively upregulated genes minus that of the downregulated subset, and normalised by the number of genes in the set. 

* **ssGSEA**: single-sample gene set enrichment score using ssGSEA; for bidirectional gene sets, the sample score is the partial score for the subset of putatively upregulated genes minus that of the downregulated subset.

These methods are very similar and, when applied to a robust gene set, are expected to yield similar results across all three methods. Empirically, a good gene set will be one that shows consistent results, both in the calculated scores and in Cohen's d or f statistics, across different methods. If the gene set is not robust, or if there is considerable noise, the results across methods may differ significantly.

The `PlotScores` function computes and visualizes gene set scores according to the chosen method and variable type:

- `method = "all"`  

    - Categorical `Variable`: produces a heatmap of Cohen's *d* and a volcano plot of all pairwise group contrasts.  
    - Numeric `Variable`: produces a heatmap of Cohen's *F* statistics and a volcano plot of associations.  

- `method != "all"`  

    - Categorical `Variable`: generates violin plots of scores per gene set.  
    - Numeric `Variable`: generates scatter plots of scores versus the numeric variable.  
    - `Variable` is `NULL`: displays a density plot of the score distribution.  

This structure clearly links each combination of method and variable type to the resulting visualization, avoiding ambiguity.


## logmedian method

The **logmedian** method computes a gene signature score per sample as follows. First, gene expression values are log2-transformed and each gene is median-centered across all samples. For a given sample, the score is the mean of the median-centered expression values of the genes in the set (De Almeida et al., 2019). Each gene’s contribution is thus evaluated relative to its baseline expression, and the resulting score quantifies the coordinated activity of the gene set.

For bidirectional gene sets, where genes are annotated by expected direction of regulation, the score is calculated as the difference between the mean of the upregulated genes and the mean of the downregulated genes. Users can compute scores for individual samples using one or more predefined gene sets.

The following example demonstrates calculation of a gene signature score using the logmedian method:

 
```{r}
df_Scores <- CalculateScores(data = counts_example,
                             metadata = metadata_example,
                             method = "logmedian",
                             gene_sets = genesets_example)


head(df_Scores$REACTOME_CELLULAR_SENESCENCE)
head(df_Scores$HernandezSegura)
head(df_Scores$LiteratureMarkers)
```

Users can directly visualize gene set scores with the plotting functions.

Effect sizes are computed via the `compute_cohen` parameter (default = `TRUE`). For a categorical metadata variable with two levels, Cohen’s d is calculated; for more than two levels, Cohen’s f is used unless a specific pairwise comparison is specified via `cond_cohend`, in which case Cohen’s d is reported for that comparison.

If `pvalcalc = TRUE` (default = `FALSE`), an associated p-value is reported (uncorrected for multiple testing). p-values are derived from a two-sample t-test for two-group or numeric-variable comparisons, and from ANOVA for multi-group comparisons.

```{r exampleScore, fig.width=8, fig.height=4, out.width="80%", warning=FALSE, message=FALSE}

senescence_triggers_colors <- c(
  "none" = "#E57373",  # Soft red   
  "Telomere shortening" = "#4FC3F7"  # Vivid sky blue  
)

cond_cohend <- list(A=c("Senescent"),  
                    B=c("Proliferative"))

PlotScores(data = counts_example, 
           metadata = metadata_example, 
           gene_sets = genesets_example,
           ColorVariable = "SenescentType", 
           Variable="Condition",  
           method ="logmedian", 
           ColorValues = senescence_triggers_colors, 
           ConnectGroups=TRUE, 
           ncol = NULL, 
           nrow = 1, 
           widthTitle=24, 
           limits = NULL, 
           legend_nrow = 1, 
           pointSize=4,
           compute_cohen=TRUE,
           cond_cohend=cond_cohend,
           title="Marthandan et al. 2016",
           labsize=9, 
           titlesize = 12)  

```


Providing gene directionality can substantially affect score interpretation. For example, in the *Literature_Senescence* signature, omitting direction of change in expression of genes upon phenotype may lead to senescent samples appearing to have lower scores than proliferative ones. Incorporating directionality aligns the scores with biological expectations. Therefore, it is **strongly recommended** to specify the expected  direction of gene expression changes in a set whenever this information is available, ensuring more accurate and meaningful interpretation of the results.

```{r exampleScore_bidirectional, fig.width=6, fig.height=4, out.width="60%", warning=FALSE, message=FALSE}
 
GeneSets_Example_Bidirectionality <- list(LiteratureMarkers=genesets_example$LiteratureMarkers,
                                          LiteratureMarkers_Unidirectional =  genesets_example$LiteratureMarkers$gene)
print(GeneSets_Example_Bidirectionality)

PlotScores(data = counts_example, 
           metadata = metadata_example, 
           gene_sets =  GeneSets_Example_Bidirectionality,
           ColorVariable = "SenescentType", 
           Variable="Condition",  
           method ="logmedian", 
           ColorValues = senescence_triggers_colors, 
           ConnectGroups=TRUE, 
           ncol = NULL, 
           nrow = NULL, 
           widthTitle=24, 
           limits = NULL, 
           legend_nrow = 1, 
           pointSize=4,
           compute_cohen=TRUE,
           cond_cohend=cond_cohend,
           title="Marthandan et al. 2016",
           labsize=9, 
           titlesize = 12)  

```


When analyzing a numeric variable, the function generates a scatter plot of the variable against gene set scores and can optionally compute an effect size using Cohen’s f. The user may select a correlation method (Pearson, Spearman, or Kendall) to quantify the association between the numeric variable and the scores. Optional p-value calculations for the association can also be included.

The following example illustrates how to configure the function for a numeric variable:



```{r examplenumeric, fig.width=8, fig.height=3.5, out.width="80%", warning=FALSE, message=FALSE}
PlotScores(data = counts_example,
           metadata = metadata_example,
           gene_sets = genesets_example,
           Variable = "DaysToSequencing",               
           method = "logmedian",            
           ColorValues = "#3B415B",             
           ConnectGroups = FALSE,          
           ncol = NULL,                    
           nrow = 1,                    
           pointSize = 6,                  
           compute_cohen = TRUE,          
           pvalcalc = TRUE,                
           title = "Marthandan et al. 2016",  
           labsize=9, 
           titlesize = 12, 
           widthTitle = 26,
           cor = "pearson")
```
 
To visualize the overall distribution of scores across gene sets, the `PlotScores` function can be used without specifying the `GroupingVariable` parameter, i.e, without grouping scores by any metadata variable.. In this case, it generates a grid of density plots, with each plot representing the score distribution for a specific gene set.


```{r plotscores_density, fig.width=8, fig.height=3, out.width="80%", warning=FALSE, message=FALSE}
 
PlotScores(data = counts_example, 
           metadata = metadata_example, 
           gene_sets = genesets_example, 
           method ="logmedian", 
           ColorValues = NULL,  
           ncol = NULL, 
           nrow = 1, 
           widthTitle=24, 
           limits = NULL,  
           title="Marthandan et al. 2016",
           labsize=9, 
           titlesize = 11)  

```

## ssGSEA method

Single sample Gene Set Enrichment Analysis (**ssGSEA**) was implemented using a modified version of the `GSVA` package’s `gsva()` function (Barbie et al., 2009), based on the original Gene Set Enrichment Analysis (GSEA) method (Subramanian et al., 2005). ssGSEA ranks all genes by their expression within each sample and computes a running-sum statistic over the ranked list. For unidirectional gene sets (i.e., with no information on the expected direction of each gene’s regulation upon phenotype), the ssGSEA sample score reflects overall coordinated expression of the genes in the set. For bidirectional gene sets, the score is calculated as the ssGSEA score computed for the upregulated subset of genes minus the score computed for the downregulated subset.

 

```{r examplessGSEA, fig.width=8, fig.height=4, out.width="80%", warning=FALSE, message=FALSE}

senescence_triggers_colors <- c(
  "none" = "#E57373",  # Soft red   
  "Telomere shortening" = "#4FC3F7"  # Vivid sky blue  
)

cond_cohend <- list(A=c("Senescent"),  
                    B=c("Proliferative"))

PlotScores(data = counts_example, 
           metadata = metadata_example, 
           gene_sets = genesets_example,
           ColorVariable = "SenescentType", 
           Variable="Condition",  
           method ="ssGSEA", 
           ColorValues = senescence_triggers_colors, 
           ConnectGroups=TRUE, 
           ncol = NULL, 
           nrow = 1, 
           widthTitle=24, 
           limits = NULL, 
           legend_nrow = 1, 
           pointSize=4,
           compute_cohen=TRUE,
           cond_cohend=cond_cohend,
           title="Marthandan et al. 2016",
           labsize=9, 
           titlesize = 12)  

```

## Ranking method

The **ranking** method calculates gene signature scores using a non-parametric approach based on the relative expression of genes within a set. For each sample, all genes are ranked by expression. The score is then calculated as the sum of the ranks of the genes in the gene set, multiplied by +1 (for upregulated genes those with unspecified direction of regulation change upon phenotype) or -1 (downregulated genes), and normalised by the number of genes in the set. 

The following example demonstrates the use of the "ranking" method for both unidirectional and bidirectional gene sets:

```{r ranking, fig.width=8, fig.height=4, out.width="80%", warning=FALSE, message=FALSE}
 
senescence_triggers_colors <- c(
  "none" = "#E57373",  # Soft red   
  "Telomere shortening" = "#4FC3F7"  # Vivid sky blue  
)

cond_cohend <- list(A=c("Senescent"),  
                    B=c("Proliferative"))

PlotScores(data = counts_example, 
           metadata = metadata_example, 
           gene_sets = genesets_example,
           ColorVariable = "SenescentType", 
           Variable="Condition",  
           method ="ranking", 
           ColorValues = senescence_triggers_colors, 
           ConnectGroups=TRUE, 
           ncol = NULL, 
           nrow = 1, 
           widthTitle=24, 
           limits = NULL, 
           legend_nrow = 1, 
           pointSize=4,
           compute_cohen=TRUE,
           cond_cohend=cond_cohend,
           title="Marthandan et al. 2016",
           labsize=9, 
           titlesize = 12)  

```

## All methods

To compare various metrics across different condition combinations, violin plots may not always be the best choice. In such cases, users can set`method = "all"` to generate a summary heatmap and volcano-like plot. The function will return one heatmap per gene set, with rows corresponding to all possible combinations of values in the `GroupingVariable`. In parenthesis is represented the p-value, adjusted using the `BH` method (i.e., the Benjamini–Hochberg false discovery rate procedure), across all combinations of contrasts and gene signatures. It will also return a volcano-like plot (Cohen's d effect sizes vs -log10(adjusted p-values)), where each dot represents a method-signature pair, faceted by contrast. The dashed lines represent user-defined thresholds for significance and effect size.

The `mode` parameter controls how contrasts are generated for categorical variables, allowing users to adjust the complexity of the analysis:

-   **"simple"**: Performs the minimal number of pairwise comparisons between group levels (e.g., for a factor with levels A, B, C and D, it generates A - B, A - C, A - D, B - C, B - D, C - D).
-   **"medium"**: Includes comparisons between one group and the union of other groups (e.g., A - (B + C + D); B - (A + C + D)), allowing for broader contrasts beyond simple pairwise comparisons. 
-   **"extensive"**: Allows for all possible algebraic combinations of group levels (e.g., (A + B) - (C + D)). 

In this example, HernandezSegura and LiteratureMarkers consistently discriminate Senescent from Proliferative samples, while REACTOME_CELLULAR_SENESCENCE shows weaker and less consistent separation.

```{r Overall_Scores, fig.width=6, fig.height=3, out.width="80%", warning=FALSE, message=FALSE}

Overall_Scores <- PlotScores(data = counts_example, 
                             metadata = metadata_example,  
                             gene_sets=genesets_example, 
                             Variable="Condition",  
                             method ="all",   
                             ncol = NULL, 
                             nrow = 1, 
                             widthTitle=30, 
                             limits = c(0,3.5),   
                             title="Marthandan et al. 2016", 
                             titlesize = 10,
                             ColorValues = list(heatmap=c("#F9F4AE", "#B44141"),
                                                volcano=signature_colors <- c(
                                                  HernandezSegura = "#A07395",               
                                                  REACTOME_CELLULAR_SENESCENCE = "#6B8E9E",  
                                                  LiteratureMarkers = "#CA7E45"             
                                                )
                             ),
                             mode="simple",
                             widthlegend=30, 
                             sig_threshold=0.05, 
                             cohen_threshold=0.6,
                             pointSize=6,
                             colorPalette="Paired")  

```

```{r Overall_Scores_heatmap, fig.width=10, fig.height=2, out.width="80%", warning=FALSE, message=FALSE}
Overall_Scores$heatmap
```

```{r Overall_Scores_volcano, fig.width=6, fig.height=3, out.width="60%", warning=FALSE, message=FALSE}
Overall_Scores$volcano
```


## Classification Potential of Gene Signatures

The `ROC_Scores` and `AUC_Scores` functions allow users to evaluate the classification potential of gene set scores based on ROC curves and their AUC values. These functions help to assess how well a given score can differentiate between conditions, based on predefined contrasts. Besides `method="all"`, these functions can also be used for each method individually.

The `ROC_Scores` function generates ROC curves for different scoring methods across contrasts, allowing users to visualize performance differences.

```{r roc_scores, fig.width=10, fig.height=3, out.width="70%", warning=FALSE, message=FALSE}
 ROC_Scores(data = counts_example, 
           metadata = metadata_example, 
           gene_sets=genesets_example, 
           method = "all", 
           variable ="Condition",
           colors = c(logmedian = "#3E5587", ssGSEA = "#B65285", ranking = "#B68C52"), 
           grid = TRUE, 
           spacing_annotation=0.3, 
           ncol=NULL, 
           nrow=1,
            mode = "simple",
            widthTitle = 28,
           titlesize = 10,  
           title="Marthandan et al. 2016") 
 
```
 
The `AUC_Scores` function generates heatmaps summarizing AUC values for each gene set, with methods as columns and contrasts as rows. The plot is returned in the `$plt` element, and the underlying data used to generate it is available in the `$data` element.

```{r auc_scores, fig.width=10, fig.height=2, out.width="80%", warning=FALSE, message=FALSE}
AUC_Scores(data = counts_example, 
           metadata = metadata_example, 
           gene_sets=genesets_example, 
           method = "all", 
           mode = "simple", 
           variable="Condition", 
           nrow = 1, 
           ncol = NULL, 
           limits = NULL, 
           widthTitle = 28, 
           titlesize = 10, 
           ColorValues = c("#F9F4AE", "#B44141"),   
           title="Marthandan et al. 2016") 
```


 HernandezSegura and LiteratureMarkers exhibit consistently high AUCs across scoring methods, while REACTOME_CELLULAR_SENESCENCE shows more heterogeneous performance.


## False Positive Rate (FPR) Calculations

The user can assess the significance of gene set scores by comparing observed effect sizes against a distribution of those originated by random gene sets with the same number of genes and matched directionality. For each original gene set, the function calculates the observed Cohen's d (and p‑value) using (`GroupingVariable`). It then generates a number of simulated gene sets (`number_of_sims`) by randomly sampling the same number of genes from a user provided gene list (`gene_list`) and computes their Cohen's d values. The simulation results are visualised as violin plots of the distribution of Cohen's d values for each method, overlaid with the observed values of the original gene sets, and a 95th percentile threshold. Significance is indicated by distinct point shapes based on the associated p‑value.

```{r FDRSim, fig.width=12, fig.height=3, out.width="100%", warning=FALSE, message=FALSE}

FPR_Simulation(data = counts_example,
               metadata = metadata_example,
               original_signatures = genesets_example,
               gene_list = row.names(counts_example),
               number_of_sims = 100,
               title = "Marthandan et al. 2016",
               widthTitle = 30,
               Variable = "Condition",
               titlesize = 12,
               pointSize = 5,
               labsize = 10,
               mode = "simple",
               ColorValues=NULL,
               ncol=NULL, 
               nrow=1 ) 

```
 
# Enrichment-Based Methods

## Differentially Expressed Genes

The `calculateDE` function in the `markeR` package leverages the `limma` framework to compute differential gene expression statistics from raw count data. This function is highly flexible and supports several modes of operation depending on the user's experimental design. In the examples below, we illustrate two common scenarios:

-   **Automatic Design Matrix with Contrasts:** In the first example, the design matrix is built automatically from the metadata using a specified variable (here, `"Condition"`). The user must explicitly define the comparisons of interest based on the levels of this variable (e.g., `Senescent - Proliferative`). Internally, this approach fits a linear model without an intercept, enabling the user to define contrasts between the levels. This is ideal for simpler experimental designs, where quick comparison between predefined groups is desired without manually specifying the full model matrix.
-   **Providing an Externally Constructed Design Matrix:** In the second example, the user manually creates the design matrix (e.g., using `model.matrix(~ 0 + Condition)` for a no-intercept model) and provides it directly to `calculateDE`. This gives full control over how the design is specified, including complex experimental setups or custom encodings. Multiple contrasts can be defined later, based on this matrix, to extract specific comparisons of interest, using the `Contrast` parameter. If this parameter is left as `NULL`, the function will return results for all conditions (i.e. columns) defined in the design matrix. This approach is recommended when the user has a complex design or has already constructed the design matrix as part of a broader analysis pipeline.

Below are the corresponding code snippets demonstrating each scenario, by answering the same question: **What are the genes differentially expressed between senescence and proliferative cells?**

```{r DEGs, out.width="80%", warning=FALSE}
# Example 1: Build design matrix from variables (Condition) and apply a contrast.
# In this case, the design matrix is constructed automatically using the variable "Condition".
DEGs <- calculateDE(data = counts_example,
                    metadata = metadata_example,
                    variables = "Condition",
                    contrasts = c("Senescent - Proliferative"))
DEGs$`Senescent-Proliferative`[1:5,]

# Example 2: Supply a custom design matrix directly.
# Here, the design matrix is created externally (using no intercept, for instance).
design <- model.matrix(~0 + Condition, data = metadata_example)
colnames(design) <- c("Proliferative","Senescent")
DEGs2 <- calculateDE(data = counts_example,
                     metadata = NULL,
                     variables = NULL,
                     modelmat = design,
                     contrasts = c("Senescent - Proliferative"))
DEGs2$`Senescent-Proliferative`[1:5,]
```

After running differential expression analysis (using the `calculateDE` function), the user can visualize their results with the `plotVolcano` function. This function provides a flexible interface for exploring their data by allowing the user to:

-   **Plot Differential Gene Expression Statistics:**  
    Display a volcano plot with chosen statistics (e.g., log fold-change on the x-axis and –log₁₀ adjusted p-value on the y-axis).
-   **Color Interesting Genes:**  
    Highlight genes that pass user-specified thresholds by adjusting `threshold_y` and `threshold_x`.
-   **Annotate Top and Bottom N Genes:**  
    Optionally, label the top (and bottom) N genes based on the chosen statistic to quickly identify the most significant genes.
-   **Highlight Gene Signatures:** If the user provides a list of gene signatures using the `genes` argument, the function can highlight these genes in the plot. The user can also specify distinct colors for putatively upregulated and downregulated if their direction is known, or a color for genes that do not have a putative direction.

Below is an example usage of `plotVolcano` that visualizes differential expression results from a `DEResultsList`. The first plot shows the default behavior, generating a basic volcano plot without thresholds or gene highlights. Subsequent examples demonstrate how to customize the plot:

-   Adding significance thresholds to highlight genes of interest,
-   Annotating the top and bottom N genes by effect size,
-   And using gene signatures to color genes across multiple plots arranged by contrast and signature.

These examples illustrate how users can customise the output plot to highlight biologically meaningful patterns or focus on specific gene sets. 

```{r volcanos_DEGs, fig.width=4, fig.height=3, out.width="40%", warning=FALSE, message=FALSE} 

# Color Interesting Genes:
plotVolcano(DEGs, genes = NULL, N = NULL,
            x = "logFC", y = "-log10(adj.P.Val)", pointSize = 2,
            color = "#6489B4", highlightcolor = "#05254A", nointerestcolor = "#B7B7B7",
            threshold_y = 0.0001, threshold_x = 1,
            xlab = NULL, ylab = NULL, ncol = NULL, nrow = NULL, title = "Marthandan et al. 2016",
            labsize = 8, widthlabs = 25, invert = FALSE)

# Annotate Top and Bottom N Genes:
plotVolcano(DEGs, genes = NULL, N = 5,
            x = "logFC", y = "-log10(adj.P.Val)", pointSize = 2,
            color = "pink", highlightcolor = "#05254A", nointerestcolor = "#B7B7B7",
            threshold_y = NULL, threshold_x = NULL,
            xlab = NULL, ylab = NULL, ncol = NULL, nrow = NULL, title = "Marthandan et al. 2016",
            labsize = 8, widthlabs = 25, invert = FALSE)

```

In this example:

* Genes in the HernandezSegura set annotated as upregulated (green) display positive log2 fold changes, whereas those annotated as downregulated (red) show negative log2 fold changes. This pattern is consistent with the annotation of the set, although these genes are not necessarily those exhibiting the largest absolute fold changes.
* In the LiteratureMarkers set, *LMNB1* and *MKI67* exhibit strongly negative log2 fold change, consistent with their roles as proliferation markers absent in senescent cells.
* Genes from the REACTOME_CELLULAR_SENESCENCE set are undirected and appear across the full range of log2 fold change values, diluting discriminatory power.

```{r volcanos_DEGs3, fig.width=10, fig.height=3, out.width="90%", warning=FALSE, message=FALSE}

# Change order: signatures in columns, contrast in rows
plotVolcano(DEGs, genes = genesets_example, 
            N = NULL,
            x = "logFC", y = "-log10(adj.P.Val)", pointSize = 2,
            color = "#6489B4", highlightcolor = "#05254A", highlightcolor_upreg = "#038C65", highlightcolor_downreg = "#8C0303",nointerestcolor = "#B7B7B7",
            threshold_y = NULL, threshold_x = NULL,
            xlab = NULL, ylab = NULL, ncol = NULL, nrow = NULL, title = "Marthandan et al. 2016",
            labsize = 10, widthlabs = 24, invert = TRUE)
```



###  Note on Continuous Variables

If the user wants to analyse continuous variables (e.g., time or dosage), they must provide a custom design matrix via the `modelmat` argument, instead of using the `variables` argument. This is because the `variables` argument is intended to be used for categorical variables only: internally, when using `variables` to specify a phenotypic variable of interest, the function constructs a design matrix without an intercept and treats each level of the variable as a discrete group. This is appropriate for defining explicit contrasts between categorical levels but is not suitable for continuous variables, where such discretization would distort the relationship.

For continuous variables, the user should instead build a design matrix manually (e.g., using `model.matrix(~ variable)`), ensuring that the variable of interest is numeric and has a corresponding column in the matrix with the same name. The `calculateDE()` function will then use this matrix directly for linear modelling.

The interpretation of the differential expression results remains consistent: the statistics reflect always the expected change in expression associated with a 1-unit increase in the variable (e.g., one day, one unit of dosage, "Senescent" to "Proliferative", etc).  


```{r example_continuousDEG_1}
 
design <- model.matrix(~1 + DaysToSequencing, data = metadata_example)

DEGs_continuous1 <- calculateDE(data = counts_example,
                                metadata = metadata_example, 
                                modelmat = design,
                                contrasts = c("DaysToSequencing"))
DEGs_continuous1$DaysToSequencing[1:3,]
```

This usage of the `modelmat` argument allows the user to combine categorical and numeric variables in a fully customized design matrix. In the example below:

-   `(Intercept)`: Baseline expression when days = 0 and Condition = Control.
-   `DaysToSequencing`: Change in expression per unit of time (day) increase.
-   `Senescent`: Average difference in expression between Senescent and Control conditions.

This approach is useful when modeling continuous effects alongside group comparisons, and it provides complete flexibility in specifying the design.
 

```{r example_continuousDEG_2}

# Manually construct the design matrix
model_matrix <- model.matrix(~ DaysToSequencing + Condition, data = metadata_example)
colnames(model_matrix) <- c("(Intercept)", "DaysToSequencing", "Senescent")

# Provide the custom design matrix to calculateDE using the `modelmat` argument
DEGs_continuous2 <- calculateDE(data = counts_example,
                                modelmat = model_matrix)

# Access results for each coefficient
DEGs_continuous2$`(Intercept)`[1:3, ]
DEGs_continuous2$DaysToSequencing[1:3, ]
DEGs_continuous2$Senescent[1:3, ]

```

## Gene Set Enrichment Analyses

To perform GSEA, the user can use the `runGSEA()` function. This function takes a named list of differential expression statistics (one per contrast) and a set of gene signatures to compute enrichment scores.

-   `DEGList`: A list of differentially expressed genes (DEGs) for each contrast.

-   `gene_sets`: A list of gene sets, where each entry can be:

    -   A vector of gene names (unidirectional analysis).
    -   A data frame where the first column is the gene name and the second column indicates the expected direction (+1 or -1, bidirectional analysis).

-   `stat`: The ranking statistic. If NULL, the ranking statistic is automatically selected:

    -   `"B"` for gene sets with **no known direction** (vectors).
    -   `"t"` for **unidirectional** or **bidirectional** gene sets (data frames).
    -   If provided, this argument overrides the automatic selection.
    
-   `ContrastCorrection`: Logical, default is `FALSE`. If `TRUE`, applies an additional multiple testing correction (Benjamini–Hochberg) across all contrasts returned in the `DEGList` results list. This accounts for the number of contrasts tested per signature and provides more stringent control of false discovery rate across multiple comparisons (similar to answering the question “Is there any signature that is significant in any of the contrasts” instead of “For each contrast, is there any signature that is significant”). If `FALSE`, the function only corrects for the number of gene sets.



```{r GSEA, warning=FALSE, message=FALSE}
GSEAresults <- runGSEA(DEGList = DEGs, 
                       gene_sets = genesets_example,
                       stat = NULL,
                        ContrastCorrection = FALSE)

GSEAresults
```


Depending on the statistic used, the interpretation of the plots changes: 

1.  **B Statistic vs. t Statistic:**
    -   The B statistic does not specify the direction (enriched or depleted) of the gene set’s alterations. It only indicates the strength of evidence for any alterations.
    -   The t statistic orders genes from stronger evidence of over-expression to stronger evidence for under-expression. 
2.  **Graph's annotation:**
    -   When using the B statistic, the plot will include “Altered Gene Set” beneath its title to reflect this focus on whether genes are altered.
    -   For the t statistic, the plot will include “Enriched/Depleted Gene Set” beneath its title, indicating its focus on the enrichment or depletion of genes. 

After running GSEA, the user can visualize enrichment plots using the `plotGSEAenrichment()` function. This function generates enrichment plots for each gene signature and contrast, displaying also the **Normalized Enrichment Scores (NES)** and **adjusted p-value** for each enrichment result. As `plotGSEAenrichment`’s relevant graphical parameters:

-   `grid = TRUE`: Arranges the plots in a grid for better visualization.
-   `titlesize`: Adjusts title font size.
-   `nrow`/`ncol`: Specifies the grid layout for arranging plots.

```{r GSEA_plotenrichment, fig.width=10, fig.height=3, out.width="90%", warning=FALSE, message=FALSE}
plotGSEAenrichment(GSEA_results=GSEAresults, 
                   DEGList=DEGs, 
                   gene_sets=genesets_example, 
                   widthTitle=40, grid = TRUE, titlesize = 10, nrow=1, ncol=3) 
```

The `plotNESlollipop()` function creates lollipop plots for visualizing Gene Set Enrichment Analysis (GSEA) results. Each plot displays gene sets on the y-axis and Normalized Enrichment Scores (NES) on the x-axis, with a color gradient proportional to the adjusted p-value. The function supports multiple contrasts and includes options for customizing the color gradient, significance threshold, and plot layout. It can also arrange individual plots into a grid layout for comparative visualization.

```{r GSEA_lollypop, fig.width=5, fig.height=4, out.width="60%", warning=FALSE, message=FALSE}
plotNESlollipop(GSEA_results=GSEAresults, 
                saturation_value=NULL, 
                nonsignif_color = "#F4F4F4", 
                signif_color = "red",
                sig_threshold = 0.05, 
                grid = FALSE, 
                nrow = NULL, ncol = NULL, 
                widthlabels=13, 
                title=NULL, titlesize=12) 
```

The `plotCombinedGSEA()` function generates a scatter plot to visualize the results of Gene Set Enrichment Analysis (GSEA) across multiple contrasts. Each point represents a pathway, with:

-   X-axis: Normalized Enrichment Score (NES)
-   Y-axis: -log10 adjusted p-value (significance)
-   Color: Gene sets
-   Shape: Contrasts
-   Dashed line: Significance threshold

This function helps compare enrichment results when the number of contrasts and the number of pathways is high, scenario of which the example herein is not too representative.

```{r GSEA_volcano, fig.width=8, fig.height=3, out.width="100%", warning=FALSE, message=FALSE}
plotCombinedGSEA(GSEAresults, sig_threshold = 0.05, PointSize=6, widthlegend = 26 )
```

### Note on GSEA for All-negative gene sets

In gene set enrichment analysis, the behaviour of gene sets annotated with regulatory direction must be interpreted carefully. In our current implementation, if a gene set includes both positive and negative regulators of a process, the statistics for the negatively annotated genes are inverted, meaning that the enrichment reflects their effect on the process rather than simply their expression direction. This ensures that the final enrichment score captures the coordinated functional outcome of the gene set. 

The same logic is preserved when a gene set contains **only negatively annotated genes** (*e.g.* negative regulators of senescence): the statistics are inverted so that a positive normalised enrichment score (NES) implies these inhibitors are less expressed, thereby supporting the activation of the process they repress (*e.g.* senescence). Conversely, a negative NES would suggest their upregulation, which could potentially suppress the process. 
 
 
 
# Comparison of both approaches

Based on these very simple analyses, the REACTOME_CELLULAR_SENESCENCE showed consistently poor performance, highlighing caution when this or other MSigDB pathways appear in common enrichment-based analyses with RNA-seq data. The HernandezSegura signature, with its moderate size (55 genes) and annotated gene direction, performed best in enrichment-based methods such as GSEA. In contrast, the LiteratureMarkers set yielded stronger results in scoring approaches, likely due to a small number of highly informative genes rather than coordinated activity across the entire set.

These findings highlight important trade-offs: while scoring methods offer per-sample resolution and are less sensitive (in terms of statistical significance) to gene set size, making them useful for classification tasks, they may be overly influenced by a small subset of genes, which could limit biological interpretability. While more robust to sample heterogeneity and better at capturing coordinated expression changes, enrichment-based methods are sensitive to gene set composition and size. Caution is warranted when interpreting results, especially from score-based approaches, as strong signals may not always reflect the intended biological process, but rather a handful of dominant genes. 


## Visualise Individual Gene Behaviour.  

As highlighted in the previous section, score-based approaches can be disproportionately influenced by a small subset of genes. To address this, `markeR` includes dedicated functions for exploring individual gene behaviour, enabling users to assess if and which genes may be driving the overall signal. We demonstrate this functionality using the `LiteratureMarkers` gene set. 
 
`markeR` provides the wrapper function `VisualiseIndividualGenes` for plotting individual genes. In this tutorial, we illustrate each visualization function separately for clarity. However, the same outputs can be generated through the wrapper by setting the `type` argument to the desired visualization strategy (i.e., `"expression"`, `"correlation"`,`"violin"`, `"roc"`, `"auc"`, `"rocauc"`, `"cohend"`, `"pca"`), and the wrapper automatically dispatches to the correct function with the appropriate parameters. 
 
The `ExpressionHeatmap` function generates a heatmap to display the expression levels of selected senescence genes across samples. Samples are annotated by a chosen condition, and expression values are color-scaled for easy visual comparison. Clustering options and customizable color palettes allow for flexible and informative visualization.

```{r example_exprheatmap, fig.width=8, fig.height=3, out.width="70%", warning=FALSE, message=FALSE}

annotation_colors <- list( 
  Condition = c(
    "Senescent"     = "#65AC7C",  # Example color: greenish
    "Proliferative" = "#5F90D4"  # Example color: blueish 
  )
)

ExpressionHeatmap(data=counts_example, 
                  metadata = metadata_example, 
                  genes=genesets_example$LiteratureMarkers$gene,  
                  annotate.by = c("Condition"),
                  annotation_colors = annotation_colors,
                  colorlist = list(low = "#3F4193", mid = "#F9F4AE", high = "#B44141"),
                  cluster_rows = TRUE, 
                  cluster_columns = FALSE,
                  title = "Senescence Genes", 
                  titlesize = 20,
                  legend_position = "right",
                  scale_position="right")
```

```{r example_exprheatmap2, fig.width=8, fig.height=3, out.width="70%", warning=FALSE, message=FALSE}
VisualiseIndividualGenes(type="expression",
                  data=counts_example, 
                  metadata = metadata_example, 
                  genes=genesets_example$LiteratureMarkers$gene,  
                  annotate.by = c("Condition"),
                  annotation_colors = annotation_colors,
                  colorlist = list(low = "#3F4193", mid = "#F9F4AE", high = "#B44141"),
                  cluster_rows = TRUE, 
                  cluster_columns = FALSE,
                  title = "Senescence Genes", 
                  titlesize = 20,
                  legend_position = "right",
                  scale_position="right")
```


The `IndividualGenes_Violins` function creates violin plots to visualize the expression distributions of selected senescence genes across conditions. Jittered points represent individual samples, and grouping (x axis, `GroupingVariable`) and color variables (`ColorVariable` and `ColorValues`) from the metadata allow for additional stratification and insight. Customization options include layout, point size, colors, and axis labeling.

```{r exampleviolins, fig.width=10, fig.height=3, out.width="100%", warning=FALSE, message=FALSE}

senescence_triggers_colors <- c(
  "none" = "#E57373",  # Soft red   
  "Telomere shortening" = "#4FC3F7"  # Vivid sky blue  
)


IndividualGenes_Violins(data = counts_example, 
                        metadata = metadata_example, 
                        genes = genesets_example$LiteratureMarkers$gene, 
                        GroupingVariable = "Condition", 
                        plot=TRUE, 
                        ncol=NULL, 
                        nrow=1, 
                        divide=NULL, 
                        invert_divide=FALSE,
                        ColorValues=senescence_triggers_colors, 
                        pointSize=2, 
                        ColorVariable="SenescentType", 
                        title="Senescence Genes", 
                        widthTitle=16,
                        y_limits = NULL,
                        legend_nrow=1, 
                        xlab="Condition",
                        colorlab="") 
```

The `CorrelationHeatmap` function displays pairwise correlations between selected genes, helping to reveal co-expression patterns within the senescence signature. Correlations can be computed separately for different conditions, and the heatmap is fully customizable with options for clustering, color scaling, and correlation method (e.g., Spearman or Pearson).


```{r example_heatmap, fig.width=8, fig.height=4, out.width="70%", warning=FALSE, message=FALSE}

CorrelationHeatmap(data=counts_example, 
                   metadata = metadata_example, 
                   genes=genesets_example$LiteratureMarkers$gene, 
                   separate.by = "Condition", 
                   method = "spearman",  
                   colorlist = list(low = "#3F4193", mid = "#F9F4AE", high = "#B44141"),
                   limits_colorscale = c(-1,0,1), 
                   widthTitle = 16, 
                   title = "Senescence Genes", 
                   cluster_rows = TRUE, 
                   cluster_columns = TRUE,  
                   detailedresults = FALSE, 
                   legend_position="right",
                   titlesize=20)
```

The `ROCandAUCplot` function evaluates the discriminatory power of individual genes in the signature by computing ROC curves and AUC values based on a binary classification (e.g., senescent vs. proliferative). If the selected grouping variable has more than two levels, the user can specify the reference group using the class parameter. For example, if a variable has levels A, B, C, and D, setting `class = c("A", "B")` will group samples from A and B together as the positive class*, while the remaining samples (C and D) are automatically grouped as the negative class. Additionally, the user can use the `group_var` parameter to split and display results separately for each level of another metadata variable — allowing for subgroup-specific ROC analyses. Outputs include individual ROC plots and an AUC heatmap, with customisable layout, color schemes, and clustering options—ideal for identifying genes with strong discriminatory ability. If `group_var`is not specified, the AUC values will be displayed in a barplot.

* For ease of interpretation, the directionality of the comparison is adjusted so that the AUC is always >=0.5, showing only discriminatory power rather than directionality.

```{r rocAUCexample, fig.width=10, fig.height=4, out.width="100%", warning=FALSE, message=FALSE}
 
ROCandAUCplot(counts_example, 
              metadata_example, 
              condition_var = "Condition", 
              class = "Senescent", 
              group_var=NULL,
              title = NULL,
              genes=genesets_example$LiteratureMarkers$gene, 
              plot_type = "all",
              auc_params =  list(colors =  "#3B415B",
                                    limits = c(0.5,1) ),
              roc_params = list(nrow=3,
                                ncol=3, 
                                colors="#3B415B"),
              commomplot_params = list(widths=c(0.5,0.3)))
 
```

The `CohenD_IndividualGenes` function computes the effect size (Cohen’s d) of the difference in expression of each gene between two conditions, given by the variable `condition_var`. If the selected condition variable has more than two levels, the `class` parameter specifies which condition will be compared to the rest. Additionally, the user can use the `group_var` parameter to split and display results separately for each level of another metadata variable. Results are visualized as a heatmap, with customizable color scales and clustering options for easy interpretation of effect sizes across genes. If `group_var` is not specified, the function will return a barplot instead.

```{r cohendexample, fig.width=6, fig.height=4, out.width="70%", warning=FALSE, message=FALSE}
 
CohenD_IndividualGenes(counts_example, 
                       metadata_example, 
                       genes=genesets_example$LiteratureMarkers$gene,
                       condition_var = "Condition", 
                       class = "Senescent", 
                       group_var = NULL,  
                       params = list(colors = "#3B415B",
                                             limits = NULL,
                                             cluster_rows=TRUE))
```

The `plotPCA` function performs a PCA on the expression of a selected set of genes to understand if they explain enough variance in the data, allowing the user to test if the genes in the signature are sufficient to separate the group of interest (given by the `ColourVariable' parameter). While previous metrics assess discriminatory power quantitatively (e.g., AUC), PCA provides a complementary, unsupervised visualization that can help assess whether the gene signature is sufficient to visually separate groups of interest. Users can customize which principal components to display using the PCs argument and adjust layout, point size, and colour annotations for comparison across conditions.

```{r pca, fig.width=7, fig.height=3, out.width="90%", warning=FALSE, message=FALSE}

annotation_colors <- c(  
  "Senescent"     = "#65AC7C",  # Example color: greenish
  "Proliferative" = "#5F90D4"  # Example color: blueish 
)


plotPCA(data = counts_example, 
        metadata = metadata_example, 
        genes=genesets_example$LiteratureMarkers$gene, 
        scale=FALSE, 
        center=TRUE, 
        PCs=list(c(1,2), c(2,3), c(3,4)), 
        ColorVariable="Condition",
        ColorValues=annotation_colors,
        pointSize=5,
        legend_nrow=1, 
        ncol=3, 
        nrow=NULL)
```

## Comment 

As demonstrated by the behaviour of individual genes in the `LiteratureMarkers` gene set, *LMNB1*, *MKI67*, and *GLB1* appear to drive the overall signal. These genes consistently show higher performance metrics (e.g., Cohen’s d, AUC), strong expression changes between conditions, and *LMNB1* and *MKI67* specifically have correlated expression patterns. This underscores the importance of examining gene-level behaviour, as a strong overall signature score may reflect the influence of only a few informative genes, rather than coordinated activity across the entire set. In this case, the strong performance of the `LiteratureMarkers` set in scoring approaches is likely driven by these genes. However, relying heavily on a few markers can be a caveat: for example, *MKI67* and *LMNB1* (proliferation-related genes) may also change in other biological contexts like quiescence or differentiation, potentially limiting their specificity for senescence. Thus, the choice of gene set and analysis strategy should be guided by the research question, and complemented with both score distributions, enrichment analyses, and individual gene behaviour. Notably, scoring with just these three genes yielded results similar to the full `LiteratureMarkers` set.

```{r example_genesdrivingresults, fig.width=6, fig.height=4, out.width="60%", warning=FALSE, message=FALSE}
 
GeneSets_Example_DrivingGenes <- list(LiteratureMarkers=genesets_example$LiteratureMarkers,
                                          LiteratureMarkers_subset=genesets_example$LiteratureMarkers[genesets_example$LiteratureMarkers$gene %in% c("LMNB1", "MKI67","GLB1"),])
print(GeneSets_Example_DrivingGenes)

PlotScores(data = counts_example, 
           metadata = metadata_example, 
           gene_sets =  GeneSets_Example_DrivingGenes,
           ColorVariable = "SenescentType", 
           Variable="Condition",  
           method ="logmedian", 
           ColorValues = senescence_triggers_colors, 
           ConnectGroups=TRUE, 
           ncol = NULL, 
           nrow = NULL, 
           widthTitle=24, 
           limits = NULL, 
           legend_nrow = 1, 
           pointSize=4,
           compute_cohen=TRUE,
           cond_cohend=cond_cohend,
           title="Marthandan et al. 2016",
           labsize=9, 
           titlesize = 12)  
```


# Session Information

```{r}
sessionInfo()
```

